---
title: "Bootstrap variable selection"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in:
# rendered as a standalone document the project's format block never applies,
# which is how server-rendered reports end up with sibling _files/ trees instead
# of being self-contained.
#
# NOTE: output-dir is a PROJECT setting and cannot be set here, so the .html
# lands next to its source. With embed-resources that is a single standalone
# file, which is the property that actually matters when a report is sent on.
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the SAS job this replaces, and paste its `%hazboot` call
     verbatim. The call is the specification: it names the candidate pool, the
     phases each candidate was offered to, the entry and stay levels, and the
     replicate count. Prose describing it is a paraphrase that drifts; the call
     does not. -->

Replaces `analyses/<job>.sas`: paste its `%hazboot` call here — the candidate
pool, which phases carried candidates, the entry and stay levels, and
`resampl=`.

This job **does not fit the final model**. It screens. The companion `hm` job
fits the multivariable model.

⚠️ `bh` and `hm` are **parallel analyses, not a pipeline.** `%hazboot` and
`%model` were parallel in SAS and these jobs reproduce that. Do not describe
this job's retained set as what `hm` fits, and do not write a handoff file for
it — an earlier study report did both, and nothing read the file. A reader who
believes the claim will assume the `hm` model was screened for reliability
first. It was not.

```{r}
#| label: setup
# This file runs either from the directory it was scaffolded into or from the
# project root. Resolve which, so that no path in this document needs editing:
# forgetting one of several ../ paths is the instantiation mistake that would
# otherwise be easy to make and quiet when made.
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
suppressPackageStartupMessages({
  library(TemporalHazard)
  library(hvtiRbootstrap)
  library(hvtiRutilities)
  library(ggplot2)
})
```

```{r}
#| label: edit-guard
#| results: asis
# The markers in this file name work a study author still has to do, and
# README.md says a job that still contains one has not been finished. This
# chunk is what makes that TRUE rather than merely stated.
#
# Without it an unedited job renders green over a meaningless analysis. In THIS
# file the danger is that EXPECT_CHUNKS and EXPECT_BOOT still hold the template's
# numbers, so the completeness check compares the run against a denominator
# nobody chose -- and passes, reporting frequencies over the wrong base.
#
# knitr::current_input() is NULL outside a render -- a study author stepping
# through chunks in RStudio -- and there is no file to scan then, so this is a
# no-op in that case. Same handling as the `set` guard below, for the same
# reason.
#
# The token is BUILT, not written literally, and that is not stylistic. Quarto
# knits through an intermediate and current_input() returns THAT file, so the
# scan reads a copy of this chunk along with everything else: a literal
# grep("<token>", .src) here matches its own source line, and the guard then
# fires on every render, finished or not. Measured, not theorised -- a probe
# found three markers in a file containing two. A guard that cries wolf on a
# finished job is a guard that gets deleted, which returns us to issue #27.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    # Report the marker TEXT. Line numbers here are the intermediate's, not
    # this .qmd's, so they need not match what the author sees in an editor.
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it.\nTo render a partly-worked draft anyway, set ",
      "HVTI_TEMPLATE_DRAFT to 1, true or yes -- the report then carries a ",
      "draft banner. Any other value, including 0, leaves this guard strict."
    )
    # Only the documented values enable draft mode. nzchar() alone would treat
    # HVTI_TEMPLATE_DRAFT=0 as ON, silently downgrading this stop() to a
    # warning for someone whose shell profile sets it to 0 meaning OFF -- a
    # switch whose enablement does not mean what it says, which is the same
    # defect this guard exists to catch. An unrecognised value falls through to
    # strict on purpose: the author then sees the stop and re-reads the docs,
    # which is the safe direction to be wrong in.
    if (tolower(Sys.getenv("HVTI_TEMPLATE_DRAFT")) %in% c("1", "true", "yes")) {
      # The banner is not optional. A draft render that looks like a finished
      # one is the same defect with an extra step: the warning scrolls past in
      # a log, while the .html is the artifact that gets sent to someone.
      warning(.msg, call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Rendered with `HVTI_TEMPLATE_DRAFT` set. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by new_job() from the endpoint/type it
     was called with, which are also the filename's first two fields. They
     name the SET this job belongs to, and every artifact path in the
     document is computed from them, so no output path below needs editing.

     Do NOT hand-edit these declarations: the render-time guard below checks
     them against this file's own name and stops if they disagree. To put
     this job in a different set, call new_job() again with the new
     endpoint/type rather than editing this copy in place.

     A set is keyed on (endpoint, analysis type), not on the endpoint alone.
     One endpoint is analysed by several methods -- parametric hazard, random
     forest survival, competing risks -- and those chains share their upstream.
     Two sets keyed on `dead_pa` alone would write their life tables to the
     same file. -->

```{r}
#| label: set
ENDPOINT <- "dead_pa"
TYPE     <- "hz"

# `new_job()` writes ENDPOINT/TYPE from the same values it put in this file's
# name, but a hand-edited declaration can drift from it afterward. set_path()
# below resolves from the declarations, not the filename, so a drifted
# declaration would silently write into ANOTHER set's artifact directory --
# exactly the collision the (endpoint, type) key exists to prevent, re-entered
# through the body instead of the name.
#
# knitr::current_input() is NULL outside a render (a study author running
# chunks interactively in RStudio), and there is no filename to check against
# yet, so this is a no-op in that case rather than a spurious error.
.current <- knitr::current_input()
if (!is.null(.current)) {
  # Quarto knits through an intermediate, so `knitr::current_input()` returns
  # `<endpoint>-<type>-<ordinal>-<prefix>.rmarkdown` here rather than the
  # `.qmd` this was scaffolded as. Strip whatever extension is actually present
  # rather than hard-coding one, so this doesn't depend on a build-tool detail
  # staying the same.
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint, "', type '",
         .name_type, "'), but declares ENDPOINT = \"", ENDPOINT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

# Resolve a path inside this set's artifact directory. `kind` is the artifact
# folder -- "estimates" for serialized results, "graphs" for figures. The set
# directory sits one layer under the kind and never two, which is the whole
# layout rule.
#
# Created on first use rather than up front, so a job that writes nothing leaves
# no empty directories behind.
set_path <- function(kind, file) {
  d <- file.path(.root, kind, paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}

# The screen this job reports on was run by a companion script, which wrote its
# chunks into this set's `estimates` directory. Reading them through set_path()
# is what keeps a screen filed against the set whose candidates it screened.
```

## The screen

```{r}
#| label: expect
# EDIT: what this run was LAUNCHED as, not what happens to be on disk.
#
# Nothing in a chunk knows how many siblings it was launched alongside, so these
# two numbers are the only thing that can tell a partial pool from a complete
# one. Without them a render at hour eight of a twelve-hour run produces a
# report that is wrong in no visible way: every health check passes, every
# frequency is honestly computed, and only the denominator is not the intended
# one.
#
# For a single unchunked run, set EXPECT_CHUNKS to 1.
EXPECT_CHUNKS <- 25L    # EDIT: how many chunks you launched
EXPECT_BOOT   <- 500L   # EDIT: %hazboot(resampl=) -- total replicates wanted

# EDIT: the file-name prefix your runner wrote its chunks under. It must match
# the runner exactly: boot_chunk_files() quotes this literally and a prefix that
# does not match finds nothing, while a prefix that matches TOO MUCH pools
# chunks from another run -- and no downstream check can see that.
BOOT_PREFIX <- "bh"
```

```{r}
#| label: load
# The screen is not run here. It is read.
#
# hzr_bootstrap() writes nothing until its final replicate and a full screen is
# days of compute, so a run that dies near the end leaves nothing at all. The
# companion runner exists to convert one all-or-nothing multi-day call into N
# independently restartable ones. That file boundary is a DURABILITY boundary,
# not tidiness, which is why this report reads chunks rather than producing
# them.
.boot_dir <- dirname(set_path("estimates", "."))
.chunk_files <- boot_chunk_files(.boot_dir, prefix = BOOT_PREFIX)
.single_file <- set_path("estimates", paste0(BOOT_PREFIX, ".rds"))

if (length(.chunk_files)) {
  # Pooling is checked, not assumed: boot_pool_chunks() refuses chunks that
  # disagree on the dataset checksum, the entry/stay levels, the base
  # parameters, the usable candidate counts, the engine version or the step
  # cap, and refuses two chunks sharing a seed. Every one of those failures is
  # silent on its own -- it shows up as a slightly different frequency, never
  # as an error -- which is why the refusal belongs here rather than in a note.
  bag <- boot_pool_chunks(lapply(.chunk_files, readRDS))
} else if (file.exists(.single_file)) {
  bag <- readRDS(.single_file)
} else {
  stop("No bootstrap output found. This screen is a long job -- days, not ",
       "minutes -- and is run from a companion script rather than inside this ",
       "render, because the run writes nothing until its final replicate.\n",
       "Run your chunks first, then render this report.", call. = FALSE)
}
```

```{r}
#| label: completeness
#| results: asis
# A callout, not a table cell. A provisional report that cannot say so is the
# failure being prevented, and it must survive someone receiving the .html
# without knowing when it was made -- which a number buried in a provenance
# table does not.
.shortfall <- boot_shortfall(bag, EXPECT_CHUNKS, EXPECT_BOOT)
if (!is.null(.shortfall)) {
  cat("\n::: {.callout-warning title=\"Provisional -- this pool is incomplete\"}\n")
  cat(.shortfall, "\n", sep = "")
  cat(":::\n\n")
}
```

## Provenance

```{r}
#| label: contract
# The runner is NOT templated -- it is a study file -- so the fields this report
# reads are a contract between two files that no shared function enforces.
# Stated here, and checked, because the alternative is a report that indexes a
# field the runner stopped writing and renders `NULL` into a provenance table as
# though it were an answer.
#
# A pooled run has already had most of these checked by boot_pool_chunks(),
# which refuses chunks that disagree. A SINGLE unchunked run has not: nothing
# ran between the runner and here. This is the only check that covers both.
.required <- c("n_boot", "seed", "slentry", "slstay", "base_params", "usable",
               "requested", "n_rows", "elapsed_mins", "manifest", "boot")
.absent <- .required[!.required %in% names(bag)]
if (length(.absent)) {
  stop("The bootstrap output is missing ", length(.absent), " field(s) this ",
       "report reads: ", paste(.absent, collapse = ", "), ".\nEither the runner ",
       "that wrote it predates them, or it is not the runner this template ",
       "expects. Reporting over it would print blanks where the screen's ",
       "criteria belong.", call. = FALSE)
}
```

```{r}
#| label: provenance
# CPU cost, NOT wall clock. boot_pool_chunks() SUMS elapsed across chunks, so on
# a chunked run this is total compute; chunks run in parallel finish in a
# fraction of it. Reported in hours because the minutes figure reads as a
# wall-clock number and is off by the chunk count.
cpu_hours <- bag$elapsed_mins / 60

# The digest travels WITH its algorithm. An md5 and a sha256 of the same file
# are different strings, and of different files may not be, so a bare digest
# recorded without its algorithm is not evidence of anything.
checksum <- NA_character_
for (algo in c("sha256", "md5")) {
  if (!is.null(bag$manifest[[algo]])) {
    checksum <- paste0(algo, ":", bag$manifest[[algo]])
    break
  }
}

# A version string alone cannot say which codebase ran: one real package version
# existed as two, one with a selection criterion and one without, and the
# selection criterion is precisely what decides what a screen selects.
engine <- if (!is.null(bag$th_sha)) {
  paste0("sha:", bag$th_sha)
} else if (!is.null(bag$th_version)) {
  paste0("version:", bag$th_version)
} else {
  NA_character_
}

# Summarised, not listed: at 25 chunks the joined seed string is a single
# 270-character cell that wrecks the table. Every seed is in the table below.
n_chunks <- if (is.null(bag$n_chunks)) 1L else as.integer(bag$n_chunks)

provenance <- data.frame(
  item = c("Replicates pooled", "Chunks pooled", "Entry level (slentry)",
           "Stay level (slstay)", "Candidates offered", "Candidates usable",
           "Rows screened", "Replicates that fitted", "Replicates that failed",
           "CPU hours (summed)", "Fitting engine", "Dataset checksum",
           "Seeds"),
  value = c(format(bag$n_boot), format(n_chunks), format(bag$slentry),
            format(bag$slstay), format(bag$requested), format(bag$usable),
            format(bag$n_rows), format(bag$boot$n_success),
            format(bag$boot$n_failed), sprintf("%.1f", cpu_hours),
            engine, checksum,
            paste0(n_chunks, " distinct (listed below)")),
  stringsAsFactors = FALSE
)
provenance
```

```{r}
#| label: seeds
# Every seed, so that a rerun is reproducible and a duplicate is visible. Two
# chunks sharing a seed contain literally the SAME replicates: pooling them
# counts each twice and reports a Monte-Carlo error smaller than the run
# actually has. boot_pool_chunks() refuses that outright, so a pooled run cannot
# reach here with a duplicate -- this table is what lets you check a single run,
# and what lets a reader reproduce either.
data.frame(chunk = seq_along(bag$seeds),
           seed  = format(bag$seeds, scientific = FALSE),
           stringsAsFactors = FALSE)
```

## Candidates that were never screened

<!-- Counts first, detail second, and that ordering is not cosmetic. The detail
     table runs to dozens of rows on a real pool, and a reader who has to scroll
     it to discover that a whole CLASS of variable was never screened will not
     discover it. -->

```{r}
#| label: dropped-summary
# A candidate the screen never saw cannot appear at any frequency, so its
# absence from the table below is indistinguishable from never having been
# selected. That is the reason this section exists at all.
if (is.null(bag$dropped) || !nrow(bag$dropped)) {
  cat("Every candidate offered was screened.\n")
} else {
  as.data.frame(table(phase = bag$dropped$phase, reason = bag$dropped$reason))
}
```

```{r}
#| label: dropped-detail
if (!is.null(bag$dropped) && nrow(bag$dropped)) {
  bag$dropped[order(bag$dropped$phase, bag$dropped$reason), ]
}
```
