---
title: "Bootstrap variable selection"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in:
# rendered as a standalone document the project's format block never applies,
# which is how server-rendered reports end up with sibling _files/ trees instead
# of being self-contained.
#
# NOTE: output-dir is a PROJECT setting and cannot be set here, so the .html
# lands next to its source. With embed-resources that is a single standalone
# file, which is the property that actually matters when a report is sent on.
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the SAS job this replaces and paste its `%hazboot` call. The
     call is the specification -- replicate count, entry and stay criteria and
     the candidate pool all read off it -- so keep it in the document rather
     than transcribing its numbers into prose. The next person to touch this
     needs to see what is being reproduced. -->

Replaces `analyses/<job>.sas`: paste its `%hazboot(...)` call here.

Each replicate resamples the cohort with replacement and runs a fresh forward
selection over the whole candidate pool. The **selection frequency** -- the
percentage of replicates in which a variable was retained -- is the deliverable,
and it is what a published *Reliability (%)* column reports.

This job **does not fit the final model**. It screens.

```{r}
#| label: setup
# This file runs either from the directory it was scaffolded into or from the
# project root. Resolve which, so that no path in this document needs editing:
# forgetting one of several ../ paths is the instantiation mistake that would
# otherwise be easy to make and quiet when made.
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
suppressPackageStartupMessages({
  library(TemporalHazard)
  library(hvtiRbootstrap)
  library(hvtiRutilities)
})
```

```{r}
#| label: edit-guard
#| results: asis
# The markers in this file name work a study author still has to do, and
# README.md says a job that still contains one has not been finished. This
# chunk is what makes that TRUE rather than merely stated.
#
# Without it an unedited job renders green over a meaningless analysis. In THIS
# file the danger is the DENOMINATOR: EXPECT_CHUNKS and EXPECT_BOOT would still
# hold the template's numbers, so the completeness check compares the run
# against a total nobody chose -- and PASSES. Every frequency below is then
# honestly computed over the wrong base, and nothing in the output looks
# wrong.
#
# knitr::current_input() is NULL outside a render -- a study author stepping
# through chunks in RStudio -- and there is no file to scan then, so this is a
# no-op in that case. Same handling as the `set` guard below, for the same
# reason.
#
# The token is BUILT, not written literally, and that is not stylistic. Quarto
# knits through an intermediate and current_input() returns THAT file, so the
# scan reads a copy of this chunk along with everything else: a literal
# grep("<token>", .src) here matches its own source line, and the guard then
# fires on every render, finished or not. Measured, not theorised -- a probe
# found three markers in a file containing two. A guard that cries wolf on a
# finished job is a guard that gets deleted, which returns us to issue #27.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    # Report the marker TEXT. Line numbers here are the intermediate's, not
    # this .qmd's, so they need not match what the author sees in an editor.
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it.\nTo render a partly-worked draft anyway, set ",
      "HVTI_TEMPLATE_DRAFT to 1, true or yes -- the report then carries a ",
      "draft banner. Any other value, including 0, leaves this guard strict."
    )
    # Only the documented values enable draft mode. nzchar() alone would treat
    # HVTI_TEMPLATE_DRAFT=0 as ON, silently downgrading this stop() to a
    # warning for someone whose shell profile sets it to 0 meaning OFF -- a
    # switch whose enablement does not mean what it says, which is the same
    # defect this guard exists to catch. An unrecognised value falls through to
    # strict on purpose: the author then sees the stop and re-reads the docs,
    # which is the safe direction to be wrong in.
    if (tolower(Sys.getenv("HVTI_TEMPLATE_DRAFT")) %in% c("1", "true", "yes")) {
      # The banner is not optional. A draft render that looks like a finished
      # one is the same defect with an extra step: the warning scrolls past in
      # a log, while the .html is the artifact that gets sent to someone.
      warning(.msg, call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Rendered with `HVTI_TEMPLATE_DRAFT` set. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by new_job() from the endpoint/type it
     was called with, which are also the filename's first two fields. They
     name the SET this job belongs to, and every artifact path in the
     document is computed from them, so no output path below needs editing.

     Do NOT hand-edit these declarations: the render-time guard below checks
     them against this file's own name and stops if they disagree. To put
     this job in a different set, call new_job() again with the new
     endpoint/type rather than editing this copy in place.

     A set is keyed on (endpoint, analysis type), not on the endpoint alone.
     One endpoint is analysed by several methods -- parametric hazard, random
     forest survival, competing risks -- and those chains share their upstream.
     Two sets keyed on `dead_pa` alone would write their life tables to the
     same file. -->


```{r}
#| label: set
ENDPOINT <- "dead_pa"
TYPE     <- "hz"

# `new_job()` writes ENDPOINT/TYPE from the same values it put in this file's
# name, but a hand-edited declaration can drift from it afterward. set_path()
# below resolves from the declarations, not the filename, so a drifted
# declaration would silently write into ANOTHER set's artifact directory --
# exactly the collision the (endpoint, type) key exists to prevent, re-entered
# through the body instead of the name.
#
# knitr::current_input() is NULL outside a render (a study author running
# chunks interactively in RStudio), and there is no filename to check against
# yet, so this is a no-op in that case rather than a spurious error.
.current <- knitr::current_input()
if (!is.null(.current)) {
  # Quarto knits through an intermediate, so `knitr::current_input()` returns
  # `<endpoint>-<type>-<ordinal>-<prefix>.rmarkdown` here rather than the
  # `.qmd` this was scaffolded as. Strip whatever extension is actually present
  # rather than hard-coding one, so this doesn't depend on a build-tool detail
  # staying the same.
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint, "', type '",
         .name_type, "'), but declares ENDPOINT = \"", ENDPOINT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

# Resolve a path inside this set's artifact directory. `kind` is the artifact
# folder -- "estimates" for serialized results, "graphs" for figures. The set
# directory sits one layer under the kind and never two, which is the whole
# layout rule.
#
# Created on first use rather than up front, so a job that writes nothing leaves
# no empty directories behind.
set_path <- function(kind, file) {
  d <- file.path(.root, kind, paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}

# `bh` reads the chunks its companion runner wrote through this, and writes its
# pooled screen back the same way, so a screen can never be filed against a
# different set than the replicates it pooled. The write is in the `save` chunk
# at the foot of this file.
```

## The screen this reports on

<!-- EDIT: nothing in this section, but read it before changing the two
     expectations below.

     This job does not RUN the screen. `hzr_bootstrap()` writes nothing until
     its final replicate and a full screen is days of compute, so a run that
     dies near the end leaves nothing at all. The run is therefore chunked from
     a companion script, and this report reads what that script wrote. The file
     boundary is a durability boundary, not tidiness. -->

```{r}
#| label: bootstrap
# EDIT: the prefix your companion runner writes its output under.
#
# The runner is not templated -- it is a study's own script -- so its filenames
# are its own choice, and this has to be told rather than guessed. One long run
# writes <prefix>.rds; a chunked run writes <prefix>.chunk01.rds, chunk02 and so
# on. Both are read the same way below.
CHUNK_PREFIX <- "bagging"

# EDIT: what this run was LAUNCHED as, not what happens to be on disk.
#
# Nothing in a chunk knows how many siblings it was launched alongside, so these
# two numbers are the only thing that can tell a partial pool from a complete
# one. Without them a render at hour eight of a twelve-hour run produces a
# report that is wrong in no visible way: every health check passes, every
# frequency is honestly computed, and only the denominator is not the one
# anybody chose.
#
# For a single unchunked run, set EXPECT_CHUNKS to 1.
EXPECT_CHUNKS <- 25L    # EDIT: how many chunks you launched
EXPECT_BOOT   <- 500L   # EDIT: %hazboot(resampl=) -- total replicates wanted

# Artifacts live under estimates/<endpoint>-<type>/, by set, so a screen can
# never be read from or filed against a different set than the one this job
# belongs to. set_path() creates the directory; dirname() is what
# boot_chunk_files() wants.
est_dir <- dirname(set_path("estimates", paste0(CHUNK_PREFIX, ".rds")))
rds     <- file.path(est_dir, paste0(CHUNK_PREFIX, ".rds"))
chunks  <- boot_chunk_files(est_dir, prefix = CHUNK_PREFIX)

if (file.exists(rds)) {
  bag <- readRDS(rds)
} else if (length(chunks)) {
  # boot_pool_chunks() REFUSES chunks that are not draws from one screen -- a
  # shared seed, a dataset rewritten mid-run, different entry criteria. So a
  # pooled frequency below is one that was checked rather than assumed, and a
  # stop() here is the check working, not an inconvenience to route around.
  bag <- boot_pool_chunks(lapply(chunks, readRDS))
} else {
  stop("No bootstrap output found in ", est_dir, ".\n",
       "This screen is a long job -- days, not minutes -- and is run from a ",
       "companion script rather than inside this render, because the run ",
       "writes nothing until its final replicate.\n",
       "Run your chunks first, then render this report.", call. = FALSE)
}
```

```{r}
#| label: completeness
#| results: asis
#| echo: false
# Rendered as a CALLOUT rather than a table cell. Looking at a screen mid-run is
# routine and worth keeping; what is not acceptable is a provisional report that
# cannot say so -- including to someone who receives the .html later, without
# the context of when it was made. A cell in a provenance table further down
# does not survive that journey. A banner does.
shortfall <- boot_shortfall(bag, EXPECT_CHUNKS, EXPECT_BOOT)
if (!is.null(shortfall)) {
  cat("::: {.callout-warning}\n",
      "## Provisional -- this screen is not the complete run\n\n",
      shortfall, "\n:::\n", sep = "")
}
```

## Provenance

```{r}
#| label: provenance
# Summarised, not listed: at 25 chunks the joined seed string is a single
# 270-character cell that wrecks the table. Every seed is in the `seeds` table
# below, so nothing is lost by compressing it here.
seeds_cell <- if (is.null(bag$seeds) || length(bag$seeds) <= 3L) {
  bag$seed
} else {
  paste0(length(bag$seeds), " seeds, ",
         format(min(bag$seeds), scientific = FALSE), "-",
         format(max(bag$seeds), scientific = FALSE))
}

knitr::kable(data.frame(
  n_boot          = bag$n_boot,
  expected_boot   = EXPECT_BOOT,
  # 1 for a single run; higher when the screen was pooled from chunks.
  n_chunks        = if (is.null(bag$n_chunks)) 1L else bag$n_chunks,
  expected_chunks = EXPECT_CHUNKS,
  seeds           = seeds_cell,
  slentry         = bag$slentry,
  slstay          = bag$slstay,
  cand_listed     = paste(bag$requested, collapse = " / "),
  cand_usable     = paste(bag$usable, collapse = " / "),
  n_rows          = bag$n_rows,
  # CPU cost, NOT wall clock. boot_pool_chunks() SUMS elapsed across chunks, so
  # on a chunked run this is the total compute; chunks run in parallel finish in
  # a fraction of it. Reported in hours because the minutes figure reads as a
  # wall-clock number and is off by the chunk count.
  cpu_hours       = round(bag$elapsed_mins / 60, 1),
  TemporalHazard  = bag$th_version
), caption = paste("Candidate counts are listed / usable, one per phase.",
                   "cpu_hours is summed across chunks and is compute cost,",
                   "not elapsed time."))

knitr::kable(bag$manifest, caption = "Dataset this screen ran against")
```

```{r}
#| label: seeds
#| code-fold: true
# Every seed, one row per chunk. A repeated seed means repeated replicates,
# which would count each twice and report a Monte-Carlo error the run does not
# have. boot_pool_chunks() already refuses duplicates, so this table is
# corroboration a reader can check rather than the gate itself.
if (!is.null(bag$seeds) && length(bag$seeds) > 1L) {
  knitr::kable(data.frame(chunk = seq_along(bag$seeds),
                          seed  = format(bag$seeds, scientific = FALSE)),
               row.names = FALSE, caption = "Seed per pooled chunk")
}
```

Unlike SAS's `seed = -1`, which seeds from the time of day, this run **is**
reproducible: same seed, same result. That is a genuine improvement over the job
being reproduced, not merely a match to it — the original run cannot be
recovered by anyone, including whoever ran it.

## Candidates that were not screened

Every candidate the job lists but the screen could not use is named here. A
candidate dropped quietly shrinks the screen without shrinking the reported
pool, and the frequencies would then be percentages of a denominator nobody
stated.

```{r}
#| label: dropped-summary
# Counts FIRST. The detail table below runs to dozens of rows on a real pool,
# and a reader who has to scroll it to discover that a whole CLASS of variable
# was never screened will not discover it.
if (nrow(bag$dropped) == 0L) {
  knitr::kable(data.frame(result = "every listed candidate was usable"))
} else {
  tab <- table(phase = bag$dropped$phase, reason = bag$dropped$reason)
  by_reason <- as.data.frame(tab, stringsAsFactors = FALSE)
  by_reason <- by_reason[by_reason$Freq > 0, ]
  knitr::kable(by_reason[order(by_reason$phase, -by_reason$Freq), ],
               row.names = FALSE,
               caption = "Why listed candidates were not screened, per phase")
}
```

```{r}
#| label: dropped-detail
#| code-fold: true
# Folded. The counts above are what a reader must not miss; this is what they
# consult once a count has raised the question.
if (nrow(bag$dropped) > 0L) {
  knitr::kable(bag$dropped[order(bag$dropped$phase, bag$dropped$variable), ],
               row.names = FALSE,
               caption = "Every candidate not screened, one row per phase")
}
```

```{r}
#| label: coercion
# What the runner did to the pool before screening it, reported because these
# repairs change WHICH VARIABLES COULD BE SELECTED and none of it is visible in
# a frequency table. Older outputs predate these fields, hence the NULL guards.
coerced_n <- if (isTRUE(bag$coerce_factors)) length(bag$coerced) else 0L
knitr::kable(data.frame(
  factors_coerced_to_numeric = coerced_n,
  pruned_to_one_form         = isTRUE(bag$prune_to_one_form),
  candidates_screened        = paste(bag$usable, collapse = " / ")
), caption = "Pool repairs applied before screening")
```

::: {.callout-important}
## Factor-coded binaries decide what the screen can see

`read_built()` returns 0/1 clinical columns as **factors**, because `vars.sas`
mean-imputes and adds `ms_*` indicators, so a two-level variable arrives as a
three-level factor. A screen that filters to numeric columns cannot see any of
them. SAS has no factors, so `%hazboot` screened them all.

Left unrepaired this is not a rounding error. On the study this template was
extracted from it removed **70 of 230 candidates per phase**, and among them
were four of the risk factors that study's paper reports. The table above says
whether your runner coerced them back; if `factors_coerced_to_numeric` is 0 and
your built dataset came from `vars.sas`, check before reading anything below.

Nothing in a frequency table shows this. The candidates are simply absent, and
the percentages are honest percentages of whatever remained.
:::

```{r}
#| label: interval-note
# Interval-censored records are treated as right-censored for this stage: the
# formula interface, which is the only one hzr_bootstrap() can resample, cannot
# carry interval censoring. Reported rather than left in a comment, because the
# reader cannot see it anywhere else -- and because for some cohorts it is a
# large enough fraction to qualify the result.
knitr::kable(data.frame(
  rows_screened              = bag$n_rows,
  interval_censored_as_right = bag$n_interval,
  pct_of_cohort              = round(100 * bag$n_interval / bag$n_rows, 2)
), caption = "Interval censoring, and how much of the cohort it affects")
```

## Did the screen actually run?

The check that has to come **before** any frequency is read. A bootstrap that
never resamples, and a screen that never selects, both produce `n_success` at
its maximum, `n_failed` at zero, no warning, and a summary table that looks
exactly like a result.

⚠️ **A screen that selected nothing is a failure, not a finding.** That is worth
stating flatly, because the artefact it produces is not obviously broken — it is
a table of variables with perfectly clean numbers beside them.

<!-- EDIT: nothing here, but check your RUNNER against this before trusting any
     number below. The runner is a study's own script and is not templated, so
     this is the only place the template can raise it.

     Write the model formula LITERALLY at the runner's call site. Not in a
     variable, not via as.formula(), not via reformulate(). hzr_bootstrap() and
     hzr_stepwise() rewrite the stored formula per replicate or per step, and a
     symbol standing where the formula should be does not survive that rewrite:
     the refit errors, the error is caught, the step reports nothing accepted,
     and the screen halts having selected nothing -- with no warning and
     n_failed = 0. The health table below is what catches it.

     A free parameter must also VARY across resamples. A bootstrap built on the
     VECTOR interface returns the original fit every replicate, with
     n_success = 500, n_failed = 0 and no warning. sd_free_base_param is the
     tell, and the stopifnot() below refuses to continue on it. -->

```{r}
#| label: health
reps       <- bag$boot$replicates
per_rep    <- tapply(reps$parameter, reps$replicate,
                     function(p) length(setdiff(p, bag$base_params)))
n_selected <- length(setdiff(unique(reps$parameter), bag$base_params))

knitr::kable(data.frame(
  n_success                 = bag$boot$n_success,
  n_failed                  = bag$boot$n_failed,
  sd_free_base_param        = signif(bag$free_sd, 4),
  distinct_covariates       = n_selected,
  replicates_selecting_none = sum(per_rep == 0),
  median_covars_per_rep     = stats::median(per_rep)
), caption = "Health of the screen. Read this before any frequency below.")

# Hard stop, not a warning. Both conditions describe a screen whose output is an
# artefact, and a warning here would scroll past in a log while the .html went
# on to be read as a result.
stopifnot(bag$free_sd > 0, n_selected > 0)
```

`sd_free_base_param` is the spread of a free base parameter across replicates.
**Zero there means the replicates all returned the original fit** — the refit
never saw resampled data — and every frequency below would be an artefact.

`replicates_selecting_none` is the companion check: replicates that ran but
retained nothing. A few are ordinary. All of them means the screen is broken
rather than strict, and the two failures look identical in a summary table.

## What a selection frequency is, and what it is not

A selection frequency is an estimate, not a count of something fixed, and it
carries Monte-Carlo error of roughly `sqrt(p(1-p)/n_boot)`. At `p = 0.5` over
500 replicates that is about **2.2 percentage points**, so a variable sitting
within a few points of the retention threshold can fall on either side of it on
resampling noise alone. The table below reports that error per variable rather
than leaving it to be assumed away.

Two consequences worth stating wherever these numbers are used:

- **Agreement on the retention decision matters more than agreement on the
  frequency.** Which side of the threshold a variable falls on is the decision
  this job exists to support.
- **A near-threshold variable is not a weak risk factor.** It is a variable
  whose selection is unstable, which is a different claim.

<!-- EDIT: if your study's entry/stay criteria disagree with anything it has
     published, say so here rather than leaving it to be rediscovered. On the
     study this template came from the `.sas` used sle = 0.07 while the paper
     stated a retention criterion of .05, and the run follows the `.sas`
     because that is the artefact that survives. A disagreement like that
     belongs in the report, not in someone's memory. Delete this note if your
     study has none. -->

```{r}
#| label: frequencies
# EDIT: the retention threshold this study reports against, and the frequency
# below which a variable is omitted from the concept table and the figure.
#
# Read RETAIN_PCT off your own `%hazboot` call and the reporting convention it
# was run under. The two studies this template was extracted from disagree on
# every knob of the screen -- 500 replicates at sle = 0.07 / sls = 0.05 against
# 1000 at 0.12 / 0.1 -- so there is no default here to fall back on, and the
# macro's own documented defaults are not what either production job ran.
#
# The criteria themselves are NOT set here. They are properties of the run that
# already happened, read off `bag` into the provenance table above; declaring
# them again in the report would let it assert a screen that was never run.
RETAIN_PCT    <- 50
DISPLAY_FLOOR <- 10

summ <- bag$boot$summary
sel  <- summ[!summ$parameter %in% bag$base_params, , drop = FALSE]

# The base parameters are excluded above and reported separately below. They
# appear at 100% by construction -- they are the model, not something the
# screen chose -- and a table that mixes them in invites reading "100%" as
# reliability.
sel$phase    <- sub("[.].*$", "", sel$parameter)
sel$variable <- sub("^[^.]*[.]", "", sel$parameter)

# Monte-Carlo standard error of the frequency itself.
p         <- sel$pct / 100
sel$se_pp <- 100 * sqrt(p * (1 - p) / bag$n_boot)
sel$retained <- sel$pct >= RETAIN_PCT

# Within two standard errors of the threshold: the DECISION is unstable, not
# the variable.
sel$near_threshold <- abs(sel$pct - RETAIN_PCT) < 2 * sel$se_pp

sel <- sel[order(sel$phase, -sel$pct), ]

freq_cols <- c("phase", "variable", "n", "pct", "se_pp", "retained", "near_threshold")
knitr::kable(sel[, freq_cols], row.names = FALSE, digits = 2,
             caption = paste0("Selection frequency over ", bag$n_boot,
                              " replicates. se_pp is the Monte-Carlo standard error in ",
                              "percentage points; it degenerates to 0 at 0% and 100%, ",
                              "where it understates the uncertainty rather than ",
                              "reporting none."))
```

```{r}
#| label: retained
# Near-threshold is counted over ALL variables, not just the retained ones. A
# variable sitting just BELOW the line is exactly as decision-unstable as one
# just above it, and counting only the retained hides it: on the study this
# template came from, two candidates at 48.8% and 46.4% were both within two
# standard errors of 50 and appeared nowhere in this summary.
by_phase <- lapply(split(sel, sel$phase), function(x) {
  data.frame(phase      = x$phase[1],
             n_retained = sum(x$retained),
             near_above = sum(x$retained & x$near_threshold),
             near_below = sum(!x$retained & x$near_threshold),
             stringsAsFactors = FALSE)
})
knitr::kable(do.call(rbind, by_phase), row.names = FALSE,
             caption = paste0("Variables retained at >= ", RETAIN_PCT,
                              "%. near_above / near_below count variables within two ",
                              "standard errors of the threshold on each side: their ",
                              "retention decision is unstable, not their effect."))
```

## Concept-level frequencies

A per-variable frequency answers "how often was *this form* selected". The
question a paper asks is "how often was *this concept* selected", and the two
differ whenever the pool offers several transformations of one variable. The gap
runs both ways: competing forms can split replicates between them, so the
concept reads weaker than it is, or two forms can both clear the threshold, so
one finding is reported twice.

⚠️ **Do not prune competing transformations from the pool before screening.**
Screen every form; group only when reading, which is what this section does.

Measured, on the study this template came from: of the 57 forms pruning removed,
**16 correlated at |r| < 0.9** with the form kept and five below 0.5. `in_zexp`
**is** `1/zexp` -- r = 0.9997 against the reciprocal -- yet correlates with
`zexp` at only **-0.195**, because `zexp` spans 0.038 to 151.9. Over a 4000-fold
range a value and its reciprocal are different information. That study's
published model uses `zexp` and `in_zexp` in the same phase, **both
significant**, a two-parameter flexible form pruning forbids.

The naming convention tells you two variables are RELATED. Only the data tells
you whether they are REDUNDANT.

```{r}
#| label: concept-map
#| code-fold: true
# The map the SCREEN used, if it used one at all. Distinct from the read-time
# grouping below, and both are worth having: this says what the screen was
# offered, that says how to read what it returned. Neither supersedes the other.
if (is.null(bag$concept_map)) {
  knitr::kable(data.frame(result = "pool not pruned: every form screened separately"))
} else {
  cm  <- bag$concept_map
  cm  <- cm[cm$phase == cm$phase[1], , drop = FALSE]
  grp <- cm[cm$concept %in% cm$concept[duplicated(cm$concept)], , drop = FALSE]
  knitr::kable(grp[order(grp$concept, grp$variable), c("concept", "variable")],
               row.names = FALSE,
               caption = "Forms the screen itself collapsed. An empty table here is the good case.")
}
```

```{r}
#| label: concept-frequencies
# EDIT: forms this study knows are one concept but no affix rule can reach.
#
# `vars.sas` truncates the stem when naming a derivative (`effic` -> `ln_effi`,
# `area_int` -> `in_arin`), and some forms carry no affix at all (`agee`,
# `sizee`). concept_of() handles truncation automatically where the stem
# prefixes exactly one plain form; the rest are declared here, because whether
# `agee` is a form of `age` is a CLINICAL judgement, not a naming rule.
#
# Leave it empty to group only what the affix and prefix rules can prove.
CONCEPT_ALIASES <- character(0)

# EDIT: the affix vocabulary, if your study's variable names do not follow
# vars.sas conventions. POOL_AFFIXES carries ln_, in_, in2, _pr and a trailing
# 2, the order in which they strip, and the deliberate refusal to reduce agee
# to age. Those are facts about this institution's names, not about statistics,
# which is why they are data you can replace rather than logic you cannot.
concept_for <- function(v) {
  # Aliases are filtered to the pool first. concept_map() refuses an alias
  # pointing outside it, and a phase whose screen never saw `age` would
  # otherwise error on an alias written for the phase that did.
  al <- CONCEPT_ALIASES[unname(CONCEPT_ALIASES) %in% v]
  concept_map(v, aliases = al, affixes = POOL_AFFIXES,
              min_stem = POOL_MIN_STEM, plain_suffix = POOL_PLAIN_SUFFIX)
}

# Per-replicate, not per-summary. A UNION across a concept's forms cannot be
# recovered from marginal percentages -- two forms at 30% each are anywhere
# between 30% and 60% of replicates depending on how often they co-occur -- so
# this reads the replicate table and counts distinct replicates.
rep_cov <- reps[!reps$parameter %in% bag$base_params, , drop = FALSE]
rep_cov$phase    <- sub("[.].*$", "", rep_cov$parameter)
rep_cov$variable <- sub("^[^.]*[.]", "", rep_cov$parameter)

cf <- do.call(rbind, lapply(split(sel, sel$phase), function(x) {
  ph <- x$phase[1]
  cm <- concept_for(x$variable)
  x$concept <- cm$concept[match(x$variable, cm$variable)]
  r <- rep_cov[rep_cov$phase == ph, , drop = FALSE]
  r$concept <- x$concept[match(r$variable, x$variable)]
  do.call(rbind, lapply(split(x, x$concept), function(g) {
    hit <- unique(r$replicate[!is.na(r$concept) & r$concept == g$concept[1]])
    data.frame(phase            = ph,
               concept          = g$concept[1],
               n_forms          = nrow(g),
               union_pct        = 100 * length(hit) / bag$n_boot,
               best_form_pct    = max(g$pct),
               n_forms_retained = sum(g$retained),
               forms            = paste(sort(g$variable), collapse = ", "),
               stringsAsFactors = FALSE)
  }))
}))

cf$spread   <- cf$union_pct - cf$best_form_pct
cf$retained <- cf$union_pct >= RETAIN_PCT
cf <- cf[order(cf$phase, -cf$union_pct), ]

shown_cf <- cf[cf$union_pct >= DISPLAY_FLOOR, , drop = FALSE]
cf_cols  <- c("phase", "concept", "n_forms", "union_pct", "best_form_pct",
              "spread", "retained", "forms")
knitr::kable(shown_cf[, cf_cols], row.names = FALSE, digits = 1,
             caption = paste0("Selection frequency per CONCEPT. union_pct is the percentage of ",
                              "replicates selecting the concept in any form; best_form_pct is its ",
                              "strongest single form. spread is how much the per-variable table ",
                              "understates the concept. Concepts below ", DISPLAY_FLOOR,
                              "% are omitted here and present in the per-variable table above."))
```

```{r}
#| label: concept-counts
counts <- lapply(split(cf, cf$phase), function(x) {
  data.frame(phase              = x$phase[1],
             concepts_retained  = sum(x$retained),
             variables_retained = sum(sel$retained[sel$phase == x$phase[1]]),
             # Concepts the per-variable table reports more than once. Uses
             # RETAINED forms, so this column actually accounts for the gap
             # between the two counts to its left.
             double_counted     = sum(x$n_forms_retained > 1),
             stringsAsFactors   = FALSE)
})
knitr::kable(do.call(rbind, counts), row.names = FALSE,
             caption = paste0("Retained at >= ", RETAIN_PCT,
                              "%, counted both ways. Where these disagree, the ",
                              "per-variable count is reporting one concept more ",
                              "than once."))
```

::: {.callout-note}
## Two things to check before quoting either number

⚠️ **A selection frequency is conditional on the candidate pool**, and that is
not a caveat this table can remove. On the study this template came from,
re-running with 226 candidates instead of 189 moved one variable from 26.6% to
95.2% and another from 93.8% to 20.2% -- same data, same seeds. Seven of nine
retained concepts were common to both runs; two differed each way. Quote a
reliability figure with the pool it came from, or it is not reproducible.

**A concept whose `spread` is large was being understated by the per-variable
table.** A concept whose `n_forms` exceeds one and which is `retained` was being
reported more than once. Both are visible above; neither is visible in a
per-variable frequency alone.
:::

```{r}
#| label: base-params
# Shown separately, and shown at all, because their absence would be the more
# suspicious thing: these are the parameters every replicate carries.
knitr::kable(summ[summ$parameter %in% bag$base_params,
                  c("parameter", "n", "pct", "mean", "sd")],
             row.names = FALSE, digits = 4,
             caption = "Base model parameters, present in every replicate by construction")
```

## Clusters: the two views a name cannot give

Concept grouping above is by NAME, and a naming convention only knows what its
affix rules can reach. Two other groupings matter and neither is derivable from
a variable's name. The SAS job this replaces ran `%cluster` for both, once per
named cluster per phase.

**Declared clusters** group by CLINICAL judgement: `Renal` is creatinine, GFR
and dialysis, which share no stem and never will. **Collinear pairs** group by
the DATA: variables that move together whatever they are called. A name tells
you two variables are related, a declaration tells you a clinician thinks they
are one thing, and a correlation tells you the pool cannot tell them apart.

```{r}
#| label: cluster-matrix
# boot_clusters() wants a replicate x term matrix, and the pooled object does
# not carry one -- boot_pool_chunks() returns replicates and summary only. It
# is rebuilt here from the replicate table, whose replicate/parameter/estimate
# columns that function validates before it will pool anything, so this is the
# one shape guaranteed to be present on both the single-run and chunked paths.
#
# One matrix per phase, with bare variable names, because a cluster is a
# statement about variables and the SAS job ran each against every phase.
phase_matrix <- function(ph, members) {
  r    <- rep_cov[rep_cov$phase == ph, , drop = FALSE]
  vars <- sort(unique(c(r$variable, members)))
  m    <- matrix(NA_real_, nrow = bag$n_boot, ncol = length(vars),
                 dimnames = list(NULL, vars))
  m[cbind(r$replicate, match(r$variable, vars))] <- r$estimate
  m
}
```

```{r}
#| label: clusters
# EDIT: your clusters, one per concept worth reporting as a unit. Which
# concepts those are is a statement about YOUR candidate pool, so there is no
# default and an empty list is a legitimate answer.
#
# The SAS job this template came from named eight -- Age, Size, BMI, race, GFR,
# iv_opyrs, Renal, blrbn_pr -- and ran each against both the early and constant
# phases. Names are the BARE variable names, without the phase prefix.
CLUSTERS <- list()

if (!length(CLUSTERS)) {
  knitr::kable(data.frame(result = "no clusters declared: see the EDIT: above"))
} else {
  members <- unique(unlist(CLUSTERS, use.names = FALSE))

  # Declared members that no replicate ever selected are carried into the
  # matrix as all-NA columns rather than dropped. boot_clusters() refuses a
  # cluster naming a term it cannot see, and dropping them silently would
  # report the union over whichever members happened to be selected -- a
  # smaller group than the one declared, under the declared group's name.
  # Carried, they contribute zero and are named below, so a typo shows up as a
  # member that was never selected instead of disappearing.
  cl <- do.call(rbind, lapply(sort(unique(rep_cov$phase)), function(ph) {
    m   <- phase_matrix(ph, members)
    out <- boot_clusters(m, CLUSTERS)
    out$phase <- ph
    out[, c("phase", "cluster", "n_any", "pct_any", "members")]
  }))
  knitr::kable(cl, row.names = FALSE, digits = 1,
               caption = paste0("Union selection over declared clusters. pct_any is the ",
                                "percentage of replicates retaining ANY member, which is ",
                                "the number a per-variable table cannot show."))

  never <- setdiff(members, unique(rep_cov$variable))
  if (length(never)) {
    knitr::kable(data.frame(never_selected = never), row.names = FALSE,
                 caption = paste0("Declared cluster members no replicate ever retained. ",
                                  "Contributing zero above. Check these are spelled as ",
                                  "the pool spells them."))
  }
}
```

```{r}
#| label: collinear
# EDIT: the correlation above which two candidates are reported as
# indistinguishable to the screen. 0.99 finds near-duplicates -- a variable and
# its rescaling, a dummy and its complement. Lower it to see the pool's
# structure rather than only its accidents.
COLLINEAR_R <- 0.99

# Read to correlate the pool, NOT to re-run anything. The screen already
# happened; this describes the data it ran against.
d <- read_built()

# A build that moved since the screen ran would have these correlations
# describing one dataset and every frequency above describing another, with
# nothing in the report to say so. n_rows is the cheapest check that catches
# it, and it is a warning rather than a stop because the frequencies above
# remain valid -- it is only this section that goes stale.
if (!is.null(bag$n_rows) && !identical(as.integer(nrow(d)), as.integer(bag$n_rows))) {
  warning("The built dataset now has ", nrow(d), " rows; this screen ran against ",
          bag$n_rows, ". The correlations below describe TODAY'S data, not the ",
          "data the frequencies above were computed on.", call. = FALSE)
}

pool  <- sort(unique(rep_cov$variable))
pairs <- pool_collinear_pairs(d, pool = pool, threshold = COLLINEAR_R)

if (!nrow(pairs)) {
  knitr::kable(data.frame(result = paste0("no pair of screened candidates correlates at |r| >= ",
                                          COLLINEAR_R)))
} else {
  knitr::kable(pairs, row.names = FALSE, digits = 4,
               caption = paste0("Screened candidates correlating at |r| >= ", COLLINEAR_R,
                                ". complement marks a pair summing to 1 -- a dummy and its ",
                                "negation, which is a coding artefact rather than a finding."))
}
```

::: {.callout-note}
## What this section is not

⚠️ **A collinear pair is not a licence to prune.** It says the two are
indistinguishable *in this cohort at this threshold*, which is a statement about
the pool, not about the concepts. The no-pruning argument above stands: the
same measurement that found 57 prunable forms found 16 of them correlating
below 0.9.

`pool_collinear_pairs()` sees only NUMERIC, non-constant columns. Factor-coded
binaries are invisible to it for the same reason they were invisible to the
screen -- check the coercion table above before reading an empty result as an
absence of collinearity.
:::
