---
title: "Multivariable hazard model"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in:
# rendered as a standalone document the project's format block never applies,
# which is how server-rendered reports end up with sibling _files/ trees instead
# of being self-contained.
#
# NOTE: output-dir is a PROJECT setting and cannot be set here, so the .html
# lands next to its source. With embed-resources that is a single standalone
# file, which is the property that actually matters when a report is sent on.
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the SAS job this replaces, and say what its `selection`
     statement screened: which phases carried candidates, and at what
     entry/stay levels. -->

Replaces `analyses/<job>.sas`: describe its `PROC HAZARD … selection` call
here — which covariates were offered to which phase, and at what entry and stay
levels.

An `hm` job builds the risk-factor model **on top of** the shape parameters the
`hz` job fitted. The shapes are not re-derived here; they are read. Patient-level
predictions from this model are an `hs` job and the bootstrap of this screen is
`bh`; neither belongs here, and **nor does SAS parity** — a parity job borrows
the ordinal of the job it checks and lives in `parity/`.

```{r}
#| label: setup
# This file runs either from the directory it was scaffolded into or from the
# project root. Resolve which, so that no path in this document needs editing:
# forgetting one of several ../ paths is the instantiation mistake that would
# otherwise be easy to make and quiet when made.
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(TemporalHazard)
  library(hvtiRutilities)
})
```

```{r}
#| label: edit-guard
#| results: asis
# The markers in this file name work a study author still has to do, and
# README.md says a job that still contains one has not been finished. This
# chunk is what makes that TRUE rather than merely stated.
#
# Without it an unedited job renders green over a meaningless analysis. In THIS
# file the danger is sharper than a wrong number: the unedited entry and stay
# levels are `hzr_stepwise()`'s defaults, which are LOOSER than the SAS job's,
# so an unedited screen admits variables the job being reproduced rejected --
# and reports them as risk factors, in a table that looks exactly like a
# result.
#
# knitr::current_input() is NULL outside a render -- a study author stepping
# through chunks in RStudio -- and there is no file to scan then, so this is a
# no-op in that case. Same handling as the `set` guard below, for the same
# reason.
#
# The token is BUILT, not written literally, and that is not stylistic. Quarto
# knits through an intermediate and current_input() returns THAT file, so the
# scan reads a copy of this chunk along with everything else: a literal
# grep("<token>", .src) here matches its own source line, and the guard then
# fires on every render, finished or not. Measured, not theorised -- a probe
# found three markers in a file containing two. A guard that cries wolf on a
# finished job is a guard that gets deleted, which returns us to issue #27.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    # Report the marker TEXT. Line numbers here are the intermediate's, not
    # this .qmd's, so they need not match what the author sees in an editor.
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it.\nTo render a partly-worked draft anyway, set ",
      "HVTI_TEMPLATE_DRAFT to 1, true or yes -- the report then carries a ",
      "draft banner. Any other value, including 0, leaves this guard strict."
    )
    # Only the documented values enable draft mode. nzchar() alone would treat
    # HVTI_TEMPLATE_DRAFT=0 as ON, silently downgrading this stop() to a
    # warning for someone whose shell profile sets it to 0 meaning OFF -- a
    # switch whose enablement does not mean what it says, which is the same
    # defect this guard exists to catch. An unrecognised value falls through to
    # strict on purpose: the author then sees the stop and re-reads the docs,
    # which is the safe direction to be wrong in.
    if (tolower(Sys.getenv("HVTI_TEMPLATE_DRAFT")) %in% c("1", "true", "yes")) {
      # The banner is not optional. A draft render that looks like a finished
      # one is the same defect with an extra step: the warning scrolls past in
      # a log, while the .html is the artifact that gets sent to someone.
      warning(.msg, call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Rendered with `HVTI_TEMPLATE_DRAFT` set. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by new_job() from the endpoint/type it
     was called with, which are also the filename's first two fields. They
     name the SET this job belongs to, and every artifact path in the
     document is computed from them, so no output path below needs editing.

     Do NOT hand-edit these declarations: the render-time guard below checks
     them against this file's own name and stops if they disagree. To put
     this job in a different set, call new_job() again with the new
     endpoint/type rather than editing this copy in place.

     A set is keyed on (endpoint, analysis type), not on the endpoint alone.
     One endpoint is analysed by several methods -- parametric hazard, random
     forest survival, competing risks -- and those chains share their upstream.
     Two sets keyed on `dead_pa` alone would write their life tables to the
     same file. -->

```{r}
#| label: set
ENDPOINT <- "dead_pa"
TYPE     <- "hz"

# `new_job()` writes ENDPOINT/TYPE from the same values it put in this file's
# name, but a hand-edited declaration can drift from it afterward. set_path()
# below resolves from the declarations, not the filename, so a drifted
# declaration would silently write into ANOTHER set's artifact directory --
# exactly the collision the (endpoint, type) key exists to prevent, re-entered
# through the body instead of the name.
#
# knitr::current_input() is NULL outside a render (a study author running
# chunks interactively in RStudio), and there is no filename to check against
# yet, so this is a no-op in that case rather than a spurious error.
.current <- knitr::current_input()
if (!is.null(.current)) {
  # Quarto knits through an intermediate, so `knitr::current_input()` returns
  # `<endpoint>-<type>-<ordinal>-<prefix>.rmarkdown` here rather than the
  # `.qmd` this was scaffolded as. Strip whatever extension is actually present
  # rather than hard-coding one, so this doesn't depend on a build-tool detail
  # staying the same.
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint, "', type '",
         .name_type, "'), but declares ENDPOINT = \"", ENDPOINT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

# Resolve a path inside this set's artifact directory. `kind` is the artifact
# folder -- "estimates" for serialized results, "graphs" for figures. The set
# directory sits one layer under the kind and never two, which is the whole
# layout rule.
#
# Created on first use rather than up front, so a job that writes nothing leaves
# no empty directories behind.
set_path <- function(kind, file) {
  d <- file.path(.root, kind, paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}

# `hm` reads the `hz` fit through this and writes its own model back, both by
# set, so a model can never be filed against a different set than the shapes it
# was built on. The write is in the `save` chunk at the foot of this file.
```
## Cohort

<!-- EDIT: this chunk, choosing ONE of the two shapes below.

     The distinction is whether the rows this job analyses are the whole study
     or a subset of it, and it is not cosmetic. `_study.yml` records the STUDY
     cohort, derived from the whole built dataset. A job that analyses a
     filtered subset has a different N, and `assert_cohort()` will gate the
     wrong number — it will pass while the job runs on a cohort nobody checked.

     Shape A — the job analyses the whole study cohort. Use the package data
     contract, and the gate is `assert_cohort()`.

     Shape B — the job analyses a filtered subset (an availability flag, a
     procedure subgroup). `read_built()` still reads the data, but the filter
     and the gate are the JOB's, and the expected counts come from the SAS
     reference rather than from `_study.yml`. -->

```{r}
#| label: cohort
# ---- Shape A: whole-study cohort ------------------------------------------
d <- read_built()
assert_cohort(d)
cc <- cohort_counts(d)

# ---- Shape B: filtered job cohort -----------------------------------------
# EDIT: delete Shape A above and uncomment this, supplying the filter and the
# counts the SAS reference reports. Do NOT reuse assert_cohort(): it checks
# _study.yml, which describes the whole study and not this job.
#
# d <- read_built()
# d <- d[!is.na(d$<flag>) & d$<flag> == 1, , drop = FALSE]
# EXPECTED <- list(n = <N>, n_events = <E>, n_censored = <C>)
# cc <- list(n          = nrow(d),
#            n_events   = sum(d[[STATUS]] == 1, na.rm = TRUE),
#            n_censored = nrow(d) - sum(d[[STATUS]] == 1, na.rm = TRUE))
# if (!identical(lapply(cc, as.integer), lapply(EXPECTED, as.integer))) {
#   stop("Cohort gate FAILED: expected ", paste(unlist(EXPECTED), collapse = "/"),
#        ", got ", paste(unlist(cc), collapse = "/"),
#        ". The analysis must not run on an unreconciled cohort.", call. = FALSE)
# }

knitr::kable(data.frame(
  quantity = c("n_analysable", "n_events", "n_censored"),
  n        = c(cc$n, cc$n_events, cc$n_censored)
))
```

A rendered report is itself evidence the gate passed. Report the counts anyway —
a gate that passes silently tells the reader nothing about what it passed on.

**An event flag that is 0/1 in SAS may arrive as `logical`**, because
`read_clinical_data()` infers binary columns as logical while `Surv()` requires
a numeric status. The `as.numeric()` coercions below are not decoration — and
see the covariate audit above for the same problem on the covariate side, where
it changes the model rather than erroring.
```{r}
#| label: numderiv-gate
# numDeriv is a Suggests of TemporalHazard, not a hard dependency. Without it,
# interval- or left-censored multiphase fits SILENTLY produce no standard
# errors: the fit succeeds, `converged` is TRUE, and vcov() is simply absent.
# A model with forty covariates and no standard errors is not a smaller result,
# it is a different one -- so this is a hard stop.
if (!requireNamespace("numDeriv", quietly = TRUE)) {
  stop("numDeriv is not installed. Interval-censored multiphase fits will ",
       "produce no standard errors, and will not tell you so. Install it ",
       "before reading anything below.", call. = FALSE)
}
```

## Covariates

The covariate blocks are **read out of the SAS job**, not transcribed. A
transcribed list drifts from the job it claims to reproduce and nothing catches
it. Only the names are taken: the `name=value` coefficients in a block are the
*other* study's answers and must not be inherited by a copy of it.

```{r}
#| label: vars
# EDIT: your time, event and interval-censoring variables -- SAS `time`,
# `event`, and the flag named by `icensor X=`.
TIME   <- "iu_dead"
EVENT  <- "idead"

# EDIT: the SAS job to read the candidate blocks from, and the macro inside it
# that holds the model. Naming the MACRO matters: a job commonly defines
# several, and they do not agree.
#
# ⚠️ A macro can be defined and never invoked. One real job carries a `%final`
# whose variable list is another study's variables entirely, pasted in and left
# there; the `.lst` shows only the macro that ran. Read the `.sas` and check
# which macro is actually called before naming it here.
SAS_JOB   <- c("analyses", "hm.dead.sas")
SAS_MACRO <- "^\\s*%macro\\s+model\\s*;"
```

```{r}
#| label: covariates
sas_lines <- readLines(do.call(sas_path, as.list(SAS_JOB)), warn = FALSE)

# EDIT: your phase-block markers. A SAS hazard model offers candidates per
# phase, and the blocks need not be identical.
COVARIATES <- list(
  early = sas_variable_block(sas_lines, "^\\s*early\\s*$", after = SAS_MACRO,
                             what = "early covariate block"),
  late  = sas_variable_block(sas_lines, "^\\s*late\\s*$",  after = SAS_MACRO,
                             what = "late covariate block")
)

knitr::kable(data.frame(
  phase     = names(COVARIATES),
  n         = vapply(COVARIATES, length, integer(1)),
  variables = vapply(COVARIATES, paste, character(1), collapse = ", "),
  row.names = NULL
))
```

### What these columns are, before they become covariates

`vars.sas` is commonly called with `%vars(missing=1, impute=1)`: it
**mean-imputes** missing values and adds a paired `ms_*` indicator, which is
why `ms_*` variables are themselves covariates in these models. So a 0/1
clinical variable arrives carrying three distinct values — 0, 1, and the cohort
mean. One real column's third value is `0.714`, the prevalence of hypertension,
not any patient's status.

In SAS these are numeric and enter **linearly**. Read into R they arrive as
**factors**, and a factor in a model formula is dummy-coded — which makes "this
value was imputed" its own category with its own coefficient. **That is a
different model, with a different parameter count, and nothing in the output
would say so.**

```{r}
#| label: audit
audit <- covariate_audit(d, unlist(COVARIATES))
knitr::kable(audit, row.names = FALSE,
             caption = "Covariate storage and the conversion applied")

# Fail rather than fit a model whose covariates are not what the job specifies.
if (any(grepl("^ERROR", audit$action))) {
  stop("covariate_audit() reports ERROR for: ",
       paste(audit$variable[grepl("^ERROR", audit$action)], collapse = ", "),
       ". Resolve these before fitting.", call. = FALSE)
}

dd <- covariates_to_numeric(d, unlist(COVARIATES))
```

**A factor left as a factor is dummy-coded on purpose** — the right outcome for
a genuinely categorical variable and the wrong one for a mean-imputed binary.
The `action` column is where that decision is visible; check it before reading
any coefficient.

```{r}
#| label: collinear
# What concept grouping structurally CANNOT catch. Grouping works on
# transformations; two candidates that are the same information under unrelated
# names share no affix relationship and are invisible to it.
#
# `male` and `female` are the case that got through a real screen: exact
# complements, both offered, and 101 of 500 replicates selected BOTH. A design
# holding both is singular, and nothing in the output said so.
collinear <- pool_collinear_pairs(dd, unlist(COVARIATES))
if (nrow(collinear)) {
  knitr::kable(collinear, row.names = FALSE, digits = 4,
               caption = "Candidate pairs at |r| >= 0.99")
} else {
  knitr::kable(data.frame(result = "no candidate pair reaches |r| = 0.99"))
}
```

An exact `complement` is a coding duplicate and one of the pair should simply
not be a candidate. A high correlation between genuinely different measurements
is a judgement call about what the model can identify — make it here, not after
reading a coefficient.

## Phases

```{r}
#| label: phases
# The shape values come from the companion hz job, NOT from a literal. This is
# what "builds on the HZ fit" means concretely: SAS holds the shapes with
# `fixthalf fixnu fixm fixeta` while it screens covariates.
hz_art <- readRDS(set_path("estimates", "hz.rds"))
hz_fit <- hz_art$deterministic

# EDIT: the shape/scale parameter names your phase specification produces.
# Named explicitly so that anything else in coef() is a covariate -- including
# a dummy column, which would otherwise be silently counted as a shape.
SHAPE_PARAMS <- c("early.log_mu", "early.log_t_half", "early.nu", "early.m",
                  "late.log_mu", "late.log_tau", "late.gamma", "late.alpha",
                  "late.eta")

# `fixed` is what makes stage 1 stage 1: shapes held, covariates free.
make_phases <- function(hold_shapes) {
  p <- hz_art$phases
  if (!hold_shapes) return(p)
  lapply(p, function(ph) {
    ph$fixed <- unique(c(ph$fixed, setdiff(names(ph$par), ph$fixed)))
    ph
  })
}
knitr::kable(data.frame(
  parameter = names(coef(hz_fit)),
  from_hz   = unname(coef(hz_fit))
), digits = 6, caption = "Shape parameters inherited from the hz job")
```

## Fitting

### Two stages, and why it is not one

Fitting the shapes and forty covariates together **from a neutral start lands in
the wrong basin, and reports `converged = TRUE` while doing it.** The stages
below mirror what the SAS job itself does: it fixes the shapes for selection and
frees them only for the final fit.

Stage 1 holds the shapes at the `hz` fit and fits the covariates. Stage 2 frees
the shapes, starting from stage 1. **The comparison is reported, not assumed.**

`n_starts > 1` perturbs its restarts randomly, so the fit is not reproducible
without a seed. The RNG state is saved and restored, so seeding this document
does not change random behaviour downstream of it.

```{r}
#| label: fit
old_seed <- if (exists(".Random.seed", .GlobalEnv)) .GlobalEnv$.Random.seed else NULL
set.seed(20260827)   # any fixed value; it must not change between renders

# EDIT: your covariate formula, and conserve = FALSE if the SAS final model
# carries `noconserve`.
rhs <- paste(unique(unlist(COVARIATES)), collapse = " + ")
form <- stats::as.formula(paste0("Surv(", TIME, ", ", EVENT, ") ~ ", rhs))

ctl1 <- list(n_starts = 5, maxit = 2000, conserve = FALSE)
ctl2 <- list(n_starts = 1, maxit = 5000, conserve = FALSE)

stage1 <- suppressWarnings(
  hazard(form,
         data = dd,
         dist = "multiphase",
         phases = make_phases(TRUE),
         fit = TRUE,
         control = ctl1)
)

stage2 <- suppressWarnings(
  hazard(form,
         data = dd,
         dist = "multiphase",
         phases = make_phases(FALSE),
         theta = stage1$fit$theta,
         fit = TRUE,
         control = ctl2)
)

if (!is.null(old_seed)) .GlobalEnv$.Random.seed <- old_seed
```

```{r}
#| label: stages
se_of <- function(f) {
  v  <- vcov(f)
  dg <- if (is.matrix(v)) diag(v) else rep(NA_real_, length(coef(f)))
  # A non-positive-definite Hessian can put NEGATIVE numbers on the diagonal.
  # sqrt() of one is NaN, which prints as a blank cell and reads as "not
  # applicable" rather than "this variance is impossible". Make it NA, and
  # count them.
  stats::setNames(sqrt(ifelse(is.finite(dg) & dg >= 0, dg, NA_real_)),
                  names(coef(f)))
}
free_of <- function(f) names(coef(f))[!as.logical(f$fit$fixed_mask)]

stages <- data.frame(
  stage       = c("1: shapes fixed at hz", "2: shapes freed from stage 1"),
  converged   = c(stage1$fit$converged, stage2$fit$converged),
  objective   = c(stage1$fit$objective, stage2$fit$objective),
  # CHARACTER, not numeric: rcond is often ~1e-6 here, and kable's `digits`
  # rounds a numeric column, so two different condition numbers both print as
  # "1e-06" and the comparison this table exists for silently disappears.
  rcond       = format(c(stage1$fit$rcond, stage2$fit$rcond),
                       digits = 3, scientific = TRUE),
  pd          = c(stage1$fit$pd, stage2$fit$pd),
  free_params = c(length(free_of(stage1)), length(free_of(stage2))),
  free_without_se = c(sum(is.na(se_of(stage1)[free_of(stage1)])),
                      sum(is.na(se_of(stage2)[free_of(stage2)])))
)
knitr::kable(stages, digits = 6)

# Named explicitly rather than left to stopifnot(), whose message is the failed
# expression. In a rendered report the reader is not looking at this source,
# and "isTRUE(stage1$fit$converged) is not TRUE" tells them nothing they can
# act on.
failed <- stages$stage[!vapply(list(stage1, stage2),
                               function(f) isTRUE(f$fit$converged), logical(1))]
if (length(failed)) {
  stop("These stages did not converge: ", paste(failed, collapse = "; "),
       ". Every estimate below would be read off a fit that never settled. ",
       "Check the shape parameters inherited from the hz job, and whether the ",
       "covariate set is identifiable on this cohort.", call. = FALSE)
}

cov_names <- setdiff(names(coef(stage1)), SHAPE_PARAMS)
delta <- abs(coef(stage1)[cov_names] - coef(stage2)[cov_names])
knitr::kable(data.frame(
  objective_gain_from_freeing_shapes = stage2$fit$objective - stage1$fit$objective,
  max_abs_covariate_change           = max(delta),
  changed_by_more_than_0.01          = sum(delta > 0.01)
), digits = 6)
```

**Read the two tables together before choosing which fit to report.** Freeing
the shapes can gain almost nothing in objective, move no covariate materially,
and still turn `pd` from `TRUE` to `FALSE` while leaving shape parameters with
no standard error — in which case the shapes were already at their optimum and
freeing them buys nothing and costs the covariance.

Reporting stage 1 in that case is a **deliberate divergence** from the SAS
`%final`, which frees them. It is legitimate when the table above is the
evidence for it. If `free_without_se` is 0 for stage 2 and `pd` is `TRUE`,
prefer stage 2.

```{r}
#| label: reported
reported <- stage1   # EDIT: stage2 if the table above favours it
```

## Estimates

```{r}
#| label: estimates
cf <- coef(reported)
se <- se_of(reported)
cov_names <- setdiff(names(cf), SHAPE_PARAMS)

# The IQR of each covariate AS FITTED -- from dd, after conversion, so it is
# the spread of the numbers that actually entered the model rather than of the
# column as stored.
iqr_of <- function(v) {
  x <- dd[[sub("^[^.]*[.]", "", v)]]
  if (is.null(x) || !is.numeric(x)) return(NA_real_)
  unname(diff(stats::quantile(x, c(0.25, 0.75), na.rm = TRUE)))
}
iqr <- vapply(cov_names, iqr_of, numeric(1))

est <- data.frame(
  parameter = cov_names,
  estimate  = unname(cf[cov_names]),
  se        = unname(se[cov_names]),
  iqr       = unname(iqr),
  # NA, not 1, on a degenerate IQR: exp(0) is 1 and would read as "no effect"
  # rather than "this scaling does not apply here".
  hr_per_iqr = unname(ifelse(is.finite(iqr) & iqr > 0,
                             exp(cf[cov_names] * iqr), NA_real_)),
  row.names = NULL
)
knitr::kable(est[order(-abs(est$estimate)), ], digits = 6, row.names = FALSE)

missing_se <- sum(is.na(est$se))
if (missing_se > 0) {
  knitr::kable(data.frame(
    warning = paste(missing_se, "covariate(s) have NO standard error.",
                    "The Hessian is not positive-definite in those directions.")
  ))
}
```

## Selection

The covariates above were taken as given. The SAS job's `%model` macro derives
them by forward selection, which is a long job and is run from a companion
script rather than inline.

⚠️ **`hzr_stepwise()`'s defaults are not SAS's.** `slentry = 0.3` and
`slstay = 0.2` against a typical `sle = 0.1`, `sls = 0.07`. Accepting the R
defaults runs a **different screen** from the job being reproduced, admits more
variables, and does not error.

⚠️ **The default `criterion = "score"` has a known defect**
([temporal_hazard#130](https://github.com/ehrlinger/temporal_hazard/issues/130)):
it declines the strongest candidates, because the observed information goes
indefinite at `beta = 0`. Consider `"wald"`, and say which you used.

```{r}
#| label: selection
# EDIT: the levels from YOUR .sas `selection` statement, and the path your
# companion runner writes to.
SLENTRY   <- 0.1     # SAS sle=
SLSTAY    <- 0.07    # SAS sls=
MAX_STEPS <- 50L     # SAS maxsteps=
sel_rds   <- set_path("estimates", "hm-selection.rds")

if (!file.exists(sel_rds)) {
  not_run <- data.frame(
    selection = "not run",
    how       = paste("Run the companion selection script, which writes",
                      basename(sel_rds)),
    slentry   = SLENTRY,
    slstay    = SLSTAY,
    max_steps = MAX_STEPS
  )
  knitr::kable(not_run)
} else {
  sel <- readRDS(sel_rds)
  summary_tbl <- data.frame(
    slentry        = sel$slentry,
    slstay         = sel$slstay,
    max_steps      = sel$max_steps,
    steps_taken    = nrow(sel$steps),
    stopped_on_cap = sel$hit_cap
  )
  keep <- intersect(c("step_num", "action", "variable", "phase",
                      "p_value", "logLik", "n_coef"), names(sel$steps))
  # print() BOTH. Only the last expression of a block is auto-printed, so a
  # kable() that is not the final line computes a table and discards it -- the
  # chunk runs, reports no error, and the table is simply absent.
  print(knitr::kable(summary_tbl,
                     caption = "Forward selection, from the companion runner"))
  print(knitr::kable(sel$steps[, keep], row.names = FALSE, digits = 6,
                     caption = "Selection path"))
}
```

```{r}
#| label: selection-monotone
# A forward selection cannot legitimately LOWER the log-likelihood: adding a
# term to a nested model can only improve the maximised likelihood. If it
# drops, the post-entry REFIT landed in a worse optimum than the model it
# started from -- which happens when collinear transformations of one variable
# enter in succession -- and every step after that is built on it.
#
# The screen reports p = 0.000 for those entries either way, so this is the
# only place a reader would find out.
if (file.exists(sel_rds)) {
  sel <- readRDS(sel_rds)
  # A screen that admitted nothing is a real outcome, not a broken run: it says
  # no candidate cleared the entry level. Report that and skip the check, since
  # which.max() on an empty vector is integer(0) and data.frame() then fails
  # with "arguments imply differing number of rows" -- an error about the
  # REPORT, in a job whose actual finding is that nothing entered.
  if (nrow(sel$steps) == 0L) {
    knitr::kable(data.frame(
      steps = 0L,
      note  = paste("No variable entered the model, so there is no path to",
                    "check. Monotonicity skipped.")
    ))
  } else if ("logLik" %in% names(sel$steps)) {
    dropped <- c(FALSE, diff(sel$steps$logLik) < -1e-6)
    mono <- data.frame(
      steps                 = nrow(sel$steps),
      steps_lowering_loglik = sum(dropped),
      best_loglik_at_step   = which.max(sel$steps$logLik),
      loglik_at_last_step   = sel$steps$logLik[nrow(sel$steps)],
      best_loglik           = max(sel$steps$logLik)
    )
    print(knitr::kable(mono, digits = 3,
                       caption = "Is the selection path monotone?"))
    if (any(dropped)) {
      warning("The selection path lowers the log-likelihood at step(s) ",
              paste(which(dropped), collapse = ", "),
              ". Those entries left the model worse than before them; the ",
              "variable list is not a nested improvement and must not be read ",
              "as one.", call. = FALSE)
    }
  }
}
```

```{r}
#| label: selection-crowding
# How much of the step budget went to a concept the model already had.
#
# The monotonicity check above catches the SYMPTOM of collinear forms entering
# in succession (a log-likelihood that drops). This reports the CAUSE, and it
# fires whether or not the likelihood misbehaved. A step budget is shared, so a
# slot spent on a second form of a concept already in the model is a slot no
# other variable could have -- and when the screen also stopped ON the cap, the
# two facts together say the model is budget-limited BY REDUNDANCY.
#
# Grouping is conservative by construction, so this is a FLOOR: a concept
# nobody listed is still counted as separate variables. It understates and
# never overstates.
if (file.exists(sel_rds)) {
  sel <- readRDS(sel_rds)
  al  <- if (is.null(sel$concept_aliases)) character(0) else sel$concept_aliases
  cr  <- selection_crowding(sel$selected, aliases = al)
  redundant <- if (nrow(cr)) sum(cr$n_forms) - nrow(cr) else 0L
  crowding <- data.frame(
    selected         = length(sel$selected),
    crowded_concepts = nrow(cr),
    redundant_slots  = redundant
  )
  print(knitr::kable(crowding,
                     caption = "Concept crowding among the selected covariates"))
  if (nrow(cr)) print(knitr::kable(cr, row.names = FALSE))
}
```

## Calibration

The model's coefficients say which factors matter. **They do not say whether the
model is right.** The SAS job answers that with `%deciles`: group patients by
predicted risk and compare observed against expected within each group.

```{r}
#| label: calibration
# EDIT: the horizon, in the time unit of TIME. SAS: %deciles(..., time=10).
DECILE_TIME <- 10

dec <- hzr_deciles(reported, time = DECILE_TIME, groups = 10)

# print(), NOT knitr::kable(). The print method carries the OVERALL chi-square,
# its degrees of freedom and its p value -- the one number that answers "does
# this model calibrate" -- and kable() renders only the per-group data frame,
# dropping it silently.
print(dec)
```

A group where observed and expected diverge is a region of risk the model gets
wrong, and it is invisible in a coefficient table.

```{r}
#| label: gof
# The parametric fit against the actuarial estimate over the whole follow-up,
# rather than at one horizon. This is the nearest available equivalent to the
# SAS `%check` step, which refits with `maxsteps=0` to compute Q-statistics for
# factors the screen may have overlooked -- that statistic has no R equivalent
# yet, so this checks the FIT rather than the omitted variables. Stated, rather
# than presented as the same thing.
gof <- hzr_gof(reported)
knitr::kable(utils::tail(gof[, intersect(c("time", "km_surv", "par_surv",
                                           "cum_observed", "cum_expected",
                                           "residual"), names(gof))], 10),
             row.names = FALSE, digits = 5,
             caption = "Observed vs parametric, last ten event times")
```

## Save

```{r}
#| label: save
# `hs` builds patient-level predictions from this model and `bh` bootstraps
# this screen, both by set.
saveRDS(list(reported = reported, stage1 = stage1, stage2 = stage2,
             covariates = COVARIATES, audit = audit, deciles = dec),
        set_path("estimates", "hm.rds"))
```
