---
title: "Bootstrap variable selection"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in:
# rendered as a standalone document the project's format block never applies,
# which is how server-rendered reports end up with sibling _files/ trees instead
# of being self-contained.
#
# NOTE: output-dir is a PROJECT setting and cannot be set here, so the .html
# lands next to its source. With embed-resources that is a single standalone
# file, which is the property that actually matters when a report is sent on.
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the SAS job this replaces and paste its `%hazboot` call. The
     call is the specification -- replicate count, entry and stay criteria and
     the candidate pool all read off it -- so keep it in the document rather
     than transcribing its numbers into prose. The next person to touch this
     needs to see what is being reproduced. -->

Replaces `analyses/<job>.sas`: paste its `%hazboot(...)` call here.

Each replicate resamples the cohort with replacement and runs a fresh forward
selection over the whole candidate pool. The **selection frequency** -- the
percentage of replicates in which a variable was retained -- is the deliverable,
and it is what a published *Reliability (%)* column reports.

This job **does not fit the final model**. It screens.

```{r}
#| label: setup
# This file runs either from the directory it was scaffolded into or from the
# project root. Resolve which, so that no path in this document needs editing:
# forgetting one of several ../ paths is the instantiation mistake that would
# otherwise be easy to make and quiet when made.
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
suppressPackageStartupMessages({
  library(TemporalHazard)
  library(hvtiRbootstrap)
  library(hvtiRutilities)
})
```

```{r}
#| label: edit-guard
#| results: asis
# The markers in this file name work a study author still has to do, and
# README.md says a job that still contains one has not been finished. This
# chunk is what makes that TRUE rather than merely stated.
#
# Without it an unedited job renders green over a meaningless analysis. In THIS
# file the danger is the DENOMINATOR: EXPECT_CHUNKS and EXPECT_BOOT would still
# hold the template's numbers, so the completeness check compares the run
# against a total nobody chose -- and PASSES. Every frequency below is then
# honestly computed over the wrong base, and nothing in the output looks
# wrong.
#
# knitr::current_input() is NULL outside a render -- a study author stepping
# through chunks in RStudio -- and there is no file to scan then, so this is a
# no-op in that case. Same handling as the `set` guard below, for the same
# reason.
#
# The token is BUILT, not written literally, and that is not stylistic. Quarto
# knits through an intermediate and current_input() returns THAT file, so the
# scan reads a copy of this chunk along with everything else: a literal
# grep("<token>", .src) here matches its own source line, and the guard then
# fires on every render, finished or not. Measured, not theorised -- a probe
# found three markers in a file containing two. A guard that cries wolf on a
# finished job is a guard that gets deleted, which returns us to issue #27.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    # Report the marker TEXT. Line numbers here are the intermediate's, not
    # this .qmd's, so they need not match what the author sees in an editor.
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it.\nTo render a partly-worked draft anyway, set ",
      "HVTI_TEMPLATE_DRAFT to 1, true or yes -- the report then carries a ",
      "draft banner. Any other value, including 0, leaves this guard strict."
    )
    # Only the documented values enable draft mode. nzchar() alone would treat
    # HVTI_TEMPLATE_DRAFT=0 as ON, silently downgrading this stop() to a
    # warning for someone whose shell profile sets it to 0 meaning OFF -- a
    # switch whose enablement does not mean what it says, which is the same
    # defect this guard exists to catch. An unrecognised value falls through to
    # strict on purpose: the author then sees the stop and re-reads the docs,
    # which is the safe direction to be wrong in.
    if (tolower(Sys.getenv("HVTI_TEMPLATE_DRAFT")) %in% c("1", "true", "yes")) {
      # The banner is not optional. A draft render that looks like a finished
      # one is the same defect with an extra step: the warning scrolls past in
      # a log, while the .html is the artifact that gets sent to someone.
      warning(.msg, call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Rendered with `HVTI_TEMPLATE_DRAFT` set. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by new_job() from the endpoint/type it
     was called with, which are also the filename's first two fields. They
     name the SET this job belongs to, and every artifact path in the
     document is computed from them, so no output path below needs editing.

     Do NOT hand-edit these declarations: the render-time guard below checks
     them against this file's own name and stops if they disagree. To put
     this job in a different set, call new_job() again with the new
     endpoint/type rather than editing this copy in place.

     A set is keyed on (endpoint, analysis type), not on the endpoint alone.
     One endpoint is analysed by several methods -- parametric hazard, random
     forest survival, competing risks -- and those chains share their upstream.
     Two sets keyed on `dead_pa` alone would write their life tables to the
     same file. -->


```{r}
#| label: set
ENDPOINT <- "dead_pa"
TYPE     <- "hz"

# `new_job()` writes ENDPOINT/TYPE from the same values it put in this file's
# name, but a hand-edited declaration can drift from it afterward. set_path()
# below resolves from the declarations, not the filename, so a drifted
# declaration would silently write into ANOTHER set's artifact directory --
# exactly the collision the (endpoint, type) key exists to prevent, re-entered
# through the body instead of the name.
#
# knitr::current_input() is NULL outside a render (a study author running
# chunks interactively in RStudio), and there is no filename to check against
# yet, so this is a no-op in that case rather than a spurious error.
.current <- knitr::current_input()
if (!is.null(.current)) {
  # Quarto knits through an intermediate, so `knitr::current_input()` returns
  # `<endpoint>-<type>-<ordinal>-<prefix>.rmarkdown` here rather than the
  # `.qmd` this was scaffolded as. Strip whatever extension is actually present
  # rather than hard-coding one, so this doesn't depend on a build-tool detail
  # staying the same.
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint, "', type '",
         .name_type, "'), but declares ENDPOINT = \"", ENDPOINT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

# Resolve a path inside this set's artifact directory. `kind` is the artifact
# folder -- "estimates" for serialized results, "graphs" for figures. The set
# directory sits one layer under the kind and never two, which is the whole
# layout rule.
#
# Created on first use rather than up front, so a job that writes nothing leaves
# no empty directories behind.
set_path <- function(kind, file) {
  d <- file.path(.root, kind, paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}

# `bh` reads the chunks its companion runner wrote through this, and writes its
# pooled screen back the same way, so a screen can never be filed against a
# different set than the replicates it pooled. The write is in the `save` chunk
# at the foot of this file.
```

## The screen this reports on

<!-- EDIT: nothing in this section, but read it before changing the two
     expectations below.

     This job does not RUN the screen. `hzr_bootstrap()` writes nothing until
     its final replicate and a full screen is days of compute, so a run that
     dies near the end leaves nothing at all. The run is therefore chunked from
     a companion script, and this report reads what that script wrote. The file
     boundary is a durability boundary, not tidiness. -->

```{r}
#| label: bootstrap
# EDIT: the prefix your companion runner writes its output under.
#
# The runner is not templated -- it is a study's own script -- so its filenames
# are its own choice, and this has to be told rather than guessed. One long run
# writes <prefix>.rds; a chunked run writes <prefix>.chunk01.rds, chunk02 and so
# on. Both are read the same way below.
CHUNK_PREFIX <- "bagging"

# EDIT: what this run was LAUNCHED as, not what happens to be on disk.
#
# Nothing in a chunk knows how many siblings it was launched alongside, so these
# two numbers are the only thing that can tell a partial pool from a complete
# one. Without them a render at hour eight of a twelve-hour run produces a
# report that is wrong in no visible way: every health check passes, every
# frequency is honestly computed, and only the denominator is not the one
# anybody chose.
#
# For a single unchunked run, set EXPECT_CHUNKS to 1.
EXPECT_CHUNKS <- 25L    # EDIT: how many chunks you launched
EXPECT_BOOT   <- 500L   # EDIT: %hazboot(resampl=) -- total replicates wanted

# Artifacts live under estimates/<endpoint>-<type>/, by set, so a screen can
# never be read from or filed against a different set than the one this job
# belongs to. set_path() creates the directory; dirname() is what
# boot_chunk_files() wants.
est_dir <- dirname(set_path("estimates", paste0(CHUNK_PREFIX, ".rds")))
rds     <- file.path(est_dir, paste0(CHUNK_PREFIX, ".rds"))
chunks  <- boot_chunk_files(est_dir, prefix = CHUNK_PREFIX)

if (file.exists(rds)) {
  bag <- readRDS(rds)
} else if (length(chunks)) {
  # boot_pool_chunks() REFUSES chunks that are not draws from one screen -- a
  # shared seed, a dataset rewritten mid-run, different entry criteria. So a
  # pooled frequency below is one that was checked rather than assumed, and a
  # stop() here is the check working, not an inconvenience to route around.
  bag <- boot_pool_chunks(lapply(chunks, readRDS))
} else {
  stop("No bootstrap output found in ", est_dir, ".\n",
       "This screen is a long job -- days, not minutes -- and is run from a ",
       "companion script rather than inside this render, because the run ",
       "writes nothing until its final replicate.\n",
       "Run your chunks first, then render this report.", call. = FALSE)
}
```

```{r}
#| label: completeness
#| results: asis
#| echo: false
# Rendered as a CALLOUT rather than a table cell. Looking at a screen mid-run is
# routine and worth keeping; what is not acceptable is a provisional report that
# cannot say so -- including to someone who receives the .html later, without
# the context of when it was made. A cell in a provenance table further down
# does not survive that journey. A banner does.
shortfall <- boot_shortfall(bag, EXPECT_CHUNKS, EXPECT_BOOT)
if (!is.null(shortfall)) {
  cat("::: {.callout-warning}\n",
      "## Provisional -- this screen is not the complete run\n\n",
      shortfall, "\n:::\n", sep = "")
}
```
