---
title: "Bootstrap variable selection"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in:
# rendered as a standalone document the project's format block never applies,
# which is how server-rendered reports end up with sibling _files/ trees instead
# of being self-contained.
#
# NOTE: output-dir is a PROJECT setting and cannot be set here, so the .html
# lands next to its source. With embed-resources that is a single standalone
# file, which is the property that actually matters when a report is sent on.
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the SAS job this replaces, and paste its `%hazboot` call
     verbatim. The call is the specification: it names the candidate pool, the
     phases each candidate was offered to, the entry and stay levels, and the
     replicate count. Prose describing it is a paraphrase that drifts; the call
     does not. -->

Replaces `analyses/<job>.sas`: paste its `%hazboot` call here — the candidate
pool, which phases carried candidates, the entry and stay levels, and
`resampl=`.

This job **does not fit the final model**. It screens. The companion `hm` job
fits the multivariable model.

⚠️ `bh` and `hm` are **parallel analyses, not a pipeline.** `%hazboot` and
`%model` were parallel in SAS and these jobs reproduce that. Do not describe
this job's retained set as what `hm` fits, and do not write a handoff file for
it — an earlier study report did both, and nothing read the file. A reader who
believes the claim will assume the `hm` model was screened for reliability
first. It was not.

```{r}
#| label: setup
# This file runs either from the directory it was scaffolded into or from the
# project root. Resolve which, so that no path in this document needs editing:
# forgetting one of several ../ paths is the instantiation mistake that would
# otherwise be easy to make and quiet when made.
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
suppressPackageStartupMessages({
  library(TemporalHazard)
  library(hvtiRbootstrap)
  library(hvtiRutilities)
  library(ggplot2)
})

# The reporting layer below -- boot_validate() and everything after it -- landed
# in 0.1.2. Without this check a study on 0.1.1 gets "could not find function
# boot_provenance" from the middle of a render: the symbol is named, the reason
# is not, and the fix is not guessable from the message.
if (utils::packageVersion("hvtiRbootstrap") < "0.1.2") {
  stop("This report needs hvtiRbootstrap >= 0.1.2 for its reporting layer; ",
       utils::packageVersion("hvtiRbootstrap"), " is installed. Update it, ",
       "then re-render.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
# The markers in this file name work a study author still has to do, and
# README.md says a job that still contains one has not been finished. This
# chunk is what makes that TRUE rather than merely stated.
#
# Without it an unedited job renders green over a meaningless analysis. In THIS
# file the danger is that EXPECT_CHUNKS and EXPECT_BOOT still hold the template's
# numbers, so the completeness check compares the run against a denominator
# nobody chose -- and passes, reporting frequencies over the wrong base.
#
# knitr::current_input() is NULL outside a render -- a study author stepping
# through chunks in RStudio -- and there is no file to scan then, so this is a
# no-op in that case. Same handling as the `set` guard below, for the same
# reason.
#
# The token is BUILT, not written literally, and that is not stylistic. Quarto
# knits through an intermediate and current_input() returns THAT file, so the
# scan reads a copy of this chunk along with everything else: a literal
# grep("<token>", .src) here matches its own source line, and the guard then
# fires on every render, finished or not. Measured, not theorised -- a probe
# found three markers in a file containing two. A guard that cries wolf on a
# finished job is a guard that gets deleted, which returns us to issue #27.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    # Report the marker TEXT. Line numbers here are the intermediate's, not
    # this .qmd's, so they need not match what the author sees in an editor.
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it.\nTo render a partly-worked draft anyway, set ",
      "HVTI_TEMPLATE_DRAFT to 1, true or yes -- the report then carries a ",
      "draft banner. Any other value, including 0, leaves this guard strict."
    )
    # Only the documented values enable draft mode. nzchar() alone would treat
    # HVTI_TEMPLATE_DRAFT=0 as ON, silently downgrading this stop() to a
    # warning for someone whose shell profile sets it to 0 meaning OFF -- a
    # switch whose enablement does not mean what it says, which is the same
    # defect this guard exists to catch. An unrecognised value falls through to
    # strict on purpose: the author then sees the stop and re-reads the docs,
    # which is the safe direction to be wrong in.
    if (tolower(Sys.getenv("HVTI_TEMPLATE_DRAFT")) %in% c("1", "true", "yes")) {
      # The banner is not optional. A draft render that looks like a finished
      # one is the same defect with an extra step: the warning scrolls past in
      # a log, while the .html is the artifact that gets sent to someone.
      warning(.msg, call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Rendered with `HVTI_TEMPLATE_DRAFT` set. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by new_job() from the endpoint/type it
     was called with, which are also the filename's first two fields. They
     name the SET this job belongs to, and every artifact path in the
     document is computed from them, so no output path below needs editing.

     Do NOT hand-edit these declarations: the render-time guard below checks
     them against this file's own name and stops if they disagree. To put
     this job in a different set, call new_job() again with the new
     endpoint/type rather than editing this copy in place.

     A set is keyed on (endpoint, analysis type), not on the endpoint alone.
     One endpoint is analysed by several methods -- parametric hazard, random
     forest survival, competing risks -- and those chains share their upstream.
     Two sets keyed on `dead_pa` alone would write their life tables to the
     same file. -->

```{r}
#| label: set
ENDPOINT <- "dead_pa"
TYPE     <- "hz"

# `new_job()` writes ENDPOINT/TYPE from the same values it put in this file's
# name, but a hand-edited declaration can drift from it afterward. set_path()
# below resolves from the declarations, not the filename, so a drifted
# declaration would silently write into ANOTHER set's artifact directory --
# exactly the collision the (endpoint, type) key exists to prevent, re-entered
# through the body instead of the name.
#
# knitr::current_input() is NULL outside a render (a study author running
# chunks interactively in RStudio), and there is no filename to check against
# yet, so this is a no-op in that case rather than a spurious error.
.current <- knitr::current_input()
if (!is.null(.current)) {
  # Quarto knits through an intermediate, so `knitr::current_input()` returns
  # `<endpoint>-<type>-<ordinal>-<prefix>.rmarkdown` here rather than the
  # `.qmd` this was scaffolded as. Strip whatever extension is actually present
  # rather than hard-coding one, so this doesn't depend on a build-tool detail
  # staying the same.
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint, "', type '",
         .name_type, "'), but declares ENDPOINT = \"", ENDPOINT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

# Resolve a path inside this set's artifact directory. `kind` is the artifact
# folder -- "estimates" for serialized results, "graphs" for figures. The set
# directory sits one layer under the kind and never two, which is the whole
# layout rule.
#
# Created on first use rather than up front, so a job that writes nothing leaves
# no empty directories behind.
set_path <- function(kind, file) {
  d <- file.path(.root, kind, paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}

# The screen this job reports on was run by a companion script, which wrote its
# chunks into this set's `estimates` directory. Reading them through set_path()
# is what keeps a screen filed against the set whose candidates it screened.
```

## The screen

```{r}
#| label: expect
# EDIT: what this run was LAUNCHED as, not what happens to be on disk.
#
# Nothing in a chunk knows how many siblings it was launched alongside, so these
# two numbers are the only thing that can tell a partial pool from a complete
# one. Without them a render at hour eight of a twelve-hour run produces a
# report that is wrong in no visible way: every health check passes, every
# frequency is honestly computed, and only the denominator is not the intended
# one.
#
# For a single unchunked run, set EXPECT_CHUNKS to 1.
EXPECT_CHUNKS <- 25L    # EDIT: how many chunks you launched
EXPECT_BOOT   <- 500L   # EDIT: %hazboot(resampl=) -- total replicates wanted

# EDIT: the file-name prefix your runner wrote its chunks under. It must match
# the runner exactly: boot_chunk_files() quotes this literally and a prefix that
# does not match finds nothing, while a prefix that matches TOO MUCH pools
# chunks from another run -- and no downstream check can see that.
BOOT_PREFIX <- "bh"
```

```{r}
#| label: load
# The screen is not run here. It is read.
#
# hzr_bootstrap() writes nothing until its final replicate and a full screen is
# days of compute, so a run that dies near the end leaves nothing at all. The
# companion runner exists to convert one all-or-nothing multi-day call into N
# independently restartable ones. That file boundary is a DURABILITY boundary,
# not tidiness, which is why this report reads chunks rather than producing
# them.
.boot_dir <- dirname(set_path("estimates", "."))
.chunk_files <- boot_chunk_files(.boot_dir, prefix = BOOT_PREFIX)
.single_file <- set_path("estimates", paste0(BOOT_PREFIX, ".rds"))

if (length(.chunk_files)) {
  # Pooling is checked, not assumed: boot_pool_chunks() refuses chunks that
  # disagree on the dataset checksum, the entry/stay levels, the base
  # parameters, the usable candidate counts, the engine version or the step
  # cap, and refuses two chunks sharing a seed. Every one of those failures is
  # silent on its own -- it shows up as a slightly different frequency, never
  # as an error -- which is why the refusal belongs here rather than in a note.
  bag <- boot_pool_chunks(lapply(.chunk_files, readRDS))
} else if (file.exists(.single_file)) {
  bag <- readRDS(.single_file)
} else {
  stop("No bootstrap output found. This screen is a long job -- days, not ",
       "minutes -- and is run from a companion script rather than inside this ",
       "render, because the run writes nothing until its final replicate.\n",
       "Run your chunks first, then render this report.", call. = FALSE)
}
```

```{r}
#| label: completeness
#| results: asis
# A callout, not a table cell. A provisional report that cannot say so is the
# failure being prevented, and it must survive someone receiving the .html
# without knowing when it was made -- which a number buried in a provenance
# table does not.
.shortfall <- boot_shortfall(bag, EXPECT_CHUNKS, EXPECT_BOOT)
if (!is.null(.shortfall)) {
  cat("\n::: {.callout-warning title=\"Provisional -- this pool is incomplete\"}\n")
  cat(.shortfall, "\n", sep = "")
  cat(":::\n\n")
}
```

## Provenance

```{r}
#| label: contract
# The runner is NOT templated -- it is a study file -- so the fields this report
# reads are a contract between two files that no shared function enforces.
# Stated here, and checked, because the alternative is a report that indexes a
# field the runner stopped writing and renders `NULL` into a provenance table as
# though it were an answer.
#
# A pooled run has already had most of these checked by boot_pool_chunks(),
# which refuses chunks that disagree. A SINGLE unchunked run has not: nothing
# ran between the runner and here. This is the only check that covers both.

# Shapes, not merely names. Checking presence is what let a length-2 `requested`
# reach a table that cannot recycle it against a length-13 item column -- the
# render-blocker that shipped in three releases and was fixed in 1.0.17. Every
# failure is reported at once, because an author fixing a runner wants the whole
# list rather than one field per re-render.
boot_validate(bag)
```

```{r}
#| label: provenance
# CPU cost, NOT wall clock. boot_pool_chunks() SUMS elapsed across chunks, so on
# a chunked run this is total compute; chunks run in parallel finish in a
# fraction of it. Reported in hours because the minutes figure reads as a
# wall-clock number and is off by the chunk count.
cpu_hours <- bag$elapsed_mins / 60

# The digest travels WITH its algorithm. An md5 and a sha256 of the same file
# are different strings, and of different files may not be, so a bare digest
# recorded without its algorithm is not evidence of anything.
checksum <- NA_character_
for (algo in c("sha256", "md5")) {
  if (!is.null(bag$manifest[[algo]])) {
    checksum <- paste0(algo, ":", bag$manifest[[algo]])
    break
  }
}

# A version string alone cannot say which codebase ran: one real package version
# existed as two, one with a selection criterion and one without, and the
# selection criterion is precisely what decides what a screen selects.
engine <- if (!is.null(bag$th_sha)) {
  paste0("sha:", bag$th_sha)
} else if (!is.null(bag$th_version)) {
  paste0("version:", bag$th_version)
} else {
  NA_character_
}

# Summarised, not listed: at 25 chunks the joined seed string is a single
# 270-character cell that wrecks the table. Every seed is in the table below.
n_chunks <- if (is.null(bag$n_chunks)) 1L else as.integer(bag$n_chunks)

# `requested` and `usable` are per PHASE, not scalars: a multiphase screen
# offers its candidate pool to each phase separately, and a real bag carries
# c(early = 230, late = 230). data.frame() does not recycle a length-2 value
# against a length-13 item column -- it errors -- so this table could not build
# on ANY multiphase bag, which is every hazard bag. Found by the first real
# render, against the study this template was written from.
#
# Collapsed to one labelled string per row rather than summed: the pool is
# OFFERED to each phase, so 230 and 230 is one pool seen twice, not 460
# candidates. Falls back to a bare value when a single-phase run makes it
# length 1, and to unlabelled values if a future runner drops the names.
.per_phase <- function(x) {
  # A zero-length value must return exactly one string, or it vanishes from the
  # `value` vector and data.frame() fails with the SAME "differing number of
  # rows" error this helper was written to fix. `format(integer(0))` is
  # character(0); `format(NULL)` is the string "NULL", so only the typed
  # zero-length case actually drops. Guarded explicitly rather than relying on
  # which of the two a future runner happens to write.
  if (length(x) == 0L) {
    return(NA_character_)
  }
  if (length(x) == 1L) {
    return(format(x))
  }
  if (is.null(names(x))) {
    return(paste(format(x), collapse = ", "))
  }
  paste0(names(x), " ", format(x), collapse = ", ")
}

provenance <- data.frame(
  item = c("Replicates pooled", "Chunks pooled", "Entry level (slentry)",
           "Stay level (slstay)", "Candidates offered", "Candidates usable",
           "Rows screened", "Replicates that fitted", "Replicates that failed",
           "CPU hours (summed)", "Fitting engine", "Dataset checksum",
           "Seeds"),
  value = c(format(bag$n_boot), format(n_chunks), format(bag$slentry),
            format(bag$slstay), .per_phase(bag$requested), .per_phase(bag$usable),
            format(bag$n_rows), format(bag$boot$n_success),
            format(bag$boot$n_failed), sprintf("%.1f", cpu_hours),
            engine, checksum,
            paste0(n_chunks, " distinct (listed below)")),
  stringsAsFactors = FALSE
)
provenance
```

```{r}
#| label: seeds
# Every seed, so that a rerun is reproducible and a duplicate is visible. Two
# chunks sharing a seed contain literally the SAME replicates: pooling them
# counts each twice and reports a Monte-Carlo error smaller than the run
# actually has. boot_pool_chunks() refuses that outright, so a pooled run cannot
# reach here with a duplicate -- this table is what lets you check a single run,
# and what lets a reader reproduce either.
#
# `seeds` is created by boot_pool_chunks(); a SINGLE unchunked run never went
# through it and carries only the scalar `seed` its runner wrote. Falling back
# is not defensive padding -- without it the one case this table exists to
# serve is the one case it cannot render.
seed_list <- if (!is.null(bag$seeds)) bag$seeds else bag$seed
data.frame(chunk = seq_along(seed_list),
           seed  = format(seed_list, scientific = FALSE),
           stringsAsFactors = FALSE)
```

## Candidates that were never screened

<!-- Counts first, detail second, and that ordering is not cosmetic. The detail
     table runs to dozens of rows on a real pool, and a reader who has to scroll
     it to discover that a whole CLASS of variable was never screened will not
     discover it. -->

```{r}
#| label: dropped-summary
# A candidate the screen never saw cannot appear at any frequency, so its
# absence from the table below is indistinguishable from never having been
# selected. That is the reason this section exists at all.
if (is.null(bag$dropped) || !nrow(bag$dropped)) {
  cat("Every candidate offered was screened.\n")
} else {
  as.data.frame(table(phase = bag$dropped$phase, reason = bag$dropped$reason))
}
```

```{r}
#| label: dropped-detail
if (!is.null(bag$dropped) && nrow(bag$dropped)) {
  bag$dropped[order(bag$dropped$phase, bag$dropped$reason), ]
}
```

## Did the screen actually run?

⚠️ **A screen that selected nothing is a failure, not a finding.** It is the
signature of a formula that did not survive the per-replicate rewrite: the refit
errors, the error is caught, the step reports nothing accepted, and the screen
halts having selected nothing — with no warning and `n_failed = 0`. The summary
then reads as a table of perfectly reliable variables.

```{r}
#| label: health
# EDIT: the candidate pool, written LITERALLY in your runner's formula.
#
# Not in a variable, not via as.formula() or reformulate(). hzr_bootstrap() and
# hzr_stepwise() rewrite the stored formula per replicate, and a symbol standing
# where the formula should be does not survive that rewrite. See the health
# check below for what the failure looks like -- it does not error.
reps <- bag$boot$replicates
candidates <- setdiff(unique(reps$parameter), bag$base_params)

# boot_pool_chunks() computes free_sd, so a SINGLE unchunked run does not carry
# it. Recomputed rather than skipped: this is the check that catches a bootstrap
# which refit nothing, and an unchunked run is the shape most likely to have
# been built that way. NA, never 0, when there is nothing to take an SD of --
# zero is the CLAIM that the parameter did not vary, which one replicate cannot
# support.
free_sd <- if (!is.null(bag$free_sd)) {
  bag$free_sd
} else {
  .est <- reps$estimate[reps$parameter == bag$base_params[[1L]]]
  if (length(.est) > 1L) stats::sd(.est) else NA_real_
}

health <- data.frame(
  check = c("Replicates that fitted", "Replicates that failed",
            "Distinct candidates ever selected",
            "SD of the first free base parameter"),
  value = c(format(bag$boot$n_success), format(bag$boot$n_failed),
            format(length(candidates)), format(free_sd)),
  stringsAsFactors = FALSE
)
health

if (!length(candidates)) {
  stop("The screen selected NOTHING: no parameter outside the base model ",
       "appears in any replicate. That is a failed screen, not a null result, ",
       "and it is what a formula held in a variable looks like from here -- ",
       "n_failed is ", bag$boot$n_failed, ", because the refit error was ",
       "caught and the step simply accepted nothing.\nWrite the model formula ",
       "literally at the call site in your runner and rerun.", call. = FALSE)
}

# A free parameter must VARY across resamples. A bootstrap built on the vector
# interface returns the original fit every replicate, with n_success = 500,
# n_failed = 0, and no warning. sd() of a free base parameter is the tell.
if (!is.na(free_sd) && free_sd == 0) {
  stop("The first free base parameter has SD exactly 0 across ", bag$n_boot,
       " replicates, so every replicate returned the SAME fit. A bootstrap ",
       "built on the vector interface does that: it refits nothing and reports ",
       "n_success = ", bag$boot$n_success, " with no warning.\nThe frequencies ",
       "below would all be 100% and mean nothing.", call. = FALSE)
}
```

## What a selection frequency is, and what it is not

A selection frequency is an estimate, not a count of something fixed, and it
carries Monte-Carlo error of roughly `sqrt(p(1-p)/n_boot)`. At `p = 0.5` over
500 replicates that is about **2.2 percentage points**, so a variable sitting
within a few points of the retention threshold can fall on either side of it on
resampling noise alone.

- **Agreement on the retention decision matters more than agreement on the
  frequency.** Which side of the threshold a variable falls on is the decision
  this job exists to support.
- **A near-threshold variable is not a weak risk factor.** It is a variable
  whose selection is unstable, which is a different claim.

```{r}
#| label: criteria
# EDIT: from YOUR .sas %hazboot call. The two studies this template was
# extracted from disagree -- 500/0.07/0.05 and 1000/0.12/0.1 -- so there is no
# default here to fall back on. Read them off the call and paste it above.
#
# RETAIN_PCT is the reliability cutoff this report calls "retained". It is a
# reporting decision, not something the run recorded: the entry and stay levels
# above governed each replicate's stepwise fit, and neither of them says how
# often a variable must survive to be worth carrying forward.
RETAIN_PCT <- 50
```

```{r}
#| label: frequencies
# The name column differs by how the screen was produced, and reading the wrong
# one yields NULL rather than an error. boot_pool_chunks() recomputes its
# summary and names the column `parameter`; boot_summary() names it `variable`.
# A single unchunked run carries whichever its runner wrote.
summ <- bag$boot$summary
name_col <- if ("parameter" %in% names(summ)) "parameter" else "variable"

freq <- data.frame(term = summ[[name_col]], pct = summ$pct, n = summ$n,
                   stringsAsFactors = FALSE)
freq <- freq[!freq$term %in% bag$base_params, , drop = FALSE]

# Terms are named `<phase>.<variable>`: one variable offered to two phases is
# two independent screening decisions and must not be pooled into one row.
freq$phase    <- sub("[.].*$", "", freq$term)
freq$variable <- sub("^[^.]*[.]", "", freq$term)

# Per-variable, not one figure for the table. Monte-Carlo error is largest at
# p = 0.5 and shrinks towards either end, so a single quoted value overstates it
# for the variables at the top and understates it for the ones in the middle --
# and the middle is where the retention decision is actually being made.
p <- freq$pct / 100
freq$mc_error <- 100 * sqrt(p * (1 - p) / bag$n_boot)

# Within two Monte-Carlo errors of the cutoff: the variables whose retention
# would not survive a rerun with different seeds.
freq$near_threshold <- abs(freq$pct - RETAIN_PCT) <= 2 * freq$mc_error

freq <- freq[order(freq$phase, -freq$pct, freq$variable), , drop = FALSE]
rownames(freq) <- NULL
freq[, c("phase", "variable", "n", "pct", "mc_error", "near_threshold")]
```

```{r}
#| label: retained
# The retained set, by phase. Reported WITH the near-threshold flag rather than
# as a bare list: a variable that cleared the cutoff by less than its own
# Monte-Carlo error cleared it on noise, and a list that does not say so invites
# the reader to treat the two kinds of retention as the same claim.
retained <- freq[freq$pct >= RETAIN_PCT, , drop = FALSE]
retained[, c("phase", "variable", "n", "pct", "mc_error", "near_threshold")]
```

## Concepts

⚠️ **Do not prune competing transformations from the pool before screening.**
Screen every form; group only when reading, which is what this section does.

Measured, on the study this template came from: of the 57 forms pruning removed,
**16 correlated at |r| < 0.9** with the form kept and five below 0.5. `in_zexp`
**is** `1/zexp` — r = 0.9997 against the reciprocal — yet correlates with
`zexp` at only **-0.195**, because `zexp` spans 0.038 to 151.9. Over a 4000-fold
range a value and its reciprocal are different information. That study's
published model uses `zexp` and `in_zexp` in the same phase, **both
significant**, a two-parameter flexible form pruning forbids.

The naming convention tells you two variables are RELATED. Only the data tells
you whether they are REDUNDANT.

```{r}
#| label: concept-map
# EDIT: the affix vocabulary, if your study's variable names do not follow
# vars.sas conventions. POOL_AFFIXES carries ln_, in_, in2, _pr and a trailing
# 2, the order in which they strip, and the deliberate refusal to reduce agee
# to age. Those are facts about this institution's names, not about statistics,
# which is why they are data you can replace rather than logic you cannot.
#
# The pool grouped here is the set of forms that reached at least one replicate.
# A candidate the screen never saw is in the unscreened table above, not here.
concept <- concept_map(unique(freq$variable), affixes = POOL_AFFIXES,
                       min_stem = POOL_MIN_STEM,
                       plain_suffix = POOL_PLAIN_SUFFIX)
concept
```

```{r}
#| label: concept-frequencies
# Grouped for READING. Every form keeps its own row and its own frequency --
# nothing is collapsed, summed or dropped -- because the whole argument above is
# that two forms of one concept may carry different information. The concept
# column is a reading aid, not an aggregation.
by_concept <- merge(freq, concept[, c("variable", "concept", "representative")],
                    by = "variable", all.x = TRUE)
by_concept <- by_concept[order(by_concept$phase, by_concept$concept,
                               -by_concept$pct), , drop = FALSE]
rownames(by_concept) <- NULL
by_concept[, c("phase", "concept", "variable", "representative", "pct",
               "mc_error")]
```

A per-form row answers "how often was *this form* selected". It cannot answer
"how often was *this concept* selected", and that is the number a paper quotes.
The gap runs both ways: competing forms split replicates between them, so the
concept reads weaker than any single figure suggests, or two forms both clear
the cutoff and one finding is reported twice.

```{r}
#| label: concept-union
# Nothing above is collapsed -- every form keeps its row and its frequency.
# This is an ADDITIONAL row per concept, not a replacement for them.
#
# Computed from the replicate table, never from the summary. A union across a
# concept's forms cannot be recovered from marginal percentages: two forms at
# 30% each are anywhere between 30% and 60% of replicates, depending entirely
# on how often the same replicate took both, and the summary does not record
# that.
cov_reps <- reps[!reps$parameter %in% bag$base_params, , drop = FALSE]
cov_reps$phase    <- sub("[.].*$", "", cov_reps$parameter)
cov_reps$variable <- sub("^[^.]*[.]", "", cov_reps$parameter)
cov_reps$concept  <- concept$concept[match(cov_reps$variable, concept$variable)]

union_rows <- lapply(split(by_concept, list(by_concept$phase, by_concept$concept), drop = TRUE),
                     function(g) {
                       ph <- g$phase[1]
                       ck <- g$concept[1]
                       in_k <- cov_reps$phase == ph & !is.na(cov_reps$concept) &
                         cov_reps$concept == ck
                       data.frame(phase         = ph,
                                  concept       = ck,
                                  n_forms       = nrow(g),
                                  union_pct     = 100 * length(unique(cov_reps$replicate[in_k])) /
                                    bag$n_boot,
                                  best_form_pct = max(g$pct),
                                  stringsAsFactors = FALSE)
                     })
concept_union <- do.call(rbind, union_rows)

# How much the per-form table understates the concept. A large spread is the
# whole reason this table exists; a spread of zero means the concept has one
# form and the two views agree.
concept_union$spread   <- concept_union$union_pct - concept_union$best_form_pct
concept_union$retained <- concept_union$union_pct >= RETAIN_PCT
concept_union <- concept_union[order(concept_union$phase, -concept_union$union_pct), ,
                               drop = FALSE]
rownames(concept_union) <- NULL
concept_union
```

⚠️ **A selection frequency is conditional on the candidate pool**, and no column
here removes that. On the study this template came from, re-running with 226
candidates instead of 189 moved one variable from 26.6% to 95.2% and another
from 93.8% to 20.2% — same data, same seeds. Quote a reliability figure with
the pool it came from.

```{r}
#| label: concept-counts
# Crowding: a phase that spent several of its slots on forms of ONE concept is
# budget-limited by redundancy, and that is invisible in a coefficient table.
# Grouping is concept_map()'s, so this is a FLOOR -- a concept nobody's affix
# vocabulary reaches is still counted as separate variables.
crowding <- selection_crowding(retained$term, affixes = POOL_AFFIXES,
                               min_stem = POOL_MIN_STEM,
                               plain_suffix = POOL_PLAIN_SUFFIX)
if (!nrow(crowding)) {
  cat("No phase retained more than one form of any concept.\n")
} else {
  crowding
}
```

## Correlation clusters

Concept grouping above is by NAME, and a naming convention only reaches what
its affix rules can reach. Two further groupings matter, and neither is
derivable from a variable's name: one by DECLARATION, which is clinical
judgement, and one by the DATA. The SAS job this replaces ran `%cluster` once
per named cluster per phase and covered both.

The three views answer different questions. A name tells you two variables are
related. A declaration tells you a clinician treats them as one thing —
`Renal` is creatinine, GFR and dialysis, which share no stem and never will. A
correlation tells you the pool cannot tell them apart.

```{r}
#| label: cluster-matrix
# boot_clusters() wants one ROW per replicate and one COLUMN per term, with an
# unselected term left NA -- that is how it counts "at least one member". The
# pooled replicates are long, so pivot rather than pass them straight in.
#
# Replicate ids are 1..n_boot across the whole pool: boot_pool_chunks() offsets
# each chunk's ids before stacking, and refuses to pool at all if any chunk's
# ids fall outside its own range. The direct indexing below relies on that.
terms <- sort(unique(reps$parameter))
coefs <- matrix(NA_real_, nrow = bag$n_boot, ncol = length(terms),
                dimnames = list(NULL, terms))
coefs[cbind(reps$replicate, match(reps$parameter, terms))] <- reps$estimate
```

```{r}
#| label: clusters
# EDIT: your clusters, one per concept you want reported, and the phase each
# runs against. The SAS job named eight -- Age, Size, BMI, race, GFR, iv_opyrs,
# Renal, blrbn_pr -- and ran each against both the early and constant phases.
# There is no sensible default: which concepts are worth clustering is a
# statement about your candidate pool.
#
# Members are PHASE-QUALIFIED, as `<phase>.<variable>`, because a variable
# offered to two phases is two screening decisions. Naming a term that no
# replicate carries is refused rather than reported as 0%: a typo that grouped
# nothing would otherwise read as a concept nobody selected.
CLUSTERS <- list(
  early.Age  = c("early.age", "early.ln_age"),
  const.Age  = c("const.age", "const.ln_age")
)

clusters <- boot_clusters(coefs, CLUSTERS)
clusters
```

The table above groups by DECLARATION, not by data: `boot_clusters()` counts
replicates retaining any member of a list you wrote, and computes no
correlation at all. The correlation is below, and the two are worth keeping
apart. A declaration says a clinician thinks these are one thing; a correlation
says the pool cannot tell them apart.

```{r}
#| label: collinear
# EDIT: the correlation above which two candidates count as indistinguishable.
# 0.99 finds near-duplicates -- a variable and its rescaling, a dummy and its
# complement. Lower it to see the pool's structure rather than only its
# accidents.
COLLINEAR_R <- 0.99

# Read to CORRELATE the pool, not to re-run anything. The screen already
# happened; this describes the data it ran against.
d <- read_built()

# A build that moved since the screen ran would leave these correlations
# describing one dataset and every frequency above describing another, with
# nothing in the report to say so. A warning rather than a stop: the
# frequencies above remain valid, and it is only this table that goes stale.
if (!is.null(bag$n_rows) && !identical(as.integer(nrow(d)), as.integer(bag$n_rows))) {
  warning("The built dataset now has ", nrow(d), " rows; this screen ran against ",
          bag$n_rows, ". The correlations below describe TODAY'S data, not the data ",
          "the frequencies above were computed on.", call. = FALSE)
}

collinear <- pool_collinear_pairs(d, pool = unique(freq$variable), threshold = COLLINEAR_R)
if (!nrow(collinear)) {
  cat("No pair of screened candidates correlates at |r| >= ", COLLINEAR_R, ".\n", sep = "")
} else {
  collinear
}
```

⚠️ **A collinear pair is not a licence to prune.** It says two candidates are
indistinguishable *in this cohort at this threshold* — a statement about the
pool, not about the concepts. The no-pruning argument above stands: the same
measurement that found 57 prunable forms found 16 of them correlating below 0.9.

`pool_collinear_pairs()` sees only NUMERIC, non-constant columns, so
factor-coded binaries are invisible to it for the same reason they were
invisible to the screen. Check the coercion table above before reading an empty
result as an absence of collinearity.

## Figure

```{r}
#| label: fig-frequencies
#| fig-height: 7
#| fig-cap: "Selection frequency by phase, with Monte-Carlo error and the retention threshold."
# The error bar is the point of the figure, not decoration. A bare dot plot
# invites the reader to rank variables by a difference of two points, which is
# inside the noise at every frequency near the middle of the range.
ggplot(freq, aes(x = stats::reorder(variable, pct), y = pct)) +
  geom_hline(yintercept = RETAIN_PCT, linetype = "dashed") +
  geom_errorbar(aes(ymin = pct - mc_error, ymax = pct + mc_error), width = 0) +
  geom_point() +
  coord_flip() +
  facet_wrap(~phase, scales = "free_y") +
  labs(x = NULL, y = "Replicates selecting the variable (%)") +
  theme_minimal()
```

## Save

```{r}
#| label: save
# Written through set_path(), so the screen is filed against the SET whose
# candidates it screened rather than against the endpoint alone.
#
# There is deliberately NO handoff file here -- no selection_bh.csv, no retained
# set written for `hm` to read. `bh` and `hm` are parallel analyses and the
# study report that wrote such a file also claimed `hm` fits what `bh` retained.
# Nothing read the file, and the claim was not true. A file whose existence
# implies a dependency that does not exist is worse than no file.
# NOT `<BOOT_PREFIX>.rds`. An unchunked run's INPUT is exactly that name, so
# writing the report there overwrites the screen it just read -- and this is
# the one job where that is unrecoverable: the input is days of compute that
# `hzr_bootstrap()` produces only on its final replicate. The second render
# then reports no bootstrap output found, having destroyed it on the first.
#
# The suffix alone is not the guard. BOOT_PREFIX is an edit point, so a study
# can set it to anything, including this name. Checked rather than assumed,
# because the failure destroys data instead of erroring.
report_file <- set_path("estimates", "bh-report.rds")
inputs <- normalizePath(c(.single_file, .chunk_files), mustWork = FALSE)
if (normalizePath(report_file, mustWork = FALSE) %in% inputs) {
  stop("This report would be written to ", basename(report_file), ", which is ",
       "the bootstrap output it just read. That file is the screen -- days of ",
       "compute, written only at the final replicate -- and overwriting it ",
       "cannot be undone.\nChange BOOT_PREFIX so the runner's output and this ",
       "report do not collide.", call. = FALSE)
}

saveRDS(list(freq = freq, retained = retained, concept = concept,
             crowding = crowding, clusters = clusters,
             provenance = provenance, retain_pct = RETAIN_PCT,
             expect_chunks = EXPECT_CHUNKS, expect_boot = EXPECT_BOOT),
        report_file)
```
