---
title: "Actuarial life table"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in:
# rendered as a standalone document the project's format block never applies,
# which is how server-rendered reports end up with sibling _files/ trees instead
# of being self-contained.
#
# NOTE: output-dir is a PROJECT setting and cannot be set here, so the .html
# lands next to its source. With embed-resources that is a single standalone
# file, which is the property that actually matters when a report is sent on.
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the SAS job this replaces, and say what its %KAPLAN calls do. -->

Replaces `distributions/<job>.sas`: describe its `%KAPLAN` calls here — overall,
and each stratification.

An `ac` job produces life tables. Figures over the same estimates are an `hp`
job, and the parametric fit is `hz`; neither belongs here.

```{r}
#| label: setup
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRutilities::study_root(if (is.null(.in)) getwd() else dirname(.in))
.provenance_data <- list()
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(TemporalHazard)
  library(hvtiRutilities)
})
```

```{r}
#| label: edit-guard
#| results: asis
# The markers in this file name work a study author still has to do, and
# README.md says a job that still contains one has not been finished. This
# chunk is what makes that TRUE rather than merely stated.
#
# Without it an unedited job renders green over a meaningless analysis. The
# derive chunk below indexes a placeholder column; when that column is absent
# `!is.na(d$<col>)` is logical(0), and assignment through a zero-length index
# is a SILENT no-op in R. Every patient keeps the initial category, every
# stratified estimate is computed over one dummy stratum, and nothing errors.
#
# knitr::current_input() is NULL outside a render -- a study author stepping
# through chunks in RStudio -- and there is no file to scan then, so this is a
# no-op in that case. Same handling as the `set` guard below, for the same
# reason.
#
# The token is BUILT, not written literally, and that is not stylistic. Quarto
# knits through an intermediate and current_input() returns THAT file, so the
# scan reads a copy of this chunk along with everything else: a literal
# grep("<token>", .src) here matches its own source line, and the guard then
# fires on every render, finished or not. Measured, not theorised -- a probe
# found three markers in a file containing two. A guard that cries wolf on a
# finished job is a guard that gets deleted, which returns us to issue #27.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    # Report the marker TEXT. Line numbers here are the intermediate's, not
    # this .qmd's, so they need not match what the author sees in an editor.
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    # A job renders as a draft by default, so an author can iterate on a
    # working report and remove markers as they go. HVTI_TEMPLATE_STRICT
    # turns the draft into a stop for a final render. Only unset, 0, false
    # and no mean "not strict": an unrecognised value stops on purpose, so a
    # mistyped switch is seen rather than quietly ignored, which is the safe
    # direction to be wrong in.
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      # The banner is not optional. A draft render that looks like a finished
      # one is the same defect with an extra step: the warning scrolls past in
      # a log, while the .html is the artifact that gets sent to someone.
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by add_job() from the subject/type it
     was called with, which are also the filename's first two fields. They
     name the SET this job belongs to, and every artifact path in the
     document is computed from them, so no output path below needs editing.

     Do NOT hand-edit these declarations: the render-time guard below checks
     them against this file's own name and stops if they disagree. To put
     this job in a different set, call add_job() again with the new
     subject/type rather than editing this copy in place.

     A set is keyed on (subject, analysis type), not on the subject alone.
     One subject may be analysed by several methods -- parametric hazard, random
     forest survival, competing risks -- and those chains share their upstream.
     Two sets keyed on `dead_pa` alone would write their life tables to the
     same file. -->

```{r}
#| label: set
SUBJECT <- "dead_pa"
TYPE    <- "hz"

# `add_job()` writes SUBJECT/TYPE from the same values it put in this file's
# name, but a hand-edited declaration can drift from it afterward. set_path()
# below resolves from the declarations, not the filename, so a drifted
# declaration would silently write into ANOTHER set's artifact directory --
# exactly the collision the (subject, type) key exists to prevent, re-entered
# through the body instead of the name.
#
# knitr::current_input() is NULL outside a render (a study author running
# chunks interactively in RStudio), and there is no filename to check against
# yet, so this is a no-op in that case rather than a spurious error.
.current <- knitr::current_input()
if (!is.null(.current)) {
  # Quarto knits through an intermediate, so `knitr::current_input()` returns
  # `...-03.01-ac.rmarkdown` here, not `...-03.01-ac.qmd`. Strip whatever
  # extension is actually present rather than hard-coding one, so this
  # doesn't depend on a build-tool detail staying the same.
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject, "', type '",
         .name_type, "'), but declares SUBJECT = \"", SUBJECT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

# Resolve a path inside this set's artifact directory. `kind` is the artifact
# folder -- "estimates" for serialized results, "graphs" for figures. The set
# directory sits one layer under the kind and never two, which is the whole
# layout rule.
#
# Created on first use rather than up front, so a job that writes nothing leaves
# no empty directories behind.
set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}

# The overall life table is saved after it is fitted. `hp` reads that exact
# handoff by set and refuses an older lineage-free artifact.
```

## Study choices

Edit these values for this study before rendering.

```{r}
#| label: study-choices
# EDIT: each derived category and its source variable.
DERIVED <- c(cat_a = "src_a")

# EDIT: stratification variables.
STRATA <- c("cat_a")

# EDIT: time and status variables.
TIME   <- "iv_dead"

STATUS <- "dead"

# EDIT: counts from the SAS reference for the rows this job analyses. Leave
# these invalid defaults in place until the job's filter and endpoint agree.
EXPECTED <- list(n = NA_integer_, n_events = NA_integer_, n_censored = NA_integer_)

# EDIT: reporting horizons and their labels.
grid <- c(30 / 365.2425, 0.5, 1, 5, 10)

labs <- c("30 days", "6 months", "1 year", "5 years", "10 years")
```

## Cohort

<!-- EDIT: apply this job's filter after reading the data, if it has one. The
     expected counts, event, and time are this job's contract with its SAS
     reference; the invalid defaults make an unedited job stop rather than
     silently accept an unreconciled cohort. -->

```{r}
#| label: cohort
.cfg <- study_config(start = .root)
.data_read <- hvtiRtemplates:::.provenance_read(
  "study", .cfg, function() read_built(cfg = .cfg)
)
d <- .data_read$value
.provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(.data_read$record))
# EDIT: apply this job's filter here, before both the observed counts and gate.
#
# d <- d[!is.na(d$<flag>) & d$<flag> == 1, , drop = FALSE]
cc <- cohort_counts(d, event = STATUS, time = TIME)
assert_cohort(d, expected = EXPECTED, event = STATUS, time = TIME)

knitr::kable(data.frame(
  quantity = c("n_analysable", "n_events", "n_censored"),
  n        = c(cc$n, cc$n_events, cc$n_censored)
))
```

A rendered report is itself evidence the gate passed. Report the counts anyway —
a gate that passes silently tells the reader nothing about what it passed on.

**An event flag that is 0/1 in SAS may arrive as `logical`**, because
`read_clinical_data()` infers binary columns as logical and `hzr_kaplan()`
requires a numeric status. The `as.numeric()` coercions below are not
decoration.

## Derived strata

<!-- EDIT: this whole chunk. The cutpoints, the variables and the number of
     strata are all study-specific. What is NOT study-specific is the pattern:
     SAS derives categories with cumulative overwrites, and the ORDER matters
     as much as the cutpoints — each later condition overwrites the earlier
     assignment. Transcribe the order from the SAS, do not reorder it.

     Delete this section entirely if the job has no derived strata. -->

```{r}
#| label: derive
# EDIT: your study's category derivations, in SAS order.
derive_cats <- function(d) {
  # Assert the source columns are present BEFORE deriving. Every condition
  # below is guarded by !is.na(d$src), and when `src` is absent that guard is
  # logical(0), so each assignment is a silent no-op: every patient keeps the
  # initial value, the counts still sum to the cohort, and the render is green
  # over a single dummy stratum. An absent column must fail here, loudly.
  .need <- "src_a"                     # EDIT: the source column(s) used below.
  .miss <- setdiff(.need, names(d))
  if (length(.miss)) {
    stop("derive_cats(): source column(s) not in the data: ",
         paste(.miss, collapse = ", "),
         ". Edit this chunk for your study before rendering.", call. = FALSE)
  }

  d$cat_a <- 4L
  d$cat_a[!is.na(d$src_a) & d$src_a <  0.00] <- 3L
  d$cat_a[!is.na(d$src_a) & d$src_a < -0.60] <- 2L
  d$cat_a[!is.na(d$src_a) & d$src_a < -1.25] <- 1L
  d
}
d <- derive_cats(d)

knitr::kable(as.data.frame(table(cat_a = d$cat_a), stringsAsFactors = FALSE))
```

```{r}
#| label: derive-check
# A cumulative-overwrite derivation fails QUIETLY: reorder the conditions and
# every patient still lands in some category, the counts still sum to the
# cohort, and every stratified estimate below is wrong while looking entirely
# plausible. So the derivation is checked against the data it came from rather
# than against an external listing.
#
# Two things are visible here and nowhere else:
#   * the source RANGE within each category. Correctly ordered, the ranges are
#     disjoint and ascend with the category. Overlapping or inverted ranges mean
#     the overwrites ran in the wrong order.
#   * src_missing. Every condition is guarded by !is.na(), so a patient whose
#     source variable is missing keeps the INITIAL value and silently joins the
#     top category. That is a missingness stratum wearing a clinical label.

# Each derived category names the variable it was derived FROM.

# DERIVED is maintained by hand and can drift from derive_cats() above -- a
# category renamed there, or a source renamed here. Either way d[[src]] is NULL,
# every row below comes back empty, and a zero-row table renders quietly beside
# a derive table showing every patient in one category. The assertion covers
# both ends: sources must exist in the data, and each derived category must
# actually have been produced above.
.miss_src <- setdiff(unname(DERIVED), names(d))
.miss_cat <- setdiff(names(DERIVED),  names(d))
if (length(.miss_src) || length(.miss_cat)) {
  stop("DERIVED does not describe this data. ",
       if (length(.miss_src))
         paste0("Source column(s) absent: ", paste(.miss_src, collapse = ", "), ". "),
       if (length(.miss_cat))
         paste0("Derived column(s) absent -- not produced by derive_cats(): ",
                paste(.miss_cat, collapse = ", "), ". "),
       call. = FALSE)
}

rng <- function(x) {
  x <- x[!is.na(x)]
  if (!length(x)) c(NA_real_, NA_real_) else range(x)
}

derived_rows <- function(cat) {
  src <- DERIVED[[cat]]
  do.call(rbind, lapply(sort(unique(d[[cat]])), function(k) {
    v <- d[[src]][d[[cat]] == k]
    r <- rng(v)
    data.frame(
      category = cat, level = k, n = length(v), source = src,
      src_min = r[1], src_max = r[2], src_missing = sum(is.na(v))
    )
  }))
}

knitr::kable(
  do.call(rbind, lapply(names(DERIVED), derived_rows)),
  row.names = FALSE, digits = 4,
  caption = "Each derived category against the variable it came from"
)
```

**Read the ranges before reading anything below.** They must not overlap between
levels of the same category, and they must ascend with the level. If a
derivation is wrong, every stratified estimate downstream is wrong in a way that
still looks entirely plausible, and this table is where that shows.

Any `src_missing` above zero is worth naming in the report: those patients are
in that stratum because their source value is unknown, not because they belong
to it clinically.

## Estimation

`hzr_kaplan()` takes vectors, not a formula with a strata term, so a stratified
fit is a split and a call per stratum.

The confidence level is SAS's `%KAPLAN` `CLEVEL` default of `0.68268948` — one
standard deviation, **not** 95%. This is a SAS convention rather than a study
choice, so it is not marked as an edit; change it only if your job overrides
`CLEVEL`, and say so in the report if you do.

```{r}
#| label: km-helpers
CLEVEL <- 0.68268948

# your time and status variables.

km_fit <- function(dd) {
  ok <- !is.na(dd[[TIME]]) & !is.na(dd[[STATUS]])
  out <- hzr_kaplan(time       = as.numeric(dd[[TIME]][ok]),
                    status     = as.numeric(dd[[STATUS]][ok]),
                    conf_level = CLEVEL)
  # End of follow-up, carried from the DATA. It cannot be recovered from the
  # life table: that has one row per EVENT time, so patients censored after the
  # last death extend follow-up without adding a row. max(out$time) is the last
  # event, which can be years short of the last observation.
  out$max_fup <- max(as.numeric(dd[[TIME]][ok]))
  out
}

# Bookkeeping columns km_fit()/km_by() attach for the code's benefit. They are
# constant down every row and must not appear in a table someone reads. Kept as
# columns rather than attributes because split()/rbind() drop attributes, and
# losing max_fup silently would restore the extrapolation bug it exists to stop.
INTERNAL_COLS <- c("max_fup", "n")
readable <- function(x) x[, setdiff(names(x), INTERNAL_COLS), drop = FALSE]

# Levels of a stratification variable that are not clean integers. A missing
# binary that was MEAN-IMPUTED upstream is stored as a third factor level: a
# value such as 0.714047380714047 is the observed proportion, not an
# observation, and one imputation constant can propagate into several derived
# variables.
#
# Splitting on such a variable produces a stratum that LOOKS like a clinical
# group, has a plausible survival curve, and is really "this covariate was
# missing".
#
# SAS %KAPLAN DROPS these rows, reporting "NOTE: n observations with invalid
# ... strata values were deleted".
#
# THIS TEMPLATE DOES NOT DROP THEM. They are carried as an explicit "Unknown"
# stratum: "we do not know this patient's value" is a fact about the cohort
# worth showing, and the stratified table then accounts for every patient
# rather than quietly shedding some. This is a DELIBERATE DIVERGENCE from the
# SAS baseline, legitimate because an R_hazard job does the analysis; matching
# SAS is a parity job's business and applies only where a stored SAS answer
# exists.
#
# Consequence to state in any report built from this file: the Unknown
# stratum's curve is NOT a clinical subgroup and must not be read as one. It is
# a missingness stratum, and a difference between it and the others is evidence
# about who had the covariate recorded, not about the covariate.
# imputed_levels() is hvtiRutilities' (>= 1.1.4). This file carried a local copy
# until then; it was identical, so the copy only meant two places to fix.
UNKNOWN <- "Unknown"

# Stratified fits, one per level, with the level carried on the result so the
# strata cannot be silently transposed when the tables are stacked.
km_by <- function(dd, var) {
  raw    <- as.character(dd[[var]])
  imp_lv <- imputed_levels(dd[[var]])

  # Both imputed levels and outright NA mean "we do not know". They are folded
  # into one Unknown stratum, but their provenance is reported separately below
  # so the fold is never silent.
  lab <- raw
  lab[raw %in% imp_lv | is.na(raw)] <- UNKNOWN

  # Unknown sorts LAST, whatever the observed levels are called.
  obs <- sort(setdiff(unique(lab), UNKNOWN))
  lab <- factor(lab, levels = c(obs, if (UNKNOWN %in% lab) UNKNOWN))

  parts <- split(dd, lab)
  out <- do.call(rbind, lapply(levels(lab), function(lv) {
    p <- parts[[lv]]
    # Carry the number of patients FITTED. Do not report n_risk[1] as the
    # stratum size: that is the number at risk at the FIRST EVENT time, and any
    # patient censored before the stratum's first death is missing from it. The
    # two agree only when a stratum has a death at t=0, which large strata do
    # and small ones do not -- so the error appears exactly where it is hardest
    # to notice.
    cbind(stratum = lv,
          n       = sum(!is.na(p[[TIME]]) & !is.na(p[[STATUS]])),
          km_fit(p))
  }))

  src <- rbind(
    if (length(imp_lv)) {
      data.frame(
        source = imp_lv,
        n = vapply(imp_lv, function(l) sum(raw == l, na.rm = TRUE), integer(1)),
        reason = "mean-imputed level, not an observed value"
      )
    },
    if (anyNA(raw)) {
      data.frame(source = "NA", n = sum(is.na(raw)), reason = "missing outright")
    }
  )
  attr(out, "unknown_sources") <- if (!is.null(src)) `rownames<-`(src, NULL) else NULL
  out
}

# Print what went into Unknown alongside the estimates, never instead of them.
report_unknown <- function(fit, var) {
  src <- attr(fit, "unknown_sources")
  if (is.null(src)) return(invisible(NULL))
  cap <- paste0(
    "What the ", var, " \"", UNKNOWN, "\" stratum is made of -- a missingness ",
    "group, NOT a clinical subgroup. SAS %KAPLAN drops these rows."
  )
  print(knitr::kable(src, row.names = FALSE, caption = cap))
}
```

## Overall

```{r}
#| label: km-overall
km <- km_fit(d)
knitr::kable(head(readable(km), 20), caption = "Overall life table (first 20 rows)")
ac_art <- hvtiRtemplates:::.attach_handoff_lineage(
  list(overall = km),
  data = .provenance_data,
  analysis = list(
    time = list(variable = TIME),
    event = list(variable = STATUS, event = 1L, censored = 0L)
  ),
  cohort = cc
)
saveRDS(ac_art, set_path("estimates", "ac.rds"))
```

Survival at reporting horizons, read off the step function as the last estimate
at or before each horizon.

```{r}
#| label: horizons
# your reporting horizons, in the time unit of TIME (years here).
# A wrong grid does not error — it produces a plausible-looking table.

at_horizon <- function(fit, g) {
  i <- which(fit$time <= g)
  # BEFORE the first event: the life table starts at the first event time, so a
  # stratum with no event before `g` has no row at or below it. Survival there
  # is 1, not missing -- NA_real_ would read as "no data" in a stratum that
  # simply had no deaths yet, and small strata are where that happens.
  beyond <- g > fit$max_fup[1]
  if (!length(i)) return(if (beyond) NA_real_ else 1)
  s <- fit$survival[max(i)]
  # BEYOND THE END OF FOLLOW-UP the KM estimate is NOT DEFINED, and carrying the
  # last value forward silently reports a horizon nobody was observed to.
  # Return NA instead -- EXCEPT when survival has already reached 0, which is
  # absorbing: everyone has died, so 0 is genuinely known thereafter.
  # Compare against max_fup, NOT max(fit$time): the latter is the last EVENT.
  if (beyond && s > 0) return(NA_real_)
  s
}

knitr::kable(data.frame(
  horizon  = labs,
  survival = round(100 * vapply(grid, at_horizon, numeric(1), fit = km), 1)
), caption = "Overall survival (%) at reporting horizons")
```

## Stratified

<!-- EDIT: one block per %KAPLAN stratification in your SAS job. Add or remove
     blocks to match, or delete the section if the job has none. -->

```{r}
#| label: km-strata
stratum_row <- function(s) {
  data.frame(
    stratum  = s$stratum[1],
    n        = s$n[1],
    n_events = sum(s$n_event),
    setNames(as.list(round(100 * vapply(grid, at_horizon, numeric(1), fit = s), 1)), labs),
    check.names = FALSE
  )
}

# your stratification variables.
for (v in STRATA) {
  fit <- km_by(d, v)
  print(knitr::kable(
    do.call(rbind, lapply(split(fit, fit$stratum), stratum_row)),
    row.names = FALSE,
    caption = paste0("Survival (%) by ", v)
  ))
  report_unknown(fit, v)
}
```

Nested stratification — one stratum variable within the levels of another — is
the same call applied to each subset.

```{r}
#| label: km-nested
#| eval: false
# EDIT: your nesting, or delete this chunk if the job has none. It is
# `eval: false` because the template ships no second category; turn it on once
# `cat_b` exists.
for (outer in sort(unique(d$cat_b))) {
  sub <- d[d$cat_b == outer, , drop = FALSE]
  fit <- km_by(sub, "cat_a")
  print(knitr::kable(
    do.call(rbind, lapply(split(fit, fit$stratum), function(s) data.frame(
      cat_a    = s$stratum[1],
      n        = s$n[1],
      n_events = sum(s$n_event),
      setNames(as.list(round(100 * vapply(grid, at_horizon, numeric(1), fit = s), 1)), labs),
      check.names = FALSE
    ))),
    row.names = FALSE,
    caption = paste0("Survival (%) by cat_a within cat_b = ", outer)
  ))
  report_unknown(fit, "cat_a")
}
```

```{r}
#| label: provenance
#| echo: false
#| results: asis
cat(hvtiRtemplates:::.embed_provenance(
  .in,
  data = .provenance_data,
  extra = list(
    subject = SUBJECT,
    type = TYPE,
    analysis = list(
      time = list(variable = TIME),
      event = list(variable = STATUS, event = 1L, censored = 0L)
    ),
    cohort = cc
  )
))
```
