---
title: "Goodness of follow-up, graph"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in.
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    page-layout: full
    grid:
      body-width: 2000px
      sidebar-width: 250px
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (an older graphs/dp.gfup job) and list
     the panels it drew: all deaths, systematic deaths, an event panel. -->

Replaces `graphs/dp.gfup`: list the follow-up panels it draws.

**Deprecated.** `dp-gfup` is deprecated in favor of `dc-gfup`, and will be
removed in a later release. For the same figure, start a `dc-gfup` job with
`add_job("dc.gfup", subject = "cohort", type = "eda")`: it draws these panels
with the same choices and the same call to `hv_followup_panels()`, beside the
follow-up tables, saved as `dc-gfup-*.png` rather than `dp-gfup-*.png`.

A `dp-gfup` job draws the goodness-of-follow-up figure. Every patient is one
point: the year of operation across, the follow-up recorded for them up, blue
while alive and red once dead. The dashed diagonal is the follow-up a patient
**could** have had, from their operation to the close date. A censored patient
far below the diagonal is someone whose follow-up stopped early, and a band of
them is a follow-up problem to settle before any time-related analysis.

The engine is `hvtiPlotR::hv_followup_panels()`, which the EDA report also
calls, so its follow-up panels are these. The tables that go with this figure
are the `dc-gfup` job, which reads the same data and rows.

```{r}
#| label: setup
# unnumbered: loads packages and checks versions only
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRtemplates:::.find_study_root(if (is.null(.in)) getwd() else dirname(.in))
.provenance_data <- list()
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(hvtiRutilities)
  library(hvtiPlotR)
  library(ggplot2)
})
if (utils::packageVersion("hvtiPlotR") < "2.8.0") {
  stop("This job needs hvtiPlotR >= 2.8.0 for hv_followup_panels() and scale_color_hv(); ",
       utils::packageVersion("hvtiPlotR"), " is installed.", call. = FALSE)
}
```

```{r}
#| label: guard-edits
#| results: asis
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

```{r}
#| label: guard-partial
#| results: asis
# unnumbered: a callout, printed only when part of the job is left out
# To render a job you have not finished, leave a chunk out with the chunk
# option skip, giving the reason in quotes, or call hvtiRtemplates::stop_here()
# in a chunk to leave out everything below it. A draft lists each one here; a
# final render refuses them, as it refuses an EDIT marker. ?stop_here has more.
hvtiRtemplates:::.guard_partial(knitr::current_input())
```

```{r}
#| label: set
SUBJECT <- "cohort"
TYPE    <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- hvtiRtemplates:::.job_name_fields(.current)
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject,
         "', type '", .name_type, "'), but declares SUBJECT = \"", SUBJECT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}

# Saves a figure as <name>.png and <name>.pdf in this set's graphs/ folder (or `kind`'s),
# under the SAVE_FIGURES and FIGURES study choices.
save_figure <- function(plot, name, width = 6, height = 4, kind = "graphs", linked = FALSE) {
  hvtiRtemplates:::.save_figure(plot, set_path(kind, paste0(name, ".png")), width, height,
                                SAVE_FIGURES, FIGURES, linked)
}
```
## Study choices

Set the values in this chunk before rendering. Left as they are, they draw the
all-deaths panel for the whole study dataset, so the job runs before it is tuned.

```{r}
#| label: edit-study-choices
# EDIT: the registered dataset this job reads ("built" is the study dataset).
DATASET <- "built"

# EDIT: an hvtiRdatabuild analysis set, or NULL to read the whole dataset.
ANALYSIS_SET <- NULL

# EDIT: rows to keep, dplyr::filter() style, or NULL to keep every row:
#   WHERE <- quote(age >= 18)
#   WHERE <- rlang::exprs(age >= 18, hx_chf == 1)
WHERE <- NULL

# EDIT: the patient identifier. Without "ccfid" the job uses MRN, then eMRN;
# name another column, such as "randid", if the study uses one.
ID <- "ccfid"

# EDIT: what makes a row unique; one row per patient unless repeated measures
# add their visit time or date, for example KEY <- c(ID, "iv_echo").
KEY <- ID

# Read the same data and rows as this study's dc-gfup job, so the figure and
# the tables describe the same patients.

# EDIT: the years-since-origin interval to the operation, and that origin. The
# origin differs between studies, and a wrong one slides every point along the
# x-axis without any other symptom, which is why the operation years are
# checked and printed below.
OPYRS <- "iv_opyrs"
ORIGIN_YEAR <- 1990

# EDIT: the close date of follow-up, as.Date("YYYY-MM-DD"). NULL estimates it
# as the latest operation date plus follow-up in the data, which is the date
# follow-up is known to reach, not the date it was closed; the report says
# which one it drew.
CLOSE_DATE <- NULL

# EDIT: one entry per death panel. `status` is the 1/0 death indicator and
# `time` its follow-up in years. A study that ascertains deaths two ways draws
# both, for example
#   systematic = list(status = "deads", time = "iv_deads", title = "Systematic deaths")
PANELS <- list(
  all = list(status = "dead", time = "iv_dead", title = "All deaths")
)

# EDIT: optional event panels, one entry per non-fatal event. `event` and
# `time` are the event's indicator and interval; `death` and `death_time` the
# death it competes with. A flagged event counts only when it comes strictly
# before death, so ties go to death. list() draws none. For example
#   reop = list(event = "ev_reop", time = "iv_reop", death = "dead",
#               death_time = "iv_dead", label = "Reoperation")
EVENTS <- list()

# EDIT: point transparency. 0.5 lets a dense cohort read as a distribution
# rather than a blob; raise it for a small one.
ALPHA <- 0.5

# EDIT: NULL draws the house colors through hvtiPlotR::scale_color_hv(): dead
# vermillion, alive blue and a non-fatal event green, all colorblind safe. Name
# all three to choose your own; c(alive = "#377EB8", dead = "#E41A1C",
# event = "#4DAF4A") is ColorBrewer Set1, what these jobs drew before 1.2.3.
COLORS <- NULL

# Each figure is saved to graphs/ as a PNG (for Word) and a PDF (for the publisher).
# SAVE_FIGURES <- FALSE saves neither; FIGURES keeps only the figures whose names
# start with one of its entries, e.g. FIGURES <- c("hp-survival"). The names are the
# file names, listed for each template in the templates README.
SAVE_FIGURES <- TRUE
FIGURES <- NULL
```

## Data

<!-- How this job finds its data. Read this before touching the chunk below.

     A study can offer a job data three ways, and `_study.yml` says which exist:

     1. The study dataset: the file named by `built:`. `DATASET <- "built"`
        with `ANALYSIS_SET <- NULL`.
     2. An additional dataset: any file declared under `additional_datasets:`,
        for example a column subset written for R. Set `DATASET` to its
        name there and `ANALYSIS_SET <- NULL`. It is read as it stands: only
        `WHERE` removes rows.
     3. An analysis set: declared under `analysis_sets:` with the columns kept
        and the rows excluded, then written once with
        `hvtiRdatabuild::write_analysis_set()`. Set `ANALYSIS_SET` to its name
        and leave `DATASET` as "built", because a set is always cut from the
        study dataset. The report shows its exclusions in order.

     A name under `additional_datasets:` is not an analysis set, so putting it
     in `ANALYSIS_SET` stops with "No analysis set". Whichever route is taken,
     the report prints what it read, the identifier it used, the rows each
     `WHERE` condition removed and the rows and patients kept, so a reader
     never has to ask which data a table describes. -->

```{r}
#| label: tbl-data
#| tbl-cap: "The data this job read"
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
.cfg <- study_config(start = .root)
job_data <- hvtiRtemplates::read_job_data(.cfg, dataset = DATASET, analysis_set = ANALYSIS_SET,
                                          where = WHERE, id = ID, key = KEY)
d <- job_data$data
.provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(job_data$provenance))
knitr::kable(job_data$record, col.names = c("Data", ""))
```

```{r}
#| label: data-attrition
#| results: asis
# unnumbered: its child chunk carries its own label and caption
if (!is.null(job_data$attrition)) {
  .fence <- strrep("`", 3)
  cat(knitr::knit_child(text = c(
    paste0(.fence, "{r}"), "#| label: tbl-data-attrition",
    paste0("#| tbl-cap: ", encodeString(paste0("Analysis set `", ANALYSIS_SET, "`: exclusions, in order"), quote = "\"")),
    "knitr::kable(job_data$attrition)", .fence
  ), envir = environment(), quiet = TRUE), sep = "\n")
}
```



## Operation years and the close date

```{r}
#| label: window
# unnumbered: computes the follow-up window; the table chunk below shows it
# hvtiPlotR::hv_followup_panels() checks every panel and event against the data,
# names every missing column at once, refuses a name used twice and the
# wrong-origin mistake, and works out the window the panels share: from
# 1 January of ORIGIN_YEAR, where hv_followup() draws its diagonal, to the last
# operation. The EDA report calls the same function, so its panels are these.
fp <- hvtiPlotR::hv_followup_panels(d, opyrs_col = OPYRS, origin_year = ORIGIN_YEAR,
                                    panels = PANELS, events = EVENTS, close_date = CLOSE_DATE)
close_date <- fp$meta$close_date
# Name the edit point a reader can find in this file, not the function's argument.
close_source <- if (is.null(CLOSE_DATE)) fp$meta$close_source else "set in CLOSE_DATE"

```

```{r}
#| label: tbl-window
#| tbl-cap: "Operation years and the close date"
knitr::kable(data.frame(
  quantity = c("patients", "operation year missing", "first operation", "last operation",
               "close date", "close date is"),
  value = c(fp$meta$n_obs, fp$meta$n_opyrs_missing, format(fp$meta$first_operation),
            format(fp$meta$study_end), format(close_date), close_source)
))
```

## Figures

A point below the diagonal is follow-up short of what the close date allows. A
dead patient's point is where the death was recorded, so it is expected below
the line; it is the **blue** points drifting below it that need an explanation.

```{r}
#| label: figures
#| results: asis
# unnumbered: each child chunk below carries its own label and caption
# The image link is relative to THIS document, which always sits in graphs/, so
# it is <set>/<file> whichever directory the render executes from. The file
# itself is written through set_path(), which resolves from the project root.
.fence <- strrep("`", 3)
.seen <- new.env()
.child <- function(label, caption, code) {
  label <- gsub("(^-+|-+$)", "", gsub("[^a-z0-9]+", "-", tolower(label)))
  if (!is.null(.seen[[label]])) stop("Two outputs would share the label ", label, ".", call. = FALSE)
  .seen[[label]] <- TRUE
  .opt <- paste0("#| ", sub("-.*$", "", label), "-cap: ", encodeString(caption, quote = "\""))
  cat(knitr::knit_child(text = c(paste0(.fence, "{r}"), paste0("#| label: ", label), .opt, code, .fence),
                        envir = parent.frame(), quiet = TRUE), sep = "\n")
  cat("\n")
}
show_figure <- function(p, fname, title, nm) {
  p <- p + labs(x = "Year of operation", y = "Follow-up (years)", color = NULL, shape = NULL) +
    theme_hv_manuscript()
  save_figure(p, sub("[.]png$", "", fname), width = 7, height = 6, linked = TRUE)
  .link <- file.path(paste0(SUBJECT, "-", TYPE), fname)
  .child(paste("fig gfup", nm), paste0("Follow-up against year of operation, ", title),
         "knitr::include_graphics(.link, error = FALSE)")
}
if (exists("COLOURS", inherits = FALSE)) stop("COLOURS is now COLORS: rename it in edit-study-choices ",
                                              "(study-choices in a job made before that rename).", call. = FALSE)
if (!is.null(COLORS) && !(is.character(COLORS) && !anyNA(COLORS) && length(COLORS) == 3L &&
                            setequal(names(COLORS), c("alive", "dead", "event")))) {
  stop("COLORS must be NULL or name exactly alive, dead and event.", call. = FALSE)
}
# Dead or Death is the event and Alive or No event the censored level, so a
# non-fatal event takes the house rule's next color, green.
status_colors <- function(levels) {
  if (is.null(COLORS)) return(scale_color_hv(event = c("Dead", "Death"), censored = c("Alive", "No event")))
  keys <- if (length(levels) == 2L) c("alive", "dead") else c("alive", "event", "dead")
  scale_color_manual(values = stats::setNames(COLORS[keys], levels))
}
plots <- plot(fp, alpha = ALPHA)
panel_counts <- list()
for (i in seq_len(nrow(fp$data))) {
  row <- fp$data[i, ]
  nm <- row$panel
  panel_counts[[nm]] <- list(drawn = row$n_drawn, excluded = row$n_excluded)
  cat("\n### ", row$title, "\n\n", row$n_drawn, " patients drawn; ", row$n_excluded,
      " missing a column this panel needs.\n\n", sep = "")
  if (row$type == "followup") {
    p <- plots[[nm]] +
      status_colors(c("Alive", "Dead")) +
      scale_shape_manual(values = c(Alive = 1, Dead = 4))
  } else {
    levels <- c("No event", row$title, "Death")
    p <- plots[[nm]] +
      status_colors(levels) +
      scale_shape_manual(values = stats::setNames(c(1, 2, 4), levels))
  }
  show_figure(p, paste0("dp-gfup-", nm, ".png"), row$title, nm)
}
```

```{r}
#| label: provenance
#| echo: false
#| results: asis
cat(hvtiRtemplates:::.embed_provenance(
  .in,
  data = .provenance_data,
  extra = list(
    subject = SUBJECT,
    type = TYPE,
    followup = list(origin_year = ORIGIN_YEAR, close_date = format(close_date),
                    close_date_source = close_source, panels = panel_counts)
  )
))
```
