---
title: "Goodness of follow-up, graph"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in.
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (an older graphs/dp.gfup job) and list
     the panels it drew: all deaths, systematic deaths, an event panel. -->

Replaces `graphs/dp.gfup`: list the follow-up panels it draws.

A `dp-gfup` job draws the goodness-of-follow-up figure. Every patient is one
point: the year of operation across, the follow-up recorded for them up, blue
while alive and red once dead. The dashed diagonal is the follow-up a patient
**could** have had, from their operation to the close date. A censored patient
far below the diagonal is someone whose follow-up stopped early, and a band of
them is a follow-up problem to settle before any time-related analysis.

The engine is `hvtiPlotR::hv_followup()`. The tables that go with this figure
are the `dc-gfup` job, which reads the same analysis set.

```{r}
#| label: setup
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRutilities::study_root(if (is.null(.in)) getwd() else dirname(.in))
.provenance_data <- list()
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(hvtiRutilities)
  library(hvtiPlotR)
  library(ggplot2)
})
if (utils::packageVersion("hvtiPlotR") < "2.7.14") {
  stop("This job needs hvtiPlotR >= 2.7.14 for hv_followup(); ",
       utils::packageVersion("hvtiPlotR"), " is installed.", call. = FALSE)
}
if (utils::packageVersion("hvtiRdatabuild") < "0.2.1") {
  stop("This job needs hvtiRdatabuild >= 0.2.1 for read_analysis_set(); ",
       utils::packageVersion("hvtiRdatabuild"), " is installed.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

```{r}
#| label: set
SUBJECT <- "cohort"
TYPE    <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject,
         "', type '", .name_type, "'), but declares SUBJECT = \"", SUBJECT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```
## Study choices

Set the values in this chunk before rendering. Left as they are, they draw the
all-deaths panel for the `eda` analysis set, so the job runs before it is tuned.

```{r}
#| label: study-choices
# EDIT: the built dataset this job reads. "study" is the file named by built:
# in _study.yml; any other name must be declared under additional_datasets:.
# Analysis sets are derived from "study" only, so another dataset needs
# ANALYSIS_SET <- NULL. Read the same set as this study's dc-gfup job, so the
# figure and the tables describe the same patients.
DATASET <- "study"
ANALYSIS_SET <- "eda"

# EDIT: the years-since-origin interval to the operation, and that origin. The
# origin differs between studies, and a wrong one slides every point along the
# x-axis without any other symptom, which is why the operation years are
# checked and printed below.
OPYRS <- "iv_opyrs"
ORIGIN_YEAR <- 1990

# EDIT: the close date of follow-up, as.Date("YYYY-MM-DD"). NULL estimates it
# as the latest operation date plus follow-up in the data, which is the date
# follow-up is known to reach, not the date it was closed; the report says
# which one it drew.
CLOSE_DATE <- NULL

# EDIT: one entry per death panel. `status` is the 1/0 death indicator and
# `time` its follow-up in years. A study that ascertains deaths two ways draws
# both, for example
#   systematic = list(status = "deads", time = "iv_deads", title = "Systematic deaths")
PANELS <- list(
  all = list(status = "dead", time = "iv_dead", title = "All deaths")
)

# EDIT: optional event panels, one entry per non-fatal event. `event` and
# `time` are the event's indicator and interval; `death` and `death_time` the
# death it competes with. A flagged event counts only when it comes strictly
# before death, so ties go to death. list() draws none. For example
#   reop = list(event = "ev_reop", time = "iv_reop", death = "dead",
#               death_time = "iv_dead", label = "Reoperation")
EVENTS <- list()

# EDIT: point transparency. 0.5 lets a dense cohort read as a distribution
# rather than a blob; raise it for a small one.
ALPHA <- 0.5

# Red is dead and blue is alive, in every job. A non-fatal event is green.
# These are ColorBrewer Set1's red, blue and green.
COLOURS <- c(alive = "#377EB8", dead = "#E41A1C", event = "#4DAF4A")
```

## Data

<!-- How this job finds its data. Read this before touching the chunk below.

     A study can offer a job data three ways, and `_study.yml` says which exist:

     1. The study dataset: the file named by `built:`. `DATASET <- "study"`
        with `ANALYSIS_SET <- NULL`.
     2. An additional dataset: any file declared under `additional_datasets:`,
        for example a column subset written for R. Set `DATASET` to its
        name there and `ANALYSIS_SET <- NULL`. It is read as it stands: no
        rows are excluded.
     3. An analysis set: declared under `analysis_sets:` with the columns kept
        and the rows excluded, then written once with
        `hvtiRdatabuild::write_analysis_set()`. Set `ANALYSIS_SET` to its name
        and leave `DATASET` as "study", because a set is always cut from the
        study dataset. The report shows its exclusions in order.

     A name under `additional_datasets:` is not an analysis set, so putting it
     in `ANALYSIS_SET` stops with "No analysis set". Whichever route is taken,
     the report prints the dataset, file and dimensions it read, so a reader
     never has to ask which data a table describes. -->

```{r}
#| label: data
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
if (is.null(ANALYSIS_SET)) {
  if (!is.character(DATASET) || length(DATASET) != 1L || is.na(DATASET) || !nzchar(DATASET)) {
    stop("DATASET must be set to one dataset registered in _study.yml, such as \"study\". ",
         "In a migrated job, see the migration report beside this job.", call. = FALSE)
  }
  .cfg <- study_config(start = .root)
  .data_read <- hvtiRtemplates:::.provenance_read(
    DATASET, .cfg, function() read_built(cfg = .cfg, dataset = DATASET)
  )
  d <- .data_read$value
  .provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(.data_read$record))
} else if (!identical(DATASET, "study")) {
  stop("An analysis set is written from the study dataset, not `", DATASET,
       "`. Set ANALYSIS_SET <- NULL to read `", DATASET, "` whole.", call. = FALSE)
} else {
  .cfg <- study_config(start = .root)
  .analysis_set_path <- file.path(study_dir("datasets", .root), paste0(ANALYSIS_SET, ".parquet"))
  .data_read <- hvtiRtemplates:::.provenance_file_read(
    paste0("analysis_set:", ANALYSIS_SET), .analysis_set_path, .cfg,
    function() hvtiRdatabuild::read_analysis_set(ANALYSIS_SET, cfg = .cfg),
    role = paste0("analysis_set:", ANALYSIS_SET)
  )
  d <- .data_read$value
  .provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(.data_read$record))
  knitr::kable(attr(d, "attrition"), caption = paste0(
    "Analysis set `", ANALYSIS_SET, "`: exclusions, in order"
  ))
}
cat("Data read: ", if (is.null(ANALYSIS_SET)) {
  paste0("dataset `", DATASET, "` (",
         basename(hvtiRutilities::built_path(dataset = DATASET)), ")")
} else {
  paste0("analysis set `", ANALYSIS_SET, "` of the study dataset")
}, ", ", nrow(d), " rows, ", ncol(d), " columns.\n", sep = "")
```



## Operation years and the close date

```{r}
#| label: window
is_name <- function(x) is.character(x) && length(x) == 1L && !is.na(x) && nzchar(x)
if (!is_name(OPYRS)) stop("OPYRS must name one column.", call. = FALSE)
if (!is.numeric(ORIGIN_YEAR) || length(ORIGIN_YEAR) != 1L || is.na(ORIGIN_YEAR) ||
      ORIGIN_YEAR != round(ORIGIN_YEAR)) {
  stop("ORIGIN_YEAR must be one whole calendar year.", call. = FALSE)
}
if (!is.list(PANELS) || !length(PANELS) || is.null(names(PANELS)) || any(!nzchar(names(PANELS)))) {
  stop("PANELS must be a named list with at least one entry.", call. = FALSE)
}
if (!is.list(EVENTS) || (length(EVENTS) && (is.null(names(EVENTS)) || any(!nzchar(names(EVENTS)))))) {
  stop("EVENTS must be a named list, or list() for none.", call. = FALSE)
}
for (nm in names(PANELS)) {
  if (!is_name(PANELS[[nm]]$status) || !is_name(PANELS[[nm]]$time)) {
    stop("PANELS entry `", nm, "` needs one `status` and one `time` column.", call. = FALSE)
  }
}
for (nm in names(EVENTS)) {
  e <- EVENTS[[nm]]
  if (!all(vapply(e[c("event", "time", "death", "death_time")], is_name, logical(1L)))) {
    stop("EVENTS entry `", nm, "` needs one `event`, `time`, `death` and `death_time` column.",
         call. = FALSE)
  }
}

# Every column named anywhere above, checked at once, so a job with three
# misspellings says so in one render rather than three.
status_cols <- unique(c(vapply(PANELS, `[[`, "", "status"),
                        unlist(lapply(EVENTS, `[`, c("event", "death")), use.names = FALSE)))
time_cols <- unique(c(vapply(PANELS, `[[`, "", "time"),
                      unlist(lapply(EVENTS, `[`, c("time", "death_time")), use.names = FALSE)))
unknown <- setdiff(c(OPYRS, status_cols, time_cols), names(d))
if (length(unknown)) {
  stop("Column(s) not in the data: ", paste(unknown, collapse = ", "), ".", call. = FALSE)
}
not_numeric <- Filter(function(v) !is.numeric(d[[v]]), c(OPYRS, time_cols))
if (length(not_numeric)) {
  stop("Interval column(s) must be numeric years: ", paste(not_numeric, collapse = ", "), ".",
       call. = FALSE)
}
not_binary <- Filter(function(v) {
  x <- d[[v]]
  (!is.numeric(x) && !is.logical(x)) || any(!is.na(x) & !x %in% c(0, 1))
}, status_cols)
if (length(not_binary)) {
  stop("Indicator column(s) must be 1/0 or logical: ", paste(not_binary, collapse = ", "), ".",
       call. = FALSE)
}

opyrs <- d[[OPYRS]]
if (all(is.na(opyrs))) stop("`", OPYRS, "` is missing for every patient.", call. = FALSE)
# hv_followup() starts the diagonal at ORIGIN_YEAR, so the study window has to
# start there too. An operation before the origin would sit left of the line.
if (any(opyrs < 0, na.rm = TRUE)) {
  stop(sum(opyrs < 0, na.rm = TRUE), " patient(s) have a negative `", OPYRS, "`, an ",
       "operation before ORIGIN_YEAR = ", ORIGIN_YEAR, ". Check the origin.", call. = FALSE)
}
# A plausible window, not a precise one: it catches the wrong-origin mistake,
# which lands decades out.
.this_year <- as.integer(format(Sys.Date(), "%Y"))
op_year <- ORIGIN_YEAR + opyrs
if (ORIGIN_YEAR < 1900 || any(op_year > .this_year + 1, na.rm = TRUE)) {
  stop("Operations fall outside 1900 to ", .this_year + 1, " (", floor(min(op_year, na.rm = TRUE)),
       " to ", floor(max(op_year, na.rm = TRUE)), "). Check ORIGIN_YEAR.", call. = FALSE)
}

year_days <- 365.2425
study_start <- as.Date(paste0(ORIGIN_YEAR, "-01-01"))
study_end <- study_start + round(max(opyrs, na.rm = TRUE) * year_days)
if (is.null(CLOSE_DATE)) {
  reach <- max(vapply(time_cols, function(v) max(opyrs + d[[v]], na.rm = TRUE), numeric(1L)))
  close_date <- max(study_end, study_start + round(reach * year_days))
  close_source <- "estimated: the latest operation plus follow-up in the data"
} else {
  close_date <- as.Date(CLOSE_DATE)
  if (length(close_date) != 1L || is.na(close_date)) {
    stop("CLOSE_DATE must be NULL or one date, as.Date(\"YYYY-MM-DD\").", call. = FALSE)
  }
  if (close_date < study_end) {
    stop("CLOSE_DATE (", close_date, ") is before the last operation (", study_end, ").",
         call. = FALSE)
  }
  close_source <- "set in CLOSE_DATE"
}

knitr::kable(data.frame(
  quantity = c("patients", "operation year missing", "first operation", "last operation",
               "close date", "close date is"),
  value = c(nrow(d), sum(is.na(opyrs)), format(study_start + round(min(opyrs, na.rm = TRUE) * year_days)),
            format(study_end), format(close_date), close_source)
))
```

## Figures

A point below the diagonal is follow-up short of what the close date allows. A
dead patient's point is where the death was recorded, so it is expected below
the line; it is the **blue** points drifting below it that need an explanation.

```{r}
#| label: figures
#| results: asis
# The image link is relative to THIS document, which always sits in graphs/, so
# it is <set>/<file> whichever directory the render executes from. The file
# itself is written through set_path(), which resolves from the project root.
show_figure <- function(p, fname, title) {
  p <- p + labs(x = "Year of operation", y = "Follow-up (years)", colour = NULL, shape = NULL) +
    theme_hv_manuscript()
  png(set_path("graphs", fname), width = 7, height = 6, units = "in", res = 150)
  print(p)
  invisible(dev.off())
  cat("![", title, "](", file.path(paste0(SUBJECT, "-", TYPE), fname), ")\n\n", sep = "")
}
followup_args <- list(data = d, iv_opyrs_col = OPYRS, origin_year = ORIGIN_YEAR,
                      study_start = study_start, study_end = study_end,
                      close_date = close_date, tolower_names = FALSE)
panel_counts <- list()

for (nm in names(PANELS)) {
  pn <- PANELS[[nm]]
  gf <- do.call(hv_followup, c(followup_args, list(
    death_col = pn$status, death_time_col = pn$time, death_levels = c("Alive", "Dead")
  )))
  title <- if (is.null(pn$title)) nm else pn$title
  panel_counts[[nm]] <- list(drawn = gf$meta$n_patients, excluded = gf$meta$n_excluded)
  cat("\n### ", title, "\n\n", gf$meta$n_patients, " of ", gf$meta$n_input,
      " patients drawn; ", gf$meta$n_excluded, " missing a column this panel needs.\n\n", sep = "")
  p <- plot(gf, alpha = ALPHA) +
    scale_colour_manual(values = c(Alive = COLOURS[["alive"]], Dead = COLOURS[["dead"]])) +
    scale_shape_manual(values = c(Alive = 1, Dead = 4))
  show_figure(p, paste0("dp-gfup-", nm, ".png"), title)
}

for (nm in names(EVENTS)) {
  e <- EVENTS[[nm]]
  label <- if (is.null(e$label)) nm else e$label
  levels <- c("No event", label, "Death")
  gf <- do.call(hv_followup, c(followup_args, list(
    death_col = e$death, death_time_col = e$death_time, event_col = e$event,
    event_time_col = e$time, event_levels = levels
  )))
  panel_counts[[nm]] <- list(drawn = gf$meta$n_event_patients, excluded = gf$meta$n_event_excluded)
  cat("\n### ", label, "\n\n", gf$meta$n_event_patients, " patients drawn; ",
      gf$meta$n_event_excluded, " missing a column this panel needs.\n\n", sep = "")
  p <- plot(gf, type = "event", alpha = ALPHA) +
    scale_colour_manual(values = stats::setNames(COLOURS[c("alive", "event", "dead")], levels)) +
    scale_shape_manual(values = stats::setNames(c(1, 2, 4), levels))
  show_figure(p, paste0("dp-gfup-", nm, ".png"), label)
}
```

```{r}
#| label: provenance
#| echo: false
#| results: asis
cat(hvtiRtemplates:::.embed_provenance(
  .in,
  data = .provenance_data,
  extra = list(
    subject = SUBJECT,
    type = TYPE,
    followup = list(origin_year = ORIGIN_YEAR, close_date = format(close_date),
                    close_date_source = close_source, panels = panel_counts)
  )
))
```
