---
title: "Trends over operation year"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in:
# rendered as a standalone document the project's format block never applies,
# which is how server-rendered reports end up with sibling _files/ trees instead
# of being self-contained.
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (an older graphs/dp.trends job) and list
     the figures it drew. A trends job usually draws several, and each one
     becomes one entry in TRENDS below. -->

Replaces `graphs/dp.trends`: list the trend figures it draws.

A `dp-trends` job shows how the cohort changed over the years of operation: the
share of patients with a characteristic, or the level of a measurement, year by
year, with a smooth through it. It **describes** the cohort. It fits no model and
tests nothing, so a trend here is a prompt for an analysis, not a result.

The engine is `hvtiPlotR::hv_trends()`. It takes one row per patient and computes
the per-year summary itself, so this job never aggregates by hand. The
`group_by()`/`summarise()` blocks and the per-series `smooth.spline()` calls that
older trends jobs carry are exactly what it replaces.

Two neighbours are **not** this job. Procedure volume per year is `dp-procs`, and
the distribution of a single variable is `dp-variable`.

```{r}
#| label: setup
# This file runs either from the directory it was scaffolded into or from the
# project root. Resolve which, so that no path in this document needs editing.
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(hvtiRutilities)
  library(hvtiPlotR)
  library(ggplot2)
})

# 2.7.7, not 2.7.6. This job reports how many rows each figure could not draw,
# and hv_trends() only counts them (meta$n_missing) from part-way through 2.7.6:
# the field arrived without a version bump, so a package reporting 2.7.6 may or
# may not have it. 2.7.7 is the first version that certainly does.
if (utils::packageVersion("hvtiPlotR") < "2.7.7") {
  stop("This job needs hvtiPlotR >= 2.7.7; ", utils::packageVersion("hvtiPlotR"),
       " is installed. Update it, then re-render.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
# The markers in this file name work a study author still has to do, and a job
# that still contains one has not been finished. This chunk is what makes that
# TRUE rather than merely stated: an unedited trends job would otherwise render
# green over the placeholder columns and the placeholder year origin below.
#
# knitr::current_input() is NULL outside a render, when a study author steps
# through chunks in RStudio, and there is no file to scan then, so this is a
# no-op in that case.
#
# The token is BUILT, not written literally, and that is not stylistic. Quarto
# knits through an intermediate and current_input() returns THAT file, so the
# scan reads a copy of this chunk along with everything else: a literal
# grep("<token>", .src) here matches its own source line, and the guard then
# fires on every render, finished or not. Measured, not theorised -- a probe
# found three markers in a file containing two. A guard that cries wolf on a
# finished job is a guard that gets deleted, which returns us to issue #27.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    # Report the marker TEXT. Line numbers here are the intermediate's, not
    # this .qmd's, so they need not match what the author sees in an editor.
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it.\nTo render a partly-worked draft anyway, set ",
      "HVTI_TEMPLATE_DRAFT to 1, true or yes -- the report then carries a ",
      "draft banner. Any other value, including 0, leaves this guard strict."
    )
    # Only the documented values enable draft mode. nzchar() alone would treat
    # HVTI_TEMPLATE_DRAFT=0 as ON, silently downgrading this stop() to a
    # warning for someone whose shell profile sets it to 0 meaning OFF -- a
    # switch whose enablement does not mean what it says, which is the same
    # defect this guard exists to catch. An unrecognised value falls through to
    # strict on purpose: the author then sees the stop and re-reads the docs,
    # which is the safe direction to be wrong in.
    if (tolower(Sys.getenv("HVTI_TEMPLATE_DRAFT")) %in% c("1", "true", "yes")) {
      # The banner is not optional. A draft render that looks like a finished
      # one is the same defect with an extra step: the warning scrolls past in
      # a log, while the .html is the artifact that gets sent to someone.
      warning(.msg, call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Rendered with `HVTI_TEMPLATE_DRAFT` set. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by new_job() from the endpoint/type it
     was called with, which are also the filename's first two fields. Every
     artifact path in this document is computed from them.

     For an EDA job the set key is (subject, eda). ENDPOINT names WHAT is
     described: `cohort` for the whole study, or a label for a sub-cohort such
     as `mv_repair`. TYPE is always `eda`. Every EDA job for one subject then
     shares a set, so its trends, follow-up and postage-stamp figures land
     together under graphs/<subject>-eda/. Decided 2026-09-10, because a trends
     plot describes a cohort rather than analysing an endpoint.

     Do NOT hand-edit these declarations: the render-time guard below checks
     them against this file's own name and stops if they disagree. -->

```{r}
#| label: set
ENDPOINT <- "cohort"
TYPE     <- "eda"

# `new_job()` writes ENDPOINT/TYPE from the same values it put in this file's
# name, but a hand-edited declaration can drift from it afterward. set_path()
# below resolves from the declarations, not the filename, so a drifted
# declaration would silently write into ANOTHER set's artifact directory --
# exactly the collision the (endpoint, type) key exists to prevent, re-entered
# through the body instead of the name.
#
# knitr::current_input() is NULL outside a render (a study author running
# chunks interactively in RStudio), and there is no filename to check against
# yet, so this is a no-op in that case rather than a spurious error.
.current <- knitr::current_input()
if (!is.null(.current)) {
  # Quarto knits through an intermediate, so `knitr::current_input()` returns
  # `...-03.01-ac.rmarkdown` here, not `...-03.01-ac.qmd`. Strip whatever
  # extension is actually present rather than hard-coding one, so this
  # doesn't depend on a build-tool detail staying the same.
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint, "', type '",
         .name_type, "'), but declares ENDPOINT = \"", ENDPOINT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

# Resolve a path inside this set's artifact directory. `kind` is the artifact
# folder -- "estimates" for serialized results, "graphs" for figures. The set
# directory sits one layer under the kind and never two, which is the whole
# layout rule.
#
# Created on first use rather than up front, so a job that writes nothing leaves
# no empty directories behind.
set_path <- function(kind, file) {
  d <- file.path(.root, kind, paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```

## Data

```{r}
#| label: data
d <- read_built()

# EDIT: the calendar year of operation. Older jobs store years since an origin
# and add the origin back, and the origin differs between studies: 1985 in one,
# 2000 in another. A wrong origin moves every figure along the x-axis with
# nothing to say so, which is why the range is checked and printed below. If
# the dataset carries the calendar year directly, use that column instead.
d$year <- d$iv_opyrs + 1985

if (!is.numeric(d$year)) {
  stop("`year` is not numeric (", class(d$year)[1L], "). Build it from a ",
       "numeric column, such as years since an origin plus that origin.",
       call. = FALSE)
}
# A plausible window, not a precise one. Its job is to catch the wrong-origin
# mistake, which lands decades or centuries out, not to police real data.
.this_year <- as.integer(format(Sys.Date(), "%Y"))
.out <- which(d$year < 1900 | d$year > .this_year + 1)
if (length(.out)) {
  stop(length(.out), " patient(s) have an operation year outside 1900 to ",
       .this_year + 1, " (for example ", d$year[.out[1L]], "). Check the ",
       "origin added to the years-since-origin column.", call. = FALSE)
}

knitr::kable(data.frame(
  quantity = c("patients", "year missing", "first year", "last year"),
  value    = c(nrow(d), sum(is.na(d$year)),
               min(d$year, na.rm = TRUE), max(d$year, na.rm = TRUE))
))
```

## What is trended

```{r}
#| label: trends
# EDIT: one entry per figure. `cols` are the columns drawn together: several
# columns become several series in one figure, which is how a set of 0/1
# indicators such as NYHA class I to IV is shown. `kind` says how to read them:
#   "percent"     0/1 or logical, drawn as the percent of patients with it
#   "continuous"  a measurement, drawn on its own scale
# `labels` names the series in the legend (default: the column names). Fixing
# `ylim` and `ybreaks` keeps figures for different subgroups comparable.
# `summary` is "mean" (the default) or "median", the per-year point drawn under
# the smooth.
TRENDS <- list(
  chf = list(cols = "hx_chf", kind = "percent", labels = "Heart failure",
             ylab = "Patients (%)", ylim = c(0, 100), ybreaks = seq(0, 100, 20)),
  lvmass = list(cols = "plvmassi", kind = "continuous", labels = "LV mass index",
                ylab = "LV mass index", ylim = NULL, ybreaks = NULL)
)

# EDIT: optional. x-axis breaks shared by every figure, e.g. seq(1985, 2025, 5).
# NULL lets ggplot choose.
XBREAKS <- NULL

# EDIT: optional. Each entry is a filter, given as a function of the data that
# returns TRUE for the rows to keep. Every trend is drawn once per entry with
# the same axes, so the figures can be read side by side. The default draws the
# whole cohort only. A second entry might be
#   complex = function(d) d$complex == 1
SUBGROUPS <- list(all = function(d) rep(TRUE, nrow(d)))
```

```{r}
#| label: helpers
# One row per patient per series: the shape hv_trends() takes. Several columns
# become several series through `series`, which hv_trends() groups on.
trend_long <- function(d, spec) {
  absent <- setdiff(spec$cols, names(d))
  if (length(absent)) {
    stop("TRENDS names column(s) not in the data: ",
         paste(absent, collapse = ", "), call. = FALSE)
  }
  labels <- if (is.null(spec$labels)) spec$cols else spec$labels
  if (length(labels) != length(spec$cols)) {
    stop("TRENDS entry has ", length(spec$cols), " column(s) but ",
         length(labels), " label(s).", call. = FALSE)
  }
  if (!spec$kind %in% c("percent", "continuous")) {
    stop("TRENDS `kind` must be \"percent\" or \"continuous\", not \"",
         spec$kind, "\".", call. = FALSE)
  }
  parts <- lapply(seq_along(spec$cols), function(i) {
    v <- d[[spec$cols[i]]]
    if (identical(spec$kind, "percent")) {
      if (is.logical(v)) v <- as.numeric(v)
      # A 1/2 or 0/1/9 code times 100 draws a plausible and wrong figure, so
      # anything other than 0/1 stops here instead of being scaled.
      bad <- setdiff(unique(v[!is.na(v)]), c(0, 1))
      if (length(bad)) {
        stop("`", spec$cols[i], "` is kind \"percent\" but holds ",
             paste(utils::head(bad, 5L), collapse = ", "),
             ". A percent column must be 0/1 or logical.", call. = FALSE)
      }
      v <- 100 * v
    }
    data.frame(year = d$year, value = v, series = labels[i],
               stringsAsFactors = FALSE)
  })
  out <- do.call(rbind, parts)
  out$series <- factor(out$series, levels = labels)
  out
}
```

## Figures

```{r}
#| label: figures
#| results: asis
# The image link is relative to THIS document, which always sits in graphs/,
# so it is <set>/<file> whichever directory the render executes from. The file
# itself is written through set_path(), which resolves from the project root.
for (sg in names(SUBGROUPS)) {
  keep <- SUBGROUPS[[sg]](d)
  keep[is.na(keep)] <- FALSE
  dd <- d[keep, , drop = FALSE]
  if (nrow(dd) == 0L) {
    stop("SUBGROUPS entry `", sg, "` selects no patients.", call. = FALSE)
  }
  cat("\n### ", sg, " (n = ", nrow(dd), ")\n\n", sep = "")
  for (nm in names(TRENDS)) {
    spec <- TRENDS[[nm]]
    tr <- hv_trends(trend_long(dd, spec), x_col = "year", y_col = "value",
                    group_col = if (length(spec$cols) == 1L) NULL else "series",
                    summary_fn = if (is.null(spec$summary)) "mean" else spec$summary)
    # Older jobs dropped incomplete rows with na.omit() and said nothing.
    # hv_trends() counts them, and the count is reported with every figure.
    cat("**", nm, "**: ", tr$meta$n_obs - tr$meta$n_missing, " of ",
        tr$meta$n_obs, " rows drawn; ", tr$meta$n_missing,
        " missing the year or the value.\n\n", sep = "")
    p <- plot(tr) +
      labs(x = "Year of operation", y = spec$ylab, colour = NULL, shape = NULL) +
      theme_hv_manuscript()
    if (!is.null(spec$ylim)) {
      p <- p + scale_y_continuous(
        limits = spec$ylim,
        breaks = if (is.null(spec$ybreaks)) waiver() else spec$ybreaks
      )
    }
    if (!is.null(XBREAKS)) p <- p + scale_x_continuous(breaks = XBREAKS)
    fname <- paste0("dp-trends-", nm, "-", sg, ".png")
    png(set_path("graphs", fname), width = 8, height = 6, units = "in", res = 150)
    print(p)
    invisible(dev.off())
    cat("![", nm, ", ", sg, "](", file.path(paste0(ENDPOINT, "-", TYPE), fname),
        ")\n\n", sep = "")
  }
}
```

The smooth is LOESS, `hv_trends()`'s default, with `span = 0.75`. Older jobs used
`smooth.spline(df = 6)`, which is not reproduced: the two differ at the ends of
the year range, where data is thinnest, so compare against an older figure by its
shape and not point by point. `plot(tr, span = ...)` changes the smoothing.
