---
title: "Concordance: every patient through every group's model"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in:
# rendered as a standalone document the project's format block never applies,
# which is how server-rendered reports end up with sibling _files/ trees instead
# of being self-contained.
#
# NOTE: output-dir is a PROJECT setting and cannot be set here, so the .html
# lands next to its source. With embed-resources that is a single standalone
# file, which is the property that actually matters when a report is sent on.
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the SAS job this replaces, and say which treatments it compared
     and at what horizon. -->

Replaces `graphs/<job>.sas`: name the treatments compared here and the horizon.

An `hs-concordance` job predicts **every patient through every treatment
group's model**. Each group's risk-factor model was fitted by its own `hm` job,
on that group's patients; this job reads them all, and asks of each patient
what the model for each treatment predicts at one horizon. The corpus calls the
result concordance and discordance: whether the treatment a patient received is
the one the models favor.

**This job reads models from other sets on purpose.** An `hs-setup` job reads
the `hm.rds` in its own set, so that a prediction can never be filed against a
model it did not come from. Here the crossing is the point, so it is declared:
`MODELS` below names the set each model comes from, and every model read is
recorded in this job's provenance.

**The job has a core and an optional decision.** The core predicts and saves.
The section headed "Decision" chooses a best treatment per patient and tables
it against the treatment received. Delete that section if the study does not
make that claim; the rest of the job does not depend on it.

The design, with the three studies it was drawn from and the failures each
guard below exists for, is in
`hvtiRtemplates:dev/specs/2026-09-30-hs-concordance-design.md`.

```{r}
#| label: setup
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRtemplates:::.find_study_root(if (is.null(.in)) getwd() else dirname(.in))
.provenance_data <- list()
.provenance_artifacts <- list()
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(TemporalHazard)
  library(hvtiRutilities)
})
```

```{r}
#| label: guard-edits
#| results: asis
# The markers in this file name work a study author still has to do, and
# README.md says a job that still contains one has not been finished. This
# chunk is what makes that TRUE rather than merely stated.
#
# Without it an unedited job renders green over a meaningless analysis. In THIS
# file that would be a table of best treatments, which reads as a clinical
# recommendation whether or not anyone finished the job.
#
# knitr::current_input() is NULL outside a render -- a study author stepping
# through chunks in RStudio -- and there is no file to scan then, so this is a
# no-op in that case. Same handling as the `set` guard below, for the same
# reason.
#
# The token is BUILT, not written literally, and that is not stylistic. Quarto
# knits through an intermediate and current_input() returns THAT file, so the
# scan reads a copy of this chunk along with everything else: a literal
# grep("<token>", .src) here matches its own source line, and the guard then
# fires on every render, finished or not. Measured, not theorised -- a probe
# found three markers in a file containing two. A guard that cries wolf on a
# finished job is a guard that gets deleted, which returns us to issue #27.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    # Report the marker TEXT. Line numbers here are the intermediate's, not
    # this .qmd's, so they need not match what the author sees in an editor.
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    # A job renders as a draft by default, so an author can iterate on a
    # working report and remove markers as they go. HVTI_TEMPLATE_STRICT
    # turns the draft into a stop for a final render. Only unset, 0, false
    # and no mean "not strict": an unrecognized value stops on purpose, so a
    # mistyped switch is seen rather than quietly ignored, which is the safe
    # direction to be wrong in.
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      # The banner is not optional. A draft render that looks like a finished
      # one is the same defect with an extra step: the warning scrolls past in
      # a log, while the .html is the artifact that gets sent to someone.
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by add_job() from the subject/type it
     was called with, which are also the filename's first two fields. They
     name the SET this job belongs to, and every artifact path in the
     document is computed from them, so no output path below needs editing.

     Do NOT hand-edit these declarations: the render-time guard below checks
     them against this file's own name and stops if they disagree. To put
     this job in a different set, call add_job() again with the new
     subject/type rather than editing this copy in place.

     A set is keyed on (subject, analysis type), not on the subject alone.
     One subject may be analyzed by several methods -- parametric hazard, random
     forest survival, competing risks -- and those chains share their upstream.
     Two sets keyed on `dead_pa` alone would write their life tables to the
     same file. -->

```{r}
#| label: set
SUBJECT <- "dead_pa"
TYPE    <- "hz"

# `add_job()` writes SUBJECT/TYPE from the same values it put in this file's
# name, but a hand-edited declaration can drift from it afterward. set_path()
# below resolves from the declarations, not the filename, so a drifted
# declaration would silently write into ANOTHER set's artifact directory --
# exactly the collision the (subject, type) key exists to prevent, re-entered
# through the body instead of the name.
#
# knitr::current_input() is NULL outside a render (a study author running
# chunks interactively in RStudio), and there is no filename to check against
# yet, so this is a no-op in that case rather than a spurious error.
.current <- knitr::current_input()
if (!is.null(.current)) {
  # Quarto knits through an intermediate, so `knitr::current_input()` returns
  # `<subject>-<type>-<prefix>.rmarkdown` here rather than the
  # `.qmd` this was scaffolded as. Strip whatever extension is actually present
  # rather than hard-coding one, so this doesn't depend on a build-tool detail
  # staying the same.
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject, "', type '",
         .name_type, "'), but declares SUBJECT = \"", SUBJECT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

# Resolve a path inside this set's artifact directory. `kind` is the artifact
# folder -- "estimates" for serialized results, "graphs" for figures. The set
# directory sits one layer under the kind and never two, which is the whole
# layout rule.
#
# Created on first use rather than up front, so a job that writes nothing leaves
# no empty directories behind.
set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}


# This job reads other sets' models by path, in the models chunk. It writes only
# into its own set, the comparison set these two values name.
```

## Study choices

Edit these values for this study before rendering.

```{r}
#| label: edit-study-choices
# The data and rows this job predicts for. Unlike hs-setup, these are this
# job's own choice, not taken from a model: each hm model was fitted on ONE
# group's rows, and this job predicts for all of them.
# EDIT: the dataset, and a WHERE that keeps every group being compared.
DATASET <- "study"
ANALYSIS_SET <- NULL
WHERE <- NULL
ID <- "ccfid"
KEY <- ID

# EDIT: the follow-up time and event columns every hm model was fitted on. The
#       models chunk stops if any model records a different time column,
#       because HORIZON below is one number applied to every model.
TIME  <- "iv_dead"
EVENT <- "dead"

# EDIT: the counts this cohort must have. The invalid defaults stop an unedited
#       job before it predicts for a cohort nobody reconciled.
EXPECTED <- list(n = NA_integer_, n_events = NA_integer_, n_censored = NA_integer_)

# EDIT: one entry per treatment group. The NAME is the group's label and must
#       be a value of the GROUP column. The VALUE is the set whose hm job
#       fitted that group's model, "<subject>-<type>".
MODELS <- NULL

# EDIT: the column recording the treatment each patient actually received.
GROUP <- NULL

# EDIT: the single horizon, in TIME's own units. If the hm jobs worked in
#       years, 30 days is 30/365.25, not 30. A wrong unit gives a plausible
#       number at the wrong time, the failure with no symptom.
HORIZON <- 10

# EDIT: columns a later plotting job needs beside each prediction, such as
#       age. They are saved once per patient, because the saved predictions
#       carry a row number, not a patient identifier.
CARRY <- character()

# EDIT: how the cohort was made comparable across groups: "none", "matched"
#       or "common_support". Say which, and why, in the text above. The three
#       studies this template was drawn from made three different choices,
#       so there is no default.
OVERLAP <- NULL

# EDIT: TRUE only when a MODELS entry is a path to another study's hm.rds.
#       Setting it is also your statement that the other study's time is in
#       TIME's unit: the saved model records its time column's name and no
#       unit, so this is the one check here that rests on your word. Write
#       the unit in the text above.
CROSS_STUDY <- FALSE
```

## Cohort

```{r}
#| label: data
# The checksum of every dataset in manifest.yaml is checked before anything is
# read, so a result can name the data that produced it. It stops on a mismatch.
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
.cfg <- study_config(start = .root)
job_data <- hvtiRtemplates::read_job_data(.cfg, dataset = DATASET, analysis_set = ANALYSIS_SET,
                                          where = WHERE, id = ID, key = KEY)
d <- job_data$data
.provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(job_data$provenance))
knitr::kable(job_data$record, col.names = c("Data", ""), caption = "The data this job read")
if (!is.null(job_data$attrition)) {
  knitr::kable(job_data$attrition, caption = paste0("Analysis set `", ANALYSIS_SET, "`: exclusions, in order"))
}
```

```{r}
#| label: cohort
cc <- cohort_counts(d, event = EVENT, time = TIME)
assert_cohort(d, expected = EXPECTED, event = EVENT, time = TIME)

# The three studies this template was drawn from made three overlap choices,
# and a fourth study made a fourth. Which one a study makes decides what its
# comparison means, so it is declared and saved, never defaulted.
.overlaps <- c("none", "matched", "common_support")
if (is.null(OVERLAP) || length(OVERLAP) != 1L || !OVERLAP %in% .overlaps) {
  stop("OVERLAP must be one of ", paste0("\"", .overlaps, "\"", collapse = ", "),
       ". Say in the text above how the cohort was made comparable across groups.", call. = FALSE)
}
```

## Models

```{r}
#| label: models
# Names first. They become model identifiers and the decision table's columns,
# so an empty or repeated one would make a winning model ambiguous.
if (is.null(MODELS) || !length(MODELS)) {
  stop("MODELS is unset. Name one hm set per treatment group, e.g. ",
       "c(cabg = \"dead-cabg\", pci = \"dead-pci\").", call. = FALSE)
}
.mn <- names(MODELS)
if (is.null(.mn) || anyNA(.mn) || any(!nzchar(.mn))) {
  stop("Every MODELS entry needs a name: the treatment group it models.", call. = FALSE)
}
if (anyDuplicated(.mn)) {
  stop("MODELS names must be unique; repeated: ", paste(unique(.mn[duplicated(.mn)]), collapse = ", "), ".",
       call. = FALSE)
}

# An entry with a "/" is a path to an hm.rds; anything else is a set in this
# study. A path is refused unless CROSS_STUDY says it was meant, so that a
# mistyped set name cannot quietly become a read from somewhere else.
.is_path <- stats::setNames(grepl("/", MODELS, fixed = TRUE), .mn)
if (any(.is_path) && !isTRUE(CROSS_STUDY)) {
  stop("MODELS entr", if (sum(.is_path) == 1L) "y " else "ies ", paste(.mn[.is_path], collapse = ", "),
       " name a path, not a set. Set CROSS_STUDY <- TRUE if a model from another study is meant.",
       call. = FALSE)
}
.set_path <- file.path(hvtiRutilities::study_dir("estimates", .root), MODELS, "hm.rds")
.model_path <- stats::setNames(ifelse(.is_path, MODELS, .set_path), .mn)
.absent_model <- !file.exists(.model_path)
if (any(.absent_model)) {
  stop("No hm.rds for ", paste0(.mn[.absent_model], " (", .model_path[.absent_model], ")", collapse = ", "),
       ". Render that set's hm job first.", call. = FALSE)
}
# Two groups read through one model would compare a model with itself and
# still produce numbers. One of the three source jobs did exactly that.
.real <- normalizePath(.model_path)
if (anyDuplicated(.real)) {
  stop("MODELS entries ", paste(.mn[.real %in% .real[duplicated(.real)]], collapse = ", "),
       " resolve to the same file. Each group needs its own model.", call. = FALSE)
}

models <- list()
for (m in .mn) {
  # A model from another study is recorded against that study's root, which
  # is where its provenance is anchored.
  .mcfg <- if (.is_path[[m]]) study_config(start = dirname(.model_path[[m]])) else .cfg
  .read <- hvtiRtemplates:::.read_handoff(.model_path[[m]], "hazard-regression-model", .mcfg,
                                          paste0("the hm job for ", m))
  .time <- .read$lineage$analysis$time$variable
  if (!identical(.time, TIME)) {
    stop("The model for ", m, " was fitted on time column \"", .time, "\", not TIME (\"", TIME, "\"). ",
         "HORIZON is one number applied to every model, so every model must share one time scale.",
         call. = FALSE)
  }
  models[[m]] <- .read$value
  .provenance_data <- c(.provenance_data, .read$lineage$data)
  .provenance_artifacts <- c(.provenance_artifacts, .read$lineage$artifacts, list(.read$record))
}
```

```{r}
#| label: groups
# Every group in the cohort must have a model: a patient whose own treatment
# has none cannot be compared with what they received. An author who means to
# leave a group out filters it with WHERE, where the choice is visible.
if (is.null(GROUP) || !GROUP %in% names(d)) {
  stop("GROUP must name the column recording each patient's treatment; got ",
       if (is.null(GROUP)) "NULL" else paste0("\"", GROUP, "\""), ".", call. = FALSE)
}
.actual <- as.character(d[[GROUP]])
if (anyNA(.actual)) {
  stop(sum(is.na(.actual)), " patient(s) have no value in ", GROUP, ". Filter them with WHERE.", call. = FALSE)
}
.no_model <- setdiff(unique(.actual), names(models))
if (length(.no_model)) {
  stop("Group(s) ", paste(.no_model, collapse = ", "), " in ", GROUP, " have no model in MODELS. ",
       "Add the model, or filter the group out with WHERE.", call. = FALSE)
}
# The reverse is allowed: a model no patient carries is a treatment nobody in
# this cohort received, which is a comparison worth making. It is reported.
knitr::kable(data.frame(model = names(models), patients = as.vector(table(factor(.actual, names(models))))),
             caption = "Patients who received each modelled treatment")
```

```{r}
#| label: covariates
# Each model's covariates are taken from its artifact, not re-declared, so the
# list cannot drift from the screen its hm job ran. The models need not share
# them: each group's model is its own, and the table shows where they differ.
covariates <- lapply(models, function(a) unique(unlist(a$covariates, use.names = FALSE)))
for (m in names(covariates)) {
  .miss <- setdiff(covariates[[m]], names(d))
  if (length(.miss)) {
    stop("The model for ", m, " was fitted on column(s) absent from this cohort: ",
         paste(.miss, collapse = ", "), ".", call. = FALSE)
  }
}
.all_cov <- sort(unique(unlist(covariates, use.names = FALSE)))
.cov_tbl <- data.frame(covariate = .all_cov,
                       vapply(covariates, function(v) ifelse(.all_cov %in% v, "yes", ""), character(length(.all_cov))),
                       check.names = FALSE)
knitr::kable(.cov_tbl, caption = "Covariates in each group's model")

# `hm` fitted on covariates_to_numeric(d, ...), so the conversion is redone
# here before any model sees the data, exactly as hs-setup does.
.clash <- intersect(.all_cov, "time")
if (length(.clash)) {
  stop("A covariate is named \"time\", which collides with the horizon column predict() needs. ",
       "Rename it in the built dataset.", call. = FALSE)
}
dd <- covariates_to_numeric(d, .all_cov)
```

## Horizon and support

```{r}
#| label: horizon
if (!is.numeric(HORIZON) || length(HORIZON) != 1L || !is.finite(HORIZON) || HORIZON <= 0) {
  stop("HORIZON must be one positive number, in TIME's units.", call. = FALSE)
}
# Each model is checked against its OWN follow-up. One of the three source jobs
# predicted a group at 10 years whose own header said it had no 10-year data.
.max_obs <- vapply(models, function(a) {
  t <- a$reported$data$frame[[TIME]]
  if (!any(is.finite(t))) NA_real_ else max(t, na.rm = TRUE)
}, numeric(1L))
.short <- is.na(.max_obs) | .max_obs < HORIZON
if (any(.short)) {
  stop("HORIZON (", HORIZON, ") is beyond the last observed time of the model for ",
       paste0(names(models)[.short], " (", signif(.max_obs[.short], 4), ")", collapse = ", "),
       ". A model's prediction past its own follow-up is its parametric form running on, not an estimate.",
       call. = FALSE)
}
```

Predicting a patient through another group's model is the purpose of this job,
so a patient outside a model's fitted range is not refused. The table below
counts them, per model and covariate, so a reader can see how far each
comparison reaches beyond the data behind it. What counts as too far is a
clinical judgment.

```{r}
#| label: support
support <- do.call(rbind, lapply(names(models), function(m) {
  frame <- models[[m]]$reported$data$frame
  do.call(rbind, lapply(covariates[[m]], function(v) {
    if (!is.numeric(frame[[v]]) || !is.numeric(dd[[v]])) return(NULL)
    lo <- min(frame[[v]], na.rm = TRUE)
    hi <- max(frame[[v]], na.rm = TRUE)
    data.frame(model = m, covariate = v, fitted_min = lo, fitted_max = hi,
               below = sum(dd[[v]] < lo, na.rm = TRUE), above = sum(dd[[v]] > hi, na.rm = TRUE))
  }))
}))
knitr::kable(support, row.names = FALSE, digits = 3,
             caption = "Patients outside each model's fitted range, by covariate")
```

## Predictions

⚠️ **The confidence limits here are about 68%, not 95%.** `CLEVEL` is
`0.68268948`, plus or minus one standard error, the SAS `HAZPRED` convention
that `hs-setup` also uses. A 95% band is nearly twice as wide. The limits travel
in the saved artifact, so **state the coverage wherever they are shown.**

```{r}
#| label: predict
# SAS convention, not a study choice: +/-1 SD, NOT 95%, with survival limits on
# the logit scale. hs-setup and hp use both; changing them here alone would put
# two coverages side by side with nothing saying so.
CLEVEL <- 0.68268948

pred <- do.call(rbind, lapply(names(models), function(m) {
  nd <- dd[, covariates[[m]], drop = FALSE]
  nd$time <- HORIZON
  p <- predict(models[[m]]$reported, newdata = nd, type = "survival",
               se.fit = TRUE, level = CLEVEL, conf.type = "logit")
  data.frame(row = seq_len(nrow(dd)), group = .actual, model = m,
             fit = p$fit, se.fit = p$se.fit, lower = p$lower, upper = p$upper)
}))

# A missing prediction is a failed one, and is caught before the range check,
# which would otherwise pass an all-NA result silently.
.bad <- !is.finite(pred$fit)
if (any(.bad)) {
  stop(sum(.bad), " of ", nrow(pred), " predictions are NA or non-finite, from model(s) ",
       paste(unique(pred$model[.bad]), collapse = ", "), ".", call. = FALSE)
}
stopifnot(all(pred$fit >= 0 & pred$fit <= 1))

# One row per patient: the treatment received and the CARRY columns a plotting
# job needs. The patient identifier is never saved: a study's ID can be the
# medical record number, and this file travels.
.id_col <- attr(job_data$record, "selection")$id
.carry_bad <- CARRY[tolower(CARRY) %in% tolower(c(.id_col, ID, "mrn", "emrn"))]
if (length(.carry_bad)) {
  stop("CARRY may not name a patient identifier: ", paste(.carry_bad, collapse = ", "), ".", call. = FALSE)
}
# `row` and `group` are the frame's own columns; a CARRY column of either name
# would sit beside them under the same name, and a reader takes the first.
.carry_clash <- intersect(CARRY, c("row", "group"))
if (length(.carry_clash)) {
  stop("CARRY may not name ", paste(.carry_clash, collapse = ", "), ": the saved frame already has ",
       "`row` and `group` columns. The treatment received is saved as `group` whatever GROUP is called.",
       call. = FALSE)
}
.carry_miss <- setdiff(CARRY, names(d))
if (length(.carry_miss)) {
  stop("CARRY names column(s) not in the cohort: ", paste(.carry_miss, collapse = ", "), ".", call. = FALSE)
}
patients <- data.frame(row = seq_len(nrow(d)), group = .actual)
patients <- cbind(patients, d[, CARRY, drop = FALSE])

summary_tbl <- stats::aggregate(fit ~ model + group, data = pred, FUN = stats::median)
names(summary_tbl)[names(summary_tbl) == "fit"] <- "median_survival"
knitr::kable(summary_tbl, row.names = FALSE, digits = 4,
             caption = paste0("Median predicted survival at ", HORIZON, ", by model and treatment received"))
```

## Decision

<!-- EDIT: keep this section only if the study claims a best treatment per
     patient. Delete it, chunk and all, if it does not; the save chunk works
     either way. -->

A patient's **best-predicted** treatment is the model giving the highest
survival. It is called **optimal** only when the choice is separated: the best
model's lower limit lies above the runner-up's upper limit. Point estimates
that differ in the fourth decimal are not a choice, and exact equality would
not catch them. The limits are about 68%, so "separated" means separated at
that coverage. The template sets no clinical tolerance of its own; what
difference in survival matters is the study's to say.

```{r}
#| label: edit-decision
# EDIT: optional. A named list, model -> the logical column saying which
#       patients were eligible for that treatment. An ineligible prediction
#       takes no part in the choice and is left unchanged in `pred`.
ELIGIBLE <- list()

.wide <- function(col) {
  w <- vapply(names(models), function(m) pred[[col]][pred$model == m], numeric(nrow(d)))
  matrix(w, nrow = nrow(d), dimnames = list(NULL, names(models)))
}
.fit <- .wide("fit")
.lo <- .wide("lower")
.hi <- .wide("upper")
for (m in names(ELIGIBLE)) {
  if (!m %in% names(models)) stop("ELIGIBLE names ", m, ", which is not in MODELS.", call. = FALSE)
  .ok <- d[[ELIGIBLE[[m]]]]
  if (!is.logical(.ok) || anyNA(.ok)) {
    stop("ELIGIBLE column for ", m, " must be logical with no missing values.", call. = FALSE)
  }
  .fit[!.ok, m] <- NA
}

.choose <- function(i) {
  f <- .fit[i, ]
  ok <- which(!is.na(f))
  if (length(ok) < 2L) {
    return(c(best = if (length(ok)) names(f)[ok] else NA_character_, optimal = NA_character_, tie = "FALSE"))
  }
  o <- ok[order(f[ok], decreasing = TRUE)]
  tie <- f[o[1L]] == f[o[2L]]
  separated <- !tie && .lo[i, o[1L]] > .hi[i, o[2L]]
  c(best = if (tie) NA_character_ else names(f)[o[1L]],
    optimal = if (separated) names(f)[o[1L]] else NA_character_,
    tie = as.character(tie))
}
.ch <- vapply(seq_len(nrow(d)), .choose, character(3L))
decision <- data.frame(row = seq_len(nrow(d)), group = .actual,
                       best = .ch["best", ], optimal = .ch["optimal", ], tie = .ch["tie", ] == "TRUE")

.levels <- names(models)
knitr::kable(table(received = factor(decision$group, .levels), best = factor(decision$best, .levels)),
             caption = "Treatment received against best-predicted treatment, every patient")
.sep <- !is.na(decision$optimal)
knitr::kable(table(received = factor(decision$group[.sep], .levels), optimal = factor(decision$optimal[.sep], .levels)),
             caption = paste0("Treatment received against optimal treatment, the ", sum(.sep),
                              " patients whose choice is separated at ~68%; ", sum(!.sep), " are not"))
cat(sum(decision$tie), "patient(s) have an exact tie between their two best models.\n")
```

## Save

```{r}
#| label: save
# Read by the plotting jobs in this set. `clevel`, `horizon` and `overlap`
# travel with the estimates because each changes what the numbers mean, and a
# reader of the file should not have to recover them from prose.
art <- list(pred = pred, patients = patients, horizon = HORIZON, time = TIME, clevel = CLEVEL,
            models = MODELS, covariates = covariates, support = support, overlap = OVERLAP,
            decision = if (exists("decision")) decision else NULL)
art <- hvtiRtemplates:::.attach_handoff_lineage(
  art,
  data = .provenance_data,
  artifacts = .provenance_artifacts,
  analysis = list(
    time = list(variable = TIME),
    event = list(variable = EVENT, event = 1L, censored = 0L)
  ),
  cohort = cc
)
saveRDS(art, set_path("estimates", "hs-concordance.rds"))
```

```{r}
#| label: provenance
#| echo: false
#| results: asis
cat(hvtiRtemplates:::.embed_provenance(
  .in,
  data = .provenance_data,
  artifacts = .provenance_artifacts,
  extra = list(
    subject = SUBJECT,
    type = TYPE,
    analysis = list(
      time = list(variable = TIME),
      event = list(variable = EVENT, event = 1L, censored = 0L)
    ),
    cohort = cc
  )
))
```
