---
title: "Patient-level predictions and expected survival"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in:
# rendered as a standalone document the project's format block never applies,
# which is how server-rendered reports end up with sibling _files/ trees instead
# of being self-contained.
#
# NOTE: output-dir is a PROJECT setting and cannot be set here, so the .html
# lands next to its source. With embed-resources that is a single standalone
# file, which is the property that actually matters when a report is sent on.
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the SAS job this replaces, and say which horizons it reported
     and against which expected-survival comparison. -->

Replaces `graphs/<job>.sas`: describe its prediction step here — which horizons
it reported, and what it compared them against.

An `hs` job turns the risk-factor model an `hm` job fitted into **predictions
for individual patients**, and sets those against the survival the general
population would have had. The model is not refitted here; it is read.

**This job is filed under `graphs/`, which reads oddly for something named
"setup".** That placement follows the corpus rather than the name: all ten
`tp.hs.*` templates in the SAS library, and ten of the eleven R `hs` jobs across
the study tree, sit in `graphs/`. An `hs` job computes what the plotting jobs
beside it consume — the `setup` / `uses_setup` pairing the corpus uses
throughout. The reasoning is in
`hvtiRtemplates:dev/specs/2026-08-29-hs-template-design.md`.

Validation metrics — C-index, Brier score — are **not** here. The corpus files
them under `hs`, but the only R implementation is a loose script on a shared
drive with no package and no version, and no supported template depends on an
unpackaged file. They arrive once it has a home.

```{r}
#| label: setup
# This file runs either from the directory it was scaffolded into or from the
# project root. Resolve which, so that no path in this document needs editing:
# forgetting one of several ../ paths is the instantiation mistake that would
# otherwise be easy to make and quiet when made.
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(TemporalHazard)
  library(hvtiRutilities)
  library(hvtiRlifetables)
})
```

```{r}
#| label: edit-guard
#| results: asis
# The markers in this file name work a study author still has to do, and
# README.md says a job that still contains one has not been finished. This
# chunk is what makes that TRUE rather than merely stated.
#
# Without it an unedited job renders green over a meaningless analysis. In THIS
# file the danger is sharper than a wrong number: the unedited entry and stay
# levels are `hzr_stepwise()`'s defaults, which are LOOSER than the SAS job's,
# so an unedited screen admits variables the job being reproduced rejected --
# and reports them as risk factors, in a table that looks exactly like a
# result.
#
# knitr::current_input() is NULL outside a render -- a study author stepping
# through chunks in RStudio -- and there is no file to scan then, so this is a
# no-op in that case. Same handling as the `set` guard below, for the same
# reason.
#
# The token is BUILT, not written literally, and that is not stylistic. Quarto
# knits through an intermediate and current_input() returns THAT file, so the
# scan reads a copy of this chunk along with everything else: a literal
# grep("<token>", .src) here matches its own source line, and the guard then
# fires on every render, finished or not. Measured, not theorised -- a probe
# found three markers in a file containing two. A guard that cries wolf on a
# finished job is a guard that gets deleted, which returns us to issue #27.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    # Report the marker TEXT. Line numbers here are the intermediate's, not
    # this .qmd's, so they need not match what the author sees in an editor.
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it.\nTo render a partly-worked draft anyway, set ",
      "HVTI_TEMPLATE_DRAFT to 1, true or yes -- the report then carries a ",
      "draft banner. Any other value, including 0, leaves this guard strict."
    )
    # Only the documented values enable draft mode. nzchar() alone would treat
    # HVTI_TEMPLATE_DRAFT=0 as ON, silently downgrading this stop() to a
    # warning for someone whose shell profile sets it to 0 meaning OFF -- a
    # switch whose enablement does not mean what it says, which is the same
    # defect this guard exists to catch. An unrecognised value falls through to
    # strict on purpose: the author then sees the stop and re-reads the docs,
    # which is the safe direction to be wrong in.
    if (tolower(Sys.getenv("HVTI_TEMPLATE_DRAFT")) %in% c("1", "true", "yes")) {
      # The banner is not optional. A draft render that looks like a finished
      # one is the same defect with an extra step: the warning scrolls past in
      # a log, while the .html is the artifact that gets sent to someone.
      warning(.msg, call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Rendered with `HVTI_TEMPLATE_DRAFT` set. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by new_job() from the endpoint/type it
     was called with, which are also the filename's first two fields. They
     name the SET this job belongs to, and every artifact path in the
     document is computed from them, so no output path below needs editing.

     Do NOT hand-edit these declarations: the render-time guard below checks
     them against this file's own name and stops if they disagree. To put
     this job in a different set, call new_job() again with the new
     endpoint/type rather than editing this copy in place.

     A set is keyed on (endpoint, analysis type), not on the endpoint alone.
     One endpoint is analysed by several methods -- parametric hazard, random
     forest survival, competing risks -- and those chains share their upstream.
     Two sets keyed on `dead_pa` alone would write their life tables to the
     same file. -->

```{r}
#| label: set
ENDPOINT <- "dead_pa"
TYPE     <- "hz"

# `new_job()` writes ENDPOINT/TYPE from the same values it put in this file's
# name, but a hand-edited declaration can drift from it afterward. set_path()
# below resolves from the declarations, not the filename, so a drifted
# declaration would silently write into ANOTHER set's artifact directory --
# exactly the collision the (endpoint, type) key exists to prevent, re-entered
# through the body instead of the name.
#
# knitr::current_input() is NULL outside a render (a study author running
# chunks interactively in RStudio), and there is no filename to check against
# yet, so this is a no-op in that case rather than a spurious error.
.current <- knitr::current_input()
if (!is.null(.current)) {
  # Quarto knits through an intermediate, so `knitr::current_input()` returns
  # `<endpoint>-<type>-<ordinal>-<prefix>.rmarkdown` here rather than the
  # `.qmd` this was scaffolded as. Strip whatever extension is actually present
  # rather than hard-coding one, so this doesn't depend on a build-tool detail
  # staying the same.
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint, "', type '",
         .name_type, "'), but declares ENDPOINT = \"", ENDPOINT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

# Resolve a path inside this set's artifact directory. `kind` is the artifact
# folder -- "estimates" for serialized results, "graphs" for figures. The set
# directory sits one layer under the kind and never two, which is the whole
# layout rule.
#
# Created on first use rather than up front, so a job that writes nothing leaves
# no empty directories behind.
set_path <- function(kind, file) {
  d <- file.path(.root, kind, paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}

# `hm` reads the `hz` fit through this and writes its own model back, both by
# set, so a model can never be filed against a different set than the shapes it
# was built on. The write is in the `save` chunk at the foot of this file.
```
## Cohort and model

<!-- EDIT: choose the cohort shape, exactly as the `hm` job did.

     The rows this job predicts for MUST be the rows the model was fitted on,
     or the predictions describe a cohort the model never saw. That failure is
     silent: `predict()` returns a number for any row you hand it, whatever
     cohort it came from. -->

```{r}
#| label: cohort
d <- read_built()
assert_cohort(d)
```

```{r}
#| label: model
# The model, not a refit. `hm.rds` is written by the hm job in this same set,
# so a prediction can never be filed against a different set than the model it
# came from.
hm_art <- readRDS(set_path("estimates", "hm.rds"))
fit    <- hm_art$reported

# The covariates the model was actually fitted on. Taken from the artifact
# rather than re-declared here: a re-declared list drifts from the model the
# moment the hm job's screen changes, and `predict()` would then either error
# on a missing column or -- worse -- silently use a stale set.
COVARIATES <- hm_art$covariates
stopifnot(is.character(COVARIATES), length(COVARIATES) > 0L)
missing_cols <- setdiff(COVARIATES, names(d))
if (length(missing_cols)) {
  stop("The model was fitted on covariate(s) absent from this cohort: ",
       paste(missing_cols, collapse = ", "),
       ". The cohort here is not the cohort hm was fitted on.", call. = FALSE)
}
```

## Predictions

<!-- EDIT: the follow-up time column, and the horizons this job reports.

     TIME must be the SAME column the hm job fitted on. It is re-declared here
     rather than read from `hm.rds`, because that artifact does not carry it --
     so the two CAN drift, and the guard below is what catches it. If you change
     one, change both.

     The horizons are in TIME's own units.

     These are reporting choices, not statistical ones, and they must match the
     job being replaced -- a nomogram at 5 years and a table at 10 answer
     different questions. The corpus uses 30-day, 5-year and 10-year most
     often. The units are whatever `hz` fitted in: if that job worked in years,
     30 days is 30/365.25, not 30. Getting this wrong produces a plausible
     number at the wrong time, which is the failure mode with no symptom. -->

```{r}
#| label: horizons
TIME     <- "iu_dead"
HORIZONS <- c(1, 5, 10)

# A `hazard` object carries `call`, `call_env`, `spec`, `data`, `fit`,
# `legacy_args` and `engine` -- there is NO `formula` slot, and `spec` holds
# only dist/control/time_windows/phases. The formula is recoverable from
# `fit$call` only when the model was built through the formula interface;
# `hazard()` also accepts time=/status=, in which case it is not there at all.
# So the time column is declared above rather than introspected.
if (!TIME %in% names(fit$data)) {
  stop("TIME is \"", TIME, "\", which is not a column of the data the model was ",
       "fitted on. It must name the same follow-up column the hm job used.",
       call. = FALSE)
}

# The model cannot extrapolate past the follow-up it was fitted on, and
# `predict()` will not say so -- it returns a number for any time you give it.
# Beyond the last observed time that number is the fitted parametric form
# running on, not an estimate the data supports.
.max_obs <- max(fit$data[[TIME]], na.rm = TRUE)
if (any(HORIZONS > .max_obs)) {
  stop("HORIZONS contains ", paste(HORIZONS[HORIZONS > .max_obs], collapse = ", "),
       ", beyond the last observed time (", signif(.max_obs, 4),
       "). Shorten the horizon or say in the text why extrapolation is defensible.",
       call. = FALSE)
}
```

`predict()` needs one row per (patient, horizon): the covariates identify the
patient and a `time` column names the horizon.

```{r}
#| label: predict
# One row per patient per horizon. Built explicitly rather than by looping
# `predict()` per horizon, so every prediction comes from ONE call and cannot
# drift between horizons.
nd <- do.call(rbind, lapply(HORIZONS, function(h) {
  x <- d[, COVARIATES, drop = FALSE]
  x$time <- h
  x$.id  <- seq_len(nrow(d))
  x$.horizon <- h
  x
}))

p <- predict(fit, newdata = nd[, c(COVARIATES, "time")], type = "survival",
             se.fit = TRUE, level = 0.95)

pred <- data.frame(id = nd$.id, horizon = nd$.horizon,
                   fit = p$fit, se.fit = p$se.fit,
                   lower = p$lower, upper = p$upper)

# Survival is a probability. A value outside [0, 1] means the delta-method
# interval was built on the wrong scale, not that a patient is 103% alive.
stopifnot(all(pred$fit >= 0 & pred$fit <= 1, na.rm = TRUE))
knitr::kable(
  do.call(rbind, lapply(split(pred, pred$horizon), function(g)
    data.frame(horizon = g$horizon[[1]], n = nrow(g),
               median = stats::median(g$fit, na.rm = TRUE),
               min = min(g$fit, na.rm = TRUE), max = max(g$fit, na.rm = TRUE)))),
  row.names = FALSE, digits = 4,
  caption = "Predicted survival by horizon, across the cohort")
```
