---
title: "Patient-level predictions and expected survival"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in:
# rendered as a standalone document the project's format block never applies,
# which is how server-rendered reports end up with sibling _files/ trees instead
# of being self-contained.
#
# NOTE: output-dir is a PROJECT setting and cannot be set here, so the .html
# lands next to its source. With embed-resources that is a single standalone
# file, which is the property that actually matters when a report is sent on.
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the SAS job this replaces, and say which horizons it reported
     and against which expected-survival comparison. -->

Replaces `graphs/<job>.sas`: describe its prediction step here — which horizons
it reported, and what it compared them against.

An `hs` job turns the risk-factor model an `hm` job fitted into **predictions
for individual patients**, and sets those against the survival the general
population would have had. The model is not refitted here; it is read.

**This job is filed under `graphs/`, which reads oddly for something named
"setup".** That placement follows the corpus rather than the name: all ten
`tp.hs.*` templates in the SAS library, and ten of the eleven R `hs` jobs across
the study tree, sit in `graphs/`. An `hs` job computes what the plotting jobs
beside it consume — the `setup` / `uses_setup` pairing the corpus uses
throughout. The reasoning is in
`hvtiRtemplates:dev/specs/2026-08-29-hs-template-design.md`.

Validation metrics — C-index, Brier score — are **not** here. The corpus files
them under `hs`, but the only R implementation is a loose script on a shared
drive with no package and no version, and no supported template depends on an
unpackaged file. They arrive once it has a home.

```{r}
#| label: setup
# This file runs either from the directory it was scaffolded into or from the
# project root. Resolve which, so that no path in this document needs editing:
# forgetting one of several ../ paths is the instantiation mistake that would
# otherwise be easy to make and quiet when made.
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(TemporalHazard)
  library(hvtiRutilities)
  library(hvtiRlifetables)
})
```

```{r}
#| label: edit-guard
#| results: asis
# The markers in this file name work a study author still has to do, and
# README.md says a job that still contains one has not been finished. This
# chunk is what makes that TRUE rather than merely stated.
#
# Without it an unedited job renders green over a meaningless analysis. In THIS
# file the danger is sharper than a wrong number: the unedited entry and stay
# levels are `hzr_stepwise()`'s defaults, which are LOOSER than the SAS job's,
# so an unedited screen admits variables the job being reproduced rejected --
# and reports them as risk factors, in a table that looks exactly like a
# result.
#
# knitr::current_input() is NULL outside a render -- a study author stepping
# through chunks in RStudio -- and there is no file to scan then, so this is a
# no-op in that case. Same handling as the `set` guard below, for the same
# reason.
#
# The token is BUILT, not written literally, and that is not stylistic. Quarto
# knits through an intermediate and current_input() returns THAT file, so the
# scan reads a copy of this chunk along with everything else: a literal
# grep("<token>", .src) here matches its own source line, and the guard then
# fires on every render, finished or not. Measured, not theorised -- a probe
# found three markers in a file containing two. A guard that cries wolf on a
# finished job is a guard that gets deleted, which returns us to issue #27.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    # Report the marker TEXT. Line numbers here are the intermediate's, not
    # this .qmd's, so they need not match what the author sees in an editor.
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it.\nTo render a partly-worked draft anyway, set ",
      "HVTI_TEMPLATE_DRAFT to 1, true or yes -- the report then carries a ",
      "draft banner. Any other value, including 0, leaves this guard strict."
    )
    # Only the documented values enable draft mode. nzchar() alone would treat
    # HVTI_TEMPLATE_DRAFT=0 as ON, silently downgrading this stop() to a
    # warning for someone whose shell profile sets it to 0 meaning OFF -- a
    # switch whose enablement does not mean what it says, which is the same
    # defect this guard exists to catch. An unrecognised value falls through to
    # strict on purpose: the author then sees the stop and re-reads the docs,
    # which is the safe direction to be wrong in.
    if (tolower(Sys.getenv("HVTI_TEMPLATE_DRAFT")) %in% c("1", "true", "yes")) {
      # The banner is not optional. A draft render that looks like a finished
      # one is the same defect with an extra step: the warning scrolls past in
      # a log, while the .html is the artifact that gets sent to someone.
      warning(.msg, call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Rendered with `HVTI_TEMPLATE_DRAFT` set. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by new_job() from the endpoint/type it
     was called with, which are also the filename's first two fields. They
     name the SET this job belongs to, and every artifact path in the
     document is computed from them, so no output path below needs editing.

     Do NOT hand-edit these declarations: the render-time guard below checks
     them against this file's own name and stops if they disagree. To put
     this job in a different set, call new_job() again with the new
     endpoint/type rather than editing this copy in place.

     A set is keyed on (endpoint, analysis type), not on the endpoint alone.
     One endpoint is analysed by several methods -- parametric hazard, random
     forest survival, competing risks -- and those chains share their upstream.
     Two sets keyed on `dead_pa` alone would write their life tables to the
     same file. -->

```{r}
#| label: set
ENDPOINT <- "dead_pa"
TYPE     <- "hz"

# `new_job()` writes ENDPOINT/TYPE from the same values it put in this file's
# name, but a hand-edited declaration can drift from it afterward. set_path()
# below resolves from the declarations, not the filename, so a drifted
# declaration would silently write into ANOTHER set's artifact directory --
# exactly the collision the (endpoint, type) key exists to prevent, re-entered
# through the body instead of the name.
#
# knitr::current_input() is NULL outside a render (a study author running
# chunks interactively in RStudio), and there is no filename to check against
# yet, so this is a no-op in that case rather than a spurious error.
.current <- knitr::current_input()
if (!is.null(.current)) {
  # Quarto knits through an intermediate, so `knitr::current_input()` returns
  # `<endpoint>-<type>-<ordinal>-<prefix>.rmarkdown` here rather than the
  # `.qmd` this was scaffolded as. Strip whatever extension is actually present
  # rather than hard-coding one, so this doesn't depend on a build-tool detail
  # staying the same.
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint, "', type '",
         .name_type, "'), but declares ENDPOINT = \"", ENDPOINT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

# Resolve a path inside this set's artifact directory. `kind` is the artifact
# folder -- "estimates" for serialized results, "graphs" for figures. The set
# directory sits one layer under the kind and never two, which is the whole
# layout rule.
#
# Created on first use rather than up front, so a job that writes nothing leaves
# no empty directories behind.
set_path <- function(kind, file) {
  d <- file.path(.root, kind, paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}

# `hm` reads the `hz` fit through this and writes its own model back, both by
# set, so a model can never be filed against a different set than the shapes it
# was built on. The write is in the `save` chunk at the foot of this file.
```
## Cohort and model

<!-- EDIT: choose the cohort shape, exactly as the `hm` job did.

     The rows this job predicts for MUST be the rows the model was fitted on,
     or the predictions describe a cohort the model never saw. That failure is
     silent: `predict()` returns a number for any row you hand it, whatever
     cohort it came from. -->

```{r}
#| label: cohort
d <- read_built()
assert_cohort(d)
```

```{r}
#| label: model
# The model, not a refit. `hm.rds` is written by the hm job in this same set,
# so a prediction can never be filed against a different set than the model it
# came from.
hm_art <- readRDS(set_path("estimates", "hm.rds"))
fit    <- hm_art$reported

# The covariates the model was actually fitted on. Taken from the artifact
# rather than re-declared here: a re-declared list drifts from the model the
# moment the hm job's screen changes, and `predict()` would then either error
# on a missing column or -- worse -- silently use a stale set.
#
# `hm` stores its screen as a LIST -- list(early = <chr>, late = <chr>) -- one
# character vector per phase, and saves it in that shape. Flatten it the same
# way hm itself does when it builds the model formula: `unique()` matters,
# because a variable offered to BOTH phases appears twice, and a duplicated
# column here makes a design matrix wider than the fitted coefficients.
COVARIATES <- unique(unlist(hm_art$covariates, use.names = FALSE))
stopifnot(is.character(COVARIATES), length(COVARIATES) > 0L)
missing_cols <- setdiff(COVARIATES, names(d))
if (length(missing_cols)) {
  stop("The model was fitted on covariate(s) absent from this cohort: ",
       paste(missing_cols, collapse = ", "),
       ". The cohort here is not the cohort hm was fitted on.", call. = FALSE)
}
```

## Predictions

<!-- EDIT: the follow-up time column, and the horizons this job reports.

     TIME must be the SAME column the hm job fitted on. It is re-declared here
     rather than read from `hm.rds`, because that artifact does not carry it --
     so the two CAN drift, and the guard below is what catches it. If you change
     one, change both.

     The horizons are in TIME's own units.

     These are reporting choices, not statistical ones, and they must match the
     job being replaced -- a nomogram at 5 years and a table at 10 answer
     different questions. The corpus uses 30-day, 5-year and 10-year most
     often. The units are whatever `hz` fitted in: if that job worked in years,
     30 days is 30/365.25, not 30. Getting this wrong produces a plausible
     number at the wrong time, which is the failure mode with no symptom. -->

```{r}
#| label: horizons
TIME     <- "iu_dead"
HORIZONS <- c(1, 5, 10)

# A `hazard` object carries `call`, `call_env`, `spec`, `data`, `fit`,
# `legacy_args` and `engine` -- there is NO `formula` slot, and `spec` holds
# only dist/control/time_windows/phases. The formula is recoverable from
# `fit$call` only when the model was built through the formula interface;
# `hazard()` also accepts time=/status=, in which case it is not there at all.
# So the time column is declared above rather than introspected.
# A `hazard` object's `data` slot is a CONTAINER -- time, time_lower,
# time_upper, status, x, weights, frame -- not the model frame. The study's own
# columns are one level down, in `fit$data$frame`. Reading `fit$data[[TIME]]`
# returns NULL, which makes this guard unsatisfiable and turns the max() below
# into -Inf. Measured, not assumed.
if (!TIME %in% names(fit$data$frame)) {
  stop("TIME is \"", TIME, "\", which is not a column of the data the model was ",
       "fitted on. It must name the same follow-up column the hm job used.",
       call. = FALSE)
}

# The model cannot extrapolate past the follow-up it was fitted on, and
# `predict()` will not say so -- it returns a number for any time you give it.
# Beyond the last observed time that number is the fitted parametric form
# running on, not an estimate the data supports.
# An all-missing time column makes max(na.rm = TRUE) return -Inf rather than
# error, and the check below would then report "beyond the last observed time
# (-Inf)", which explains nothing. Catch the empty case where it can still be
# named.
if (!any(is.finite(fit$data$frame[[TIME]]))) {
  stop("Column \"", TIME, "\" in the fitted data has no finite values, so there ",
       "is no observed follow-up to compare HORIZONS against.", call. = FALSE)
}
# HORIZONS must be sorted. `us_matched()` enforces it too, but three chunks
# later and about `times`, not about the value the author actually edited --
# and the expected-survival join below pairs `us_cohort_curve()`'s output
# (times in order of first appearance) with `split()`'s (numeric order), which
# agree only while HORIZONS is sorted. Fail at the edit, not downstream of it.
if (is.unsorted(HORIZONS, strictly = TRUE)) {
  stop("HORIZONS must be in increasing order with no repeats; got ",
       paste(HORIZONS, collapse = ", "), ".", call. = FALSE)
}

.max_obs <- max(fit$data$frame[[TIME]], na.rm = TRUE)
if (any(HORIZONS > .max_obs)) {
  stop("HORIZONS contains ", paste(HORIZONS[HORIZONS > .max_obs], collapse = ", "),
       ", beyond the last observed time (", signif(.max_obs, 4),
       "). Shorten the horizon or say in the text why extrapolation is defensible.",
       call. = FALSE)
}
```

`predict()` needs one row per (patient, horizon): the covariates identify the
patient and a `time` column names the horizon.

```{r}
#| label: predict
# One row per patient per horizon. Built explicitly rather than by looping
# `predict()` per horizon, so every prediction comes from ONE call and cannot
# drift between horizons.
# `time`, `.id` and `.horizon` are constructed below. A covariate with one of
# those names would be silently overwritten by the construction rather than
# erroring, so the collision is checked rather than assumed away.
# `hm` fitted on covariates_to_numeric(d, ...), not on the raw built data: that
# conversion turns numeric-looking FACTORS into numerics, and a model fitted
# after it has a different parameter count from one fitted before. hm.rds does
# not carry the converted frame, so the conversion is redone here. Predicting
# from the raw data instead would dummy-code a factor the model fitted
# linearly -- a different design matrix, and nothing in the output would say so.
dd <- covariates_to_numeric(d, COVARIATES)

.clash <- intersect(COVARIATES, c("time", ".id", ".horizon"))
if (length(.clash)) {
  stop("Covariate(s) ", paste(.clash, collapse = ", "), " collide with the ",
       "column names this chunk builds. Rename the column in the built ",
       "dataset, or this job will predict from overwritten values.",
       call. = FALSE)
}
nd <- do.call(rbind, lapply(HORIZONS, function(h) {
  x <- dd[, COVARIATES, drop = FALSE]
  x$time <- h
  x$.id  <- seq_len(nrow(dd))
  x$.horizon <- h
  x
}))

# CLEVEL and conf.type are SAS conventions recovered from the HAZPRED source
# during parity work, not study choices -- and NEITHER errors when wrong.
#
# CLEVEL is 0.68268948, which is pnorm(1) - pnorm(-1): plus or minus one
# standard error, about 68%, NOT 95%. The published papers state it. On the
# same fit and grid a 95% band is 1.97x the width, so a reader who takes these
# for 95% limits reads an interval roughly half as wide as they think.
#
# Survival limits are formed on the logit scale. predict.hazard() defaults to
# "log-log", the survfit standard, so conf.type must be passed explicitly.
#
# `06.01-hp.qmd` uses both, and this job's estimates are read by plotting jobs
# that do. Changing either here alone puts two coverages on one figure with
# nothing saying so. Label the coverage wherever these numbers are published.
CLEVEL <- 0.68268948   # SAS convention, not a study choice. +/-1 SD, NOT 95%.

p <- predict(fit, newdata = nd[, c(COVARIATES, "time")], type = "survival",
             se.fit = TRUE, level = CLEVEL, conf.type = "logit")

pred <- data.frame(id = nd$.id, horizon = nd$.horizon,
                   fit = p$fit, se.fit = p$se.fit,
                   lower = p$lower, upper = p$upper)

# Two checks, and the order matters. A prediction that came back NA is a
# FAILED prediction, not an out-of-range one -- and `all(..., na.rm = TRUE)`
# would strip them and then return TRUE for an empty vector, so a run where
# EVERY prediction failed would pass a range check silently and render a table
# of NA. Catch the missing case first, loudly, then range-check what remains
# without na.rm so a partial failure cannot hide either.
.bad <- !is.finite(pred$fit)
if (any(.bad)) {
  stop(sum(.bad), " of ", nrow(pred), " predictions are NA or non-finite. ",
       "predict() returned no usable value for those rows -- check that the ",
       "cohort's covariates match the ones the model was fitted on, and that ",
       "TIME and HORIZONS are in the units the hz job used.", call. = FALSE)
}
# Survival is a probability. A value outside [0, 1] means the delta-method
# interval was built on the wrong scale, not that a patient is 103% alive.
stopifnot(all(pred$fit >= 0 & pred$fit <= 1))
# Split out rather than nested inside kable(): a table built inside its own
# printing call is hard to inspect when a number looks wrong, and the nesting
# is what tripped the indentation and brace linters this repo keeps on for
# templates on purpose.
by_horizon <- split(pred, pred$horizon)
summary_tbl <- do.call(rbind, lapply(by_horizon, function(g) {
  data.frame(horizon = g$horizon[[1L]], n = nrow(g),
             median = stats::median(g$fit, na.rm = TRUE),
             min = min(g$fit, na.rm = TRUE),
             max = max(g$fit, na.rm = TRUE))
}))
knitr::kable(summary_tbl, row.names = FALSE, digits = 4,
             caption = "Predicted survival by horizon, across the cohort")
```

## Expected population survival

<!-- EDIT: the columns naming each patient's age, sex and race, and the life
     table vintage.

     `vintage` HAS NO DEFAULT HERE, ON PURPOSE. The SAS macro `%usmatchd` took
     no vintage argument, so every `hs.uslife` job silently inherited the macro
     default AS OF ITS RUN DATE -- and that default moved twice, table84 ->
     table2008 -> table2023, with no signal at either move. The vintages are
     structurally different fits, not perturbations, so a wrong one is wrong by
     orders of magnitude. A parity check on one such job could not read the
     vintage off it at all and had to RECOVER it, by fitting all three and
     seeing which landed at machine precision. A job that does not state its
     vintage cannot be reproduced.

     `us_lifetable_vintages()` lists the valid values AND what each one's
     non-white stratum actually means -- which is not what its code says.
     Under `table2023`, stratum `"b"` is a risk-weighted average of Black,
     Asian, American Indian and Hispanic death rates, NOT Black, despite the
     code and despite the macro's own comment. Name strata by meaning in your
     prose, never by the letter code.

     TABLE selects which strata the life table itself is built from. Use
     "sexrace" when race is usable, "sex" when it is not, "overall" when
     neither is. This is not the axis you group your REPORT by -- that is a
     separate concern the template does not automate.

     SCALE must match the units TIME and HORIZONS are in. It is NOT
     independently checkable: pass "years" for a job fitted in months and you
     get expected survival at 6, 12 and 24 YEARS against observed values at 6,
     12 and 24 months, with a plausible-looking ratio and no error anywhere. -->

```{r}
#| label: expected
AGE_COL   <- "age"
MALE_COL  <- "male"
OTHER_COL <- "other"
VINTAGE   <- NULL
TABLE     <- "sexrace"
SCALE     <- "years"

# Two failure modes, both caught here rather than inside us_matched(): unset,
# and set to something that is not a vintage. The second matters because the
# valid values are not guessable -- "table2023" is not "2023" -- and an error
# raised three calls down names neither the argument nor the alternatives.
#
# us_lifetable_vintages() returns a DATA FRAME -- vintage, n_strata,
# nonwhite_code, nonwhite_meaning, added -- not a character vector. The
# `$vintage` is deliberate; pasting the frame prints column-wise nonsense.
valid_vintages <- us_lifetable_vintages()$vintage
if (is.null(VINTAGE)) {
  stop("VINTAGE is unset. Choose one of: ", paste(valid_vintages, collapse = ", "),
       " -- and record WHY in the text above. There is no safe default; see the ",
       "comment on this chunk.", call. = FALSE)
}
if (!VINTAGE %in% valid_vintages) {
  stop("VINTAGE is \"", VINTAGE, "\", which is not a life table vintage. ",
       "Valid values: ", paste(valid_vintages, collapse = ", "), ".",
       call. = FALSE)
}

# Named columns are checked before use, as the model chunk does for covariates.
# Without this a typo surfaces as "`id` must be the same length as `age`; got
# N and 0" from inside us_matched(), which never names the column at fault.
.absent <- setdiff(c(AGE_COL, MALE_COL, OTHER_COL), names(d))
if (length(.absent)) {
  stop("Column(s) not in the cohort: ", paste(.absent, collapse = ", "),
       ". Set AGE_COL, MALE_COL and OTHER_COL to this study's own column names.",
       call. = FALSE)
}

# Read this before choosing, and print it in the report:
knitr::kable(us_lifetable_vintages(), row.names = FALSE,
             caption = "Life table vintages, and what each one's non-white stratum means")

expected <- us_matched(
  age = d[[AGE_COL]], male = d[[MALE_COL]], other = d[[OTHER_COL]],
  times = HORIZONS, id = seq_len(nrow(d)),
  vintage = VINTAGE, table = TABLE, scale = SCALE, individual = TRUE
)
```

```{r}
#| label: obs-vs-exp
# The cohort average at each horizon, which is what gets overlaid on an
# actuarial curve. This was derived inline until hvtiRlifetables 0.1.3 shipped
# `us_cohort_curve()` (hvtiRlifetables#16); it is now one call, and the numbers
# are unchanged -- checked to 1e-12 against the derivation it replaces before
# the swap, because the comment that stood here promised exactly that.
#
# Prefer the function over re-deriving it. The averaging is not arbitrary: it
# follows `%usmatchd` (lines 338-350), taking the UNWEIGHTED cohort mean of
# survival, over SURVIVAL rather than cumulative hazard, with no patient
# weighted by follow-up. Two studies that each roll their own average
# differently is the drift that issue was filed to stop.
#
# ⚠️ It deliberately does NOT drop NA. Every value `us_matched()` returns is
# finite by construction, so an NA arriving here is a defect rather than a
# datum, and averaging around it would hide it. The inline version used
# `na.rm = TRUE` and would have.
#
# `us_cohort_curve()` returns `time`, `smatched` and `hmatched`. `smatched` is
# the matched SURVIVAL column -- `agesurv` in `us_matched()`'s output is the
# age-matched baseline, a different quantity, and taking it would be a
# plausible number for a different question.
#
# EDIT: to report expected survival within groups -- age band, treatment arm,
#       era -- pass `by =`. It carries VALUES, not column names, and `expected`
#       holds one row per patient per time, so a patient-level grouping has to
#       be expanded to it. Grouping is not stratification: `table =` above
#       chose which strata the life TABLE is built from; `by` chooses how the
#       resulting curves are grouped for reporting, and a report broken down by
#       age band is normally still matched on "sexrace".
exp_curve <- us_cohort_curve(expected)
exp_avg <- stats::setNames(exp_curve$smatched, exp_curve$time)
obs_avg <- vapply(split(pred$fit, pred$horizon), mean, numeric(1), na.rm = TRUE)

stopifnot(identical(names(exp_avg), names(obs_avg)))
obs_vs_exp <- data.frame(
  horizon  = as.numeric(names(obs_avg)),
  observed = unname(obs_avg),
  expected = unname(exp_avg)
)
obs_vs_exp$ratio <- obs_vs_exp$observed / obs_vs_exp$expected

knitr::kable(obs_vs_exp, row.names = FALSE, digits = 4,
             caption = "Model-predicted vs age/sex/race-matched population survival")
```

A ratio below 1 says the cohort does worse than the matched general population
at that horizon. It is a comparison, not a test — no interval is attached,
because the expected curve carries no sampling error of its own here.

## Save

```{r}
#| label: save
# Read by the plotting jobs in this set. `vintage` travels WITH the estimates
# rather than being recorded only in prose: an expected curve whose vintage has
# to be recovered later is the defect this template's vintage marker exists to
# prevent, and it would be reintroduced here by omitting it.
#
# `clevel` travels with the estimates for the same reason `vintage` does: a
# confidence limit whose coverage has to be inferred later is a number waiting
# to be misread, and 0.68268948 is not the coverage anyone assumes.
saveRDS(list(pred = pred, expected = expected, obs_vs_exp = obs_vs_exp,
             horizons = HORIZONS, covariates = COVARIATES, vintage = VINTAGE,
             clevel = CLEVEL),
        set_path("estimates", "hs.rds"))
```
