---
title: "Multiphase parametric hazard fit"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in:
# rendered as a standalone document the project's format block never applies,
# which is how server-rendered reports end up with sibling _files/ trees instead
# of being self-contained.
#
# NOTE: output-dir is a PROJECT setting and cannot be set here, so the .html
# lands next to its source. With embed-resources that is a single standalone
# file, which is the property that actually matters when a report is sent on.
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the SAS job this replaces, and say what its PROC HAZARD call
     fits: which phases, which parameters are free, and which are fixed. -->

Replaces `distributions/<job>.sas`: describe its `PROC HAZARD` call here — the
phase structure, what is estimated and what is held fixed.

An `hz` job produces a parametric fit and persists it. The life table it is read
against is an `ac` job, the figures over it are `hp`, and the covariate model
built on its shape parameters is `hm`; none of those belong here. **Nor does SAS
parity** — a parity job borrows the ordinal of the job it checks and lives in
`parity/`.

```{r}
#| label: setup
# This file runs either from the directory it was scaffolded into or from the
# project root. Resolve which, so that no path in this document needs editing:
# forgetting one of several ../ paths is the instantiation mistake that would
# otherwise be easy to make and quiet when made.
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(TemporalHazard)
  library(hvtiRutilities)
})
```

```{r}
#| label: edit-guard
#| results: asis
# The markers in this file name work a study author still has to do, and
# README.md says a job that still contains one has not been finished. This
# chunk is what makes that TRUE rather than merely stated.
#
# Without it an unedited job renders green over a meaningless analysis. The
# derive chunk below indexes a placeholder column; when that column is absent
# `!is.na(d$<col>)` is logical(0), and assignment through a zero-length index
# is a SILENT no-op in R. Every patient keeps the initial category, every
# stratified estimate is computed over one dummy stratum, and nothing errors.
#
# knitr::current_input() is NULL outside a render -- a study author stepping
# through chunks in RStudio -- and there is no file to scan then, so this is a
# no-op in that case. Same handling as the `set` guard below, for the same
# reason.
#
# The token is BUILT, not written literally, and that is not stylistic. Quarto
# knits through an intermediate and current_input() returns THAT file, so the
# scan reads a copy of this chunk along with everything else: a literal
# grep("<token>", .src) here matches its own source line, and the guard then
# fires on every render, finished or not. Measured, not theorised -- a probe
# found three markers in a file containing two. A guard that cries wolf on a
# finished job is a guard that gets deleted, which returns us to issue #27.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    # Report the marker TEXT. Line numbers here are the intermediate's, not
    # this .qmd's, so they need not match what the author sees in an editor.
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it.\nTo render a partly-worked draft anyway, set ",
      "HVTI_TEMPLATE_DRAFT to 1, true or yes -- the report then carries a ",
      "draft banner. Any other value, including 0, leaves this guard strict."
    )
    # Only the documented values enable draft mode. nzchar() alone would treat
    # HVTI_TEMPLATE_DRAFT=0 as ON, silently downgrading this stop() to a
    # warning for someone whose shell profile sets it to 0 meaning OFF -- a
    # switch whose enablement does not mean what it says, which is the same
    # defect this guard exists to catch. An unrecognised value falls through to
    # strict on purpose: the author then sees the stop and re-reads the docs,
    # which is the safe direction to be wrong in.
    if (tolower(Sys.getenv("HVTI_TEMPLATE_DRAFT")) %in% c("1", "true", "yes")) {
      # The banner is not optional. A draft render that looks like a finished
      # one is the same defect with an extra step: the warning scrolls past in
      # a log, while the .html is the artifact that gets sent to someone.
      warning(.msg, call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Rendered with `HVTI_TEMPLATE_DRAFT` set. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by new_job() from the endpoint/type it
     was called with, which are also the filename's first two fields. They
     name the SET this job belongs to, and every artifact path in the
     document is computed from them, so no output path below needs editing.

     Do NOT hand-edit these declarations: the render-time guard below checks
     them against this file's own name and stops if they disagree. To put
     this job in a different set, call new_job() again with the new
     endpoint/type rather than editing this copy in place.

     A set is keyed on (endpoint, analysis type), not on the endpoint alone.
     One endpoint is analysed by several methods -- parametric hazard, random
     forest survival, competing risks -- and those chains share their upstream.
     Two sets keyed on `dead_pa` alone would write their life tables to the
     same file. -->

```{r}
#| label: set
ENDPOINT <- "dead_pa"
TYPE     <- "hz"

# `new_job()` writes ENDPOINT/TYPE from the same values it put in this file's
# name, but a hand-edited declaration can drift from it afterward. set_path()
# below resolves from the declarations, not the filename, so a drifted
# declaration would silently write into ANOTHER set's artifact directory --
# exactly the collision the (endpoint, type) key exists to prevent, re-entered
# through the body instead of the name.
#
# knitr::current_input() is NULL outside a render (a study author running
# chunks interactively in RStudio), and there is no filename to check against
# yet, so this is a no-op in that case rather than a spurious error.
.current <- knitr::current_input()
if (!is.null(.current)) {
  # Quarto knits through an intermediate, so `knitr::current_input()` returns
  # `...-03.01-ac.rmarkdown` here, not `...-03.01-ac.qmd`. Strip whatever
  # extension is actually present rather than hard-coding one, so this
  # doesn't depend on a build-tool detail staying the same.
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint, "', type '",
         .name_type, "'), but declares ENDPOINT = \"", ENDPOINT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

# Resolve a path inside this set's artifact directory. `kind` is the artifact
# folder -- "estimates" for serialized results, "graphs" for figures. The set
# directory sits one layer under the kind and never two, which is the whole
# layout rule.
#
# Created on first use rather than up front, so a job that writes nothing leaves
# no empty directories behind.
set_path <- function(kind, file) {
  d <- file.path(.root, kind, paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}

# An `hz` job ALWAYS persists: `hp` plots what it writes and `hm` fixes its
# shape parameters at these converged values, both by set. The write itself is
# in the `save` chunk at the foot of this file, not here.
```
## Cohort

<!-- EDIT: this chunk, choosing ONE of the two shapes below.

     The distinction is whether the rows this job analyses are the whole study
     or a subset of it, and it is not cosmetic. `_study.yml` records the STUDY
     cohort, derived from the whole built dataset. A job that analyses a
     filtered subset has a different N, and `assert_cohort()` will gate the
     wrong number — it will pass while the job runs on a cohort nobody checked.

     Shape A — the job analyses the whole study cohort. Use the package data
     contract, and the gate is `assert_cohort()`.

     Shape B — the job analyses a filtered subset (an availability flag, a
     procedure subgroup). `read_built()` still reads the data, but the filter
     and the gate are the JOB's, and the expected counts come from the SAS
     reference rather than from `_study.yml`. -->

```{r}
#| label: cohort
# ---- Shape A: whole-study cohort ------------------------------------------
d <- read_built()
assert_cohort(d)
cc <- cohort_counts(d)

# ---- Shape B: filtered job cohort -----------------------------------------
# EDIT: delete Shape A above and uncomment this, supplying the filter and the
# counts the SAS reference reports. Do NOT reuse assert_cohort(): it checks
# _study.yml, which describes the whole study and not this job.
#
# d <- read_built()
# d <- d[!is.na(d$<flag>) & d$<flag> == 1, , drop = FALSE]
# EXPECTED <- list(n = <N>, n_events = <E>, n_censored = <C>)
# cc <- list(n          = nrow(d),
#            n_events   = sum(d[[STATUS]] == 1, na.rm = TRUE),
#            n_censored = nrow(d) - sum(d[[STATUS]] == 1, na.rm = TRUE))
# if (!identical(lapply(cc, as.integer), lapply(EXPECTED, as.integer))) {
#   stop("Cohort gate FAILED: expected ", paste(unlist(EXPECTED), collapse = "/"),
#        ", got ", paste(unlist(cc), collapse = "/"),
#        ". The analysis must not run on an unreconciled cohort.", call. = FALSE)
# }

knitr::kable(data.frame(
  quantity = c("n_analysable", "n_events", "n_censored"),
  n        = c(cc$n, cc$n_events, cc$n_censored)
))
```

A rendered report is itself evidence the gate passed. Report the counts anyway —
a gate that passes silently tells the reader nothing about what it passed on.

**An event flag that is 0/1 in SAS may arrive as `logical`**, because
`read_clinical_data()` infers binary columns as logical while `Surv()` requires
a numeric status. The `as.numeric()` coercions below are not decoration.

```{r}
#| label: numderiv-gate
# numDeriv is a Suggests of TemporalHazard, not a hard dependency. Without it,
# interval- or left-censored multiphase fits SILENTLY produce no standard
# errors: the fit succeeds, `converged` is TRUE, and vcov() is simply absent.
# A model reported without uncertainty is not a smaller result, it is a
# different one -- so this is a hard stop rather than a warning.
if (!requireNamespace("numDeriv", quietly = TRUE)) {
  stop("numDeriv is not installed. Interval-censored multiphase fits will ",
       "produce no standard errors, and will not tell you so. Install it ",
       "before reading anything below.", call. = FALSE)
}
```

## Model

<!-- EDIT: this prose. Say which parameters are free and which are fixed, and
     name the branch. With TAU = GAMMA = ALPHA = 1 the late phase is the pure
     Weibull `G3 = t^eta`; say so if that is what your SAS job did, because a
     reader cannot tell a fixed 1 from an estimated 1 in the output. -->

Describe the phase structure here: how many phases, which of `THALF`, `NU`,
`M`, `MUE`, `TAU`, `GAMMA`, `ALPHA`, `ETA`, `MUL` are estimated, and which are
held.

⚠️ **A SAS hazard fit is always three-phase.** An absent phase sits in the model
at zero rather than being removed, so classify from the sums table, never from
the parameter block — a phase that contributed nothing still prints nothing.

```{r}
#| label: phases
# EDIT: your phases, and the `fixed` set on each.
phases <- list(
  early = hzr_phase("cdf", t_half = 1, nu = 1, m = 1),
  late  = hzr_phase("g3", tau = 1, gamma = 1, alpha = 1, eta = 1,
                    fixed = c("tau", "gamma", "alpha"))
)

# The names, IN ORDER, that theta must supply. Read from the package rather
# than written as a literal: the count and the order are properties of the
# phase specification above, and a hand-written vector silently mis-assigns
# every value if the specification changes.
#
# ⚠️ .hzr_phase_theta_names() is INTERNAL to TemporalHazard (temporal_hazard#186
# asks for it to be exported). Reached through getFromNamespace() rather than
# `:::` so that if it is ever renamed this job stops with a sentence a study
# author can act on, instead of an object-not-found from a namespace they have
# no reason to know about.
theta_names <- try(
  getFromNamespace(".hzr_phase_theta_names", "TemporalHazard")(phases),
  silent = TRUE
)
if (inherits(theta_names, "try-error")) {
  stop("TemporalHazard no longer provides .hzr_phase_theta_names(). The order ",
       "and count of `theta` below cannot be checked automatically. See ",
       "temporal_hazard#186 and pin the parameter order by hand before ",
       "trusting any estimate in this file.", call. = FALSE)
}
theta_table <- data.frame(
  position  = seq_along(theta_names),
  parameter = theta_names
)
knitr::kable(theta_table)
```

⚠️ **`theta` is positional, and the scales are not uniform.** For an
`early` + `late` pair the order is

```
early.log_mu, early.log_t_half, early.nu, early.m,
late.log_mu,  late.log_tau,     late.gamma, late.alpha, late.eta
```

Note the asymmetry: the late phase logs `mu` and `tau` but carries `gamma`,
`alpha` and `eta` on the natural scale. Wrapping the wrong element in `log()`
produces a fit, not an error.

## Starting values

<!-- EDIT: this chunk, choosing ONE of the two shapes below.

     The distinction is whether this job replaces an existing SAS job, and it
     is not cosmetic.

     Shape A — a SAS job exists. Seed from ITS converged estimates, so that any
     disagreement is about the likelihood rather than about where the search
     began. This is the only seed that makes the comparison clean.

     Shape B — no SAS predecessor. Shapes start neutral at 1 and only the TIME
     SCALE is taken from the data. A rough neighbourhood, not an estimate; its
     one job is to be the right order of magnitude. Whether the choice mattered
     is checked by the multi-start probes below, not assumed. -->

```{r}
#| label: start
# ---- Shape A: seed from the SAS job's converged estimates -------------------
# EDIT: your SAS estimates, on their NATURAL scale, wrapped per `theta_names`
# above -- log() for the log-scale positions, bare for the rest.
theta0 <- c(log(1), log(1), 1, 1,
            log(1), log(1), 1, 1, 1)

# ---- Shape B: neutral shapes, data-derived time scale -----------------------
# EDIT: delete Shape A above and uncomment this.
#
# ev_times <- sort(as.numeric(d[[TIME]][as.numeric(d[[STATUS]]) == 1]))
# t_half_0 <- as.numeric(stats::quantile(ev_times, 0.25))
# theta0   <- c(log(1), log(t_half_0), 1, 1,
#               log(1), log(1), 1, 1, 1)
# knitr::kable(data.frame(
#   n_events_used = length(ev_times), t_half_start = t_half_0,
#   note = "derived from the data; shapes start neutral at 1"))

if (length(theta0) != length(theta_names)) {
  stop("theta0 has ", length(theta0), " values but the phase specification ",
       "needs ", length(theta_names), ": ",
       paste(theta_names, collapse = ", "), call. = FALSE)
}
```

## Response

<!-- EDIT: this chunk, choosing ONE of the two shapes below, from what the SAS
     job's `time` / `event` / `icensor` statements actually say. -->

```{r}
#| label: response
# ---- Shape A: no interval censoring -----------------------------------------
# SAS `time <t>; event <e>;` maps straight onto Surv(time, status). Nothing
# below needs the interval machinery and `objective = "sas"` does not apply.
# EDIT: your time and status variables.
resp <- data.frame(time   = as.numeric(d[[TIME]]),
                   status = as.numeric(d[[STATUS]]))
stopifnot(all(resp$status %in% c(0, 1)), !anyNA(resp$time))

# ---- Shape B: interval-censored ---------------------------------------------
# EDIT: delete Shape A above and uncomment this, from SAS's
#   time <upper>; icensor <flag>=<lower>;
# Expressed as Surv(type = "interval2"): an exact event is (t, t), a
# right-censored row is (t, NA), an interval-censored row is (lo, hi).
#
# ⚠️ Do NOT pass the bounds through time_lower/time_upper. time_lower is an
# ENTRY time for left truncation, not an interval bound, and using it that way
# fits a different model without complaining.
#
# status_sas <- ifelse(!is.na(d$<flag>) & d$<flag> == 1, 2,
#               ifelse(d$<event> == 1, 1, 0))
# lo   <- ifelse(status_sas == 2, d$<lower>, d$<upper>)
# hi   <- ifelse(status_sas == 0, NA,        d$<upper>)
# resp <- data.frame(lo = lo, hi = hi)
# knitr::kable(as.data.frame(table(`SAS status` = status_sas)))
```

```{r}
#| label: response-check
# The response build is the step that fails SILENTLY. Get it wrong -- mark an
# interval-censored death as right-censored at its upper bound, the opposite
# claim -- and the fit still converges, on a number that is simply wrong.
#
# So the two event counts are compared here rather than left for the reader to
# compare. A WARNING, not a stop: the cohort gate is defined on its own
# variables, and a study where those legitimately differ from the fitted pair
# should say so here rather than be blocked.
n_fitted <- if ("status" %in% names(resp)) {
  sum(resp$status == 1, na.rm = TRUE)
} else {
  sum(!is.na(resp$hi), na.rm = TRUE)
}
knitr::kable(data.frame(
  source = c("cohort gate", "response as built"),
  n_events = c(cc$n_events, n_fitted)
))
if (!identical(as.integer(n_fitted), as.integer(cc$n_events))) {
  warning("The response carries ", n_fitted, " events but the cohort gate ",
          "counted ", cc$n_events, ". Reconcile before reading the fit.",
          call. = FALSE)
}
```

```{r}
#| label: guard
# The objective is a log-likelihood. A positive one is impossible: it signals a
# response that was mis-specified rather than a fit. Assert it, instead of
# reporting a plateau as a result.
check_fit <- function(f, label) {
  o <- f$fit$objective
  if (!is.finite(o) || o > 0) {
    stop(label, ": objective is ", o, ", which is not a log-likelihood.",
         call. = FALSE)
  }
  invisible(f)
}
```

## Fit

### Fit 1 — deterministic

The reported fit. One start, from `theta0`, so the result is reproducible from
the file alone.

```{r}
#| label: fit-deterministic
# EDIT: the response formula must match the shape chosen above --
# Surv(time, status) for Shape A, Surv(lo, hi, type = "interval2") for Shape B.
fit_det <- hazard(Surv(time, status) ~ 1, data = resp, dist = "multiphase",
                  phases = phases, theta = theta0, fit = TRUE,
                  control = list(conserve = TRUE, condition = 14,
                                 n_starts = 1, maxit = 2000))
check_fit(fit_det, "deterministic")
summary(fit_det)
```

```{r}
#| label: convergence
# There is no logLik() method for class "hazard": the log-likelihood is
# fit$fit$objective.
#
# rcond and pd are read OFF THE FIT rather than recomputed from vcov(): vcov()
# includes every FIXED parameter as an all-NA row and column, and rcond()
# cannot take an NA matrix. The fit's own values are computed on the free
# parameters, which is also what SAS's condition number refers to.
knitr::kable(data.frame(
  quantity = c("log-likelihood", "converged", "iterations", "rcond", "pd"),
  value    = c(format(fit_det$fit$objective, digits = 10),
               fit_det$fit$converged, fit_det$fit$iterations,
               format(fit_det$fit$rcond, digits = 6), fit_det$fit$pd)
))
```

### Fit 2 — did the starting point matter?

```{r}
#| label: multistart
# The probes run at n_starts = 1 ON PURPOSE. Multi-start is what makes the
# reported fit robust -- it perturbs around its start until it escapes small
# basins -- which is exactly what would HIDE a second basin from this check. A
# single start goes where its starting point leads, so it is the thing that
# actually finds another optimum if one is there.
#
# EDIT: widen or add probes if your model has more phases or free parameters.
# Every row must differ from every other row: a duplicated start is a fit that
# costs time and tests nothing.
probes <- rbind(theta0, theta0 + 0.5, theta0 - 0.5)
if (anyDuplicated(probes)) {
  stop("Two probe rows are identical: that fit costs time and tests nothing.",
       call. = FALSE)
}
probe_ll <- vapply(seq_len(nrow(probes)), function(i) {
  f <- try(hazard(Surv(time, status) ~ 1, data = resp, dist = "multiphase",
                  phases = phases, theta = probes[i, ], fit = TRUE,
                  control = list(conserve = TRUE, condition = 14,
                                 n_starts = 1, maxit = 2000)), silent = TRUE)
  if (inherits(f, "try-error")) NA_real_ else f$fit$objective
}, numeric(1))

probe_table <- data.frame(
  probe       = seq_len(nrow(probes)),
  logLik      = probe_ll,
  vs_reported = probe_ll - fit_det$fit$objective
)
knitr::kable(probe_table, digits = 6)
```

A probe that beats the reported fit means the reported fit is in the wrong
basin, and the starting values need revisiting before anything below is read.

### Fit 3 — `noconserve` sensitivity

⚠️ **This is a sensitivity, never a parity number.** A SAS `noconserve` fit is
only a conservation reference if it changes conservation *and nothing else* —
and in practice these fits often add a covariate at the same time, which
confounds the two changes.

```{r}
#| label: noconserve
fit_nc <- hazard(Surv(time, status) ~ 1, data = resp, dist = "multiphase",
                 phases = phases, theta = theta0, fit = TRUE,
                 control = list(conserve = FALSE, condition = 14,
                                n_starts = 1, maxit = 2000))
check_fit(fit_nc, "noconserve")
```

```{r}
#| label: conservation-binding
# Sum_i Lambda(t_i) is what `conserve` pins to the event count. Reporting it
# under BOTH settings is the only way a reader can tell "the control did
# nothing" from "the control had almost nothing to do" -- the two are identical
# in the likelihood column and mean opposite things.
sum_lambda <- c(
  sum(predict(fit_det, type = "cumulative_hazard")),
  sum(predict(fit_nc, type = "cumulative_hazard"))
)
conservation <- data.frame(
  conserve   = c(TRUE, FALSE),
  sum_lambda = sum_lambda,
  events     = cc$n_events
)
conservation$excess <- conservation$sum_lambda - conservation$events
knitr::kable(conservation, digits = 6)
```

## Estimates

```{r}
#| label: estimates
cf <- coef(fit_det)

# vcov() returns a scalar NA rather than a matrix when the Hessian could not be
# inverted, so diag() on it errors. That is a result ABOUT the fit, not a bug
# to route around: report it and carry on with what does not need curvature.
V <- vcov(fit_det)
se_available <- is.matrix(V) && nrow(V) == length(cf) && any(is.finite(diag(V)))

# NOTE the brace. At top level `if (a) x` <newline> `else y` is a PARSE ERROR:
# `else` must not start a line outside a block. knitr captures that error into
# the document and carries on, so the chunk "runs" and silently emits nothing.
se <- if (se_available) {
  setNames(sqrt(diag(V)), names(cf))
} else {
  setNames(rep(NA_real_, length(cf)), names(cf))
}

# EDIT: map your coefficients to their natural scale, and to the SAS names a
# reader will be comparing against. exp() where R reports a log, identity where
# it already reports the natural value.
est <- data.frame(
  parameter     = c("MUE", "THALF", "NU", "M", "MUL", "ETA"),
  estimate      = c(exp(cf[["early.log_mu"]]), exp(cf[["early.log_t_half"]]),
                    cf[["early.nu"]], cf[["early.m"]],
                    exp(cf[["late.log_mu"]]), cf[["late.eta"]]),
  se_on_r_scale = c(se[["early.log_mu"]], se[["early.log_t_half"]],
                    se[["early.nu"]], se[["early.m"]],
                    se[["late.log_mu"]], se[["late.eta"]]),
  r_scale       = c("log", "log", "natural", "natural", "log", "natural"),
  row.names     = NULL
)
knitr::kable(est, digits = 6)

if (!se_available) {
  no_se <- data.frame(
    warning = "The Hessian could not be inverted: NO standard errors are available."
  )
  knitr::kable(no_se)
}
```

**A caveat that must travel with these standard errors.** Under a near-singular
Hessian — check `rcond` above — the marginal standard errors on weakly
identified shape parameters are unstable while the fitted curve is not. The data
can pin the survival function tightly and still leave individual shape
parameters loose, because `se(S(t))² = gᵀVg` can be well determined when the
diagonal of `V` is not.

So report per-parameter standard errors with that caveat attached, and treat the
survival curve and its band — computed in the companion `hp` job — as the
quantity that actually reaches a clinician.

## Save

```{r}
#| label: save
# `hp` plots what this writes and `hm` fixes its shape parameters at these
# values, both by set. A job that persists nothing cannot be chained from.
saveRDS(list(deterministic = fit_det, noconserve = fit_nc,
             phases = phases, theta0 = theta0),
        set_path("estimates", "hz.rds"))
```
