---
title: "Migrate a legacy study to R jobs"
author: "John Ehrlinger"
vignette: >
  %\VignetteIndexEntry{Migrate a legacy study to R jobs}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    embed-resources: true
---

Most studies already have a `datasets/` folder, SAS programs in `descriptive/`
and `graphs/`, and output in `documents/`. We will start there. Adoption adds
the study contract to that existing tree; the legacy programs and their
evidence remain in place.

This example creates its own disposable study with 40 synthetic observations
and a named 24-row subset. Every study file and job output stays inside that
temporary directory, which is removed when the R session ends. Run the chunks
in order in one session to inspect the files yourself. For the statistical
mapping behind these jobs, see [From SAS descriptive jobs to
R](sas-to-r-descriptive.html).

The render steps require Quarto and the packages declared by the job templates,
including `hvtiRdatabuild >= 0.2.1`, `hvtiRtables >= 1.0.1`, and
`hvtiPlotR >= 2.7.14`. A missing or older dependency stops the render and names
the required version.

```{r}
#| label: synthetic-setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(hvtiRutilities)
library(hvtiRtemplates)
root <- withr::local_tempdir(.local_envir = environment())
folders <- c("datasets", "descriptive", "distributions", "analyses",
             "graphs", "documents", "estimates")
for (folder in folders) dir.create(file.path(root, folder))
writeLines(c("project:", "  type: default"), file.path(root, "_quarto.yml"))
i <- seq_len(40L)
built <- data.frame(
  dead = i %% 2L, iv_dead = i / 10, iv_fup = i / 10 + 1,
  year = 2000L + i %% 10L, female = i %% 2L, race_grp = 1L + i %% 3L,
  repair = as.integer(i %% 3L == 0L), age = 40 + i, bmi = 20 + i / 10,
  hx_chf = as.integer(i %% 4L == 0L), lvmassi = 80 + i,
  iv_opyrs = i + 0.25
)
utils::write.csv(built, file.path(root, "datasets", "built.csv"), row.names = FALSE)
utils::write.csv(built[1:24, ], file.path(root, "datasets", "complete_cases.csv"), row.names = FALSE)
writeLines(c(
  'title "Synthetic descriptive tables";',
  '%desc_tab(vartype=category, input=built, by=,',
  '  varlist=/* Demography */ female race_grp /* Procedure */ repair);',
  '%desc_tab(vartype=continuous, input=built, by=,',
  '  varlist=/* Demography */ age bmi /* Follow-up */ iv_dead);'
), file.path(root, "descriptive", "dc.tables.sas"))
writeLines(c("Synthetic descriptive tables", "Demography", "Procedure", "Follow-up"),
           file.path(root, "descriptive", "dc.tables.lst"))
writeLines("NOTE: There were 40 observations read from the data set BUILT.",
           file.path(root, "descriptive", "dc.tables.log"))
writeLines("{\\rtf1\\ansi Synthetic descriptive tables\\par Demography\\par Procedure\\par Follow-up}",
           file.path(root, "documents", "general.rtf"))
writeLines(c(
  "data followup; set built; run;",
  "proc sort data=followup; by dead; run;",
  "proc means data=followup; var iv_dead iv_fup; by dead; run;",
  "proc print data=followup; var iv_dead iv_fup; if dead=0; run;"
), file.path(root, "descriptive", "dc.gfup.sas"))
writeLines(c(
  "data trends; set built; year=floor(iv_opyrs)+1985; run;",
  "%let percent = hx_chf;", "%let continuous = lvmassi;",
  "label hx_chf = 'Heart failure';", "label lvmassi = 'LV mass index';",
  "axis1 order=(1985 to 2025 by 5);", "axis2 order=(0 to 100 by 20);",
  "axis3 order=(80 to 140 by 10);", "proc gplot data=trends;",
  "plot hx_chf*year / haxis=axis1 vaxis=axis2;",
  "plot lvmassi*year / haxis=axis1 vaxis=axis3; run; quit;"
), file.path(root, "graphs", "dp.trends.sas"))
writeLines(c(
  "set complete_cases;", "%let pref_time_var=year;",
  "%let variables=age bmi female repair;", "%let exclude=;",
  "%let ncol=2;", "%let nrow=2;", "%let unique_limit=6;",
  "%let show_percent=0;"
), file.path(root, "descriptive", "dp.postage.sas"))
legacy_files <- list.files(root, recursive = TRUE, full.names = TRUE)
legacy_checksums <- tools::md5sum(legacy_files)
```

## Start with the legacy tree

The synthetic programs state the variables and dataset each job uses. The
listing, log, and small RTF supply evidence to read beside the first migration
report. They are newly written examples, so no patient records are needed.

```{r}
#| label: before-adoption
list.files(root, recursive = TRUE)
```

## Adopt the study and register its data

The Study Tracker ID identifies the study. Here `42L` is part of the synthetic
example. Adoption keeps the bare folder names, then adds the package-owned
identity and environment files. Recovery of missing study state is a separate
operation.

```{r}
#| label: adopt-register
study_setup(root, study = "Synthetic legacy study",
            study_tracker_id = 42L, adopt = TRUE)
register_data(root, built = "built.csv", event = "dead", time = "iv_dead",
              role = "study", population = "Synthetic full cohort")
register_data(root, built = "complete_cases.csv", event = "dead",
              time = "iv_dead", dataset = "complete_cases",
              role = "named", population = "Synthetic complete cases")
study_status(root)
stopifnot(identical(unname(tools::md5sum(legacy_files)),
                    unname(legacy_checksums)))
```

Registration records the file checksum and cohort counts for each dataset.
The study cohort has 40 observations; the named subset has 24. We use its first
24 rows only to make this example deterministic. A real complete-case dataset
needs a stated exclusion rule in its data build.

```{r}
#| label: cohort-counts
cfg <- study_config(root)
study_data <- read_built(cfg, dataset = "study")
subset_data <- read_built(cfg, dataset = "complete_cases")
study_counts <- cohort_counts(study_data, cfg, dataset = "study")
subset_counts <- cohort_counts(subset_data, cfg, dataset = "complete_cases")
cohorts <- data.frame(
  dataset = c("study", "complete_cases"),
  rows = c(study_counts$n, subset_counts$n),
  events = c(study_counts$n_events, subset_counts$n_events)
)
cohorts
```

The named registration preserves the study-wide counts. A job using
`complete_cases` must check that dataset's contract; checking the 40-row study
contract cannot establish that a 24-row job cohort is correct. Further
filtering inside a job needs its own reference counts and checks.

## Migrate one job at a time

Each call selects one template and its source-specific adapter. The source
supplies analysis choices; logs, listings and RTF/DOCX references add evidence.
Migration writes an editable QMD and a report beside it. Existing jobs or
reports stop the call, so keep your edited job when you repeat this workflow.

### Descriptive tables

The `%desc_tab` calls supply the variable groups and types. The migrated job
will write one editable Word table. Review the SAS presentation choices before
accepting its summary statistics and display.

```{r}
#| label: migrate-tables
tables <- migrate_job(
  "descriptive/dc.tables.sas", "cohort", "eda", "dc", qualifier = "tables",
  lst = "descriptive/dc.tables.lst", log = "descriptive/dc.tables.log",
  reference = "documents/general.rtf", dir = root
)
```

### Goodness of follow-up

This job reads the registered event and follow-up fields. Confirm the time
units from the data build. Its recorded-interval checks do not establish
completeness against an administrative close date, and patient identifiers
remain disabled.

```{r}
#| label: migrate-gfup
gfup <- migrate_job(
  "descriptive/dc.gfup.sas", "cohort", "eda", "dc", qualifier = "gfup",
  dir = root
)
```

### Trends

The explicit year origin and interval expression define whole operation years.
`hv_trends()` owns the summaries and smoothing; compare these figures with
legacy output before treating a method change as equivalent.

```{r}
#| label: migrate-trends
trends <- migrate_job(
  "graphs/dp.trends.sas", "cohort", "eda", "dp", qualifier = "trends",
  dir = root
)
```

### Postage stamps on the named subset

This source says `set complete_cases;`. The adapter resolves that name to the
registered subset, so this job reads 24 rows. It makes a panel for each chosen
variable and arranges the four panels on one PNG page.

```{r}
#| label: migrate-postage
postage <- migrate_job(
  "descriptive/dp.postage.sas", "cohort", "eda", "dp", qualifier = "postage",
  dir = root
)
jobs <- c(tables = tables, gfup = gfup, trends = trends, postage = postage)
reports <- sub("[.]qmd$", "-migration.md", jobs)
```

## Read the reports and resolve the markers

Read all four reports before editing the jobs. They record translated values,
unresolved decisions, ignored source sections, checksums, and evidence line
numbers. This first report includes the synthetic listing, log and RTF.

```{r}
#| label: migration-report
#| results: asis
cat(paste(readLines(reports[["tables"]]), collapse = "\n"))
```

For this synthetic example we accept the migrated groups, the whole cohort for
the first three jobs, and the named subset for postage. We choose no group
comparison, median summaries with the 15th and 85th percentiles, no table
abbreviations or correlations, and `dc-tables.docx` as the Word filename.
Follow-up intervals are years and no extra event cross-tabs are needed.
The trend figures use the stated year origin, breaks and variable labels with
no subgroup filter. Postage uses the four stated variables and a 2-by-2 grid.

The following helper acknowledges only exact, reviewed comment lines. It is a
way to repeat these synthetic decisions while rendering this article. In a
study job, edit each marked choice against the source and protocol, then remove
that marker. Replacing every marker without review can turn an incomplete
migration into a misleading finished report.

```{r}
#| label: review-helper
acknowledge <- function(job, reviewed_lines) {
  text <- readLines(job, warn = FALSE)
  stopifnot(all(reviewed_lines %in% text))
  selected <- text %in% reviewed_lines
  text[selected] <- gsub("EDIT:", "REVIEWED:", text[selected], fixed = TRUE)
  writeLines(text, job)
}
```

```{r}
#| label: review-synthetic-choices
#| code-fold: true
acknowledge(tables, c(
  "<!-- EDIT: name the job this replaces (descriptive/dc.tables.*) and say whether it",
  "# EDIT: the analysis set this job reads: the columns and excluded rows declared",
  "# EDIT: the variables to report, grouped. Each name is a section heading and",
  "# EDIT: review SAS presentation choices in the migration report.",
  "# EDIT: confirm comparison, summaries, percentiles, abbreviations and Word filename.",
  "# EDIT: force a bucket where the data guess wrong, e.g. an ordinal score you",
  "# EDIT: NULL skips the correlation variant (dc.tables.ods_<topic>.sas). To run"
))
acknowledge(gfup, c(
  "<!-- EDIT: name the job this replaces (descriptive/dc.gfup). -->",
  "# EDIT: the analysis set this job reads: declare its columns and excluded rows",
  "# EDIT: review unresolved migration evidence and confirm follow-up units in years.",
  "# EDIT: optional event-coding cross-tabs, each a vector of column names."
))
acknowledge(trends, c(
  "<!-- EDIT: name the job this replaces (an older graphs/dp.trends job) and list",
  "# EDIT: the calendar year of operation, as a WHOLE year. hv_trends() puts one",
  "# EDIT: one entry per figure. `cols` are the columns drawn together: several",
  "# EDIT: optional. x-axis breaks shared by every figure, e.g. seq(1985, 2025, 5).",
  "# EDIT: optional. Each entry is a filter, given as a function of the data that",
  "# EDIT: review source filters and subgroup intent before accepting the whole cohort."
))
acknowledge(postage, c(
  "<!-- EDIT: name the job this replaces (tp.dp.EDA_barplots_scatterplots*.R). -->",
  "# EDIT: select the registered dataset or analysis set this job reads.",
  "# EDIT: review unresolved postage source choices in the migration report."
))
```

```{r}
#| label: check-reviewed-jobs
remaining <- vapply(jobs, function(job) {
  sum(grepl("EDIT:", readLines(job), fixed = TRUE))
}, integer(1L))
remaining
stopifnot(all(remaining == 0L))
```

## Render and inspect the output

Render each job from its own directory so its guard can read the authored QMD.
The guard stops if a marker remains. The commands below use the normal render
path with every analysis guard active.

```{r}
#| label: render-jobs
for (job in jobs) {
  quarto::quarto_render(job, execute_dir = dirname(job), quiet = FALSE)
}
```

The authored jobs and migration reports stay flat in their legacy folders.
Generated figures and the Word table go under the `cohort-eda/` output set.
Check both the artifact paths and their contents: a file existing is only the
first check. The Word structural check catches document-format defects; it
does not prove numerical agreement with SAS.

```{r}
#| label: inspect-outputs
outputs <- list.files(root, pattern = "[.](html|docx|png)$", recursive = TRUE)
outputs
word <- file.path(root, "documents", "cohort-eda", "dc-tables.docx")
png <- file.path(root, "graphs", "cohort-eda", "dp-postage-page-01.png")
stopifnot(file.exists(word), file.exists(png))
hvtiRtables::hv_check_docx(word)
knitr::include_graphics(png)
stopifnot(identical(unname(tools::md5sum(legacy_files)),
                    unname(legacy_checksums)))
```

Open the Word file to compare row order, labels, counts, summaries, precision
and footnotes with the legacy reference. Inspect each trend and postage figure
for variable selection, axes and cohort. Record any methodological differences
alongside the migration report before using the output in a study document.
