---
title: "Start a new study from a delivered dataset"
author: "John Ehrlinger"
vignette: >
  %\VignetteIndexEntry{Start a new study}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    embed-resources: true
---

This tutorial starts with an empty directory and an analysis dataset that has
just been delivered. It takes the study through setup, data registration and a
first descriptive job, then through one analysis chain, so you can see how the
templates hand work to one another.

If the study directory already exists and holds jobs and data, start with
[Adopt an existing study](study-setup.html) instead. That guide keeps the
existing layout and files.

The executable example uses a 40-row synthetic dataset generated below. Its
files stay in a temporary directory and disappear after the article renders.

## Before you start

A delivered dataset here means one analysis-ready file, CSV or SAS transport,
already built from the source extract. Building that file is a separate step;
this guide begins once it exists. Keep it on the study share: patient data
never enter a git repository.

The study itself must exist in Study Tracker before its directory does. Create
the study record there first and note its Study Tracker ID: every later step
takes its identity from that record.

```{r}
#| label: synthetic-setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(hvtiRutilities)
library(hvtiRtemplates)
workspace <- withr::local_tempdir(.local_envir = environment())
new_root <- file.path(workspace, "new-study")
i <- seq_len(40L)
delivered <- data.frame(
  ccfid = i,
  dead = i %% 2L,
  iv_dead = i / 10,
  age = 40 + i,
  female = i %% 2L
)
```

## Create the study

Create the directory with the `study-setup` command from `qhsprograms`, giving
it the Study Tracker ID and the directory name. Preview first; the dry run
writes nothing:

```sh
study-setup --dry-run 42 example_study
study-setup 42 example_study
```

`study-setup` reads the Tracker record, read-only, and turns the new directory
into a study. It writes `_study.yml`, which carries the study's name and
Tracker ID, a README linking the Tracker record, an RStudio project file,
environment defaults, `renv`, and the numbered folders every job is placed in:

| folder | holds |
|---|---|
| `00_datasets` | the registered data |
| `10_descriptive` | descriptive tables and checks |
| `20_distributions` | life tables and hazard models |
| `30_analyses` | multivariable and machine-learning models |
| `40_graphs` | figures |
| `50_documents` | Word tables and manuscript material |
| `90_estimates` | saved output that other jobs read |

`90_estimates` is numbered last on purpose. It holds results, not jobs, and a
job that needs another job's result reads it from there.

Now open the directory in RStudio as a Project (**File > New Project >
Existing Directory**) and work from its `.Rproj` from then on. Select the R
version and load the hvtiR packages as described in
[Adopt an existing study](study-setup.html#open-the-study-as-an-rstudio-project):

```r
library(hvtiRutilities)
library(hvtiRtemplates)
```

Underneath, `study-setup` calls `hvtiRutilities::study_setup()` with the
Tracker record's values. This article cannot reach Study Tracker while it
renders, so the executable example calls `study_setup()` directly, in a
temporary directory:

```{r}
#| label: create-study
study_setup(new_root, study = "Synthetic new study", study_tracker_id = 42L)
sort(list.files(new_root))
```

## Register the dataset

Copy the delivered file into `00_datasets`, then register it. Registration
records the file's size, checksum and the population it describes in
`manifest.yaml`:

```r
register_data(
  built = "built.csv",
  role = "study",
  population = "Patients meeting the study inclusion criteria"
)
```

Every job reads data through this registration rather than through a file
path. That is what lets a rendered job record exactly which bytes it read: a
job that opens a path directly cannot tell you, a year later, which version of
the file it saw.

```{r}
#| label: register-data
utils::write.csv(
  delivered, file.path(study_dir("datasets", new_root), "built.csv"), row.names = FALSE
)
register_data(
  new_root,
  built = "built.csv",
  role = "study",
  population = "Synthetic full cohort"
)
```

Registration names no endpoint and no cohort. Those belong to the jobs that
analyse them, so one registered dataset can serve every job in the study.

## Scaffold the first EDA job

Start with the general descriptive checks:

```r
general <- open_job("dc", subject = "cohort", type = "eda", qualifier = "general")
```

`open_job()` creates the job and opens it in RStudio. The file is named
`<subject>-<type>-<prefix>-<qualifier>.qmd`, here
`10_descriptive/cohort-eda-dc-general.qmd`. The subject names what the job is
about and need not be a statistical endpoint: `cohort` describes the whole
registered population. The type names the stage, and the pair keeps a set of
related jobs and their outputs together.

The executable example scaffolds without opening an editor:

```{r}
#| label: scaffold-eda
general <- add_job("dc", subject = "cohort", type = "eda", dir = new_root, qualifier = "general")
sub(paste0(new_root, "/"), "", general, fixed = TRUE)
```

Each job carries `EDIT:` markers where the template cannot know the study's
answer: variables, labels, groups. Work through them from the top. A job that
still contains one is unfinished.

The first `add_job()` in a study also installs its provenance hooks. It adds
pre-render and post-render entries to `_quarto.yml`, keeping any settings
already there, and copies the hook scripts into `.hvtiR/hooks/`:

```{r}
#| label: hooks
cat(readLines(file.path(new_root, "_quarto.yml")), sep = "\n")
list.files(file.path(new_root, ".hvtiR"), recursive = TRUE)
```

## Render and read the provenance

Render the job with the **Render** button, or from the Console:

```r
render_job(general, final = TRUE)
```

Both run the same hooks. While the job executes it records what it read; once
Quarto has written the HTML, the post-render hook publishes a sidecar beside
it, `cohort-eda-dc-general.provenance.json`. An abridged example from a
synthetic study:

```json
{
  "job": "cohort-eda-dc-general",
  "rendered": "2026-09-23T14:42:04Z",
  "study": { "name": "Synthetic new study", "file": "_study.yml", "sha256": "35f29fb1..." },
  "r": { "version": "4.6.1", "platform": "aarch64-apple-darwin23" },
  "data": [
    {
      "dataset": "study",
      "path": "00_datasets/built.csv",
      "role": "study",
      "bytes": 1051,
      "sha256": "3d6f6b0c..."
    }
  ],
  "artifacts": [],
  "source": "10_descriptive/cohort-eda-dc-general.qmd",
  "subject": "cohort",
  "type": "eda",
  "output": { "file": "cohort-eda-dc-general.html", "sha256": "95455dd3..." }
}
```

Read it as a receipt. `data` names the registered file and the checksum of the
bytes the job read. `packages` and `renv_lock`, omitted here, record every
package version and the study's lockfile.
`output` carries the checksum of the HTML it describes, so a sidecar can be
matched to its report and to nothing else.

If a render fails, the report and sidecar already on disk are left as they
were. A new sidecar is only ever published beside the HTML it describes. One
job renders at a time in a study; a second render started while one is running
stops and asks you to wait.

## A first analysis chain: death

Most analyses are chains: one job saves a result that the next reads. The
hazard chain has three jobs. `ac` computes the actuarial life table, `hz`
fits the parametric hazard model, and `hp` plots the two together.

```r
open_job("ac", subject = "death", type = "hz")
open_job("hz", subject = "death", type = "hz")
open_job("hp", subject = "death", type = "hz")
```

All three share one subject and one type, and that is what connects them. The
pair `("death", "hz")` names a set, and each job writes and reads its saved
results in the set's own folder, `90_estimates/death-hz/`. `hp` finds `ac.rds`
and `hz.rds` there without a path to edit. A second analysis of the same
endpoint, a random survival forest say, takes a different type and so a
different folder; the two chains cannot overwrite each other's life table.

```{r}
#| label: scaffold-chain
chain <- c(
  ac = add_job("ac", subject = "death", type = "hz", dir = new_root),
  hz = add_job("hz", subject = "death", type = "hz", dir = new_root),
  hp = add_job("hp", subject = "death", type = "hz", dir = new_root)
)
sub(paste0(new_root, "/"), "", chain, fixed = TRUE)
```

The filenames read `death-hz-ac.qmd`, `death-hz-hz.qmd` and `death-hz-hp.qmd`:
subject, type, then the template. `hz` appears twice in the second because the
set is named for its method and the job is that method's template.

```{=html}
<figure style="margin: 1.5rem 0;">
<svg viewBox="0 0 640 210" role="img" aria-labelledby="chain-title" style="width: 100%; height: auto; color: inherit;"
     xmlns="http://www.w3.org/2000/svg" font-family="system-ui, sans-serif" font-size="13">
  <title id="chain-title">The death hazard chain: ac and hz save handoffs that hp reads</title>
  <defs>
    <marker id="arrow" viewBox="0 0 10 10" refX="9" refY="5" markerWidth="7" markerHeight="7" orient="auto">
      <path d="M0,0 L10,5 L0,10 z" fill="currentColor"/>
    </marker>
  </defs>
  <g fill="none" stroke="currentColor" stroke-width="1.5">
    <rect x="10" y="20" width="150" height="46" rx="6"/>
    <rect x="10" y="130" width="150" height="46" rx="6"/>
    <rect x="245" y="20" width="150" height="46" rx="6" stroke-dasharray="4 3"/>
    <rect x="245" y="130" width="150" height="46" rx="6" stroke-dasharray="4 3"/>
    <rect x="480" y="75" width="150" height="46" rx="6"/>
    <path d="M160,43 L243,43" marker-end="url(#arrow)"/>
    <path d="M160,153 L243,153" marker-end="url(#arrow)"/>
    <path d="M395,43 C440,43 440,90 478,95" marker-end="url(#arrow)"/>
    <path d="M395,153 C440,153 440,106 478,101" marker-end="url(#arrow)"/>
  </g>
  <g fill="currentColor" text-anchor="middle">
    <text x="85" y="41">death-hz-ac.qmd</text>
    <text x="85" y="58" font-size="11">actuarial life table</text>
    <text x="85" y="151">death-hz-hz.qmd</text>
    <text x="85" y="168" font-size="11">parametric hazard</text>
    <text x="320" y="41">ac.rds</text>
    <text x="320" y="58" font-size="11">90_estimates/death-hz/</text>
    <text x="320" y="151">hz.rds</text>
    <text x="320" y="168" font-size="11">90_estimates/death-hz/</text>
    <text x="555" y="96">death-hz-hp.qmd</text>
    <text x="555" y="113" font-size="11">overlay plot</text>
    <text x="320" y="200" font-size="11">each job also publishes a .provenance.json beside its HTML</text>
  </g>
</svg>
<figcaption>Solid boxes are jobs; dashed boxes are the saved handoffs they share through the set's folder.</figcaption>
</figure>
```

Render in chain order: `ac`, then `hz`, then `hp`. Each saved handoff carries
its own lineage, the registered data it was built from, so `hp` can check its
inputs before it plots them. It stops, with a message saying which job to
rerun, when a handoff has no lineage or when `ac` and `hz` were built from
different data. Rerender both upstream jobs after the registered data change.
`hp`'s sidecar then lists the data both handoffs were built from and each
handoff's checksum.

## Where to go next

`template_list()` lists every template with its prefix, qualifier and folder:

```{r}
#| label: templates
head(template_list()[, c("name", "prefix", "qualifier", "folder")], 10)
```

The other chains follow the same pattern: a fitting job saves a handoff that
later jobs read. The random forest jobs (`rfc`, `rfr`, `rfs`) pair a `-fit`
job with an `-explain` job; the logistic family (`lm`) fits and validates
outcome and propensity models; `bh` reports bootstrap variable selection.
