---
title: "Start a new study from a delivered dataset"
author: "John Ehrlinger"
vignette: >
  %\VignetteIndexEntry{Start a new study}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    embed-resources: true
---

This tutorial starts with an empty directory and an analysis dataset that has
just been delivered. It takes the study through setup, data registration and a
first descriptive job, then through one analysis chain, so you can see how the
templates hand work to one another.

If the study directory already exists and holds jobs and data, start with
[Adopt an existing study](study-setup.html) instead. That guide keeps the
existing layout and files.

The executable example uses a 40-row synthetic dataset generated below. Its
files stay in a temporary directory and disappear after the article renders.

## Before you start

A delivered dataset here means one analysis-ready file, CSV or SAS transport,
already built from the source extract. Building that file is a separate step;
this guide begins once it exists. Keep it on the study share: patient data
never enter a git repository.

Create the study directory, open it in RStudio as a Project
(**File > New Project > Existing Directory**) and work from its `.Rproj` from
then on. Select the R version and load the hvtiR packages as described in
[Adopt an existing study](study-setup.html#open-the-study-as-an-rstudio-project).
Those steps are the same for a new study, so they are kept in one place.

```r
library(hvtiRutilities)
library(hvtiRtemplates)
```

```{r}
#| label: synthetic-setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(hvtiRutilities)
library(hvtiRtemplates)
workspace <- withr::local_tempdir(.local_envir = environment())
new_root <- file.path(workspace, "new-study")
i <- seq_len(40L)
delivered <- data.frame(
  ccfid = i,
  dead = i %% 2L,
  iv_dead = i / 10,
  age = 40 + i,
  female = i %% 2L
)
```

## Create the study

`study_setup()` turns an empty directory into a study. It writes `_study.yml`,
which names the study and its Study Tracker ID, an RStudio project file,
environment defaults, and the numbered folders every job is placed in:

| folder | holds |
|---|---|
| `00_datasets` | the registered data |
| `10_descriptive` | descriptive tables and checks |
| `20_distributions` | life tables and hazard models |
| `30_analyses` | multivariable and machine-learning models |
| `40_graphs` | figures |
| `50_documents` | Word tables and manuscript material |
| `90_estimates` | saved output that other jobs read |

`90_estimates` is numbered last on purpose. It holds results, not jobs, and a
job that needs another job's result reads it from there.

```r
study_setup(
  root = ".",
  study = "Study title from Study Tracker",
  study_tracker_id = 42L
)
```

The executable example uses its temporary directory:

```{r}
#| label: create-study
study_setup(new_root, study = "Synthetic new study", study_tracker_id = 42L)
sort(list.files(new_root))
```

## Register the dataset

Copy the delivered file into `00_datasets`, then register it. Registration
records the file's size, checksum and the population it describes in
`manifest.yaml`:

```r
register_data(
  built = "built.csv",
  role = "study",
  population = "Patients meeting the study inclusion criteria"
)
```

Every job reads data through this registration rather than through a file
path. That is what lets a rendered job record exactly which bytes it read: a
job that opens a path directly cannot tell you, a year later, which version of
the file it saw.

```{r}
#| label: register-data
utils::write.csv(
  delivered, file.path(study_dir("datasets", new_root), "built.csv"), row.names = FALSE
)
register_data(
  new_root,
  built = "built.csv",
  role = "study",
  population = "Synthetic full cohort"
)
```

Registration names no endpoint and no cohort. Those belong to the jobs that
analyse them, so one registered dataset can serve every job in the study.

## Scaffold the first EDA job

Start with the general descriptive checks:

```r
general <- open_job("dc", subject = "cohort", type = "eda", qualifier = "general")
```

`open_job()` creates the job and opens it in RStudio. The file is named
`<subject>-<type>-<prefix>-<qualifier>.qmd`, here
`10_descriptive/cohort-eda-dc-general.qmd`. The subject names what the job is
about and need not be a statistical endpoint: `cohort` describes the whole
registered population. The type names the stage, and the pair keeps a set of
related jobs and their outputs together.

The executable example scaffolds without opening an editor:

```{r}
#| label: scaffold-eda
general <- add_job("dc", subject = "cohort", type = "eda", dir = new_root, qualifier = "general")
sub(paste0(new_root, "/"), "", general, fixed = TRUE)
```

Each job carries `EDIT:` markers where the template cannot know the study's
answer: variables, labels, groups. Work through them from the top. A job that
still contains one is unfinished.

The first `add_job()` in a study also installs its provenance hooks. It adds
pre-render and post-render entries to `_quarto.yml`, keeping any settings
already there, and copies the hook scripts into `.hvtiR/hooks/`:

```{r}
#| label: hooks
cat(readLines(file.path(new_root, "_quarto.yml")), sep = "\n")
list.files(file.path(new_root, ".hvtiR"), recursive = TRUE)
```

## Render and read the provenance

Render the job with the **Render** button, or from the Console:

```r
render_job(general, final = TRUE)
```

Both run the same hooks. While the job executes it records what it read; once
Quarto has written the HTML, the post-render hook publishes a sidecar beside
it, `cohort-eda-dc-general.provenance.json`. An abridged example from a
synthetic study:

```json
{
  "job": "cohort-eda-dc-general",
  "rendered": "2026-09-23T14:42:04Z",
  "study": { "name": "Synthetic new study", "file": "_study.yml", "sha256": "35f29fb1..." },
  "r": { "version": "4.6.1", "platform": "aarch64-apple-darwin23" },
  "data": [
    {
      "dataset": "study",
      "path": "00_datasets/built.csv",
      "role": "study",
      "bytes": 1051,
      "sha256": "3d6f6b0c..."
    }
  ],
  "artifacts": [],
  "source": "10_descriptive/cohort-eda-dc-general.qmd",
  "subject": "cohort",
  "type": "eda",
  "output": { "file": "cohort-eda-dc-general.html", "sha256": "95455dd3..." }
}
```

Read it as a receipt. `data` names the registered file and the checksum of the
bytes the job read. `packages`, omitted here, lists every package version.
`output` carries the checksum of the HTML it describes, so a sidecar can be
matched to its report and to nothing else.

If a render fails, the report and sidecar already on disk are left as they
were. A new sidecar is only ever published beside the HTML it describes. One
job renders at a time in a study; a second render started while one is running
stops and asks you to wait.
