---
title: "From SAS descriptive jobs to R"
author: "John Ehrlinger"
date: today
vignette: >
  %\VignetteIndexEntry{From SAS descriptive jobs to R}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Who this is for

This guide is for a SAS biostatistician who knows the `tp.dc.*` templates and
wants to run the same descriptive work in an hvtiR study. "R equivalent" means
the same numbers when the method is the same, and an explicit difference when
the R job makes a different choice. It does not mean translating every SAS
statement line by line.

## Install once

Install the hvtiR package set, then check it before starting a study job:

```r
hvtiR::install()
hvtiR::status()
```

## The map

| SAS job | R job | template | engine |
|:--|:--|:--|:--|
| `descriptive/dc.tables.ods.sas` (`%desc_tab`) | `dc-tables` | `10_descriptive/dc-tables.qmd` | `hv_tbl_summary()`; `hv_correlation_table()` and `hv_correlation_matrix()` |
| `descriptive/dc.gfup.sas` | `dc-gfup` | `10_descriptive/dc-gfup.qmd` | `proc_means()` |
| `graphs/dp.trends.R` | `dp-trends` | `40_graphs/dp-trends.qmd` | `hv_trends()` |
| `tp.dp.EDA_barplots_scatterplots*.R` | `dp-postage` | `10_descriptive/dp-postage.qmd` | `hv_eda()` |

## Five steps for every job

1. Scaffold the job. For example,
   `new_job("dc", "cohort", "eda", qualifier = "tables")` writes
   `descriptive/cohort-eda-dc-tables.qmd`.
2. Work every `EDIT:` marker. The edit guard stops a render while any remain.
3. Render from the study root with
   `quarto render descriptive/cohort-eda-dc-tables.qmd`.
4. Compare the result with the SAS `.lst`, section by section.
5. Keep the authored job flat in `descriptive/`; its generated artifacts are
   filed beneath `descriptive/cohort-eda/`.

## Descriptive tables and correlations: `dc-tables`

The SAS job commonly calls `%desc_tab` once for categorical variables and once
for continuous variables. The R job supplies both sets to one table. SAS
comment banners become the names in `GROUPS`; `by=` and `byvalue=` become `BY`;
and `perctile=15 50 85` becomes `continuous_stat = "both"` with percentiles 15
and 85. The `dc.tables.ods_<topic>.sas` correlation variant becomes `CORR` in
the same job.

```sas
%desc_tab(vartype=category,   varlist=female diabetes nyha_pr);
%desc_tab(vartype=continuous, varlist=age bmi, perctile=15 50 85);
```

```{r dc-tables}
if (requireNamespace("hvtiRtables", quietly = TRUE) &&
      requireNamespace("hvtiPlotR", quietly = TRUE)) {
  d <- hvtiPlotR::sample_correlation_data(n = 80)
  d$female <- rep(0:1, length.out = nrow(d))
  groups <- list(
    Demography = c("female"),
    Measurements = c("a1c", "glucose", "creatinine")
  )
  hvtiRtables::hv_tbl_summary(
    d,
    groups = groups,
    binary = "female",
    continuous = c("a1c", "glucose", "creatinine"),
    continuous_stat = "both",
    percentiles = c(15, 85),
    compare = "none"
  )
  hvtiRtables::hv_correlation_table(
    d,
    vars = c("glucose", "creatinine"),
    with = "a1c"
  )
}
```

## Goodness of follow-up: `dc-gfup`

The SAS `proc means` and `proc univariate` calls become `proc_means()`. The R
job reports follow-up distributions overall, by vital status, and among
survivors. It deliberately omits the patient-level listing: a rendered report
should answer the completeness question without exposing an identifier.

```sas
proc means data=built n mean std min max;
  class dead;
  var iv_dead iv_fup;
run;
```

```{r dc-gfup}
if (requireNamespace("hvtiRutilities", quietly = TRUE)) {
  d <- hvtiRutilities::generate_survival_data(n = 80)
  d$iv_fup <- pmin(d$iv_dead, 0.85 * d$iv_dead)
  hvtiRutilities::proc_means(
    d,
    vars = c("iv_dead", "iv_fup"),
    class = "dead",
    stats = c("n", "nmiss", "mean", "std", "min", "p15", "median", "p85", "max")
  )
}
```

## Trends over operation year: `dp-trends`

The shipped `dp-trends` template replaces hand-built annual summaries and
per-series `smooth.spline(df = 6)` calls with `hv_trends()`. Its smoother is
LOESS, not the SAS-era smoothing spline, so compare the old and new figures by
shape rather than pointwise equality.

```r
fit <- smooth.spline(year, annual_mean, df = 6)
```

```{r dp-trends}
if (requireNamespace("hvtiPlotR", quietly = TRUE)) {
  d <- hvtiPlotR::sample_eda_data(n = 120)
  trend_data <- data.frame(year = d$year, value = d$ef, group = "EF")
  plot(hvtiPlotR::hv_trends(trend_data))
}
```

## EDA postage stamps: `dp-postage`

The older `Function_DataPlotting()` loops become one `hv_eda()` call per entry
in `VARS`. Character and factor columns produce bars. A numeric column produces
bars only when it has at most six distinct non-missing values and every value
is a non-negative whole number no greater than six; otherwise it produces a
scatter panel. The template lays those panels out in printable pages.

```r
for (v in variables) Function_DataPlotting(data, v, "year")
```

```{r dp-postage}
if (requireNamespace("hvtiPlotR", quietly = TRUE)) {
  d <- hvtiPlotR::sample_eda_data(n = 120)
  variables <- c(male = "Male", ef = "Ejection fraction")
  panels <- lapply(names(variables), function(v) {
    plot(hvtiPlotR::hv_eda(d, x_col = "year", y_col = v, y_label = variables[[v]]))
  })
  panels[[1]]
  panels[[2]]
}
```

## Where the numbers differ

Percentiles use SAS `QNTLDEF=5`, equivalent to R's `type = 2`;
`proc_means()` and the table engine already make that choice. Pin `type = 2`
in any hand-written `quantile()` call. `N` is the non-missing count, not always
the number of rows. Continuous-variable p-values are non-parametric.
Correlation intervals are 68% Fisher intervals by default. Finally,
`%desc_tab` wrote separate categorical and continuous RTF files; the R job
writes one DOCX table.

## What is not ported

The templates do not reproduce RTF output; they produce DOCX and self-contained
HTML instead. They also do not reproduce identified patient listings. Use the
distribution summaries to assess completeness, then investigate any necessary
patient-level records inside the governed study environment rather than in a
portable report.
