---
title: "Adopt an existing study for descriptive and EDA work"
author: "John Ehrlinger"
vignette: >
  %\VignetteIndexEntry{Adopt a study for descriptive and EDA work}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    embed-resources: true
---

This tutorial starts with a study that already has working directories and an
existing dataset. It may also contain older SAS or R jobs, but those are not
required. We will open that directory as an RStudio
Project, remove obsolete setup files, adopt the study into the hvtiR contract,
register its data, write an `eda` analysis set and create the descriptive and
EDA jobs used before a specific analysis begins.

The result is not an analysis result yet. It is a reproducible starting point:
the study has a known root, a recorded data contract and jobs whose remaining
study decisions are visible as `EDIT:` markers.

The executable example uses 40 synthetic observations in one existing study
dataset. Every file stays in a temporary directory. During rendering, that
directory is removed after the R chunks run; the displayed figure is embedded
before cleanup. Run the chunks in order in one interactive session to inspect
the files yourself. For the statistical mapping behind the jobs, see
[From SAS descriptive jobs to R](sas-to-r-descriptive.html).

The render steps require Quarto and the packages declared by the job templates,
including `hvtiRdatabuild >= 0.2.1`, `hvtiRtables >= 1.0.1`, and
`hvtiPlotR >= 2.7.14`. A missing or older dependency stops the render and names
the required version.

## Open the existing study in RStudio

In RStudio, choose **File > New Project > Existing Directory** and select the
study directory. Open the resulting `.Rproj` file whenever you work on the
study. RStudio starts the session in that directory, so the project supplies
the study root. Do not change the working directory manually.

Before changing anything, confirm that R is looking at the intended directory:

```r
project_root <- getwd()
normalizePath(project_root)
```

`getwd()` reads the location RStudio established; it does not change it. The
tutorial's executable chunks use `adopted_root` only because they build a
disposable study while this page renders. In a real study, pass
`project_root` to the initial adoption call. After `_study.yml` exists, hvtiR
functions such as `study_root()`, `open_job()` and `render_job()` find the root
for themselves.

### Choose the R version before adoption

If the study already has an `renv.lock`, its `R` `Version` entry records the R
version the study is already pinned to. Select that version in RStudio and keep
it for the adoption. Do not silently move an already pinned study to a new R
version.

If there is no `renv.lock`, select **R 4.6** in RStudio before continuing. The
first `renv::snapshot()` will then record R 4.6 for the study instead of the
older 4.4.1 default. Restart the R session after changing versions and confirm
the active version:

```r
R.version.string
```

```{r}
#| label: synthetic-setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(hvtiRutilities)
library(hvtiRtemplates)
workspace <- withr::local_tempdir(.local_envir = environment())
adopted_root <- file.path(workspace, "legacy-study")
folders <- c("datasets", "descriptive", "distributions", "analyses",
             "graphs", "documents", "estimates")
dir.create(adopted_root)
for (folder in folders) dir.create(file.path(adopted_root, folder))
i <- seq_len(40L)
built <- data.frame(
  ccfid = i,
  dead = i %% 2L, iv_dead = i / 10, iv_fup = i / 10 + 1,
  year = 2000L + i %% 10L, female = i %% 2L, race_grp = 1L + i %% 3L,
  repair = as.integer(i %% 3L == 0L), age = 40 + i, bmi = 20 + i / 10,
  hx_chf = as.integer(i %% 4L == 0L), lvmassi = 80 + i,
  iv_opyrs = i + 0.25
)
utils::write.csv(built, file.path(adopted_root, "datasets", "built.csv"), row.names = FALSE)
writeLines(c(
  'title "Synthetic descriptive tables";',
  "%desc_tab(vartype=category, input=built, by=,",
  "  varlist=/* Demography */ female race_grp /* Procedure */ repair);",
  "%desc_tab(vartype=continuous, input=built, by=,",
  "  varlist=/* Demography */ age bmi /* Follow-up */ iv_dead);"
), file.path(adopted_root, "descriptive", "dc.tables.sas"))
writeLines(c("Synthetic descriptive tables", "Demography", "Procedure", "Follow-up"),
           file.path(adopted_root, "descriptive", "dc.tables.lst"))
writeLines("NOTE: There were 40 observations read from the data set BUILT.",
           file.path(adopted_root, "descriptive", "dc.tables.log"))
writeLines("{\\rtf1\\ansi Synthetic descriptive tables\\par Demography\\par Procedure\\par Follow-up}",
           file.path(adopted_root, "documents", "general.rtf"))
writeLines(c(
  "data followup; set built; run;",
  "proc sort data=followup; by dead; run;",
  "proc means data=followup; var iv_dead iv_fup; by dead; run;",
  "proc print data=followup; var iv_dead iv_fup; if dead=0; run;"
), file.path(adopted_root, "descriptive", "dc.gfup.sas"))
writeLines(c(
  "data trends; set built; year=floor(iv_opyrs)+1985; run;",
  "%let percent = hx_chf;", "%let continuous = lvmassi;",
  "label hx_chf = 'Heart failure';", "label lvmassi = 'LV mass index';",
  "axis1 order=(1985 to 2025 by 5);", "axis2 order=(0 to 100 by 20);",
  "axis3 order=(80 to 140 by 10);", "proc gplot data=trends;",
  "plot hx_chf*year / haxis=axis1 vaxis=axis2;",
  "plot lvmassi*year / haxis=axis1 vaxis=axis3; run; quit;"
), file.path(adopted_root, "graphs", "dp.trends.sas"))
writeLines(c(
  "set built;", "%let pref_time_var=year;",
  "%let variables=age bmi female repair;", "%let exclude=;",
  "%let ncol=2;", "%let nrow=2;", "%let unique_limit=6;",
  "%let show_percent=0;"
), file.path(adopted_root, "descriptive", "dp.postage.sas"))
writeLines(c(
  "proc univariate data=built; var age; run;"
), file.path(adopted_root, "distributions", "ac.age.sas"))
dir.create(file.path(adopted_root, ".git"))
writeLines("ref: refs/heads/main", file.path(adopted_root, ".git", "HEAD"))
writeLines("obsolete copied template", file.path(adopted_root, "tp.shared.sas"))
```

## Clean up the old scaffolding

An adopted study keeps its bare working-directory names and all study-authored
programs, logs, listings, and output. Before adoption, inventory two pieces of
old scaffolding that should not become part of the study contract:

- a local `.git` repository left by the former setup process; and
- obsolete copied template files beginning with `tp` anywhere in the study
  working directories.

Do not remove either with a broad wildcard. Confirm the study root and inspect
the exact matches first. The final call below permanently deletes only those
paths; there is no backup step. Stop before `unlink()` if the inventory contains
anything that should remain.

```{r}
#| label: inventory-adoption-cleanup
normalizePath(adopted_root)
legacy_git <- file.path(adopted_root, ".git")
legacy_tp <- list.files(
  adopted_root, pattern = "^tp", recursive = TRUE, full.names = TRUE,
  include.dirs = TRUE, all.files = TRUE, no.. = TRUE
)
cleanup_targets <- c(if (dir.exists(legacy_git)) legacy_git, legacy_tp)
cleanup_targets

if (length(cleanup_targets)) {
  cleanup_status <- unlink(cleanup_targets, recursive = TRUE)
  stopifnot(cleanup_status == 0L)
}
stopifnot(
  !file.exists(legacy_git),
  !any(file.exists(cleanup_targets)),
  length(list.files(
    adopted_root, pattern = "^tp", recursive = TRUE, include.dirs = TRUE,
    all.files = TRUE, no.. = TRUE
  )) == 0L
)
```

This cleanup concerns obsolete copied scaffolding. It does not remove
study-authored jobs such as `descriptive/dp.postage.sas`. Inventory the
remaining study files and record their checksums before adoption.

```{r}
#| label: before-adoption
legacy_files <- list.files(adopted_root, recursive = TRUE, full.names = TRUE)
legacy_checksums <- tools::md5sum(legacy_files)
list.files(adopted_root, recursive = TRUE)
```

`adopt = TRUE` is additive. It records identity and environment files while
preserving the legacy layout and study-authored files. In the real RStudio
session, use the root you confirmed at the start:

```r
study_setup(
  root = project_root,
  study = "Study title from Study Tracker",
  study_tracker_id = 42L,
  adopt = TRUE
)
study_root()
```

The executable example passes its temporary root explicitly:

```{r}
#| label: adopt-register
study_setup(adopted_root, study = "Synthetic legacy study",
            study_tracker_id = 42L, adopt = TRUE)
register_data(adopted_root, built = "built.csv", event = "dead", time = "iv_dead",
              role = "study", population = "Synthetic full cohort")
study_status(adopted_root)
stopifnot(identical(unname(tools::md5sum(legacy_files)),
                    unname(legacy_checksums)))
```

## Pin the package environment

`study_setup()` reports a missing `renv.lock` until the study initializes its
package environment. If the study was not already pinned, do this after
selecting R 4.6 and restarting RStudio. It is not executed in the temporary
example because `renv::init()` changes the active R project. Snapshot again
after installing the packages used by the study and whenever an accepted
result needs a new reproducible package state.

```r
renv::init()

# Install or update the hvtiR packages used by the study, then record them.
renv::snapshot()
```

## Register and verify the data

Registration turns an existing data file into a checked study input. It records
the file checksum, its role and the expected cohort counts. The synthetic study
has one 40-row study dataset. A real EDA study may not
yet have an event or follow-up-time field; both `event` and `time` may be
`NULL`. Supply them only when they are part of the data contract.

```{r}
#| label: cohort-counts
root <- adopted_root
cfg <- study_config(root)
study_data <- read_built(cfg, dataset = "study")
study_counts <- cohort_counts(study_data, cfg, dataset = "study")
cohorts <- data.frame(
  dataset = "study",
  rows = study_counts$n,
  events = study_counts$n_events
)
cohorts
```

The counts are a checkpoint, not decoration. Compare them with the data build
or protocol before proceeding. Further filtering inside a job needs its own
reference counts and checks.

## Write the EDA analysis set

An analysis set is a declared, reproducible view of the registered study
dataset. Add its definition under `analysis_sets:` in `_study.yml`. Name the
patient identifier, the columns to keep, ordered exclusions, and expected
counts. This synthetic set keeps every row:

```yaml
analysis_sets:
  eda:
    id: ccfid
    vars: [ccfid, dead, iv_dead, iv_fup, year, female, race_grp, repair,
           age, bmi, hx_chf, lvmassi, iv_opyrs]
    exclude: []
    expect:
      n: 40
      n_events: 20
```

Write the set after its declaration. `write_analysis_set()` records its link
to the registered dataset and stops later reads if either the source data or
the declaration changes. The helper below preserves other analysis sets and
refuses to replace an existing `eda` definition; review and rename an existing
set deliberately instead of overwriting it.

```{r}
#| label: write-eda-set
declare_eda <- function(root) {
  cfg_file <- file.path(root, "_study.yml")
  raw <- yaml::read_yaml(cfg_file)
  if (is.null(raw$analysis_sets)) raw$analysis_sets <- list()
  if (!is.list(raw$analysis_sets)) {
    stop("analysis_sets in _study.yml must be a mapping.", call. = FALSE)
  }
  if (!is.null(raw$analysis_sets$eda)) {
    stop("This study already has an eda analysis set; review it before changing it.",
         call. = FALSE)
  }
  raw$analysis_sets$eda <- list(
    id = "ccfid",
    vars = names(built),
    exclude = list(),
    expect = list(n = 40L, n_events = 20L)
  )
  replacement <- tempfile(
    pattern = "_study-", tmpdir = dirname(cfg_file), fileext = ".yml"
  )
  on.exit(unlink(replacement))
  yaml::write_yaml(raw, replacement)
  fs::file_move(replacement, cfg_file)
  hvtiRdatabuild::write_analysis_set("eda", study_config(root))
}
declare_eda(adopted_root)
```

## Add the pre-analysis jobs

Three jobs form the general descriptive and EDA starting point. Add the other
two only when the required study fields exist. `open_job()` is the interactive
entry point: it creates a missing job and opens it, or opens an existing job
without overwriting edits. The executable example uses `add_job()` so
rendering this article does not open editor windows.

| job | purpose | output |
|:--|:--|:--|
| `dc-general` | contents, frequencies, distributions, correlations | HTML report |
| `dc-tables` | formatted descriptive table and optional correlations | HTML and DOCX |
| `dp-postage` | one small EDA panel per selected variable | HTML and PNG pages |
| `dc-gfup` (optional) | follow-up distributions and event checks | HTML report |
| `dp-trends` (optional) | cohort characteristics over calendar or operation time | HTML and PNG |

### Name the EDA job set

`open_job()` needs four pieces of a job name. Two select the template, and two
keep the study's related jobs together:

- `prefix` is the broad job family: `dc` for descriptive computation or `dp`
  for a descriptive plot.
- `qualifier` selects the particular template within that family, such as
  `tables` or `postage`.
- The next two values name the output set. During EDA there may be no clinical
  endpoint or analysis method yet. Use a subject label such as `cohort` and the
  stage label `eda`; they are organizational names, not claims that an outcome
  model has been chosen.

For example, the call for descriptive tables creates
`descriptive/cohort-eda-dc-tables.qmd` in an adopted study. The authored QMD
stays flat in `descriptive/`; generated files are filed beneath the
`cohort-eda/` output directory. Use the same subject and stage for all related
jobs so their filenames and outputs form one coherent set.

From the RStudio Console, create the three general jobs:

```r
general <- open_job("dc", "cohort", "eda", qualifier = "general")
tables <- open_job("dc", "cohort", "eda", qualifier = "tables")
postage <- open_job("dp", "cohort", "eda", qualifier = "postage")
```

If the registered data have a meaningful event indicator and follow-up time,
add `dc-gfup`. If they have a meaningful calendar or operation-time field, add
`dp-trends`:

```r
gfup <- open_job("dc", "cohort", "eda", qualifier = "gfup")
trends <- open_job("dp", "cohort", "eda", qualifier = "trends")
```

In a real study, use `open_job()` for each row, work its `EDIT:` markers from
top to bottom, and render drafts as you go. Use `render_job(job, final = TRUE)`
only after every marker is resolved. The descriptive, follow-up, and postage
jobs can read the `eda` analysis set. The trends job reads a registered dataset
directly. Choose the whole study dataset or a named dataset only when it is the
intended population.

### When registered data change

Do not overwrite a registered dataset and run `update_manifest()` alone.
Registration also records cohort metadata in `_study.yml`, so that shortcut
can leave the two authoritative files disagreeing. Keep the existing extract
unchanged until `hvtiRutilities` provides a coordinated update operation, then
re-write every dependent analysis set and re-render these jobs.

## Migrate one job at a time

Each call selects one template and its source-specific adapter. The source
supplies analysis choices; logs, listings and RTF/DOCX references add evidence.
Migration writes an editable QMD and a report beside it. Existing jobs or
reports stop the call, so keep your edited job when you repeat this workflow.

### General descriptive and EDA checks

The adopted study has no single legacy source that represents the full
`dc-general` template. Scaffold that job directly, then review its editable
choices against the registered data and protocol.

```{r}
#| label: add-adopted-general
general <- add_job("dc", "cohort", "eda", root, "general")
```

### Descriptive tables

The `%desc_tab` calls supply the variable groups and types. The migrated job
will write one editable Word table. Review the SAS presentation choices before
accepting its summary statistics and display.

```{r}
#| label: migrate-tables
tables <- migrate_job(
  file.path(root, "descriptive", "dc.tables.sas"), "cohort", "eda",
  reference = file.path(root, "documents", "general.rtf")
)
```

### Goodness of follow-up

This job reads the registered event and follow-up fields. Confirm the time
units from the data build. Its recorded-interval checks do not establish
completeness against an administrative close date, and patient identifiers
remain disabled.

```{r}
#| label: migrate-gfup
gfup <- migrate_job(file.path(root, "descriptive", "dc.gfup.sas"), "cohort", "eda")
```

### Trends

The explicit year origin and interval expression define whole operation years.
`hv_trends()` owns the summaries and smoothing; compare these figures with
legacy output before treating a method change as equivalent.

```{r}
#| label: migrate-trends
trends <- migrate_job(file.path(root, "graphs", "dp.trends.sas"), "cohort", "eda")
```

### Postage stamps

This source says `set built;`, so the adapter keeps the registered study
dataset. It makes a panel for each chosen variable and arranges the four panels
on one PNG page.

```{r}
#| label: migrate-postage
postage <- migrate_job(file.path(root, "descriptive", "dp.postage.sas"), "cohort", "eda")
jobs <- c(tables = tables, gfup = gfup, trends = trends, postage = postage)
adopted_jobs <- c(general = general, jobs)
reports <- sub("[.]qmd$", "-migration.md", jobs)
```

## Read the reports and resolve the markers

Read all four reports before editing the jobs. They record translated values,
unresolved decisions, ignored source sections, checksums, and evidence line
numbers. This first report includes the synthetic listing, log and RTF.

```{r}
#| label: migration-report
#| results: asis
cat(paste(readLines(reports[["tables"]]), collapse = "\n"))
```

For this synthetic example we accept the migrated groups and the whole cohort
for all four jobs. We choose no group
comparison, median summaries with the 15th and 85th percentiles, no table
abbreviations or correlations, and `dc-tables.docx` as the Word filename.
Follow-up intervals are years and no extra event cross-tabs are needed.
The trend figures use the stated year origin, breaks and variable labels with
no subgroup filter. Postage uses the four stated variables and a 2-by-2 grid.

The following helper acknowledges only exact, reviewed comment lines. It is a
way to repeat these synthetic decisions while rendering this article. In a
study job, edit each marked choice against the source and protocol, then remove
that marker. Replacing every marker without review can turn an incomplete
migration into a misleading finished report.

```{r}
#| label: review-helper
acknowledge <- function(job, reviewed_lines) {
  text <- readLines(job, warn = FALSE)
  stopifnot(all(reviewed_lines %in% text))
  selected <- text %in% reviewed_lines
  text[selected] <- gsub("EDIT:", "REVIEWED:", text[selected], fixed = TRUE)
  writeLines(text, job)
}
```

```{r}
#| label: review-synthetic-choices
#| code-fold: true
acknowledge(tables, c(
  "<!-- EDIT: name the job this replaces (descriptive/dc.tables.*) and say whether it",
  "# EDIT: the built dataset this job reads. \"study\" is the file named by built:",
  "# EDIT: the analysis set this job reads: the columns and excluded rows declared",
  "# EDIT: the variables to report, grouped. Each name is a section heading and",
  "# EDIT: review SAS presentation choices in the migration report.",
  "# EDIT: confirm comparison, summaries, percentiles, abbreviations and Word filename.",
  "# EDIT: force a bucket where the data guess wrong, e.g. an ordinal score you",
  "# EDIT: NULL skips the correlation variant (dc.tables.ods_<topic>.sas). To run"
))
acknowledge(gfup, c(
  "<!-- EDIT: name the job this replaces (descriptive/dc.gfup). -->",
  "# EDIT: the built dataset this job reads. \"study\" is the file named by built:",
  "# EDIT: the analysis set this job reads: declare its columns and excluded rows",
  "# EDIT: review unresolved migration evidence and confirm follow-up units in years.",
  "# EDIT: optional event-coding cross-tabs, each a vector of column names."
))
acknowledge(trends, c(
  "<!-- EDIT: name the job this replaces (an older graphs/dp.trends job) and list",
  "# EDIT: choose a dataset registered in _study.yml.",
  "# EDIT: the calendar year of operation, as a WHOLE year. hv_trends() puts one",
  "# EDIT: one entry per figure. `cols` are the columns drawn together: several",
  "# EDIT: optional. x-axis breaks shared by every figure, e.g. seq(1985, 2025, 5).",
  "# EDIT: optional. Each entry is a filter, given as a function of the data that",
  "# EDIT: review source filters and subgroup intent before accepting the whole cohort."
))
acknowledge(postage, c(
  "<!-- EDIT: name the job this replaces (tp.dp.EDA_barplots_scatterplots*.R). -->",
  "# EDIT: the built dataset this job reads. \"study\" is the file named by built:",
  "# EDIT: the analysis set this job reads: declare its columns and excluded rows",
  "# EDIT: review unresolved postage source choices in the migration report."
))
```

```{r}
#| label: check-reviewed-jobs
remaining <- vapply(jobs, function(job) {
  sum(grepl("EDIT:", readLines(job), fixed = TRUE))
}, integer(1L))
remaining
stopifnot(all(remaining == 0L))
```

## Render and inspect the output

`render_job()` renders each job from its own directory so its guard can read
the authored QMD. A job with a marker left renders as a draft by default;
`final = TRUE` makes an unresolved marker stop the render instead, which is
what an accepted result requires.

```{r}
#| label: render-jobs
for (job in jobs) {
  render_job(job, final = TRUE, quiet = FALSE)
}
```

The authored jobs and migration reports stay flat in their legacy folders.
Generated figures and the Word table go under the `cohort-eda/` output set.
Check both the artifact paths and their contents: a file existing is only the
first check. The Word structural check catches document-format defects; it
does not prove numerical agreement with SAS.

```{r}
#| label: inspect-outputs
outputs <- list.files(root, pattern = "[.](html|docx|png)$", recursive = TRUE)
outputs
word <- file.path(root, "documents", "cohort-eda", "dc-tables.docx")
png <- file.path(root, "graphs", "cohort-eda", "dp-postage-page-01.png")
stopifnot(file.exists(word), file.exists(png))
hvtiRtables::hv_check_docx(word)
# Embed while the PNG exists; study cleanup runs before HTML conversion.
knitr::asis_output(paste0(
  '<img src="', knitr::image_uri(png),
  '" alt="Postage stamp plots for the study cohort">'
))
stopifnot(identical(unname(tools::md5sum(legacy_files)),
                    unname(legacy_checksums)))
```

Open the Word file to compare row order, labels, counts, summaries, precision
and footnotes with the legacy reference. Inspect each trend and postage figure
for variable selection, axes and cohort. Record any methodological differences
alongside the migration report before using the output in a study document.

## A job with no converter

Not every prefix has a migration adapter yet. `migrate_job()` still scaffolds
the template, keeps every `EDIT:` marker for manual review, and its report
says so plainly.

```{r}
#| label: migrate-no-converter
distribution <- migrate_job(file.path(root, "distributions", "ac.age.sas"), "cohort", "eda")
distribution_report <- sub("[.]qmd$", "-migration.md", distribution)
cat(head(readLines(distribution_report), 6L), sep = "\n")
```
