---
title: "The template catalog"
author: "John Ehrlinger"
date: today
vignette: >
  %\VignetteIndexEntry{The template catalog}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE)
library(hvtiRtemplates)
```

## What is here

Every analysis job an HVTI study runs starts from a template: a Quarto file
with the study-specific lines marked `EDIT:`. This page lists them all, what
each one does, and how close each is to being delivered. It is built from the
package's own catalog when the page is rendered, so a template added, renamed
or shipped shows up here without anyone remembering to update the page.

A template becomes a job in your study with `add_job()`. The first argument is
the template's prefix, then the subject and type that name the job's set, and
the qualifier when the prefix has one:

```r
add_job("dp", "cohort", "eda", qualifier = "gfup")
```

That writes `graphs/cohort-eda-dp-gfup.qmd` in the study. The job refuses to
overwrite an existing file, because a job accumulates a study's edits.

## Delivery

```{r}
#| label: delivery
catalog <- template_catalog()
on_disk <- template_list()
key <- function(prefix, qualifier) paste(prefix, ifelse(is.na(qualifier), "", qualifier))
catalog <- catalog[catalog$kind == "job", , drop = FALSE]

light <- ifelse(
  catalog$status == "shipped", "green",
  ifelse(catalog$status %in% c("revisit", "in-flight") |
           (catalog$status == "queued" & is.na(catalog$blocked_on) & !is.na(catalog$batch)),
         "yellow", "red")
)
catalog$light <- c(green = "\U0001F7E2 Shipped", yellow = "\U0001F7E1 In progress",
                   red = "\U0001F534 Not yet on the way")[light]
catalog$template <- ifelse(is.na(catalog$qualifier), catalog$prefix,
                           paste0(catalog$prefix, "-", catalog$qualifier))
catalog$template <- paste0("`", catalog$template, "`")

counts <- table(factor(light, levels = c("green", "yellow", "red")))
```

Each template carries a light.

- 🟢 **Shipped** (`r counts[["green"]]`): on disk and supported. Scaffold it
  and use it.
- 🟡 **In progress** (`r counts[["yellow"]]`): on disk but being reworked, or
  scheduled into a delivery batch with nothing blocking it.
- 🔴 **Not yet on the way** (`r counts[["red"]]`): waiting on a function in
  another package, or not yet scheduled. The job type is known and counted;
  for now, write it by hand.

## Available now

Every template on disk, grouped by the study folder a job lands in.
`add_job()` places each one; you do not choose the folder.

```{r}
#| label: shipped
#| results: asis
folders <- c("datasets", "descriptive", "distributions", "analyses", "graphs",
             "documents", "estimates")
here <- catalog[key(catalog$prefix, catalog$qualifier) %in%
                  key(on_disk$prefix, on_disk$qualifier), , drop = FALSE]
here$scaffold <- ifelse(
  is.na(here$qualifier),
  sprintf("`add_job(\"%s\", subject, type)`", here$prefix),
  sprintf("`add_job(\"%s\", subject, type, qualifier = \"%s\")`", here$prefix, here$qualifier)
)
for (f in intersect(folders, here$folder)) {
  rows <- here[here$folder == f, , drop = FALSE]
  rows <- rows[order(rows$template), , drop = FALSE]
  cat("\n### ", f, "\n\n", sep = "")
  cat(knitr::kable(data.frame(
    Delivery = rows$light, Template = rows$template,
    Description = rows$description, Scaffold = rows$scaffold
  ), row.names = FALSE), sep = "\n")
  cat("\n")
}
```

## Coming

Job types the SAS library and the study corpus carry that have no template
yet. **SAS jobs** counts the distinct studies with a SAS job of that type, so a
larger number is a template more studies are waiting for. *Unmeasured* means
nobody has counted it yet, which is not the same as zero.

```{r}
#| label: coming
coming <- catalog[!key(catalog$prefix, catalog$qualifier) %in%
                    key(on_disk$prefix, on_disk$qualifier), , drop = FALSE]
coming <- coming[order(coming$light != "\U0001F7E1 In progress",
                       -ifelse(is.na(coming$sas_breadth_jobs), -1, coming$sas_breadth_jobs)), ]
what <- ifelse(is.na(coming$description), coming$name, coming$description)
knitr::kable(data.frame(
  Delivery = coming$light, Template = coming$template, Description = what,
  `SAS jobs` = ifelse(is.na(coming$sas_breadth_jobs), "unmeasured", coming$sas_breadth_jobs),
  `Waiting on` = ifelse(is.na(coming$blocked_on), "", coming$blocked_on),
  check.names = FALSE
), row.names = FALSE)
```

## Where to read more

- `template_catalog()` returns this table with every field the catalog keeps,
  including the design note behind each row.
- `inst/templates/README.md` in the package source explains the naming and
  the folder layout.
- The *From SAS descriptive jobs to R* vignette walks through running the
  descriptive templates against a study.
