---
title: "Work a job, from template to final report"
author: "John Ehrlinger"
vignette: >
  %\VignetteIndexEntry{Work a job}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    embed-resources: true
---

Every analysis in a study is a *job*: one Quarto file that reads the study's
data, runs one piece of the analysis, and renders a report. You start a job
from a template, change the lines the template cannot know about your study,
and render it. This guide walks that loop once,
in the order you meet it, and says what to do when the job stops you.

It assumes the study already exists and has its dataset registered. If it does
not, start with [Start a new study](new-study.html), or with
[Adopt an existing study](study-setup.html) for a study that already has a
folder of SAS jobs. Everything below is typed in the R console of the study's
RStudio Project.

```{r}
#| include: false
knitr::opts_chunk$set(eval = FALSE)
```

## Which template do I want?

Start at the [template gallery](https://ehrlinger.github.io/hvtiRtemplates/gallery/).
It lists every template the package ships, says in a sentence what each one is
for, and opens the report each one produces on a synthetic study, so you can see
the tables and figures before you commit to a job. If you knew the SAS job by its
name, `dc.gfup.sas` say, look for the same letters: the R template keeps the
prefix (`dc`) and the job type (`gfup`).

The same list is in R. `template_list()` returns one row per template:

```{r}
library(hvtiRtemplates)
tl <- template_list()
tl[, c("name", "folder", "call")]
```

The `call` column is the line that adds the template to your study. To find the
follow-up jobs, ask for their job type:

```{r}
subset(tl, qualifier == "gfup", c(name, folder, call))
```

Two templates come back. `dc-gfup` is the one to use; `dp-gfup` still works but
is deprecated, and says so with a warning when you add it.

## Add it to the study

Copy the call, and change two words before you run it:

```{r}
add_job("dc-gfup", subject = "cohort", type = "eda")
```

`subject` and `type` name *your* job, not the template. Together they say which
set of work the job belongs to: here, exploratory checks (`eda`) on the whole
cohort (`cohort`). A job modelling death might be `subject = "dead"`,
`type = "hz"`. Use letters, digits and underscores only, because the two words
become part of the filename.

The job lands in the folder the template belongs in, named
`<subject>-<type>-<template>.qmd`, so the line above writes
`10_descriptive/cohort-eda-dc-gfup.qmd` (or `descriptive/...` in a study that
kept the older, unnumbered folder names). `add_job()` never overwrites a job
that already exists, because a job collects your edits. To open a job, creating
it first if it is not there yet, use `open_job()` with the same arguments; it
opens the file in RStudio.

```{r}
job <- open_job("dc-gfup", subject = "cohort", type = "eda")
```

## Find what you have to change

A new job is not finished. Every line the template cannot fill in for your study
carries an `EDIT:` marker: a comment saying what to set and why it matters. The
dataset to read, the event and follow-up columns, the origin year of the
operation-year interval, and so on. The job is finished when no marker is left.

There are three ways to find them.

**The report tells you.** Render the job (next section) and the top of the
report carries a DRAFT banner listing every marker still in the file, with the
first words of each line. A new `dc-gfup` job starts with sixteen:

```
16 unresolved EDIT: marker(s) remain in this job:
  - <!-- EDIT: name the job this replaces (descriptive/dc.gfup). -->
  - # EDIT: the registered dataset this job reads ("study" is the built dataset).
  - # EDIT: an hvtiRdatabuild analysis set, or NULL to read the whole dataset.
  ...
```

**RStudio finds them.** With the job open, press Ctrl+F (Cmd+F on a Mac), type
`EDIT:` and step through the matches with Enter. In most templates the bulk of
them sit together under the heading *Study choices*, near the top, so most of
the work is in one place.

**The console lists them, with line numbers.** This prints the line number and
text of each marker left in the job, which is handy for a long job:

```{r}
lines <- readLines(job)
hits <- grep("EDIT:", lines, fixed = TRUE)
data.frame(line = hits, text = trimws(lines[hits]))
```

Work each marker: set the value, then delete the word `EDIT:` (or the whole
comment, if it no longer says anything you need). Read the comment before you
delete it. Several say why a choice matters, and a few name a mistake that
otherwise fails quietly, such as an origin year that slides every point along
the axis without any other symptom.

## Render as you go

You do not have to finish the markers before you render. Render early and often:

```{r}
render_job(job)
```

That is a *draft* render. It runs the whole job from top to bottom in a clean R
session, writes the HTML report beside the job, and shows the DRAFT banner while
any marker remains. The numbers in a draft are not a result, and the banner says
so.

Run All and Render do different things. **Run All** in RStudio
runs the chunks in your current R session, so an object you made by hand an hour
ago can quietly stand in for one the job no longer makes. **Render** starts from
nothing. Use Run All while you are setting the choices and want to look at the
data. Use `render_job()` to find out whether the job works on its own.

When the markers are gone and the report says what it should, render the final:

```{r}
render_job(job, final = TRUE)
```

A final render is the accepted result. It refuses to finish while any `EDIT:`
marker remains, rather than producing a report that looks done and is not. It
also records the job's *provenance*, which data it read and a fingerprint of the
job file that read it, and writes it beside the report as
`cohort-eda-dc-gfup.provenance.json`, so the numbers can be traced back later.

## Render part of a job

Sometimes a job is half built. The cohort tables are right, but the model below
them waits on a corrected variable, and you want a report of what works. The SAS
jobs did this by commenting the unfinished part out. Here there are two ways to
leave part of a job out, and both keep it visible.

**Leave one chunk out** with the chunk option `skip`, giving the reason in quotes:

````
```{r}
#| label: tbl-model
#| skip: "waiting on the corrected NYHA coding"
...
```
````

The chunk is neither run nor shown. Use it for a chunk nothing below depends on.
If a later chunk needs what the skipped one makes, that later chunk fails; use
the next form instead.

**Stop the render at a point** by calling `stop_here()` in a chunk on its own:

````
```{r}
hvtiRtemplates::stop_here()
```
````

Everything above it renders; everything below is left out. This suits building
a job from the top down, a section at a time.

Both follow the rules the `EDIT:` markers follow:

- A draft lists every skip and every stop, by line, in a PARTIAL box at the top
  of the report, so a partial report cannot pass for a whole one.
- A final render refuses both. Remove them before `render_job(job, final = TRUE)`.
- A skip must say why. `skip: true` with no reason is an error.
- When you run chunks by hand, `stop_here()` does nothing, so the chunks below it
  still run in your session.

## When the job stops you

Most stops are the package refusing to guess. The message says what it needs;
these are the ones you meet first.

**"prefix 'dc' carries 3 templates; name one with `qualifier`, or by its full
name."** Several templates share the `dc` prefix. Use the template's full name,
`add_job("dc-gfup", ...)`, as the `call` column prints it.

**"... already exists; refusing to overwrite."** The job is there already. Open
it with `open_job()` instead of adding it again.

**"... is deprecated in favor of ..."** A warning, not an error: the job is
still written. The message names the template that replaces it; use that one
for new work.

**"N unresolved EDIT: marker(s) remain in this job"**, from
`render_job(final = TRUE)`. The job is not finished. Find the markers as above,
or render a draft to see them listed.

**"This job is rendered in part"**, from a final render. A `skip` or
`stop_here()` is still in the job. Remove it, or render a draft instead.

**"The figure was not drawn"**, in a `dc-gfup` report. The follow-up figure
needs a sensible operation-year column and origin. The box names the problem,
usually `OPYRS` or `ORIGIN_YEAR` under *Study choices*. The follow-up tables
above it do not depend on the figure, so they are still right.

## Where to go next

- The [template gallery](https://ehrlinger.github.io/hvtiRtemplates/gallery/),
  for what each template produces.
- [From SAS descriptive jobs to R](sas-to-r-descriptive.html), for the
  descriptive templates job by job, beside the SAS each one replaces.
- [The template catalog](template-catalog.html), for every template, shipped and
  coming.
- `?add_job`, `?render_job` and `?stop_here` in R, for every argument.
