---
title: "Bootstrap variable selection: logistic"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in:
# rendered as a standalone document the project's format block never applies,
# which is how server-rendered reports end up with sibling _files/ trees instead
# of being self-contained.
#
# NOTE: output-dir is a PROJECT setting and cannot be set here, so the .html
# lands next to its source. With embed-resources that is a single standalone
# file, which is the property that actually matters when a report is sent on.
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the SAS job this replaces, and paste its `%bootreg` call
     verbatim. The call is the specification: it names the candidate pool, the
     entry and stay levels, the replicate count and `proc=`. Prose describing
     it is a paraphrase that drifts; the call does not. -->

Replaces `analyses/<job>.sas`: paste its `%bootreg` call here — the candidate
pool, `sle=`, `sls=`, `resampl=` and `proc=logistic`.

This job reports a bootstrap variable selection screen over a **logistic**
model: how often each candidate survived stepwise selection across resamples,
how much of that figure is Monte-Carlo noise, and which candidates the screen
never saw. **It does not run the screen and it does not fit the final model.**
A companion runner screens; the companion `lm` job fits the multivariable
model.

⚠️ `bl` and `lm` are **parallel analyses, not a pipeline.** `%bootreg` and the
model job were parallel in SAS and these jobs reproduce that. Do not describe
this job's retained set as what `lm` fits, and do not write a handoff file for
it — an earlier study report did both for the hazard pair, and nothing read the
file. A reader who believes the claim will assume the `lm` model was screened
for reliability first. It was not.

A selection frequency is an estimate, not a count of something fixed. Every
table below is built to keep that visible, because the decision this job exists
to support is which side of a retention threshold a variable falls on, and that
is the part resampling noise moves.

```{r}
#| label: setup
# This file runs either from the directory it was scaffolded into or from the
# project root. Resolve which, so that no path in this document needs editing:
# forgetting one of several ../ paths is the instantiation mistake that would
# otherwise be easy to make and quiet when made.
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
# TemporalHazard is deliberately absent: a logistic screen never touches it.
# The runner calls hvtiRbootstrap::boot_select() with fit_logistic() and
# converts the result with boot_bag().
suppressPackageStartupMessages({
  library(hvtiRbootstrap)
  library(hvtiRutilities)
  library(ggplot2)
})

# boot_bag() lands in 0.9.2, and without it a boot_select() screen cannot be
# read by anything below. A study on 0.9.1 otherwise gets "could not find
# function boot_bag" from inside its runner, one file away from this one, and
# the message names the symbol without naming the release that carries it.
#
# The check lives HERE rather than only in the runner because this report is
# what a reader opens. A bag written by an older runner would reach these
# chunks missing the provenance fields, and the tables would print blanks where
# the screen's criteria belong.
#
# THE FLOOR IS 0.9.3, NOT 0.9.2, AND THE EXTRA DISTANCE IS THE POINT. Below
# 0.9.3 boot_select() recorded `sle` and `sls` on the screen and then selected
# on AIC regardless, so the provenance table this report prints names an entry
# and a stay criterion the screen never applied. A reader takes those two rows
# as the rule that produced the frequencies beneath them. A blank would be
# obvious; a plausible number that describes nothing is not. bh.qmd stays at
# 0.9.0 on purpose -- its screen comes from TemporalHazard, whose own stepwise
# always honoured them.
if (utils::packageVersion("hvtiRbootstrap") < "0.9.3") {
  stop("This report needs hvtiRbootstrap >= 0.9.3; ",
       utils::packageVersion("hvtiRbootstrap"), " is installed. boot_bag() ",
       "converts a boot_select() screen into the bag this report reads, and ",
       "nothing below 0.9.2 has such a function; below 0.9.3 the screen ",
       "recorded `sle` and `sls` but selected on AIC, so the provenance ",
       "table below would name criteria it never used.\nUpdate it, then ",
       "re-render.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
# The markers in this file name work a study author still has to do, and
# README.md says a job that still contains one has not been finished. This
# chunk is what makes that TRUE rather than merely stated.
#
# Without it an unedited job renders green over a meaningless analysis. In THIS
# file the danger is that EXPECT_BOOT still holds the template's number, so the
# completeness check compares the run against a denominator nobody chose -- and
# passes, reporting frequencies over the wrong base.
#
# knitr::current_input() is NULL outside a render -- a study author stepping
# through chunks in RStudio -- and there is no file to scan then, so this is a
# no-op in that case. Same handling as the `set` guard below, for the same
# reason.
#
# The token is BUILT, not written literally, and that is not stylistic. Quarto
# knits through an intermediate and current_input() returns THAT file, so the
# scan reads a copy of this chunk along with everything else: a literal
# grep("<token>", .src) here matches its own source line, and the guard then
# fires on every render, finished or not. Measured, not theorised -- a probe
# found three markers in a file containing two. A guard that cries wolf on a
# finished job is a guard that gets deleted, which returns us to issue #27.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    # Report the marker TEXT. Line numbers here are the intermediate's, not
    # this .qmd's, so they need not match what the author sees in an editor.
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it.\nTo render a partly-worked draft anyway, set ",
      "HVTI_TEMPLATE_DRAFT to 1, true or yes -- the report then carries a ",
      "draft banner. Any other value, including 0, leaves this guard strict."
    )
    # Only the documented values enable draft mode. nzchar() alone would treat
    # HVTI_TEMPLATE_DRAFT=0 as ON, silently downgrading this stop() to a
    # warning for someone whose shell profile sets it to 0 meaning OFF -- a
    # switch whose enablement does not mean what it says, which is the same
    # defect this guard exists to catch. An unrecognised value falls through to
    # strict on purpose: the author then sees the stop and re-reads the docs,
    # which is the safe direction to be wrong in.
    if (tolower(Sys.getenv("HVTI_TEMPLATE_DRAFT")) %in% c("1", "true", "yes")) {
      # The banner is not optional. A draft render that looks like a finished
      # one is the same defect with an extra step: the warning scrolls past in
      # a log, while the .html is the artifact that gets sent to someone.
      warning(.msg, call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Rendered with `HVTI_TEMPLATE_DRAFT` set. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by new_job() from the endpoint/type it
     was called with, which are also the filename's first two fields. They
     name the SET this job belongs to, and every artifact path in the
     document is computed from them, so no output path below needs editing.

     Do NOT hand-edit these declarations: the render-time guard below checks
     them against this file's own name and stops if they disagree. To put
     this job in a different set, call new_job() again with the new
     endpoint/type rather than editing this copy in place.

     A set is keyed on (endpoint, analysis type), not on the endpoint alone.
     One endpoint is analysed by several methods -- parametric hazard, random
     forest survival, competing risks -- and those chains share their upstream.
     Two sets keyed on `dead_pa` alone would write their life tables to the
     same file. -->

```{r}
#| label: set
ENDPOINT <- "dead_pa"
TYPE     <- "hz"

# `new_job()` writes ENDPOINT/TYPE from the same values it put in this file's
# name, but a hand-edited declaration can drift from it afterward. set_path()
# below resolves from the declarations, not the filename, so a drifted
# declaration would silently write into ANOTHER set's artifact directory --
# exactly the collision the (endpoint, type) key exists to prevent, re-entered
# through the body instead of the name.
#
# knitr::current_input() is NULL outside a render (a study author running
# chunks interactively in RStudio), and there is no filename to check against
# yet, so this is a no-op in that case rather than a spurious error.
.current <- knitr::current_input()
if (!is.null(.current)) {
  # Quarto knits through an intermediate, so `knitr::current_input()` returns
  # `<endpoint>-<type>-<ordinal>-<prefix>.rmarkdown` here rather than the
  # `.qmd` this was scaffolded as. Strip whatever extension is actually present
  # rather than hard-coding one, so this doesn't depend on a build-tool detail
  # staying the same.
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint, "', type '",
         .name_type, "'), but declares ENDPOINT = \"", ENDPOINT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

# Resolve a path inside this set's artifact directory. `kind` is the artifact
# folder -- "estimates" for serialized results, "graphs" for figures. The set
# directory sits one layer under the kind and never two, which is the whole
# layout rule.
#
# Created on first use rather than up front, so a job that writes nothing leaves
# no empty directories behind.
set_path <- function(kind, file) {
  d <- file.path(.root, kind, paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}

# The screen this job reports on was run by a companion script, which wrote its
# chunks into this set's `estimates` directory. Reading them through set_path()
# is what keeps a screen filed against the set whose candidates it screened.
```

## The screen

```{r}
#| label: expect
# EDIT: what this run was LAUNCHED as, not what happens to be on disk.
#
# Nothing in this file knows how many replicates the runner was asked for, so
# this number is the only thing that can tell a partial screen from a complete
# one. Without it a render against a run that stopped early produces a report
# that is wrong in no visible way: every health check passes, every frequency
# is honestly computed, and only the denominator is not the intended one.
EXPECT_BOOT <- 1000L   # EDIT: %bootreg(resampl=) -- valid models wanted

# EDIT: the file your runner wrote its screen to, under this set's estimates
# directory. It must match the runner exactly.
BOOT_FILE <- "bagging.rds"
```

```{r}
#| label: load
# The screen is not run here. It is read.
#
# boot_select() writes nothing until it returns, so a run that dies partway
# leaves nothing at all. Keeping the run in a companion file means a failed
# render costs a re-render rather than a re-run, and it is what lets this
# report be re-read months later against the screen it actually described.
.single_file <- set_path("estimates", BOOT_FILE)
if (!file.exists(.single_file)) {
  stop("No screen found at ", .single_file, ". This report reads a bag your ",
       "runner wrote with boot_bag(); it does not run the screen.\nRun the ",
       "companion runner first, or correct BOOT_FILE.", call. = FALSE)
}
bag <- readRDS(.single_file)

# Bound empty so the `save` chunk's overwrite guard has both names to compare
# against. A logistic screen is not chunked by default; a chunked one is pooled
# in the RUNNER, not here.
.chunk_files <- character(0)

# A run split across chunks is still supported, and belongs upstream: pool with
# hvtiRbootstrap::boot_pool_chunks() in your runner and write the pooled bag to
# BOOT_FILE. Only the runner knows how many chunks it launched, and
# boot_pool_chunks() refuses chunks that disagree on the dataset checksum, the
# entry and stay levels, the base parameters or the engine version -- every one
# of which is silent on its own, showing up as a slightly different frequency
# rather than as an error.
```

```{r}
#| label: completeness
#| results: asis
# A callout, not a table cell. A provisional report that cannot say so is the
# failure being prevented, and it must survive someone receiving the .html
# without knowing when it was made -- which a number buried in a provenance
# table does not.
.shortfall <- boot_shortfall(bag, expect_chunks = 1L,
                             expect_boot = EXPECT_BOOT)
if (!is.null(.shortfall)) {
  cat("\n::: {.callout-warning title=\"Provisional -- this pool is incomplete\"}\n")
  cat(.shortfall, "\n", sep = "")
  cat(":::\n\n")
}
```

## Provenance

```{r}
#| label: contract
# The runner is NOT templated -- it is a study file -- so the fields this report
# reads are a contract between two files that no shared function enforces.
# Stated here, and checked, because the alternative is a report that indexes a
# field the runner stopped writing and renders `NULL` into a provenance table as
# though it were an answer.
#
# A pooled run has already had most of these checked by boot_pool_chunks(),
# which refuses chunks that disagree. A SINGLE unchunked run has not: nothing
# ran between the runner and here. This is the only check that covers both.

# Shapes, not merely names. Checking presence is what let a length-2 `requested`
# reach a table that cannot recycle it against a length-13 item column -- the
# render-blocker that shipped in three releases and was fixed in 1.0.17. Every
# failure is reported at once, because an author fixing a runner wants the whole
# list rather than one field per re-render.
boot_validate(bag)
```

```{r}
#| label: provenance
# CPU cost, NOT wall clock. boot_pool_chunks() SUMS elapsed across chunks, so on
# a chunked run this is total compute; chunks run in parallel finish in a
# fraction of it. Reported in hours because the minutes figure reads as a
# wall-clock number and is off by the chunk count.

# The digest travels WITH its algorithm. An md5 and a sha256 of the same file
# are different strings, and of different files may not be, so a bare digest
# recorded without its algorithm is not evidence of anything.

# A version string alone cannot say which codebase ran: one real package version
# existed as two, one with a selection criterion and one without, and the
# selection criterion is precisely what decides what a screen selects.

# `requested` and `usable` are scalars on a single-phase screen: the pool is
# offered once. boot_provenance() also accepts the per-phase vectors a
# multiphase hazard bag carries, and collapses them to one labelled string per
# row rather than summing -- a pool offered to two phases is one pool seen
# twice, not twice as many candidates. Nothing here needs that, and nothing
# here breaks if a future runner supplies it.
provenance <- boot_provenance(bag)
provenance
```

```{r}
#| label: seeds
# Every seed, so that a rerun is reproducible and a duplicate is visible. Two
# chunks sharing a seed contain literally the SAME replicates: pooling them
# counts each twice and reports a Monte-Carlo error smaller than the run
# actually has. boot_pool_chunks() refuses that outright, so a pooled run cannot
# reach here with a duplicate -- this table is what lets you check a single run,
# and what lets a reader reproduce either.
#
# `seeds` is created by boot_pool_chunks(); a SINGLE unchunked run never went
# through it and carries only the scalar `seed` its runner wrote. Falling back
# is not defensive padding -- without it the one case this table exists to
# serve is the one case it cannot render.
boot_seeds(bag)
```

## Candidates that were never screened

<!-- Counts first, detail second, and that ordering is not cosmetic. The detail
     table runs to dozens of rows on a real pool, and a reader who has to scroll
     it to discover that a whole CLASS of variable was never screened will not
     discover it. -->

```{r}
#| label: dropped-summary
# A candidate the screen never saw cannot appear at any frequency, so its
# absence from the table below is indistinguishable from never having been
# selected. That is the reason this section exists at all.
dropped <- boot_dropped(bag)
if (!nrow(dropped)) {
  cat("Every candidate offered was screened.\n")
} else {
  # No phase column: boot_dropped() returns whatever the runner wrote, and a
  # single-phase runner writes none. Tabulating an absent column errors.
  as.data.frame(table(reason = dropped$reason))
}
```

```{r}
#| label: dropped-detail
if (nrow(dropped)) {
  dropped
}
```

## Did the screen actually run?

⚠️ **A screen that selected nothing is a failure, not a finding.** Nothing
about it looks like an error: `boot_select()` keeps drawing until it has the
replicates it was asked for, so `n_success` is the full count and `n_failed` is
whatever the fitter genuinely refused. The frequency table is then empty, or
holds only the base model, and reads as a screen that found no risk factors
rather than as a screen that never selected.

```{r}
#| label: health
# A formula held in a variable is FINE here, unlike in the hazard sibling.
# boot_select()'s fitters evaluate through .fit_in_env(), which binds `formula`
# and `data` into the environment step() updates in, so `f <- y ~ a + b` then
# boot_select(d, f, ...) gives byte-identical results to writing the formula at
# the call site. Verified against 0.9.1 rather than assumed; bh warns the
# opposite because hzr_bootstrap() rewrites the stored formula per replicate
# and a symbol does not survive that. Do not carry bh's warning across.
#
# `reps` is bound here, and not only for this chunk: `cluster-matrix` below pivots
# these same replicates into the wide matrix boot_clusters() wants. Dropping the
# binding when this body moved into the package would fail 250 lines away,
# naming `reps` and nothing about the cause.
reps <- bag$boot$replicates

health <- boot_health(bag)
health

# The refusals below select their check by NAME, so a rename upstream would
# make both `%in%` tests FALSE and silently delete the guards rather than
# break them. Checked here so that drift stops the render instead of quietly
# passing a failed screen.
.want <- c("Distinct candidates ever selected", "SD of the first free base parameter")
if (!all(.want %in% health$check)) {
  stop("boot_health() no longer reports the check(s): ",
       paste(setdiff(.want, health$check), collapse = ", "),
       ".\nThe refusals below match on that name, so they would silently stop ",
       "firing. Update them to the new name in hvtiRbootstrap.", call. = FALSE)
}

# boot_health() REPORTS; it does not refuse. Both rows below are failures whose
# reports read as healthy -- a screen that selected nothing has n_failed = 0,
# and a bootstrap that refit nothing has n_success = 500 -- so a table cell is
# not enough. The refusal stays in the template, where the render stops.
.failed <- health$check[!is.na(health$ok) & !health$ok]

if (.want[[1L]] %in% .failed) {
  stop("The screen selected NOTHING: no parameter outside the base model ",
       "appears in any replicate. That is a failed screen, not a null result. ",
       "n_failed is ", bag$boot$n_failed, ", so the fits themselves ran.\n",
       "The usual causes, in the order worth checking: the runner passed ",
       "select = \"none\" and every term was forced rather than screened, so ",
       "no term was ever a candidate; sle is far too strict for the pool; or ",
       "base_params in the boot_bag() call names the whole model, which ",
       "subtracts every term from the frequencies.", call. = FALSE)
}

if (.want[[2L]] %in% .failed) {
  stop("The first free base parameter has SD exactly 0 across ", bag$n_boot,
       " replicates, so every replicate returned the SAME fit. A bootstrap ",
       "that resamples nothing does that, and it reports n_success = ",
       bag$boot$n_success, " with no warning.\nCheck that the runner passed ",
       "the cohort rather than a single row, and that boot_bag() was given the ",
       "screen it thinks it was.\nThe frequencies below would all be 100% and ",
       "mean nothing.", call. = FALSE)
}
```

## What a selection frequency is, and what it is not

A selection frequency is an estimate, not a count of something fixed, and it
carries Monte-Carlo error of roughly `sqrt(p(1-p)/n_boot)`. At `p = 0.5` over
500 replicates that is about **2.2 percentage points**, so a variable sitting
within a few points of the retention threshold can fall on either side of it on
resampling noise alone.

- **Agreement on the retention decision matters more than agreement on the
  frequency.** Which side of the threshold a variable falls on is the decision
  this job exists to support.
- **A near-threshold variable is not a weak risk factor.** It is a variable
  whose selection is unstable, which is a different claim.

```{r}
#| label: criteria
# EDIT: from YOUR .sas %bootreg call. Its documented defaults are
# resampl=1000, sle=0.10, sls=0.05, but real jobs override them -- and the
# entry and stay levels decide what each replicate selected, so a report
# quoting the defaults when the run used something else is describing a
# different screen. Read them off the call and paste it above.
#
# RETAIN_PCT is the reliability cutoff this report calls "retained". It is a
# reporting decision, not something the run recorded: the entry and stay levels
# above governed each replicate's stepwise fit, and neither of them says how
# often a variable must survive to be worth carrying forward.
RETAIN_PCT <- 50

# A logistic screen has no phases: a candidate is offered once, so one term is
# one screening decision and there is nothing to split a term name on.
#
# boot_frequencies() and boot_concepts() take phase = NULL for exactly this
# case, and it is the same code path bh drives with a term-splitting rule. One
# implementation serves both, which is the whole reason this file is thin.
#
# Do NOT supply a rule here to "be safe". A rule that matches nothing still
# adds a phase column of empty strings, and every grouped table below would
# then group by a column that says nothing -- silently, since no count changes
# and no error is raised.
PHASE_OF <- NULL
```

```{r}
#| label: frequencies
# mc_error is the per-variable Monte-Carlo error described above; near_threshold
# flags a variable within two Monte-Carlo errors of RETAIN_PCT, whose retention
# would not survive a rerun with different seeds.
freq <- boot_frequencies(bag, phase = PHASE_OF, threshold = RETAIN_PCT)
freq[, c("variable", "n", "pct", "mc_error", "near_threshold")]
```

```{r}
#| label: retained
# The retained set. Reported WITH the near-threshold flag rather than
# as a bare list: a variable that cleared the cutoff by less than its own
# Monte-Carlo error cleared it on noise, and a list that does not say so invites
# the reader to treat the two kinds of retention as the same claim.
retained <- freq[freq$retained, , drop = FALSE]
retained[, c("variable", "n", "pct", "mc_error", "near_threshold")]
```

## Concepts

⚠️ **Do not prune competing transformations from the pool before screening.**
Screen every form; group only when reading, which is what this section does.

Measured, on the study this template came from: of the 57 forms pruning removed,
**16 correlated at |r| < 0.9** with the form kept and five below 0.5. `in_zexp`
**is** `1/zexp` — r = 0.9997 against the reciprocal — yet correlates with
`zexp` at only **-0.195**, because `zexp` spans 0.038 to 151.9. Over a 4000-fold
range a value and its reciprocal are different information. That study's
published model uses `zexp` and `in_zexp` **together and both significant**, a
two-parameter flexible form pruning forbids. The measurement is from a hazard
study, and it is quoted here because the argument is about variable NAMES and
correlation, which a logistic screen shares.

The naming convention tells you two variables are RELATED. Only the data tells
you whether they are REDUNDANT.

```{r}
#| label: concept-map
# EDIT: the affix vocabulary, if your study's variable names do not follow
# vars.sas conventions. POOL_AFFIXES carries ln_, in_, in2, _pr and a trailing
# 2, the order in which they strip, and the deliberate refusal to reduce agee
# to age. Those are facts about this institution's names, not about statistics,
# which is why they are data you can replace rather than logic you cannot.
#
# The pool grouped here is the set of forms that reached at least one replicate.
# A candidate the screen never saw is in the unscreened table above, not here.
concept <- concept_map(unique(freq$variable), affixes = POOL_AFFIXES,
                       min_stem = POOL_MIN_STEM,
                       plain_suffix = POOL_PLAIN_SUFFIX)
concept
```

```{r}
#| label: concept-frequencies
# Grouped for READING. Every form keeps its own row and its own frequency --
# nothing is collapsed, summed or dropped -- because the whole argument above is
# that two forms of one concept may carry different information. The concept
# column is a reading aid, not an aggregation.
by_concept <- merge(freq, concept[, c("variable", "concept", "representative")],
                    by = "variable", all.x = TRUE)
by_concept <- by_concept[order(by_concept$concept, -by_concept$pct), ,
                         drop = FALSE]
rownames(by_concept) <- NULL
by_concept[, c("concept", "variable", "representative", "pct", "mc_error")]
```

A per-form row answers "how often was *this form* selected". It cannot answer
"how often was *this concept* selected", and that is the number a paper quotes.
The gap runs both ways: competing forms split replicates between them, so the
concept reads weaker than any single figure suggests, or two forms both clear
the cutoff and one finding is reported twice.

```{r}
#| label: concept-union
# Nothing above is collapsed -- every form keeps its row and its frequency.
# This is an ADDITIONAL row per concept, not a replacement for them.
#
# Computed from the replicate table, never from the summary. A union across a
# concept's forms cannot be recovered from marginal percentages: two forms at
# 30% each are anywhere between 30% and 60% of replicates, depending entirely
# on how often the same replicate took both, and the summary does not record
# that.
#
# How much the per-form table understates the concept. A large spread is the
# whole reason this table exists; a spread of zero means the concept has one
# form and the two views agree.
#
# This table's at-least-one figure is `pct_any`, not `union_pct`; every value
# is unchanged.
concept_union <- boot_concepts(bag, concept_map = concept[, c("variable", "concept")],
                               phase = PHASE_OF, threshold = RETAIN_PCT)
concept_union[, c("concept", "n_forms", "pct_any", "best_form_pct",
                  "spread", "retained")]
```

⚠️ **A selection frequency is conditional on the candidate pool**, and no column
here removes that. On the study this template came from, re-running with 226
candidates instead of 189 moved one variable from 26.6% to 95.2% and another
from 93.8% to 20.2% — same data, same seeds. Quote a reliability figure with
the pool it came from.

```{r}
#| label: concept-counts
# Crowding: a model that spent several of its slots on forms of ONE concept is
# budget-limited by redundancy, and that is invisible in a coefficient table.
# Grouping is concept_map()'s, so this is a FLOOR -- a concept nobody's affix
# vocabulary reaches is still counted as separate variables.
crowding <- selection_crowding(retained$term, affixes = POOL_AFFIXES,
                               min_stem = POOL_MIN_STEM,
                               plain_suffix = POOL_PLAIN_SUFFIX)
if (!nrow(crowding)) {
  cat("No concept had more than one form retained.\n")
} else {
  crowding
}
```

## Correlation clusters

Concept grouping above is by NAME, and a naming convention only reaches what
its affix rules can reach. Two further groupings matter, and neither is
derivable from a variable's name: one by DECLARATION, which is clinical
judgement, and one by the DATA. The SAS job this replaces ran `%cluster` once
per named cluster and covered both.

The three views answer different questions. A name tells you two variables are
related. A declaration tells you a clinician treats them as one thing —
`Renal` is creatinine, GFR and dialysis, which share no stem and never will. A
correlation tells you the pool cannot tell them apart.

```{r}
#| label: cluster-matrix
# boot_clusters() wants one ROW per replicate and one COLUMN per term, with an
# unselected term left NA -- that is how it counts "at least one member". The
# pooled replicates are long, so pivot rather than pass them straight in.
#
# Replicate ids are 1..n_boot across the whole pool: boot_pool_chunks() offsets
# each chunk's ids before stacking, and refuses to pool at all if any chunk's
# ids fall outside its own range. The direct indexing below relies on that.
terms <- sort(unique(reps$parameter))
coefs <- matrix(NA_real_, nrow = bag$n_boot, ncol = length(terms),
                dimnames = list(NULL, terms))
coefs[cbind(reps$replicate, match(reps$parameter, terms))] <- reps$estimate
```

```{r}
#| label: clusters
# EDIT: your clusters, one per concept you want reported. The SAS jobs this
# replaces named around eight -- Age, Size, BMI, race, GFR, Renal and the like.
# There is no sensible default: which concepts are worth clustering is a
# statement about your candidate pool, not about logistic regression.
#
# Members are BARE variable names, not phase-qualified: this screen offers each
# candidate once, so a term is one screening decision. Naming a term that no
# replicate carries is refused rather than reported as 0%: a typo that grouped
# nothing would otherwise read as a concept nobody selected.
CLUSTERS <- list(
  Age = c("age", "ln_age")
)

clusters <- boot_clusters(coefs, CLUSTERS)
clusters
```

The table above groups by DECLARATION, not by data: `boot_clusters()` counts
replicates retaining any member of a list you wrote, and computes no
correlation at all. The correlation is below, and the two are worth keeping
apart. A declaration says a clinician thinks these are one thing; a correlation
says the pool cannot tell them apart.

```{r}
#| label: collinear
# EDIT: the correlation above which two candidates count as indistinguishable.
# 0.99 finds near-duplicates -- a variable and its rescaling, a dummy and its
# complement. Lower it to see the pool's structure rather than only its
# accidents.
COLLINEAR_R <- 0.99

# Read to CORRELATE the pool, not to re-run anything. The screen already
# happened; this describes the data it ran against.
d <- read_built()

# A build that moved since the screen ran would leave these correlations
# describing one dataset and every frequency above describing another, with
# nothing in the report to say so. A warning rather than a stop: the
# frequencies above remain valid, and it is only this table that goes stale.
if (!is.null(bag$n_rows) && !identical(as.integer(nrow(d)), as.integer(bag$n_rows))) {
  warning("The built dataset now has ", nrow(d), " rows; this screen ran against ",
          bag$n_rows, ". The correlations below describe TODAY'S data, not the data ",
          "the frequencies above were computed on.", call. = FALSE)
}

collinear <- pool_collinear_pairs(d, pool = unique(freq$variable), threshold = COLLINEAR_R)
if (!nrow(collinear)) {
  cat("No pair of screened candidates correlates at |r| >= ", COLLINEAR_R, ".\n", sep = "")
} else {
  collinear
}
```

⚠️ **A collinear pair is not a licence to prune.** It says two candidates are
indistinguishable *in this cohort at this threshold* — a statement about the
pool, not about the concepts. The no-pruning argument above stands: the same
measurement that found 57 prunable forms found 16 of them correlating below 0.9.

`pool_collinear_pairs()` sees only NUMERIC, non-constant columns, so
factor-coded binaries are invisible to it for the same reason they were
invisible to the screen. Check the coercion table above before reading an empty
result as an absence of collinearity.

## Figure

```{r}
#| label: fig-frequencies
#| fig-height: 7
#| fig-cap: "Selection frequency, with Monte-Carlo error and the retention threshold."
# The error bar is the point of the figure, not decoration. A bare dot plot
# invites the reader to rank variables by a difference of two points, which is
# inside the noise at every frequency near the middle of the range.
ggplot(freq, aes(x = stats::reorder(variable, pct), y = pct)) +
  geom_hline(yintercept = RETAIN_PCT, linetype = "dashed") +
  geom_errorbar(aes(ymin = pct - mc_error, ymax = pct + mc_error), width = 0) +
  geom_point() +
  coord_flip() +
  labs(x = NULL, y = "Replicates selecting the variable (%)") +
  theme_minimal()
```

## Save

```{r}
#| label: save
# Written through set_path(), so the screen is filed against the SET whose
# candidates it screened rather than against the endpoint alone.
#
# There is deliberately NO handoff file here -- no selection_bl.csv, no
# retained set written for `lm` to read. `bl` and `lm` are parallel analyses,
# and the study report that wrote such a file for the hazard pair also claimed
# the model fits what the screen retained. Nothing read the file, and the claim
# was not true. A file whose existence implies a dependency that does not exist
# is worse than no file.

# NOT BOOT_FILE. The run's INPUT is that name, so writing the report there
# overwrites the screen it just read. BOOT_FILE is an edit point, so a study
# can set it to anything, including this name -- checked rather than assumed,
# because the failure destroys the run instead of erroring, and a screen is
# hours of compute that boot_select() produces only on return.
report_file <- set_path("estimates", "bl-report.rds")
inputs <- normalizePath(c(.single_file, .chunk_files), mustWork = FALSE)
if (normalizePath(report_file, mustWork = FALSE) %in% inputs) {
  stop("This report would be written to ", basename(report_file), ", which is ",
       "the bootstrap output it just read. That file is the screen itself, and ",
       "overwriting it cannot be undone.\nChange BOOT_FILE so the runner's ",
       "output and this report do not collide.", call. = FALSE)
}

saveRDS(list(freq = freq, retained = retained, concept = concept,
             crowding = crowding, clusters = clusters,
             provenance = provenance, retain_pct = RETAIN_PCT,
             expect_boot = EXPECT_BOOT),
        report_file)
```
