---
title: "Boosted multivariate trees: boostmtree"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in:
# rendered as a standalone document the project's format block never applies,
# which is how server-rendered reports end up with sibling _files/ trees instead
# of being self-contained.
#
# NOTE: output-dir is a PROJECT setting and cannot be set here, so the .html
# lands next to its source. With embed-resources that is a single standalone
# file, which is the property that actually matters when a report is sent on.
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    page-layout: full
    grid:
      body-width: 2000px
      sidebar-width: 250px
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (an `nb.<variable>.Boostmtree` job), and
     say which response, visit time and predictors it used. -->

Replaces `analyses/<job>`: describe its response, visit time and predictors here.

An `nb-boostmtree` job fits boosted multivariate trees to a response measured
repeatedly over follow-up, one row per visit, and reports how well it fits and
what drives it: the cross-validated error path, observed against fitted values
over time, variable importance, partial effects over time, and a sample of
patients' fitted trajectories.

```{r}
#| label: setup
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRtemplates:::.find_study_root(if (is.null(.in)) getwd() else dirname(.in))
.provenance_data <- list()
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
# boostmtree must be the CCF fork, 2.0.2 or later. Every study fits with
# cv.flag = TRUE, and in boostmtree 2.0.0 (the CRAN release) that flag leaves the
# in-sample fit frozen, so the stored coefficients diverge with M and every
# non-cross-validated result is wrong: partial and marginal effects, importance,
# and predict(use.cv.flag = FALSE). Reported upstream as kogalur/boostmtree#2;
# the fix is kogalur/boostmtree#3. The version floor separates the fork from
# CRAN today. Should upstream release a 2.0.2 without the fix, this check would
# pass it; read the installed package's NEWS if in doubt.
if (!requireNamespace("boostmtree", quietly = TRUE) || utils::packageVersion("boostmtree") < "2.0.2") {
  stop("This job needs boostmtree >= 2.0.2 from the CCF fork. Install it with\n",
       "  remotes::install_github(\"ehrlinger/boostmtree_src\", subdir = \"boostmtree\", ref = \"v2.0.3-ccf\")\n",
       "then re-render.", call. = FALSE)
}
if (!requireNamespace("ggBoostedTrees", quietly = TRUE) || utils::packageVersion("ggBoostedTrees") < "0.9.0") {
  stop("This job needs ggBoostedTrees >= 0.9.0. Install it with\n",
       "  remotes::install_github(\"ehrlinger/ggBoostedTrees\")\nthen re-render.", call. = FALSE)
}
suppressPackageStartupMessages({
  library(boostmtree)
  library(ggBoostedTrees)
  library(hvtiRutilities)
})
```

```{r}
#| label: guard-edits
#| results: asis
# The markers in this file name work a study author still has to do, and
# README.md says a job that still contains one has not been finished. This
# chunk is what makes that TRUE rather than merely stated.
#
# Without it an unedited job renders green over a meaningless analysis. Here
# that is a boosted fit on the template's placeholder response, predictors and
# iteration count, whose error path and effect plots look exactly like a result.
#
# knitr::current_input() is NULL outside a render -- a study author stepping
# through chunks in RStudio -- and there is no file to scan then, so this is a
# no-op in that case. Same handling as the `set` guard below, for the same
# reason.
#
# The token is BUILT, not written literally, and that is not stylistic. Quarto
# knits through an intermediate and current_input() returns THAT file, so the
# scan reads a copy of this chunk along with everything else: a literal
# grep("<token>", .src) here matches its own source line, and the guard then
# fires on every render, finished or not. Measured, not theorised -- a probe
# found three markers in a file containing two. A guard that cries wolf on a
# finished job is a guard that gets deleted, which returns us to issue #27.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    # Report the marker TEXT. Line numbers here are the intermediate's, not
    # this .qmd's, so they need not match what the author sees in an editor.
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    # A job renders as a draft by default, so an author can iterate on a
    # working report and remove markers as they go. HVTI_TEMPLATE_STRICT
    # turns the draft into a stop for a final render. Only unset, 0, false
    # and no mean "not strict": an unrecognized value stops on purpose, so a
    # mistyped switch is seen rather than quietly ignored, which is the safe
    # direction to be wrong in.
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      # The banner is not optional. A draft render that looks like a finished
      # one is the same defect with an extra step: the warning scrolls past in
      # a log, while the .html is the artifact that gets sent to someone.
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by add_job() from the subject/type it
     was called with, which are also the filename's first two fields. They
     name the SET this job belongs to, and every artifact path in the
     document is computed from them, so no output path below needs editing.

     Do NOT hand-edit these declarations: the render-time guard below checks
     them against this file's own name and stops if they disagree. To put
     this job in a different set, call add_job() again with the new
     subject/type rather than editing this copy in place.

     A set is keyed on (subject, analysis type), not on the subject alone.
     One subject may be analyzed by several methods -- parametric hazard, random
     forest survival, competing risks -- and those chains share their upstream.
     Two sets keyed on `dead_pa` alone would write their life tables to the
     same file. -->

```{r}
#| label: set
SUBJECT <- "lvef"
TYPE    <- "boost"

# `add_job()` writes SUBJECT/TYPE from the same values it put in this file's
# name, but a hand-edited declaration can drift from it afterward. set_path()
# below resolves from the declarations, not the filename, so a drifted
# declaration would silently write into ANOTHER set's artifact directory --
# exactly the collision the (subject, type) key exists to prevent, re-entered
# through the body instead of the name.
#
# knitr::current_input() is NULL outside a render (a study author running
# chunks interactively in RStudio), and there is no filename to check against
# yet, so this is a no-op in that case rather than a spurious error.
.current <- knitr::current_input()
if (!is.null(.current)) {
  # Quarto knits through an intermediate, so `knitr::current_input()` returns
  # `<subject>-<type>-<prefix>.rmarkdown` here rather than the
  # `.qmd` this was scaffolded as. Strip whatever extension is actually present
  # rather than hard-coding one, so this doesn't depend on a build-tool detail
  # staying the same.
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject, "', type '",
         .name_type, "'), but declares SUBJECT = \"", SUBJECT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

# Resolve a path inside this set's artifact directory. `kind` is the artifact
# folder -- "estimates" for serialized results, "graphs" for figures. The set
# directory sits one layer under the kind and never two, which is the whole
# layout rule.
#
# Created on first use rather than up front, so a job that writes nothing leaves
# no empty directories behind.
set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}

# The fit is cached and saved through set_path(), by set, so a later render or job finds this set's fit and no other's.
```

## Study choices

Edit these values for this study before rendering.

```{r}
#| label: edit-study-choices
# EDIT: the registered dataset this job reads ("study" is the built dataset).
DATASET <- "study"

# EDIT: an hvtiRdatabuild analysis set, or NULL to read the whole dataset.
ANALYSIS_SET <- NULL

# EDIT: rows to keep, dplyr::filter() style, or NULL to keep every row:
#   WHERE <- quote(age >= 18)
#   WHERE <- rlang::exprs(age >= 18, hx_chf == 1)
WHERE <- NULL

# EDIT: the patient identifier. Without "ccfid" the job uses MRN, then eMRN;
# name another column, such as "randid", if the study uses one.
ID <- "ccfid"

# EDIT: the visit time, in years. The data hold one row per visit, built
# upstream, so the time is part of what makes a row unique.
TIME <- "iv_echo"

# EDIT: what makes a row unique. The ID alone does not: a patient has a row per
# visit. A duplicated visit then stops the read rather than double-counting.
KEY <- c(ID, TIME)

# EDIT: the response measured at each visit, and its kind: "continuous",
# "binary", "ordinal" or "nominal".
RESPONSE <- "<response>"
FAMILY   <- "continuous"

# EDIT: the predictors, or NULL for every other column, which is what the
# studies this template came from all did. NULL leaves out the ID, every KEY
# column, the time, the response, and any ccfid, MRN or eMRN column; it takes
# everything else, dates included. Name the predictors when the built dataset
# carries dates or derived outcomes that should not drive the fit.
PREDICTORS <- NULL

# EDIT: boosting iterations and the learning rate. Studies used 0.01 to 0.05; a
# smaller NU needs a larger M. The error path below says whether M was enough:
# the cross-validated best M should sit well inside it, not at its right edge.
M  <- 1000
NU <- 0.05

# EDIT: seeds the fit and the sample of patients the trajectory plot draws, so
# the report reproduces. Part of the cache key.
SEED <- 2026

# EDIT: the variables to draw partial effects for, or NULL for the six most
# important. N_TRACES patients are sampled for the trajectory plot; Inf draws
# every one.
EFFECT_VARIABLES <- NULL
N_TRACES <- 100

# Set TRUE after changing a choice above. A cache whose inputs changed stops the
# render instead of returning the old fit; TRUE recomputes it.
REFIT <- FALSE
```

## Cohort

```{r}
#| label: data
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
.cfg <- study_config(start = .root)
job_data <- hvtiRtemplates::read_job_data(.cfg, dataset = DATASET, analysis_set = ANALYSIS_SET,
                                          where = WHERE, id = ID, key = KEY)
d <- job_data$data
.provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(job_data$provenance))
knitr::kable(job_data$record, col.names = c("Data", ""), caption = "The data this job read")
if (!is.null(job_data$attrition)) {
  knitr::kable(job_data$attrition, caption = paste0("Analysis set `", ANALYSIS_SET, "`: exclusions, in order"))
}
.selection <- attr(job_data$record, "selection")
.id <- .selection$id
# The data's own spelling of TIME and RESPONSE, matched ignoring case as the
# ID and KEY are, so "IV_ECHO" finds iv_echo. Later chunks use these names.
.resolve <- function(x) {
  hit <- match(tolower(x), tolower(names(d)))
  ifelse(is.na(hit), x, names(d)[hit])
}
.time <- .resolve(TIME)
.response <- .resolve(RESPONSE)
# The fit groups visits by patient, so the ID, the rest of KEY and the visit
# time are its structure, not its predictors; a predictor named for any of them
# would let the model learn who a patient is or when they were seen. A known
# identifier column is never a predictor either, whatever ID names: boostmtree
# keeps its predictors, so one would reach the cache and the saved fit.
.not_predictors <- unique(c(.id, .selection$key, .time, "ccfid", hvtiRtemplates:::.job_identifier_names))
if (is.null(PREDICTORS)) {
  .identifiers <- names(d)[tolower(names(d)) %in% c("ccfid", hvtiRtemplates:::.job_identifier_names) & names(d) != .id]
  .predictors <- names(d)[!tolower(names(d)) %in% tolower(c(.not_predictors, .response))]
  if (length(.identifiers)) {
    cat("Identifier columns left out of the predictors: ", paste(.identifiers, collapse = ", "), "\n", sep = "")
  }
} else {
  .predictors <- PREDICTORS
}
.duplicated <- unique(.predictors[duplicated(.predictors)])
if (length(.duplicated)) {
  stop("PREDICTORS names a variable more than once: ", paste(.duplicated, collapse = ", "), ".", call. = FALSE)
}
.structural <- .predictors[tolower(.predictors) %in% tolower(.not_predictors)]
if (length(.structural)) {
  stop("PREDICTORS names an identifier, a KEY column or the visit time: ", paste(.structural, collapse = ", "),
       ".", call. = FALSE)
}
# The response predicting itself would fit perfectly and mean nothing.
if (tolower(.response) %in% tolower(.predictors)) {
  stop("PREDICTORS names the response (", .response, ").", call. = FALSE)
}
# Nor may an identifier be the response or the visit time: boostmtree keeps
# both, so one would reach the cache and the saved fit. That covers the ID this
# job uses and any other identifier column the data carry, such as a ccfid
# beside an ID of randid.
.identifier_cols <- c(.id, "ccfid", hvtiRtemplates:::.job_identifier_names)
if (tolower(.response) %in% tolower(.identifier_cols)) {
  stop("A patient identifier (", .response, ") cannot be the response.", call. = FALSE)
}
if (tolower(.time) %in% tolower(.identifier_cols)) {
  stop("A patient identifier (", .time, ") cannot be the visit time.", call. = FALSE)
}
.missing <- setdiff(c(.time, .response, .predictors), names(d))
if (length(.missing)) {
  stop("Not in the data this job read: ", paste(.missing, collapse = ", "), call. = FALSE)
}
d <- d[, c(.id, .time, .response, .predictors), drop = FALSE]
# boostmtree imputes a missing predictor but cannot learn from a visit with no
# response, so those visits are dropped here, and counted.
.no_response <- is.na(d[[.response]])
if (any(.no_response)) {
  cat(sum(.no_response), " visit(s) have no ", .response, " and are left out.\n", sep = "")
  d <- d[!.no_response, , drop = FALSE]
}
.chr <- .predictors[vapply(d[.predictors], is.character, logical(1L))]
if (length(.chr)) {
  d[.chr] <- lapply(d[.chr], factor)
  cat("Text predictors converted to factors: ", paste(.chr, collapse = ", "), "\n", sep = "")
}
```

## Fit

```{r}
#| label: fit
#| message: true
# The fit groups visits by patient through `id`. It is given the study-keyed
# digest of each patient's ID, not the ID: a boostmtree fit stores the ID in
# several places, its base learners included, and every one of them is saved
# with it. A digest still groups the visits; it cannot be read back.
CACHE_DIR <- dirname(set_path("estimates", "nb-boostmtree.rds"))
.patient <- hvtiRtemplates:::.id_digest(d[[.id]], hvtiRtemplates:::.study_id_key(.root))
fit <- cache_fit(
  "nb-boostmtree-fit",
  boostmtree(x = d[.predictors], tm = d[[.time]], id = .patient, y = d[[.response]],
             family = FAMILY, M = M, nu = NU, mod.grad = TRUE, cv.flag = TRUE, verbose = FALSE),
  seed = SEED, dir = CACHE_DIR, refit = REFIT
)
```

`cv.flag = TRUE` and `mod.grad = TRUE` are not settings. Every study fitted with
both, and cross-validation is how the job chooses its best number of iterations.

## Does the fit work?

```{r}
#| label: fit-summary
knitr::kable(data.frame(
  quantity = c("Family", "Iterations (M)", "Learning rate (nu)", "Patients", "Visits"),
  value = c(FAMILY, M, NU, length(unique(.patient)), nrow(d))
), col.names = c("", ""), caption = "The fit")
# A continuous or binary fit has one response component; an ordinal or nominal
# fit has one per level after the first, each with its own cross-validated best
# M and error path (fit$err.rate is then a list). Read the error at the best M
# from the standardized l2 column, the scale gg_boost_error() plots.
.err <- if (is.matrix(fit$err.rate)) list(fit$err.rate) else fit$err.rate
# Each component is named by the level that defines it, which boostmtree keeps
# in fit$q.set: a threshold for an ordinal fit, a class compared against the
# reference class for a binary or nominal one. partial.plot() selects a
# component by the same label ("response" for a continuous fit).
.labels <- if (identical(fit$family, "continuous")) "response" else as.character(fit$q.set)
.reference <- setdiff(as.character(fit$y.levels), .labels)
# An ordinal component is a cumulative logit at its threshold, but boostmtree's
# documentation does not say which side of the threshold it models, so the
# label names the threshold only. (Its 2.0.2 source codes P(y <= k); read the
# installed package before relying on that.)
.components <- switch(fit$family,
  continuous = .response,
  binary = ,
  nominal = paste(.response, "=", .labels, "against", paste(.reference, collapse = ", ")),
  ordinal = paste(.response, "threshold", .labels)
)
fit_summary <- data.frame(
  component = .components,
  best_m = as.integer(fit$m.opt),
  cv_error = mapply(function(e, m) e[m, "l2"], .err, fit$m.opt)
)
knitr::kable(fit_summary, col.names = c("Response component", "Best M (cross-validated)",
                                        "Cross-validated error there (standardized)"),
             caption = "Cross-validated fit, by response component")
```

```{r}
#| label: error-path
# The cross-validated error by iteration, and the fitted rho, phi and lambda.
# Best M at the right edge means M was too small.
p_error <- plot(gg_boost_error(fit))
p_path  <- plot(gg_boost_path(fit))
p_error
p_path
```

```{r}
#| label: calibration
# Observed against fitted values over follow-up, in time bins.
p_calibration <- plot(gg_boost_calibration(fit))
p_calibration
```

```{r}
#| label: importance
# Importance of each predictor's main effect and its interaction with time; the
# interaction is where a predictor whose effect changes over follow-up shows.
vimp <- vimp.boostmtree(fit)
p_vimp <- plot(gg_boost_vimp(vimp))
p_vimp
```

```{r}
#| label: effects
# Partial effects over time, for EFFECT_VARIABLES or the six most important.
# An ordinal or nominal fit reports importance once per response component, so
# a variable is ranked by its largest importance across them.
.vimp_main <- gg_boost_vimp(vimp, components = "main")
.top <- sort(tapply(.vimp_main$importance, as.character(.vimp_main$variable), max), decreasing = TRUE)
.effect_vars <- if (is.null(EFFECT_VARIABLES)) head(names(.top), 6L) else EFFECT_VARIABLES
.unknown <- setdiff(.effect_vars, .predictors)
if (length(.unknown)) stop("EFFECT_VARIABLES names no predictor: ", paste(.unknown, collapse = ", "), call. = FALSE)
# One partial.plot() covers every response component: ggBoostedTrees 0.9.0 and
# later take an ordinal or nominal one whole and facet its figure by component.
# Each component is renamed as the fit summary names it, not by its bare level.
.effect <- gg_boost_effect(partial.plot(fit, x.var.names = .effect_vars, output = "data", verbose = FALSE))
.relabel <- .components[match(levels(.effect$response), .labels)]
# A component ggBoostedTrees names differently from the fit would otherwise be
# drawn under a blank label; stop and say so instead.
if (anyNA(.relabel)) {
  stop("ggBoostedTrees labels a response component this job does not recognize: ",
       paste(levels(.effect$response)[is.na(.relabel)], collapse = ", "), ".", call. = FALSE)
}
levels(.effect$response) <- .relabel
# What ggBoostedTrees 0.9.0 still cannot do: a continuous covariate is drawn as
# a curve and a factor as a point per level, on axes ggplot2 cannot share, so
# plot() refuses an effect holding both. Each kind is drawn in its own figure,
# which is why p_effects is a list.
.discrete <- tapply(!is.na(.effect$x_label), .effect$variable, any)
.discrete <- .discrete[!is.na(.discrete)]
p_effects <- lapply(unname(split(names(.discrete), .discrete)), function(v) {
  part <- .effect[.effect$variable %in% v, , drop = FALSE]
  part$variable <- droplevels(part$variable)
  plot(part)
})
for (.p in p_effects) print(.p)
```

```{r}
#| label: traces
# Fitted trajectories for a sample of patients, with their observed values as
# points, and the cohort's mean fitted value over time drawn across them. The
# patients are labeled by digest, so none is identified.
traces <- gg_boost_trajectory(fit)
# Visit times are irregular, so most exact times belong to one patient, and a
# mean taken at each time would join single values. The mean is taken instead
# within ten equal-count time bins, over every patient's fitted values, and
# drawn at each bin's mean time. A bin is per response component when there
# are several.
.edges <- unique(stats::quantile(traces$time, seq(0, 1, length.out = 11L), names = FALSE))
# plot() reports that it chose the patients at random; SEED makes the choice
# the same on every render.
p_traces <- withr::with_seed(SEED, plot(traces, n_max = N_TRACES)) + ggplot2::labs(x = .time, y = .response)
if (length(.edges) < 2L) {
  # Every visit at one time leaves no span to bin, and no line to draw.
  trace_means <- NULL
  cat("Every visit is at one time, so no mean line is drawn.\n")
} else {
  traces$.bin <- cut(traces$time, .edges, include.lowest = TRUE)
  .by <- intersect(c("response", ".bin"), names(traces))
  trace_means <- do.call(rbind, lapply(split(traces, traces[.by], drop = TRUE), function(b) {
    data.frame(b[1L, setdiff(.by, ".bin"), drop = FALSE], time = mean(b$time), fitted = mean(b$fitted),
               n_patients = length(unique(b$id)))
  }))
  p_traces <- p_traces +
    ggplot2::geom_line(data = trace_means, ggplot2::aes(x = .data[["time"]], y = .data[["fitted"]]),
                       linewidth = 1.2, inherit.aes = FALSE)
}
p_traces
```

```{r}
#| label: save
# The product of this job. It is separate from the cache on purpose: the cache
# is this job's internals and this file its interface, which carries the
# selection a later job would rebuild the rows from.
fit <- hvtiRtemplates:::.attach_handoff_lineage(
  fit,
  data = .provenance_data,
  analysis = list(response = .response, time = .time, family = FAMILY, M = M, nu = NU),
  cohort = list(n_patients = length(unique(.patient)), n_visits = nrow(d)),
  selection = attr(job_data$record, "selection")
)
saveRDS(fit, set_path("estimates", "nb-boostmtree.rds"))
```

```{r}
#| label: provenance
#| echo: false
#| results: asis
cat(hvtiRtemplates:::.embed_provenance(
  .in,
  data = .provenance_data,
  extra = list(
    subject = SUBJECT, type = TYPE,
    analysis = list(response = .response, time = .time, family = FAMILY, M = M, nu = NU),
    cohort = list(n_patients = length(unique(.patient)), n_visits = nrow(d))
  )
))
```
