---
title: "Random forest classification: fit"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in:
# rendered as a standalone document the project's format block never applies,
# which is how server-rendered reports end up with sibling _files/ trees instead
# of being self-contained.
#
# NOTE: output-dir is a PROJECT setting and cannot be set here, so the .html
# lands next to its source. With embed-resources that is a single standalone
# file, which is the property that actually matters when a report is sent on.
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (an `rfsrc.*` or `rf.*` job with a
     classification or binary outcome), and say which outcome and candidate
     predictors it used. -->

Replaces `analyses/<job>`: describe its outcome and candidate predictors here.

An `rfc-fit` job grows a classification forest and checks that it learned
something: out-of-bag error by number of trees, and the ROC curve with its
AUC. **Importance, VarPro and dependence plots are not here.** They are the
`rfc-explain` job in this same set, which reads the forest this job saves
instead of growing its own, so the explanations always describe this forest.

Old `rfsrc.*` and `rf.*` jobs map onto three prefixes by outcome, read from the
job's fit call and not from its name: a survival outcome is `rfs`, a
classification or binary outcome `rfc`, a continuous outcome `rfr`.

```{r}
#| label: setup
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRutilities::study_root(if (is.null(.in)) getwd() else dirname(.in))
.provenance_data <- list()
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(randomForestSRC)
  library(ggRandomForests)
  library(hvtiRutilities)
})
# The versions this template was verified against, end to end, on 2026-09-19.
# Below them a call here can fail or change meaning, and the message would
# arrive mid-render rather than here.
if (utils::packageVersion("randomForestSRC") < "3.7.0") {
  stop("This job needs randomForestSRC >= 3.7.0; ", utils::packageVersion("randomForestSRC"),
       " is installed.\nUpdate it, then re-render.", call. = FALSE)
}
if (utils::packageVersion("ggRandomForests") < "4.0.0") {
  stop("This job needs ggRandomForests >= 4.0.0; ", utils::packageVersion("ggRandomForests"),
       " is installed.\nUpdate it, then re-render.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
# The markers in this file name work a study author still has to do, and
# README.md says a job that still contains one has not been finished. This
# chunk is what makes that TRUE rather than merely stated.
#
# Without it an unedited job renders green over a meaningless analysis. Here
# that is a forest grown on the template's placeholder outcome, predictors and
# tree count, whose error curve and ROC curve look exactly like a result.
#
# knitr::current_input() is NULL outside a render -- a study author stepping
# through chunks in RStudio -- and there is no file to scan then, so this is a
# no-op in that case. Same handling as the `set` guard below, for the same
# reason.
#
# The token is BUILT, not written literally, and that is not stylistic. Quarto
# knits through an intermediate and current_input() returns THAT file, so the
# scan reads a copy of this chunk along with everything else: a literal
# grep("<token>", .src) here matches its own source line, and the guard then
# fires on every render, finished or not. Measured, not theorised -- a probe
# found three markers in a file containing two. A guard that cries wolf on a
# finished job is a guard that gets deleted, which returns us to issue #27.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    # Report the marker TEXT. Line numbers here are the intermediate's, not
    # this .qmd's, so they need not match what the author sees in an editor.
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    # A job renders as a draft by default, so an author can iterate on a
    # working report and remove markers as they go. HVTI_TEMPLATE_STRICT
    # turns the draft into a stop for a final render. Only unset, 0, false
    # and no mean "not strict": an unrecognised value stops on purpose, so a
    # mistyped switch is seen rather than quietly ignored, which is the safe
    # direction to be wrong in.
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      # The banner is not optional. A draft render that looks like a finished
      # one is the same defect with an extra step: the warning scrolls past in
      # a log, while the .html is the artifact that gets sent to someone.
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by add_job() from the subject/type it
     was called with, which are also the filename's first two fields. They
     name the SET this job belongs to, and every artifact path in the
     document is computed from them, so no output path below needs editing.

     Do NOT hand-edit these declarations: the render-time guard below checks
     them against this file's own name and stops if they disagree. To put
     this job in a different set, call add_job() again with the new
     subject/type rather than editing this copy in place.

     A set is keyed on (subject, analysis type), not on the subject alone.
     One subject may be analysed by several methods -- parametric hazard, random
     forest survival, competing risks -- and those chains share their upstream.
     Two sets keyed on `dead_pa` alone would write their life tables to the
     same file. -->

```{r}
#| label: set
SUBJECT <- "dead"
TYPE    <- "rfc"

# `add_job()` writes SUBJECT/TYPE from the same values it put in this file's
# name, but a hand-edited declaration can drift from it afterward. set_path()
# below resolves from the declarations, not the filename, so a drifted
# declaration would silently write into ANOTHER set's artifact directory --
# exactly the collision the (subject, type) key exists to prevent, re-entered
# through the body instead of the name.
#
# knitr::current_input() is NULL outside a render (a study author running
# chunks interactively in RStudio), and there is no filename to check against
# yet, so this is a no-op in that case rather than a spurious error.
.current <- knitr::current_input()
if (!is.null(.current)) {
  # Quarto knits through an intermediate, so `knitr::current_input()` returns
  # `<subject>-<type>-<prefix>.rmarkdown` here rather than the
  # `.qmd` this was scaffolded as. Strip whatever extension is actually present
  # rather than hard-coding one, so this doesn't depend on a build-tool detail
  # staying the same.
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject, "', type '",
         .name_type, "'), but declares SUBJECT = \"", SUBJECT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

# Resolve a path inside this set's artifact directory. `kind` is the artifact
# folder -- "estimates" for serialized results, "graphs" for figures. The set
# directory sits one layer under the kind and never two, which is the whole
# layout rule.
#
# Created on first use rather than up front, so a job that writes nothing leaves
# no empty directories behind.
set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}

# `rfc-fit` writes the forest through this and `rfc-explain` reads it back,
# both by set, so an explanation can never be filed against a different set
# than the forest it explains. The write is the `save` chunk at the foot of
# this file.
```

## Study choices

Edit these values for this study before rendering.

```{r}
#| label: study-choices
# EDIT: the outcome to classify. Any coding works: 0/1, "yes"/"no" or a label.
# The read step makes it a factor, which is what makes this a classification
# forest rather than a regression on the codes.
RESPONSE <- "dead"

# EDIT: the class whose one-versus-rest ROC curve and AUC this report should
# show. Use its label after RESPONSE becomes a factor: "1" for a 0/1 outcome,
# "yes" for no/yes, or the clinical category of interest for multiclass data.
ROC_CLASS <- NA_character_

# EDIT: the candidate predictors. Name them rather than taking every column: a
# built dataset carries identifiers, dates and derived outcomes, and a forest
# will split on a date that quietly encodes follow-up.
PREDICTORS <- c("age", "female", "hx_chf")

# EDIT: forest size and seed. The error-by-trees plot below says whether NTREE
# was enough: a curve still falling at its right edge was not. SEED makes the
# forest reproducible and is part of its cache key.
NTREE <- 1000
SEED  <- 2026

# EDIT: "na.omit" drops a patient missing any predictor; "na.impute" keeps them
# and imputes inside the forest. The classification and regression exemplars
# imputed. Dropping is simpler to report and can shrink the cohort a great deal.
NA_ACTION <- "na.omit"

# Set TRUE after changing a choice above. A cache whose inputs changed stops the
# render instead of returning the old forest; TRUE recomputes only the caches
# that are out of date, so it is safe to leave on while iterating.
REFIT <- FALSE
```

## Cohort

```{r}
#| label: read
.cfg <- study_config(start = .root)
.data_read <- hvtiRtemplates:::.provenance_read(
  "study", .cfg, function() read_built(cfg = .cfg)
)
d <- .data_read$value
.provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(.data_read$record))
.outcomes <- intersect(RESPONSE, PREDICTORS)
if (length(.outcomes)) {
  stop("Do not include the outcome variable in PREDICTORS: ", .outcomes, ".", call. = FALSE)
}
.duplicated <- unique(PREDICTORS[duplicated(PREDICTORS)])
if (length(.duplicated)) {
  stop("PREDICTORS names a variable more than once: ", paste(.duplicated, collapse = ", "), ".", call. = FALSE)
}
.missing <- setdiff(c(RESPONSE, PREDICTORS), names(d))
if (length(.missing)) {
  stop("Not in the built dataset: ", paste(.missing, collapse = ", "), call. = FALSE)
}
d <- d[, c(RESPONSE, PREDICTORS), drop = FALSE]

# A forest cannot learn from a patient with no outcome, and na.impute would
# impute one. So the outcome is required whatever NA_ACTION says.
.no_outcome <- is.na(d[[RESPONSE]])
if (any(.no_outcome)) {
  stop(sum(.no_outcome), " patient(s) have no ", RESPONSE, ". Resolve them in the ",
       "dataset build; the forest will not.", call. = FALSE)
}

# A numeric 0/1 outcome grows a REGRESSION forest, whose "probabilities" can
# fall outside 0 and 1 and whose error is a mean squared error. Nothing warns.
d[[RESPONSE]] <- factor(d[[RESPONSE]])
if (!is.character(ROC_CLASS) || length(ROC_CLASS) != 1L || is.na(ROC_CLASS) || !nzchar(ROC_CLASS) ||
      !ROC_CLASS %in% levels(d[[RESPONSE]])) {
  stop("ROC_CLASS must name one observed level of ", RESPONSE, "; ", deparse(ROC_CLASS),
       " is not an observed level. Choose one of: ", paste(levels(d[[RESPONSE]]), collapse = ", "), ".", call. = FALSE)
}
ROC_OUTCOME <- match(ROC_CLASS, levels(d[[RESPONSE]]))
```

## Forest

```{r}
#| label: fit
#| message: true
# cache_fit() writes <dir>/<name>.rds and its default dir is the study's
# estimates folder, not this set's, so every call here names the set's.
CACHE_DIR <- dirname(set_path("estimates", "rfc.rds"))
model <- stats::reformulate(".", response = RESPONSE)
forest <- cache_fit(
  "rfc-forest",
  rfsrc(model, data = d, ntree = NTREE, na.action = NA_ACTION, importance = "none"),
  seed = SEED, dir = CACHE_DIR, refit = REFIT
)
forest
```

`importance = "none"` is deliberate. Importance is the explain job's work, and
cached there on its own, so it can be recomputed without regrowing the forest.

## Does the forest work?

```{r}
#| label: diagnostics
# Out-of-bag error by number of trees, overall and per class.
err <- gg_error(forest)
plot(err)

# Out-of-bag one-versus-rest ROC curve, and the area under it, for the class
# named explicitly above. ggRandomForests selects by factor-level position,
# so ROC_OUTCOME maps the study's label to its position instead of letting
# factor ordering decide what this report means.
roc <- gg_roc(forest, which_outcome = ROC_OUTCOME)
plot(roc)
auc <- calc_auc(roc)
auc
```

`gg_brier()` is not here: it supports survival forests only.

```{r}
#| label: save
# The product of this job, and what rfc-explain reads. It is separate from the
# cache on purpose: cache_fit() stores a keyed record rather than the bare
# forest, so the cache is this job's internals and this file is its interface.
# The price is the forest on disk twice.
saveRDS(forest, set_path("estimates", "rfc.rds"))
```

```{r}
#| label: provenance
#| echo: false
#| results: asis
cat(hvtiRtemplates:::.embed_provenance(
  .in,
  data = .provenance_data,
  extra = list(
    subject = SUBJECT,
    type = TYPE,
    analysis = list(
      outcome = list(
        variable = RESPONSE,
        kind = "classification",
        observed_levels = levels(forest$yvar),
        target_level = ROC_CLASS
      )
    ),
    cohort = list(n = as.integer(length(forest$yvar)))
  )
))
```
