---
title: "General descriptive checks"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    page-layout: full
    grid:
      body-width: 2000px
      sidebar-width: 250px
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (descriptive/dc.general). -->

Replaces `descriptive/dc.general`.

A `dc-general` job is the first look at a built cohort, before any table or
model: what the data contain, how each categorical variable breaks down, how
each continuous variable is distributed, and which variables move together. It
uses base procedures only. The formatted manuscript table is a `dc-tables` job.

```{r}
#| label: setup
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRtemplates:::.find_study_root(if (is.null(.in)) getwd() else dirname(.in))
.provenance_data <- list()
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages(library(hvtiRutilities))
```

```{r}
#| label: guard-edits
#| results: asis
# The markers in this file name work a study author still has to do, and a job
# that still contains one has not been finished. This chunk is what makes that
# TRUE rather than merely stated: an unedited job would otherwise render green
# over placeholder columns.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

```{r}
#| label: guard-partial
#| results: asis
# unnumbered: a callout, printed only when part of the job is left out
# To render a job you have not finished, leave a chunk out with the chunk
# option skip, giving the reason in quotes, or call hvtiRtemplates::stop_here()
# in a chunk to leave out everything below it. A draft lists each one here; a
# final render refuses them, as it refuses an EDIT marker. ?stop_here has more.
hvtiRtemplates:::.guard_partial(knitr::current_input())
```

```{r}
#| label: set
SUBJECT <- "cohort"
TYPE    <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject,
         "', type '", .name_type, "'), but declares SUBJECT = \"", SUBJECT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```

## Study choices

Edit these values for this study before rendering.

```{r}
#| label: edit-study-choices
# EDIT: the registered dataset this job reads ("study" is the built dataset).
DATASET <- "study"

# EDIT: an hvtiRdatabuild analysis set, or NULL to read the whole dataset.
ANALYSIS_SET <- NULL

# EDIT: rows to keep, dplyr::filter() style, or NULL to keep every row:
#   WHERE <- quote(age >= 18)
#   WHERE <- rlang::exprs(age >= 18, hx_chf == 1)
WHERE <- NULL

# EDIT: the patient identifier. Without "ccfid" the job uses MRN, then eMRN;
# name another column, such as "randid", if the study uses one.
ID <- "ccfid"

# EDIT: what makes a row unique; one row per patient unless repeated measures
# add their visit time or date, for example KEY <- c(ID, "iv_echo").
KEY <- ID

# EDIT: grouped categorical variables. Each name is a section heading, in this
# order. Carry the /* Demography */ banners across from the SAS %macro freq
# tables list. Missing values are tabulated as their own level (missprint).
CATEGORICAL <- list(
  Demography = c("female")
)

# EDIT: continuous variables, grouped the same way (the SAS %macro cdfs var
# list).
CONTINUOUS <- list(
  Demography = c("age")
)

# EDIT: variables for the pairwise correlation sweep (SAS proc corr nosimple
# rank). Defaults to every continuous variable; character() skips the section.
CORR_VARS <- unlist(CONTINUOUS, use.names = FALSE)

# EDIT: leave NULL to keep identifiers out of reports. A column name (the SAS
# job used id ccfid) labels the extreme values with it, which puts patient
# identifiers in the report: a report rendered that way must not be circulated.
ID_COL <- NULL
```

## Data

<!-- How this job finds its data. Read this before touching the chunk below.

     A study can offer a job data three ways, and `_study.yml` says which exist:

     1. The study dataset: the file named by `built:`. `DATASET <- "study"`
        with `ANALYSIS_SET <- NULL`.
     2. An additional dataset: any file declared under `additional_datasets:`,
        for example a column subset written for R. Set `DATASET` to its
        name there and `ANALYSIS_SET <- NULL`. It is read as it stands: only
        `WHERE` removes rows.
     3. An analysis set: declared under `analysis_sets:` with the columns kept
        and the rows excluded, then written once with
        `hvtiRdatabuild::write_analysis_set()`. Set `ANALYSIS_SET` to its name
        and leave `DATASET` as "study", because a set is always cut from the
        study dataset. The report shows its exclusions in order.

     A name under `additional_datasets:` is not an analysis set, so putting it
     in `ANALYSIS_SET` stops with "No analysis set". Whichever route is taken,
     the report prints what it read, the identifier it used, the rows each
     `WHERE` condition removed and the rows and patients kept, so a reader
     never has to ask which data a table describes. -->

```{r}
#| label: tbl-data
#| tbl-cap: "The data this job read"
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
.cfg <- study_config(start = .root)
job_data <- hvtiRtemplates::read_job_data(.cfg, dataset = DATASET, analysis_set = ANALYSIS_SET,
                                          where = WHERE, id = ID, key = KEY)
d <- job_data$data
.provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(job_data$provenance))
knitr::kable(job_data$record, col.names = c("Data", ""))
```

```{r}
#| label: data-attrition
#| results: asis
# unnumbered: its child chunk carries its own label and caption
if (!is.null(job_data$attrition)) {
  .fence <- strrep("`", 3)
  cat(knitr::knit_child(text = c(
    paste0(.fence, "{r}"), "#| label: tbl-data-attrition",
    paste0("#| tbl-cap: ", encodeString(paste0("Analysis set `", ANALYSIS_SET, "`: exclusions, in order"), quote = "\"")),
    "knitr::kable(job_data$attrition)", .fence
  ), envir = environment(), quiet = TRUE), sep = "\n")
}
```

```{r}
#| label: derive
# The identifier column the data step used: ID as named, or MRN when the data
# have no ccfid. It is kept out of every summary, because the overall minimum
# and maximum would each be one patient's identifier.
.id <- attr(job_data$record, "selection")$id
PROBS <- c(0, 0.01, 0.05, 0.10, 0.25, 0.50, 0.75, 0.90, 0.95, 0.99, 1)
check_group_names <- function(x, setting) {
  if (!length(x)) return(invisible())
  nm <- names(x)
  if (is.null(nm) || any(nm == "") || anyDuplicated(nm)) {
    stop(setting, " must be a list with a unique, non-empty name for every group.", call. = FALSE)
  }
}
check_group_names(CATEGORICAL, "CATEGORICAL")
check_group_names(CONTINUOUS, "CONTINUOUS")
if (!is.null(ID_COL) && !(is.character(ID_COL) && length(ID_COL) == 1L)) {
  stop("ID_COL must be NULL or a single column name.", call. = FALSE)
}
cat_vars <- unlist(CATEGORICAL, use.names = FALSE)
con_vars <- unlist(CONTINUOUS, use.names = FALSE)
unknown <- setdiff(unique(c(cat_vars, con_vars, CORR_VARS, ID_COL, .id)), names(d))
if (length(unknown)) {
  stop("Unknown column(s): ", paste(unknown, collapse = ", "), call. = FALSE)
}
numeric_vars <- unique(c(con_vars, CORR_VARS))
not_numeric <- numeric_vars[!vapply(numeric_vars, function(v) is.numeric(d[[v]]), logical(1))]
if (length(not_numeric)) {
  stop("Not numeric, so not summarized as continuous or correlated: ",
       paste(not_numeric, collapse = ", "), call. = FALSE)
}
key_overlap <- intersect(unique(c(cat_vars, con_vars, CORR_VARS)), .id)
if (length(key_overlap)) {
  stop("The ID column cannot be summarized: ", paste(key_overlap, collapse = ", "), call. = FALSE)
}
overall_vars <- setdiff(names(d)[vapply(d, is.numeric, logical(1))], c(.id, ID_COL))

freqs <- lapply(CATEGORICAL, function(vars) {
  stats::setNames(lapply(vars, function(v) {
    tab <- table(d[[v]], useNA = "ifany")
    is_na_level <- is.na(names(tab))
    n_int <- as.integer(tab)
    total_nonmissing <- sum(n_int[!is_na_level])
    percent <- rep(NA_real_, length(tab))
    if (total_nonmissing > 0L) {
      percent[!is_na_level] <- round(100 * n_int[!is_na_level] / total_nonmissing, 1)
    }
    data.frame(
      level = ifelse(is_na_level, "(missing)", names(tab)),
      n = n_int,
      percent = percent
    )
  }), vars)
})

cdfs <- lapply(CONTINUOUS, function(vars) {
  stats::setNames(lapply(vars, function(v) {
    x <- d[[v]]
    seen <- which(!is.na(x))
    lowest <- seen[order(x[seen])][seq_len(min(5L, length(seen)))]
    highest <- seen[order(-x[seen])][seq_len(min(5L, length(seen)))]
    extremes <- data.frame(
      end = rep(c("lowest", "highest"), c(length(lowest), length(highest))),
      value = x[c(lowest, highest)]
    )
    if (!is.null(ID_COL)) extremes[[ID_COL]] <- d[[ID_COL]][c(lowest, highest)]
    list(
      summary = data.frame(n = length(seen), nmiss = sum(is.na(x))),
      quantiles = data.frame(
        percent = 100 * PROBS,
        value = if (length(seen)) stats::quantile(x[seen], PROBS, type = 2, names = FALSE) else NA_real_
      ),
      extremes = extremes
    )
  }), vars)
})

corr_vars <- unique(CORR_VARS)
corrs <- data.frame(var1 = character(), var2 = character(), r = numeric(), n = integer())
if (length(corr_vars) >= 2L) {
  pairs <- utils::combn(corr_vars, 2L)
  pair_stats <- apply(pairs, 2L, function(p) {
    ok <- stats::complete.cases(d[[p[[1L]]]], d[[p[[2L]]]])
    r <- if (sum(ok) >= 2L) stats::cor(d[[p[[1L]]]][ok], d[[p[[2L]]]][ok]) else NA_real_
    c(r = r, n = sum(ok))
  })
  corrs <- data.frame(
    var1 = pairs[1L, ], var2 = pairs[2L, ],
    r = pair_stats["r", ], n = as.integer(pair_stats["n", ])
  )
  corrs <- corrs[order(-abs(corrs$r)), , drop = FALSE]
  rownames(corrs) <- NULL
}
```

## Other investigations

<!-- EDIT: study-specific checks go here, above the standard sections (the SAS
job's O T H E R   I N V E S T I G A T I O N S banner). Delete this marker when
there are none. -->

## Overall statistics

```{r}
#| label: tbl-overall-contents
#| tbl-cap: "Contents of the data: each column, its type and its label"
proc_contents(d)
```

```{r}
#| label: overall-means
#| results: asis
# unnumbered: its child chunk carries its own label and caption
if (length(overall_vars)) {
  .fence <- strrep("`", 3)
  cat(knitr::knit_child(text = c(
    paste0(.fence, "{r}"), "#| label: tbl-overall-means",
    "#| tbl-cap: \"Summary statistics of the overall variables\"",
    "proc_means(d, vars = overall_vars, stats = c(\"n\", \"nmiss\", \"mean\", \"std\", \"min\", \"max\", \"sum\"))",
    .fence
  ), envir = environment(), quiet = TRUE), sep = "\n")
}
```

## Contingency tables for categorical variables

```{r}
#| label: freqs
#| results: asis
# unnumbered: each child chunk below carries its own label and caption
.kable_na <- options(knitr.kable.NA = "")
.fence <- strrep("`", 3)
.seen <- new.env()
.child <- function(kind, key, caption, code) {
  label <- paste(kind, "freqs", gsub("(^-+|-+$)", "", gsub("[^a-z0-9]+", "-", tolower(key))), sep = "-")
  if (!is.null(.seen[[label]])) stop("Two outputs would share the label ", label, ".", call. = FALSE)
  .seen[[label]] <- TRUE
  cat(knitr::knit_child(text = c(paste0(.fence, "{r}"), paste0("#| label: ", label),
                                 paste0("#| ", kind, "-cap: ", encodeString(caption, quote = "\"")), code, .fence),
                        envir = parent.frame(), quiet = TRUE), sep = "\n")
}
# One table per group, a block of rows per variable, rather than a table per
# variable: a 0/1 variable is two rows, and a table of two rows each is most of
# a report spent on captions.
for (group in names(freqs)) {
  .freq <- do.call(rbind, lapply(names(freqs[[group]]), function(v) {
    f <- freqs[[group]][[v]]
    cbind(variable = ifelse(seq_len(nrow(f)) == 1L, v, ""), f)
  }))
  .child("tbl", group, paste0("Frequencies, ", group), "knitr::kable(.freq, row.names = FALSE)")
  cat("\n")
}
options(.kable_na)
```

## Cumulative distributions for continuous variables

```{r}
#| label: cdfs
#| results: asis
# unnumbered: each child chunk below carries its own label and caption
.fence <- strrep("`", 3)
.seen <- new.env()
.child <- function(kind, key, caption, code) {
  label <- paste(kind, "cdfs", gsub("(^-+|-+$)", "", gsub("[^a-z0-9]+", "-", tolower(key))), sep = "-")
  if (!is.null(.seen[[label]])) stop("Two outputs would share the label ", label, ".", call. = FALSE)
  .seen[[label]] <- TRUE
  cat(knitr::knit_child(text = c(paste0(.fence, "{r}"), paste0("#| label: ", label),
                                 paste0("#| ", kind, "-cap: ", encodeString(caption, quote = "\"")), code, .fence),
                        envir = parent.frame(), quiet = TRUE), sep = "\n")
}
# One quantile table per group, a row per variable with its n and nmiss beside
# its quantiles, rather than a one-row count table and a quantile table for
# every variable. The extremes stay one table per variable, since ID_COL can
# add a column to them.
for (group in names(cdfs)) {
  cat("\n### ", group, "\n\n", sep = "")
  .quant <- do.call(rbind, lapply(names(cdfs[[group]]), function(v) {
    .cdf <- cdfs[[group]][[v]]
    row <- as.data.frame(as.list(stats::setNames(.cdf$quantiles$value, paste0(.cdf$quantiles$percent, "%"))),
                         check.names = FALSE)
    cbind(variable = v, .cdf$summary, row)
  }))
  .child("tbl", paste(group, "quantiles"), paste0("Quantiles, ", group, " (SAS QNTLDEF=5)"),
         "knitr::kable(.quant, row.names = FALSE)")
  cat("\n")
  for (v in names(cdfs[[group]])) {
    .cdf <- cdfs[[group]][[v]]
    .child("tbl", paste(group, v, "extremes"), paste0("Five lowest and five highest values, ", v),
           "knitr::kable(.cdf$extremes)")
    cat("\n")
  }
}
```

## Pair-wise correlations

```{r}
#| label: corrs
#| results: asis
# unnumbered: its child chunk carries its own label and caption
if (nrow(corrs)) {
  .fence <- strrep("`", 3)
  cat(knitr::knit_child(text = c(
    paste0(.fence, "{r}"), "#| label: tbl-corrs",
    "#| tbl-cap: \"Pearson correlations, pairwise complete, strongest first\"",
    "knitr::kable(corrs, digits = 3, row.names = FALSE)", .fence
  ), envir = environment(), quiet = TRUE), sep = "\n")
} else {
  cat("No correlation sweep: `CORR_VARS` names fewer than two variables.\n")
}
```

`proc_means()` covers every numeric column other than `ID` and `ID_COL`.
Percentages in the contingency tables are over non-missing rows, as SAS
`MISSPRINT` computes them; the `(missing)` row shows its count with a blank
percent. Quantiles are SAS `QNTLDEF=5`, the `proc univariate` default. The
highest extremes are listed largest first (SAS lists them ascending). The SAS
job printed `ccfid` beside every extreme value so an author could look the
patient up; this job prints the values alone unless `ID_COL` is set under
Study choices.

```{r}
#| label: provenance
#| echo: false
#| results: asis
cat(hvtiRtemplates:::.embed_provenance(
  .in,
  data = .provenance_data,
  extra = list(
    subject = SUBJECT,
    type = TYPE
  )
))
```
