---
title: "Goodness of follow-up"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    page-layout: full
    grid:
      body-width: 2000px
      sidebar-width: 250px
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (descriptive/dc.gfup). -->

Replaces `descriptive/dc.gfup`.

A `dc-gfup` job checks the recorded follow-up intervals before time-related
analyses. We summarize the full cohort and the event and censored subsets,
then review missing, negative and zero intervals. Follow-up units must be
years. Derive any close-date or potential-follow-up measures in the data build
and register them before selecting them here. These tables describe recorded
follow-up; they do not establish completeness against a close date.
The figure, each patient's follow-up against the diagonal a close date allows,
is the `dp-gfup` job; read the same data and rows there.

```{r}
#| label: setup
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRtemplates:::.find_study_root(if (is.null(.in)) getwd() else dirname(.in))
.provenance_data <- list()
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages(library(hvtiRutilities))
if (utils::packageVersion("hvtiRutilities") < "1.4.1") {
  stop("This job needs hvtiRutilities >= 1.4.1 for followup_check(); ",
       utils::packageVersion("hvtiRutilities"), " is installed.", call. = FALSE)
}
```

```{r}
#| label: guard-edits
#| results: asis
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

```{r}
#| label: set
SUBJECT <- "cohort"
TYPE    <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject,
         "', type '", .name_type, "'), but declares SUBJECT = \"", SUBJECT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```

## Study choices

Set the values in this chunk before rendering.

```{r}
#| label: edit-study-choices
# EDIT: the registered dataset this job reads ("study" is the built dataset).
# MIGRATE-BEGIN: dc-gfup-data
DATASET <- "study"
# MIGRATE-END: dc-gfup-data

# EDIT: an hvtiRdatabuild analysis set, or NULL to read the whole dataset.
ANALYSIS_SET <- NULL

# EDIT: rows to keep, dplyr::filter() style, or NULL to keep every row:
#   WHERE <- quote(age >= 18)
#   WHERE <- rlang::exprs(age >= 18, hx_chf == 1)
WHERE <- NULL

# EDIT: the patient identifier. Without "ccfid" the job uses MRN, then eMRN;
# name another column, such as "randid", if the study uses one.
ID <- "ccfid"

# EDIT: what makes a row unique; one row per patient unless repeated measures
# add their visit time or date, for example KEY <- c(ID, "iv_echo").
KEY <- ID

# MIGRATE-BEGIN: dc-gfup-config
EVENT <- "dead"       # EDIT: binary event indicator, 1=event and 0=censored
FOLLOWUP <- "iv_dead" # EDIT: follow-up interval(s) this job checks, in years
IDENTIFIER <- NULL    # EDIT: opt in only for local review
MAX_REVIEW_ROWS <- 25L
# MIGRATE-END: dc-gfup-config
# EDIT: optional event-coding cross-tabs, each a vector of column names.
# Missing values are shown. Use list() when no additional checks are needed.
CHECKS <- list()
```

## Data

<!-- How this job finds its data. Read this before touching the chunk below.

     A study can offer a job data three ways, and `_study.yml` says which exist:

     1. The study dataset: the file named by `built:`. `DATASET <- "study"`
        with `ANALYSIS_SET <- NULL`.
     2. An additional dataset: any file declared under `additional_datasets:`,
        for example a column subset written for R. Set `DATASET` to its
        name there and `ANALYSIS_SET <- NULL`. It is read as it stands: only
        `WHERE` removes rows.
     3. An analysis set: declared under `analysis_sets:` with the columns kept
        and the rows excluded, then written once with
        `hvtiRdatabuild::write_analysis_set()`. Set `ANALYSIS_SET` to its name
        and leave `DATASET` as "study", because a set is always cut from the
        study dataset. The report shows its exclusions in order.

     A name under `additional_datasets:` is not an analysis set, so putting it
     in `ANALYSIS_SET` stops with "No analysis set". Whichever route is taken,
     the report prints what it read, the identifier it used, the rows each
     `WHERE` condition removed and the rows and patients kept, so a reader
     never has to ask which data a table describes. -->

```{r}
#| label: data
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
.cfg <- study_config(start = .root)
job_data <- hvtiRtemplates::read_job_data(.cfg, dataset = DATASET, analysis_set = ANALYSIS_SET,
                                          where = WHERE, id = ID, key = KEY)
d <- job_data$data
.provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(job_data$provenance))
knitr::kable(job_data$record, col.names = c("Data", ""), caption = "The data this job read")
if (!is.null(job_data$attrition)) {
  knitr::kable(job_data$attrition, caption = paste0("Analysis set `", ANALYSIS_SET, "`: exclusions, in order"))
}
```


## Follow-up checks

Missing event status is counted separately and excluded from both event and
censored subsets. Suspicious rows have missing event status or at least one
missing, negative or zero interval. The local review prints at most
`MAX_REVIEW_ROWS` rows, in source order, and includes only selected fields.
Identifiers remain disabled until you explicitly select one. Check the output
before sharing a report that contains an identifier.

```{r}
#| label: qc
#| results: asis
# Tables, not console output: a report reader should not have to parse print().
show_table <- function(x, caption) {
  cat("\n", knitr::kable(x, caption = caption, row.names = FALSE), "\n", sep = "\n")
}
# The checks and tables are hvtiRutilities::followup_check(), the function the
# EDA report calls too, so the two reports cannot disagree. It checks every
# choice above before computing anything, and names every missing column at once.
fc <- hvtiRutilities::followup_check(d, event = EVENT, followup = FOLLOWUP,
                                     identifier = IDENTIFIER, max_rows = MAX_REVIEW_ROWS)
cohort_counts <- fc$cohort
followup_qc <- fc$intervals
followup_means <- fc$means
review_rows <- fc$review
show_table(cohort_counts, "Cohort: event, censored and missing event status")
show_table(followup_qc, "Follow-up intervals: missing, negative and zero values")
for (nm in names(followup_means)) {
  show_table(followup_means[[nm]], paste0("Follow-up, ", nm, " (proc_means)"))
}
show_table(review_rows, paste0("Suspicious rows, first ", MAX_REVIEW_ROWS, " at most"))
```

The QC quartiles use R's default interpolated quantiles. `proc_means()` uses
SAS `QNTLDEF=5`, so its quartiles can differ. The migration report preserves
legacy sort orders, date arithmetic and includes for review; none establishes
an equivalent follow-up definition by itself.

## Event-coding consistency

```{r}
#| label: checks
#| results: asis
unknown_checks <- setdiff(unlist(CHECKS, use.names = FALSE), names(d))
if (length(unknown_checks)) {
  stop("Unknown event-check column(s): ", paste(unknown_checks, collapse = ", "), call. = FALSE)
}
for (cc in CHECKS) {
  cat("\n", knitr::kable(as.data.frame(table(d[cc], useNA = "ifany")),
                         caption = paste(cc, collapse = " by ")), "\n", sep = "\n")
}
```

```{r}
#| label: provenance
#| echo: false
#| results: asis
cat(hvtiRtemplates:::.embed_provenance(
  .in,
  data = .provenance_data,
  extra = list(
    subject = SUBJECT,
    type = TYPE,
    analysis = list(
      event = list(variable = EVENT, event = 1L, censored = 0L),
      followup = list(variable = FOLLOWUP)
    ),
    cohort = as.list(cohort_counts[1L, , drop = FALSE])
  )
))
```
