---
title: "Goodness of follow-up"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    page-layout: full
    grid:
      body-width: 2000px
      sidebar-width: 250px
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (descriptive/dc.gfup). -->

Replaces `descriptive/dc.gfup`.

A `dc-gfup` job checks follow-up before time-related analyses. We summarize
the full cohort and the event and censored subsets, review missing, negative
and zero intervals, then draw the goodness-of-follow-up figure. Follow-up
units must be years. Derive any potential-follow-up measures in the data build
and register them before selecting them here.

The tables describe recorded follow-up. The figure measures it against a close
date: every patient is one point, the year of operation across and the
follow-up recorded for them up, blue while alive and red once dead. The dashed
diagonal is the follow-up a patient **could** have had, from their operation
to the close date. A censored patient far below the diagonal is someone whose
follow-up stopped early, and a band of them is a follow-up problem to settle
before any time-related analysis.

The figure is `hvtiPlotR::hv_followup_panels()`, which the EDA report also
calls, so its follow-up panels are these. It was the separate `dp-gfup` job,
now deprecated in favor of this one.

```{r}
#| label: setup
# unnumbered: loads packages and checks versions only
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRtemplates:::.find_study_root(if (is.null(.in)) getwd() else dirname(.in))
.provenance_data <- list()
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(hvtiRutilities)
  library(hvtiPlotR)
  library(ggplot2)
})
if (utils::packageVersion("hvtiRutilities") < "1.4.1") {
  stop("This job needs hvtiRutilities >= 1.4.1 for followup_check(); ",
       utils::packageVersion("hvtiRutilities"), " is installed.", call. = FALSE)
}
if (utils::packageVersion("hvtiPlotR") < "2.8.0") {
  stop("This job needs hvtiPlotR >= 2.8.0 for hv_followup_panels() and scale_color_hv(); ",
       utils::packageVersion("hvtiPlotR"), " is installed.", call. = FALSE)
}
```

```{r}
#| label: guard-edits
#| results: asis
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

```{r}
#| label: set
SUBJECT <- "cohort"
TYPE    <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject,
         "', type '", .name_type, "'), but declares SUBJECT = \"", SUBJECT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```

## Study choices

Set the values in this chunk before rendering.

```{r}
#| label: edit-study-choices
# EDIT: the registered dataset this job reads ("study" is the built dataset).
# MIGRATE-BEGIN: dc-gfup-data
DATASET <- "study"
# MIGRATE-END: dc-gfup-data

# EDIT: an hvtiRdatabuild analysis set, or NULL to read the whole dataset.
ANALYSIS_SET <- NULL

# EDIT: rows to keep, dplyr::filter() style, or NULL to keep every row:
#   WHERE <- quote(age >= 18)
#   WHERE <- rlang::exprs(age >= 18, hx_chf == 1)
WHERE <- NULL

# EDIT: the patient identifier. Without "ccfid" the job uses MRN, then eMRN;
# name another column, such as "randid", if the study uses one.
ID <- "ccfid"

# EDIT: what makes a row unique; one row per patient unless repeated measures
# add their visit time or date, for example KEY <- c(ID, "iv_echo").
KEY <- ID

# MIGRATE-BEGIN: dc-gfup-config
EVENT <- "dead"       # EDIT: binary event indicator, 1=event and 0=censored
FOLLOWUP <- "iv_dead" # EDIT: follow-up interval(s) this job checks, in years
IDENTIFIER <- NULL    # EDIT: opt in only for local review
MAX_REVIEW_ROWS <- 25L
# MIGRATE-END: dc-gfup-config
# EDIT: optional event-coding cross-tabs, each a vector of column names.
# Missing values are shown. Use list() when no additional checks are needed.
CHECKS <- list()

# The figure. These are dp-eda's follow-up choices, and mean what they mean there.
# EDIT: the years-since-origin interval to the operation, and that origin, a
# calendar year such as 1990. The origin differs between studies, and a wrong
# one slides every point along the x-axis without any other symptom, which is
# why the operation years are checked and printed below.
OPYRS <- "iv_opyrs"
ORIGIN_YEAR <- 1990

# EDIT: the close date of follow-up, as.Date("YYYY-MM-DD"). NULL estimates it
# as the latest operation date plus follow-up in the data, which is the date
# follow-up is known to reach, not the date it was closed; the report says
# which one it drew.
CLOSE_DATE <- NULL

# EDIT: one entry per death panel. `status` is the 1/0 death indicator and
# `time` its follow-up in years; the default draws the same columns as the
# EVENT and FOLLOWUP defaults above. A study that ascertains deaths two ways draws both,
# for example
#   systematic = list(status = "deads", time = "iv_deads", title = "Systematic deaths")
PANELS <- list(
  all = list(status = "dead", time = "iv_dead", title = "All deaths")
)

# EDIT: optional event panels, one entry per non-fatal event. `event` and
# `time` are the event's indicator and interval; `death` and `death_time` the
# death it competes with. A flagged event counts only when it comes strictly
# before death, so ties go to death. list() draws none. For example
#   reop = list(event = "ev_reop", time = "iv_reop", death = "dead",
#               death_time = "iv_dead", label = "Reoperation")
EVENTS <- list()

# EDIT: point transparency. 0.5 lets a dense cohort read as a distribution
# rather than a blob; raise it for a small one.
ALPHA <- 0.5

# EDIT: NULL draws the house colors through hvtiPlotR::scale_color_hv(): dead
# vermillion, alive blue and a non-fatal event green, all colorblind safe. Name
# all three to choose your own; c(alive = "#377EB8", dead = "#E41A1C",
# event = "#4DAF4A") is ColorBrewer Set1, what these jobs drew before 1.2.3.
COLORS <- NULL
```

## Data

<!-- How this job finds its data. Read this before touching the chunk below.

     A study can offer a job data three ways, and `_study.yml` says which exist:

     1. The study dataset: the file named by `built:`. `DATASET <- "study"`
        with `ANALYSIS_SET <- NULL`.
     2. An additional dataset: any file declared under `additional_datasets:`,
        for example a column subset written for R. Set `DATASET` to its
        name there and `ANALYSIS_SET <- NULL`. It is read as it stands: only
        `WHERE` removes rows.
     3. An analysis set: declared under `analysis_sets:` with the columns kept
        and the rows excluded, then written once with
        `hvtiRdatabuild::write_analysis_set()`. Set `ANALYSIS_SET` to its name
        and leave `DATASET` as "study", because a set is always cut from the
        study dataset. The report shows its exclusions in order.

     A name under `additional_datasets:` is not an analysis set, so putting it
     in `ANALYSIS_SET` stops with "No analysis set". Whichever route is taken,
     the report prints what it read, the identifier it used, the rows each
     `WHERE` condition removed and the rows and patients kept, so a reader
     never has to ask which data a table describes. -->

```{r}
#| label: tbl-data
#| tbl-cap: "The data this job read"
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
.cfg <- study_config(start = .root)
job_data <- hvtiRtemplates::read_job_data(.cfg, dataset = DATASET, analysis_set = ANALYSIS_SET,
                                          where = WHERE, id = ID, key = KEY)
d <- job_data$data
.provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(job_data$provenance))
knitr::kable(job_data$record, col.names = c("Data", ""))
```

```{r}
#| label: data-attrition
#| results: asis
# unnumbered: its child chunk carries its own label and caption
if (!is.null(job_data$attrition)) {
  .fence <- strrep("`", 3)
  cat(knitr::knit_child(text = c(
    paste0(.fence, "{r}"), "#| label: tbl-data-attrition",
    paste0("#| tbl-cap: ", encodeString(paste0("Analysis set `", ANALYSIS_SET, "`: exclusions, in order"), quote = "\"")),
    "knitr::kable(job_data$attrition)", .fence
  ), envir = environment(), quiet = TRUE), sep = "\n")
}
```


## Follow-up checks

Missing event status is counted separately and excluded from both event and
censored subsets. Suspicious rows have missing event status or at least one
missing, negative or zero interval. The local review prints at most
`MAX_REVIEW_ROWS` rows, in source order, and includes only selected fields.
Identifiers remain disabled until you explicitly select one. Check the output
before sharing a report that contains an identifier.

```{r}
#| label: qc
# The checks and tables are hvtiRutilities::followup_check(), the function the
# EDA report calls too, so the two reports cannot disagree. It checks every
# choice above before computing anything, and names every missing column at once.
fc <- hvtiRutilities::followup_check(d, event = EVENT, followup = FOLLOWUP,
                                     identifier = IDENTIFIER, max_rows = MAX_REVIEW_ROWS)
cohort_counts <- fc$cohort
followup_qc <- fc$intervals
followup_means <- fc$means
review_rows <- fc$review
```

```{r}
#| label: tbl-qc-cohort
#| tbl-cap: "Cohort: event, censored and missing event status"
knitr::kable(cohort_counts, row.names = FALSE)
```

```{r}
#| label: tbl-qc-intervals
#| tbl-cap: "Follow-up intervals: missing, negative and zero values"
knitr::kable(followup_qc, row.names = FALSE)
```

```{r}
#| label: qc-means
#| results: asis
# unnumbered: each child chunk below carries its own label and caption
.fence <- strrep("`", 3)
.seen <- new.env()
.child <- function(kind, key, caption, code) {
  label <- paste(kind, "qc", gsub("(^-+|-+$)", "", gsub("[^a-z0-9]+", "-", tolower(key))), sep = "-")
  if (!is.null(.seen[[label]])) stop("Two outputs would share the label ", label, ".", call. = FALSE)
  .seen[[label]] <- TRUE
  cat(knitr::knit_child(text = c(paste0(.fence, "{r}"), paste0("#| label: ", label),
                                 paste0("#| ", kind, "-cap: ", encodeString(caption, quote = "\"")), code, .fence),
                        envir = parent.frame(), quiet = TRUE), sep = "\n")
}
for (nm in names(followup_means)) {
  .means <- followup_means[[nm]]
  .child("tbl", paste("means", nm), paste0("Follow-up, ", nm, " (proc_means)"), "knitr::kable(.means, row.names = FALSE)")
  cat("\n")
}
```

```{r}
#| label: tbl-qc-review
#| tbl-cap: !expr 'paste0("Suspicious rows, first ", MAX_REVIEW_ROWS, " at most")'
knitr::kable(review_rows, row.names = FALSE)
```

The QC quartiles use R's default interpolated quantiles. `proc_means()` uses
SAS `QNTLDEF=5`, so its quartiles can differ. The migration report preserves
legacy sort orders, date arithmetic and includes for review; none establishes
an equivalent follow-up definition by itself.

## Operation years and the close date

```{r}
#| label: window
# unnumbered: computes the follow-up window; the table chunk below shows it
# hvtiPlotR::hv_followup_panels() checks every panel and event against the data,
# names every missing column at once, refuses a name used twice and the
# wrong-origin mistake, and works out the window the panels share: from
# 1 January of ORIGIN_YEAR, where hv_followup() draws its diagonal, to the last
# operation. The EDA report calls the same function, so its panels are these.
# A refusal does not stop the job: the tables above are the follow-up check and
# do not depend on the figure, so the refusal is reported where the figure would
# be, and the render carries on. It is also raised as a warning, so a console
# render says so too.
panel_counts <- list()
fp <- tryCatch(
  hvtiPlotR::hv_followup_panels(d, opyrs_col = OPYRS, origin_year = ORIGIN_YEAR,
                                panels = PANELS, events = EVENTS, close_date = CLOSE_DATE),
  error = function(e) e
)
figure_drawn <- !inherits(fp, "error")
if (figure_drawn) {
  close_date <- fp$meta$close_date
  # Name the edit point a reader can find in this file, not the function's argument.
  close_source <- if (is.null(CLOSE_DATE)) fp$meta$close_source else "set in CLOSE_DATE"
} else {
  warning("The goodness-of-follow-up figure was not drawn: ", conditionMessage(fp), call. = FALSE)
  close_date <- NA
  close_source <- paste("not drawn:", conditionMessage(fp))
}
```

```{r}
#| label: tbl-window
#| tbl-cap: "Operation years and the close date"
#| eval: !expr figure_drawn
knitr::kable(data.frame(
  quantity = c("patients", "operation year missing", "first operation", "last operation",
               "close date", "close date is"),
  value = c(fp$meta$n_obs, fp$meta$n_opyrs_missing, format(fp$meta$first_operation),
            format(fp$meta$study_end), format(close_date), close_source)
))
```

## Goodness of follow-up

A point below the diagonal is follow-up short of what the close date allows. A
dead patient's point is where the death was recorded, so it is expected below
the line; it is the **blue** points drifting below it that need an explanation.

```{r}
#| label: figures
#| results: asis
# unnumbered: each child chunk below carries its own label and caption
# This document sits in descriptive/ and its figures are written to graphs/
# through set_path(), so the image link is the figure's path relative to this
# document, worked out at render time, as dp-eda's is.
.fence <- strrep("`", 3)
.seen <- new.env()
.child <- function(label, caption, code) {
  label <- gsub("(^-+|-+$)", "", gsub("[^a-z0-9]+", "-", tolower(label)))
  if (!is.null(.seen[[label]])) stop("Two outputs would share the label ", label, ".", call. = FALSE)
  .seen[[label]] <- TRUE
  .opt <- paste0("#| ", sub("-.*$", "", label), "-cap: ", encodeString(caption, quote = "\""))
  cat(knitr::knit_child(text = c(paste0(.fence, "{r}"), paste0("#| label: ", label), .opt, code, .fence),
                        envir = parent.frame(), quiet = TRUE), sep = "\n")
  cat("\n")
}
if (!figure_drawn) {
  cat("\n::: {.callout-warning title=\"The figure was not drawn\"}\n",
      conditionMessage(fp), "\n\nThe tables above do not depend on it. Fix the choice the ",
      "message names, usually `OPYRS` or `ORIGIN_YEAR`, and render again.\n:::\n\n", sep = "")
}
figure_link <- function(file) {
  xfun::relative_path(file, if (exists(".in") && !is.null(.in)) dirname(.in) else getwd(), error = FALSE)
}
if (exists("COLOURS", inherits = FALSE)) stop("COLOURS is now COLORS: rename it in edit-study-choices ",
                                              "(study-choices in a job made before that rename).", call. = FALSE)
if (!is.null(COLORS) && !(is.character(COLORS) && !anyNA(COLORS) && length(COLORS) == 3L &&
                            setequal(names(COLORS), c("alive", "dead", "event")))) {
  stop("COLORS must be NULL or name exactly alive, dead and event.", call. = FALSE)
}
# Dead or Death is the event and Alive or No event the censored level, so a
# non-fatal event takes the house rule's next color, green.
status_colors <- function(levels) {
  if (is.null(COLORS)) return(scale_color_hv(event = c("Dead", "Death"), censored = c("Alive", "No event")))
  keys <- if (length(levels) == 2L) c("alive", "dead") else c("alive", "event", "dead")
  scale_color_manual(values = stats::setNames(COLORS[keys], levels))
}
plots <- if (figure_drawn) plot(fp, alpha = ALPHA) else list()
for (i in seq_len(if (figure_drawn) nrow(fp$data) else 0L)) {
  row <- fp$data[i, ]
  nm <- row$panel
  panel_counts[[nm]] <- list(drawn = row$n_drawn, excluded = row$n_excluded)
  cat("\n### ", row$title, "\n\n", row$n_drawn, " patients drawn; ", row$n_excluded,
      " missing a column this panel needs.\n\n", sep = "")
  if (row$type == "followup") {
    p <- plots[[nm]] +
      status_colors(c("Alive", "Dead")) +
      scale_shape_manual(values = c(Alive = 1, Dead = 4))
  } else {
    levels <- c("No event", row$title, "Death")
    p <- plots[[nm]] +
      status_colors(levels) +
      scale_shape_manual(values = stats::setNames(c(1, 2, 4), levels))
  }
  p <- p + labs(x = "Year of operation", y = "Follow-up (years)", color = NULL, shape = NULL) +
    theme_hv_manuscript()
  file <- set_path("graphs", paste0("dc-gfup-", nm, ".png"))
  ggplot2::ggsave(file, p, width = 7, height = 6, units = "in", dpi = 150)
  .link <- figure_link(file)
  .child(paste("fig gfup", nm), paste0("Follow-up against year of operation, ", row$title),
         "knitr::include_graphics(.link, error = FALSE)")
}
```

## Event-coding consistency

```{r}
#| label: checks
#| results: asis
# unnumbered: each child chunk below carries its own label and caption
unknown_checks <- setdiff(unlist(CHECKS, use.names = FALSE), names(d))
if (length(unknown_checks)) {
  stop("Unknown event-check column(s): ", paste(unknown_checks, collapse = ", "), call. = FALSE)
}
.fence <- strrep("`", 3)
.seen <- new.env()
.child <- function(kind, key, caption, code) {
  label <- paste(kind, "checks", gsub("(^-+|-+$)", "", gsub("[^a-z0-9]+", "-", tolower(key))), sep = "-")
  if (!is.null(.seen[[label]])) stop("Two outputs would share the label ", label, ".", call. = FALSE)
  .seen[[label]] <- TRUE
  cat(knitr::knit_child(text = c(paste0(.fence, "{r}"), paste0("#| label: ", label),
                                 paste0("#| ", kind, "-cap: ", encodeString(caption, quote = "\"")), code, .fence),
                        envir = parent.frame(), quiet = TRUE), sep = "\n")
}
for (cc in CHECKS) {
  .check <- as.data.frame(table(d[cc], useNA = "ifany"))
  .child("tbl", paste(cc, collapse = " by "), paste(cc, collapse = " by "), "knitr::kable(.check)")
  cat("\n")
}
```

```{r}
#| label: provenance
#| echo: false
#| results: asis
cat(hvtiRtemplates:::.embed_provenance(
  .in,
  data = .provenance_data,
  extra = list(
    subject = SUBJECT,
    type = TYPE,
    analysis = list(
      event = list(variable = EVENT, event = 1L, censored = 0L),
      followup = list(variable = FOLLOWUP)
    ),
    cohort = as.list(cohort_counts[1L, , drop = FALSE]),
    followup = list(origin_year = ORIGIN_YEAR,
                    close_date = if (figure_drawn) format(close_date) else NA_character_,
                    close_date_source = close_source, panels = panel_counts)
  )
))
```
