---
title: "General descriptive checks"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (descriptive/dc.general). -->

Replaces `descriptive/dc.general`.

A `dc-general` job is the first look at a built cohort, before any table or
model: what the data contain, how each categorical variable breaks down, how
each continuous variable is distributed, and which variables move together. It
uses base procedures only. The formatted manuscript table is a `dc-tables` job.

```{r}
#| label: setup
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages(library(hvtiRutilities))
if (utils::packageVersion("hvtiRdatabuild") < "0.2.1") {
  stop("This job needs hvtiRdatabuild >= 0.2.1 for read_analysis_set(); ",
       utils::packageVersion("hvtiRdatabuild"), " is installed.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it.\nTo render a partly-worked draft anyway, set ",
      "HVTI_TEMPLATE_DRAFT to 1, true or yes -- the report then carries a ",
      "draft banner. Any other value, including 0, leaves this guard strict."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_DRAFT")) %in% c("1", "true", "yes")) {
      warning(.msg, call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Rendered with `HVTI_TEMPLATE_DRAFT` set. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, call. = FALSE)
    }
  }
}
```

```{r}
#| label: set
ENDPOINT <- "cohort"
TYPE     <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint,
         "', type '", .name_type, "'), but declares ENDPOINT = \"", ENDPOINT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```

## Data

```{r}
#| label: data
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
# EDIT: the analysis set this job reads: declare its columns and excluded rows
# under analysis_sets: in _study.yml, then write it with
# hvtiRdatabuild::write_analysis_set(). Use NULL only when the whole built
# cohort is the intended population.
ANALYSIS_SET <- "eda"
if (is.null(ANALYSIS_SET)) {
  d <- read_built()
} else {
  d <- hvtiRdatabuild::read_analysis_set(ANALYSIS_SET)
  knitr::kable(attr(d, "attrition"), caption = paste0(
    "Analysis set `", ANALYSIS_SET, "`: exclusions, in order"
  ))
}
```

```{r}
#| label: spec
# EDIT: categorical variables, grouped. Each name is a section heading, in this
# order. Carry the /* Demography */ banners across from the SAS %macro freq
# tables list. Missing values are tabulated as their own level (missprint).
CATEGORICAL <- list(
  Demography = c("female")
)
# EDIT: continuous variables, grouped the same way (the SAS %macro cdfs var
# list).
CONTINUOUS <- list(
  Demography = c("age")
)
# EDIT: variables for the pairwise correlation sweep (SAS proc corr nosimple
# rank). Defaults to every continuous variable; character() skips the section.
CORR_VARS <- unlist(CONTINUOUS, use.names = FALSE)
# EDIT: identifier columns, kept out of every summary. The overall statistics
# would otherwise print their minimum and maximum, and each of those is one
# patient's identifier. Name every key column the data carry; character()
# only if there are none.
KEY_COLS <- "ccfid"
# EDIT: leave NULL. A column name (the SAS job used id ccfid) labels the extreme
# values with it, which puts patient identifiers in the report: a report
# rendered that way must not be circulated.
ID_COL <- NULL
```

```{r}
#| label: derive
PROBS <- c(0, 0.01, 0.05, 0.10, 0.25, 0.50, 0.75, 0.90, 0.95, 0.99, 1)
check_group_names <- function(x, setting) {
  if (!length(x)) return(invisible())
  nm <- names(x)
  if (is.null(nm) || any(nm == "") || anyDuplicated(nm)) {
    stop(setting, " must be a list with a unique, non-empty name for every group.", call. = FALSE)
  }
}
check_group_names(CATEGORICAL, "CATEGORICAL")
check_group_names(CONTINUOUS, "CONTINUOUS")
if (!is.null(ID_COL) && !(is.character(ID_COL) && length(ID_COL) == 1L)) {
  stop("ID_COL must be NULL or a single column name.", call. = FALSE)
}
cat_vars <- unlist(CATEGORICAL, use.names = FALSE)
con_vars <- unlist(CONTINUOUS, use.names = FALSE)
unknown <- setdiff(unique(c(cat_vars, con_vars, CORR_VARS, ID_COL, KEY_COLS)), names(d))
if (length(unknown)) {
  stop("Unknown column(s): ", paste(unknown, collapse = ", "), call. = FALSE)
}
numeric_vars <- unique(c(con_vars, CORR_VARS))
not_numeric <- numeric_vars[!vapply(numeric_vars, function(v) is.numeric(d[[v]]), logical(1))]
if (length(not_numeric)) {
  stop("Not numeric, so not summarised as continuous or correlated: ",
       paste(not_numeric, collapse = ", "), call. = FALSE)
}
key_overlap <- intersect(unique(c(cat_vars, con_vars, CORR_VARS)), KEY_COLS)
if (length(key_overlap)) {
  stop("Key column(s) cannot be summarised: ", paste(key_overlap, collapse = ", "), call. = FALSE)
}
overall_vars <- setdiff(names(d)[vapply(d, is.numeric, logical(1))], c(KEY_COLS, ID_COL))

freqs <- lapply(CATEGORICAL, function(vars) {
  stats::setNames(lapply(vars, function(v) {
    tab <- table(d[[v]], useNA = "ifany")
    is_na_level <- is.na(names(tab))
    n_int <- as.integer(tab)
    total_nonmissing <- sum(n_int[!is_na_level])
    percent <- rep(NA_real_, length(tab))
    if (total_nonmissing > 0L) {
      percent[!is_na_level] <- round(100 * n_int[!is_na_level] / total_nonmissing, 1)
    }
    data.frame(
      level = ifelse(is_na_level, "(missing)", names(tab)),
      n = n_int,
      percent = percent
    )
  }), vars)
})

cdfs <- lapply(CONTINUOUS, function(vars) {
  stats::setNames(lapply(vars, function(v) {
    x <- d[[v]]
    seen <- which(!is.na(x))
    lowest <- seen[order(x[seen])][seq_len(min(5L, length(seen)))]
    highest <- seen[order(-x[seen])][seq_len(min(5L, length(seen)))]
    extremes <- data.frame(
      end = rep(c("lowest", "highest"), c(length(lowest), length(highest))),
      value = x[c(lowest, highest)]
    )
    if (!is.null(ID_COL)) extremes[[ID_COL]] <- d[[ID_COL]][c(lowest, highest)]
    list(
      summary = data.frame(n = length(seen), nmiss = sum(is.na(x))),
      quantiles = data.frame(
        percent = 100 * PROBS,
        value = if (length(seen)) stats::quantile(x[seen], PROBS, type = 2, names = FALSE) else NA_real_
      ),
      extremes = extremes
    )
  }), vars)
})

corr_vars <- unique(CORR_VARS)
corrs <- data.frame(var1 = character(), var2 = character(), r = numeric(), n = integer())
if (length(corr_vars) >= 2L) {
  pairs <- utils::combn(corr_vars, 2L)
  pair_stats <- apply(pairs, 2L, function(p) {
    ok <- stats::complete.cases(d[[p[[1L]]]], d[[p[[2L]]]])
    r <- if (sum(ok) >= 2L) stats::cor(d[[p[[1L]]]][ok], d[[p[[2L]]]][ok]) else NA_real_
    c(r = r, n = sum(ok))
  })
  corrs <- data.frame(
    var1 = pairs[1L, ], var2 = pairs[2L, ],
    r = pair_stats["r", ], n = as.integer(pair_stats["n", ])
  )
  corrs <- corrs[order(-abs(corrs$r)), , drop = FALSE]
  rownames(corrs) <- NULL
}
```

## Other investigations

<!-- EDIT: study-specific checks go here, above the standard sections (the SAS
job's O T H E R   I N V E S T I G A T I O N S banner). Delete this marker when
there are none. -->

## Overall statistics

```{r}
#| label: overall
proc_contents(d)
if (length(overall_vars)) {
  proc_means(d, vars = overall_vars, stats = c("n", "nmiss", "mean", "std", "min", "max", "sum"))
}
```

## Contingency tables for categorical variables

```{r}
#| label: freqs
#| results: asis
.kable_na <- options(knitr.kable.NA = "")
for (group in names(freqs)) {
  cat("\n### ", group, "\n\n", sep = "")
  for (v in names(freqs[[group]])) {
    cat(knitr::kable(freqs[[group]][[v]], caption = v), sep = "\n")
    cat("\n")
  }
}
options(.kable_na)
```

## Cumulative distributions for continuous variables

```{r}
#| label: cdfs
#| results: asis
for (group in names(cdfs)) {
  cat("\n### ", group, "\n\n", sep = "")
  for (v in names(cdfs[[group]])) {
    cdf <- cdfs[[group]][[v]]
    cat("\n#### ", v, "\n\n", sep = "")
    cat(knitr::kable(cdf$summary), sep = "\n")
    cat("\n")
    cat(knitr::kable(cdf$quantiles, caption = "Quantiles (SAS QNTLDEF=5)"), sep = "\n")
    cat("\n")
    cat(knitr::kable(cdf$extremes, caption = "Five lowest and five highest values"), sep = "\n")
    cat("\n")
  }
}
```

## Pair-wise correlations

```{r}
#| label: corrs
#| results: asis
if (nrow(corrs)) {
  cat(knitr::kable(corrs, digits = 3, row.names = FALSE,
                   caption = "Pearson, pairwise complete, strongest first"), sep = "\n")
} else {
  cat("No correlation sweep: `CORR_VARS` names fewer than two variables.\n")
}
```

`proc_means()` covers every numeric column other than `KEY_COLS` and `ID_COL`.
Percentages in the contingency tables are over non-missing rows, as SAS
`MISSPRINT` computes them; the `(missing)` row shows its count with a blank
percent. Quantiles are SAS `QNTLDEF=5`, the `proc univariate` default. The
highest extremes are listed largest first (SAS lists them ascending). The SAS
job printed `ccfid` beside every extreme value so an author could look the
patient up; this job prints the values alone unless `ID_COL` is set in the
spec chunk.
