---
title: "General descriptive checks"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (descriptive/dc.general). -->

Replaces `descriptive/dc.general`.

A `dc-general` job is the first look at a built cohort, before any table or
model: what the data contain, how each categorical variable breaks down, how
each continuous variable is distributed, and which variables move together. It
uses base procedures only. The formatted manuscript table is a `dc-tables` job.

```{r}
#| label: setup
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRutilities::study_root(if (is.null(.in)) getwd() else dirname(.in))
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages(library(hvtiRutilities))
if (utils::packageVersion("hvtiRdatabuild") < "0.2.1") {
  stop("This job needs hvtiRdatabuild >= 0.2.1 for read_analysis_set(); ",
       utils::packageVersion("hvtiRdatabuild"), " is installed.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
# The markers in this file name work a study author still has to do, and a job
# that still contains one has not been finished. This chunk is what makes that
# TRUE rather than merely stated: an unedited job would otherwise render green
# over placeholder columns.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

```{r}
#| label: set
SUBJECT <- "cohort"
TYPE    <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject,
         "', type '", .name_type, "'), but declares SUBJECT = \"", SUBJECT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```

## Study choices

Edit these values for this study before rendering.

```{r}
#| label: study-choices
# EDIT: the built dataset this job reads.
DATASET <- "study"

# EDIT: the analysis set, or NULL for a whole additional dataset.
ANALYSIS_SET <- "eda"

# EDIT: grouped categorical variables.
CATEGORICAL <- list(
  Demography = c("female")
)

# EDIT: grouped continuous variables.
CONTINUOUS <- list(
  Demography = c("age")
)

# EDIT: variables in the correlation sweep.
CORR_VARS <- unlist(CONTINUOUS, use.names = FALSE)

# EDIT: identifier columns excluded from summaries.
KEY_COLS <- "ccfid"

# EDIT: leave NULL to keep identifiers out of reports.
ID_COL <- NULL
```

## Data

<!-- How this job finds its data. Read this before touching the chunk below.

     A study can offer a job data three ways, and `_study.yml` says which exist:

     1. The study dataset: the file named by `built:`. `DATASET <- "study"`
        with `ANALYSIS_SET <- NULL`.
     2. An additional dataset: any file declared under `additional_datasets:`,
        for example a column subset written for R. Set `DATASET` to its
        name there and `ANALYSIS_SET <- NULL`. It is read as it stands: no
        rows are excluded.
     3. An analysis set: declared under `analysis_sets:` with the columns kept
        and the rows excluded, then written once with
        `hvtiRdatabuild::write_analysis_set()`. Set `ANALYSIS_SET` to its name
        and leave `DATASET` as "study", because a set is always cut from the
        study dataset. The report shows its exclusions in order.

     A name under `additional_datasets:` is not an analysis set, so putting it
     in `ANALYSIS_SET` stops with "No analysis set". Whichever route is taken,
     the report prints the dataset, file and dimensions it read, so a reader
     never has to ask which data a table describes. -->

```{r}
#| label: data
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
# The built dataset "study" is the file named by built:
# in _study.yml; any other name must be declared under additional_datasets:,
# such as a column subset written for R. Analysis sets are derived from
# "study" only, so another dataset needs ANALYSIS_SET <- NULL.
# Declare the analysis set's columns and excluded rows
# under analysis_sets: in _study.yml, then write it with
# hvtiRdatabuild::write_analysis_set(). Use NULL only when the whole built
# cohort is the intended population.
if (is.null(ANALYSIS_SET)) {
  if (!is.character(DATASET) || length(DATASET) != 1L || is.na(DATASET) || !nzchar(DATASET)) {
    stop("DATASET must be set to one dataset registered in _study.yml, such as \"study\".",
         call. = FALSE)
  }
  d <- read_built(dataset = DATASET)
} else if (!identical(DATASET, "study")) {
  stop("An analysis set is written from the study dataset, not `", DATASET,
       "`. Set ANALYSIS_SET <- NULL to read `", DATASET, "` whole.", call. = FALSE)
} else {
  d <- hvtiRdatabuild::read_analysis_set(ANALYSIS_SET)
  knitr::kable(attr(d, "attrition"), caption = paste0(
    "Analysis set `", ANALYSIS_SET, "`: exclusions, in order"
  ))
}
cat("Data read: ", if (is.null(ANALYSIS_SET)) {
  paste0("dataset `", DATASET, "` (",
         basename(hvtiRutilities::built_path(dataset = DATASET)), ")")
} else {
  paste0("analysis set `", ANALYSIS_SET, "` of the study dataset")
}, ", ", nrow(d), " rows, ", ncol(d), " columns.\n", sep = "")
```

```{r}
#| label: spec
# categorical variables, grouped. Each name is a section heading, in this
# order. Carry the /* Demography */ banners across from the SAS %macro freq
# tables list. Missing values are tabulated as their own level (missprint).
# continuous variables, grouped the same way (the SAS %macro cdfs var
# list).
# variables for the pairwise correlation sweep (SAS proc corr nosimple
# rank). Defaults to every continuous variable; character() skips the section.
# Identifier columns are kept out of every summary. The overall statistics
# would otherwise print their minimum and maximum, and each of those is one
# patient's identifier. Name every key column the data carry; character()
# only if there are none.
# A column name (the SAS job used id ccfid) labels the extreme
# values with it, which puts patient identifiers in the report: a report
# rendered that way must not be circulated.
```

```{r}
#| label: derive
PROBS <- c(0, 0.01, 0.05, 0.10, 0.25, 0.50, 0.75, 0.90, 0.95, 0.99, 1)
check_group_names <- function(x, setting) {
  if (!length(x)) return(invisible())
  nm <- names(x)
  if (is.null(nm) || any(nm == "") || anyDuplicated(nm)) {
    stop(setting, " must be a list with a unique, non-empty name for every group.", call. = FALSE)
  }
}
check_group_names(CATEGORICAL, "CATEGORICAL")
check_group_names(CONTINUOUS, "CONTINUOUS")
if (!is.null(ID_COL) && !(is.character(ID_COL) && length(ID_COL) == 1L)) {
  stop("ID_COL must be NULL or a single column name.", call. = FALSE)
}
cat_vars <- unlist(CATEGORICAL, use.names = FALSE)
con_vars <- unlist(CONTINUOUS, use.names = FALSE)
unknown <- setdiff(unique(c(cat_vars, con_vars, CORR_VARS, ID_COL, KEY_COLS)), names(d))
if (length(unknown)) {
  stop("Unknown column(s): ", paste(unknown, collapse = ", "), call. = FALSE)
}
numeric_vars <- unique(c(con_vars, CORR_VARS))
not_numeric <- numeric_vars[!vapply(numeric_vars, function(v) is.numeric(d[[v]]), logical(1))]
if (length(not_numeric)) {
  stop("Not numeric, so not summarised as continuous or correlated: ",
       paste(not_numeric, collapse = ", "), call. = FALSE)
}
key_overlap <- intersect(unique(c(cat_vars, con_vars, CORR_VARS)), KEY_COLS)
if (length(key_overlap)) {
  stop("Key column(s) cannot be summarised: ", paste(key_overlap, collapse = ", "), call. = FALSE)
}
overall_vars <- setdiff(names(d)[vapply(d, is.numeric, logical(1))], c(KEY_COLS, ID_COL))

freqs <- lapply(CATEGORICAL, function(vars) {
  stats::setNames(lapply(vars, function(v) {
    tab <- table(d[[v]], useNA = "ifany")
    is_na_level <- is.na(names(tab))
    n_int <- as.integer(tab)
    total_nonmissing <- sum(n_int[!is_na_level])
    percent <- rep(NA_real_, length(tab))
    if (total_nonmissing > 0L) {
      percent[!is_na_level] <- round(100 * n_int[!is_na_level] / total_nonmissing, 1)
    }
    data.frame(
      level = ifelse(is_na_level, "(missing)", names(tab)),
      n = n_int,
      percent = percent
    )
  }), vars)
})

cdfs <- lapply(CONTINUOUS, function(vars) {
  stats::setNames(lapply(vars, function(v) {
    x <- d[[v]]
    seen <- which(!is.na(x))
    lowest <- seen[order(x[seen])][seq_len(min(5L, length(seen)))]
    highest <- seen[order(-x[seen])][seq_len(min(5L, length(seen)))]
    extremes <- data.frame(
      end = rep(c("lowest", "highest"), c(length(lowest), length(highest))),
      value = x[c(lowest, highest)]
    )
    if (!is.null(ID_COL)) extremes[[ID_COL]] <- d[[ID_COL]][c(lowest, highest)]
    list(
      summary = data.frame(n = length(seen), nmiss = sum(is.na(x))),
      quantiles = data.frame(
        percent = 100 * PROBS,
        value = if (length(seen)) stats::quantile(x[seen], PROBS, type = 2, names = FALSE) else NA_real_
      ),
      extremes = extremes
    )
  }), vars)
})

corr_vars <- unique(CORR_VARS)
corrs <- data.frame(var1 = character(), var2 = character(), r = numeric(), n = integer())
if (length(corr_vars) >= 2L) {
  pairs <- utils::combn(corr_vars, 2L)
  pair_stats <- apply(pairs, 2L, function(p) {
    ok <- stats::complete.cases(d[[p[[1L]]]], d[[p[[2L]]]])
    r <- if (sum(ok) >= 2L) stats::cor(d[[p[[1L]]]][ok], d[[p[[2L]]]][ok]) else NA_real_
    c(r = r, n = sum(ok))
  })
  corrs <- data.frame(
    var1 = pairs[1L, ], var2 = pairs[2L, ],
    r = pair_stats["r", ], n = as.integer(pair_stats["n", ])
  )
  corrs <- corrs[order(-abs(corrs$r)), , drop = FALSE]
  rownames(corrs) <- NULL
}
```

## Other investigations

<!-- EDIT: study-specific checks go here, above the standard sections (the SAS
job's O T H E R   I N V E S T I G A T I O N S banner). Delete this marker when
there are none. -->

## Overall statistics

```{r}
#| label: overall
proc_contents(d)
if (length(overall_vars)) {
  proc_means(d, vars = overall_vars, stats = c("n", "nmiss", "mean", "std", "min", "max", "sum"))
}
```

## Contingency tables for categorical variables

```{r}
#| label: freqs
#| results: asis
.kable_na <- options(knitr.kable.NA = "")
for (group in names(freqs)) {
  cat("\n### ", group, "\n\n", sep = "")
  for (v in names(freqs[[group]])) {
    cat(knitr::kable(freqs[[group]][[v]], caption = v), sep = "\n")
    cat("\n")
  }
}
options(.kable_na)
```

## Cumulative distributions for continuous variables

```{r}
#| label: cdfs
#| results: asis
for (group in names(cdfs)) {
  cat("\n### ", group, "\n\n", sep = "")
  for (v in names(cdfs[[group]])) {
    cdf <- cdfs[[group]][[v]]
    cat("\n#### ", v, "\n\n", sep = "")
    cat(knitr::kable(cdf$summary), sep = "\n")
    cat("\n")
    cat(knitr::kable(cdf$quantiles, caption = "Quantiles (SAS QNTLDEF=5)"), sep = "\n")
    cat("\n")
    cat(knitr::kable(cdf$extremes, caption = "Five lowest and five highest values"), sep = "\n")
    cat("\n")
  }
}
```

## Pair-wise correlations

```{r}
#| label: corrs
#| results: asis
if (nrow(corrs)) {
  cat(knitr::kable(corrs, digits = 3, row.names = FALSE,
                   caption = "Pearson, pairwise complete, strongest first"), sep = "\n")
} else {
  cat("No correlation sweep: `CORR_VARS` names fewer than two variables.\n")
}
```

`proc_means()` covers every numeric column other than `KEY_COLS` and `ID_COL`.
Percentages in the contingency tables are over non-missing rows, as SAS
`MISSPRINT` computes them; the `(missing)` row shows its count with a blank
percent. Quantiles are SAS `QNTLDEF=5`, the `proc univariate` default. The
highest extremes are listed largest first (SAS lists them ascending). The SAS
job printed `ccfid` beside every extreme value so an author could look the
patient up; this job prints the values alone unless `ID_COL` is set in the
spec chunk.

```{r}
#| label: provenance
#| include: false
if (is.null(.in) || !is.character(.in) || length(.in) != 1L || is.na(.in) || !nzchar(.in)) {
  stop("Cannot record provenance without the current render input.", call. = FALSE)
}
.job_stem <- tools::file_path_sans_ext(basename(.in))
.output <- file.path(dirname(.in), paste0(.job_stem, ".html"))
hvtiRutilities::record_provenance(
  .output,
  extra = list(
    subject = SUBJECT,
    type = TYPE
  )
)
```
