---
title: "Goodness of follow-up"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (descriptive/dc.gfup). -->

Replaces `descriptive/dc.gfup`.

A `dc-gfup` job asks whether follow-up is complete enough to trust the
time-related analyses: how long patients were followed for death and for
non-fatal events, against how long they could have been followed by the close
date. `hvtiPlotR::hv_followup()` supplies the companion figure; this job is the
tables. A scaffolded `dp-gfup` job is not part of this template wave.

```{r}
#| label: setup
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages(library(hvtiRutilities))
if (utils::packageVersion("hvtiRdatabuild") < "0.2.1") {
  stop("This job needs hvtiRdatabuild >= 0.2.1 for read_analysis_set(); ",
       utils::packageVersion("hvtiRdatabuild"), " is installed.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it.\nTo render a partly-worked draft anyway, set ",
      "HVTI_TEMPLATE_DRAFT to 1, true or yes -- the report then carries a ",
      "draft banner. Any other value, including 0, leaves this guard strict."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_DRAFT")) %in% c("1", "true", "yes")) {
      warning(.msg, call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Rendered with `HVTI_TEMPLATE_DRAFT` set. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, call. = FALSE)
    }
  }
}
```

```{r}
#| label: set
ENDPOINT <- "cohort"
TYPE     <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint,
         "', type '", .name_type, "'), but declares ENDPOINT = \"", ENDPOINT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(.root, kind, paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```

## Data

```{r}
#| label: data
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
# EDIT: the analysis set this job reads: declare its columns and excluded rows
# under analysis_sets: in _study.yml, then write it with
# hvtiRdatabuild::write_analysis_set(). Use NULL only when the whole built
# cohort is the intended population.
ANALYSIS_SET <- "eda"
if (is.null(ANALYSIS_SET)) {
  d <- read_built()
} else {
  d <- hvtiRdatabuild::read_analysis_set(ANALYSIS_SET)
  knitr::kable(attr(d, "attrition"), caption = paste0(
    "Analysis set `", ANALYSIS_SET, "`: exclusions, in order"
  ))
}
```

```{r}
#| label: spec
# EDIT: the close date of follow-up (SAS: current=mdy(mm,dd,yyyy)). Potential
# follow-up is measured to this date, so a wrong date makes follow-up look
# better or worse than it is.
CLOSE_DATE <- as.Date("2026-01-01")
# EDIT: column names in the built data. The SAS template used dt_surg, iv_dead,
# dead and iv_fup; change any that differ.
COLS <- c(dt_surg = "dt_surg", iv_dead = "iv_dead", dead = "dead", iv_fup = "iv_fup")
# EDIT: event-coding consistency checks, each a set of columns cross-tabulated
# with missing values shown (SAS: proc freq; table a*b*c / list missprint).
# list() for none.
CHECKS <- list(c("dead", "hdeath"))
```

```{r}
#| label: derive
needed <- unique(c(unname(COLS), unlist(CHECKS, use.names = FALSE)))
unknown <- setdiff(needed, names(d))
if (length(unknown)) {
  stop("Unknown follow-up column(s): ", paste(unknown, collapse = ", "), call. = FALSE)
}
g <- data.frame(
  dt_surg = as.Date(d[[COLS[["dt_surg"]]]]),
  iv_dead = d[[COLS[["iv_dead"]]]],
  dead = d[[COLS[["dead"]]]],
  iv_fup = d[[COLS[["iv_fup"]]]]
)
yrs <- function(days) as.numeric(days) / 365.2425
g$maxfup <- ifelse(g$dead == 1, g$iv_dead, yrs(CLOSE_DATE - g$dt_surg))
g$potfup <- ifelse(
  g$dead == 1,
  NA_real_,
  yrs(CLOSE_DATE - (g$dt_surg + g$iv_fup * 365.2425))
)
g$pctfup <- g$iv_fup / g$maxfup
g$pctvit <- g$iv_fup / g$iv_dead
survivors <- g[!is.na(g$dead) & g$dead == 0, , drop = FALSE]
```

## Data checks

```{r}
#| label: checks
for (cc in CHECKS) print(as.data.frame(table(d[cc], useNA = "ifany")))
```

## Follow-up for survival

```{r}
#| label: survival
st <- c("n", "nmiss", "mean", "std", "min", "p15", "median", "p85", "max")
proc_means(g, vars = "iv_dead", class = "dead", stats = st)
proc_means(survivors, vars = c("iv_dead", "maxfup", "pctfup"), stats = st)
```

## Follow-up for non-fatal events

```{r}
#| label: events
proc_means(g, vars = "iv_fup", class = "dead", stats = st)
proc_means(
  survivors,
  vars = c("iv_fup", "potfup", "pctfup", "pctvit"),
  stats = st
)
```

The SAS job printed every patient, sorted by `dead` and operation date, so an
author could find the incompletely followed ones by `ccfid`. This job prints
none: the percentiles of `pctfup` among survivors answer the same question
without an identifier. Quantiles here are SAS `QNTLDEF=5`; the SAS template's
`proc univariate def=4` gives slightly different tail percentiles.
