---
title: "Goodness of follow-up"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (descriptive/dc.gfup). -->

Replaces `descriptive/dc.gfup`.

A `dc-gfup` job checks the recorded follow-up intervals before time-related
analyses. We summarize the full cohort and the event and censored subsets,
then review missing, negative and zero intervals. Follow-up units must be
years. Derive any close-date or potential-follow-up measures in the data build
and register them before selecting them here. These tables describe recorded
follow-up; they do not establish completeness against a close date.

```{r}
#| label: setup
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRutilities::study_root(if (is.null(.in)) getwd() else dirname(.in))
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages(library(hvtiRutilities))
if (utils::packageVersion("hvtiRdatabuild") < "0.2.1") {
  stop("This job needs hvtiRdatabuild >= 0.2.1 for read_analysis_set(); ",
       utils::packageVersion("hvtiRdatabuild"), " is installed.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

```{r}
#| label: set
ENDPOINT <- "cohort"
TYPE     <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint,
         "', type '", .name_type, "'), but declares ENDPOINT = \"", ENDPOINT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```

## Data

<!-- How this job finds its data. Read this before touching the chunk below.

     A study can offer a job data three ways, and `_study.yml` says which exist:

     1. The study dataset: the file named by `built:`. `DATASET <- "study"`
        with `ANALYSIS_SET <- NULL`.
     2. An additional dataset: any file declared under `additional_datasets:`,
        for example a column subset written for R. Set `DATASET` to its
        name there and `ANALYSIS_SET <- NULL`. It is read as it stands: no
        rows are excluded.
     3. An analysis set: declared under `analysis_sets:` with the columns kept
        and the rows excluded, then written once with
        `hvtiRdatabuild::write_analysis_set()`. Set `ANALYSIS_SET` to its name
        and leave `DATASET` as "study", because a set is always cut from the
        study dataset. The report shows its exclusions in order.

     A name under `additional_datasets:` is not an analysis set, so putting it
     in `ANALYSIS_SET` stops with "No analysis set". Whichever route is taken,
     the report prints the dataset, file and dimensions it read, so a reader
     never has to ask which data a table describes. -->

```{r}
#| label: data
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
# EDIT: the built dataset this job reads. "study" is the file named by built:
# in _study.yml; any other name must be declared under additional_datasets:,
# such as a column subset written for R. Analysis sets are derived from
# "study" only, so another dataset needs ANALYSIS_SET <- NULL.
# EDIT: the analysis set this job reads: declare its columns and excluded rows
# under analysis_sets: in _study.yml, then write it with
# hvtiRdatabuild::write_analysis_set(). Use NULL only when the whole built
# cohort is the intended population.
# MIGRATE-BEGIN: dc-gfup-data
DATASET <- "study"
ANALYSIS_SET <- "eda"
# MIGRATE-END: dc-gfup-data
if (is.null(ANALYSIS_SET)) {
  d <- read_built(dataset = DATASET)
} else if (!identical(DATASET, "study")) {
  stop("An analysis set is written from the study dataset, not `", DATASET,
       "`. Set ANALYSIS_SET <- NULL to read `", DATASET, "` whole.", call. = FALSE)
} else {
  d <- hvtiRdatabuild::read_analysis_set(ANALYSIS_SET)
  knitr::kable(attr(d, "attrition"), caption = paste0(
    "Analysis set `", ANALYSIS_SET, "`: exclusions, in order"
  ))
}
cat("Data read: ", if (is.null(ANALYSIS_SET)) {
  paste0("dataset `", DATASET, "` (",
         basename(hvtiRutilities::built_path(dataset = DATASET)), ")")
} else {
  paste0("analysis set `", ANALYSIS_SET, "` of the study dataset")
}, ", ", nrow(d), " rows, ", ncol(d), " columns.\n", sep = "")
```

```{r}
#| label: spec
# MIGRATE-BEGIN: dc-gfup-config
EVENT <- "dead"       # EDIT: binary event indicator, 1=event and 0=censored
FOLLOWUP <- "iv_dead" # EDIT: registered follow-up interval(s), in years
IDENTIFIER <- NULL    # EDIT: opt in only for local review
MAX_REVIEW_ROWS <- 25L
# MIGRATE-END: dc-gfup-config
# EDIT: optional event-coding cross-tabs, each a vector of column names.
# Missing values are shown. Use list() when no additional checks are needed.
CHECKS <- list()
```

## Follow-up checks

Missing event status is counted separately and excluded from both event and
censored subsets. Suspicious rows have missing event status or at least one
missing, negative or zero interval. The local review prints at most
`MAX_REVIEW_ROWS` rows, in source order, and includes only selected fields.
Identifiers remain disabled until you explicitly select one. Check the output
before sharing a report that contains an identifier.

```{r}
#| label: qc
field_names <- function(x) is.character(x) && length(x) > 0L && !anyNA(x) && all(nzchar(x)) && !anyDuplicated(x)
if (!field_names(EVENT) || length(EVENT) != 1L) stop("EVENT must name one column.", call. = FALSE)
if (!field_names(FOLLOWUP)) stop("FOLLOWUP must name one or more distinct columns.", call. = FALSE)
if (!is.null(IDENTIFIER) && (!field_names(IDENTIFIER) || length(IDENTIFIER) != 1L)) {
  stop("IDENTIFIER must be NULL or one column name.", call. = FALSE)
}
if (!is.numeric(MAX_REVIEW_ROWS) || length(MAX_REVIEW_ROWS) != 1L || is.na(MAX_REVIEW_ROWS) ||
      !is.finite(MAX_REVIEW_ROWS) || MAX_REVIEW_ROWS < 1 || MAX_REVIEW_ROWS != floor(MAX_REVIEW_ROWS)) {
  stop("MAX_REVIEW_ROWS must be one positive integer.", call. = FALSE)
}
needed <- unique(c(EVENT, FOLLOWUP, IDENTIFIER))
unknown <- setdiff(needed, names(d))
if (length(unknown)) {
  stop("Unknown follow-up column(s): ", paste(unknown, collapse = ", "), call. = FALSE)
}
event <- d[[EVENT]]
if ((!is.numeric(event) && !is.logical(event)) || any(!is.na(event) & !event %in% c(0, 1))) {
  stop("EVENT must be binary (0=censored, 1=event; missing is allowed).", call. = FALSE)
}
if (!all(vapply(d[FOLLOWUP], is.numeric, logical(1L)))) stop("FOLLOWUP columns must be numeric.", call. = FALSE)
cohort_counts <- data.frame(full = nrow(d), event = sum(event == 1, na.rm = TRUE),
                            censored = sum(event == 0, na.rm = TRUE), missing_event = sum(is.na(event)))
print(cohort_counts, row.names = FALSE)
followup_qc <- do.call(rbind, lapply(FOLLOWUP, function(variable) {
  x <- d[[variable]]
  observed <- x[!is.na(x)]
  q <- if (length(observed)) stats::quantile(observed, c(0, .25, .5, .75, 1), names = FALSE) else rep(NA_real_, 5L)
  data.frame(interval = variable, missing = sum(is.na(x)), negative = sum(x < 0, na.rm = TRUE),
             zero = sum(x == 0, na.rm = TRUE), min = q[[1L]], q1 = q[[2L]], median = q[[3L]], q3 = q[[4L]],
             mean = if (length(observed)) mean(observed) else NA_real_, sd = stats::sd(observed), max = q[[5L]])
}))
print(followup_qc, row.names = FALSE)
subsets <- list(full = d, event = d[!is.na(event) & event == 1, , drop = FALSE],
                censored = d[!is.na(event) & event == 0, , drop = FALSE])
st <- c("n", "nmiss", "mean", "std", "min", "p25", "median", "p75", "max")
followup_means <- lapply(subsets, function(data) hvtiRutilities::proc_means(data, vars = FOLLOWUP, stats = st))
print(followup_means)
suspicious <- is.na(event) | Reduce(`|`, lapply(d[FOLLOWUP], function(x) is.na(x) | x <= 0))
review_rows <- d[head(which(suspicious), MAX_REVIEW_ROWS), needed, drop = FALSE]
rownames(review_rows) <- NULL
print(review_rows, row.names = FALSE)
```

The QC quartiles use R's default interpolated quantiles. `proc_means()` uses
SAS `QNTLDEF=5`, so its quartiles can differ. The migration report preserves
legacy sort orders, date arithmetic and includes for review; none establishes
an equivalent follow-up definition by itself.

## Event-coding consistency

```{r}
#| label: checks
unknown_checks <- setdiff(unlist(CHECKS, use.names = FALSE), names(d))
if (length(unknown_checks)) {
  stop("Unknown event-check column(s): ", paste(unknown_checks, collapse = ", "), call. = FALSE)
}
for (cc in CHECKS) print(as.data.frame(table(d[cc], useNA = "ifany")))
```
