---
title: "Descriptive tables"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (descriptive/dc.tables.*) and say whether it
     was the overall tables, the by-group tables, or the correlation variant. -->

Replaces `descriptive/dc.tables`: say which variant.

A `dc-tables` job is the formatted descriptive table: every variable the study
reports, grouped under the section headings the SAS `%desc_tab` lists carried as
comments, categorical as `n (%)` and continuous as mean±SD and median (15th,
85th percentile). `%desc_tab` wrote the categorical and continuous tables as two
RTF files; `hv_tbl_summary()` writes one table with both. It describes; with
`BY` set it also tests, and a p-value here is a prompt for an analysis, not a
result.

```{r}
#| label: setup
# This file runs either from the directory it was scaffolded into or from the
# project root. Resolve which, so that no path in this document needs editing.
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(hvtiRutilities)
  library(hvtiRtables)
  library(hvtiPlotR)
  library(ggplot2)
})
if (utils::packageVersion("hvtiRdatabuild") < "0.2.1") {
  stop("This job needs hvtiRdatabuild >= 0.2.1 for read_analysis_set(); ",
       utils::packageVersion("hvtiRdatabuild"), " is installed.", call. = FALSE)
}
if (utils::packageVersion("hvtiRtables") < "1.0.1") {
  stop("This job needs hvtiRtables >= 1.0.1 for hv_correlation_table(); ",
       utils::packageVersion("hvtiRtables"), " is installed.", call. = FALSE)
}
if (utils::packageVersion("hvtiPlotR") < "2.7.14") {
  stop("This job needs hvtiPlotR >= 2.7.14 for hv_correlation_matrix(); ",
       utils::packageVersion("hvtiPlotR"), " is installed.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
# The markers in this file name work a study author still has to do, and a job
# that still contains one has not been finished. This chunk is what makes that
# TRUE rather than merely stated: an unedited job would otherwise render green
# over placeholder columns.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it.\nTo render a partly-worked draft anyway, set ",
      "HVTI_TEMPLATE_DRAFT to 1, true or yes -- the report then carries a ",
      "draft banner. Any other value, including 0, leaves this guard strict."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_DRAFT")) %in% c("1", "true", "yes")) {
      warning(.msg, call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Rendered with `HVTI_TEMPLATE_DRAFT` set. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by add_job() from the endpoint/type it
     was called with, which are also the filename's first two fields. Every
     artifact path in this document is computed from them. -->

```{r}
#| label: set
ENDPOINT <- "cohort"
TYPE     <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint, "', type '",
         .name_type, "'), but declares ENDPOINT = \"", ENDPOINT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

set_path <- function(kind, file) {
  dir <- file.path(hvtiRutilities::study_dir(kind, .root),
                   paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(dir)) dir.create(dir, recursive = TRUE)
  file.path(dir, file)
}
```

## Data

<!-- How this job finds its data. Read this before touching the chunk below.

     A study can offer a job data three ways, and `_study.yml` says which exist:

     1. The study dataset: the file named by `built:`. `DATASET <- "study"`
        with `ANALYSIS_SET <- NULL`.
     2. An additional dataset: any file declared under `additional_datasets:`,
        for example a column subset written for R. Set `DATASET` to its
        name there and `ANALYSIS_SET <- NULL`. It is read as it stands: no
        rows are excluded.
     3. An analysis set: declared under `analysis_sets:` with the columns kept
        and the rows excluded, then written once with
        `hvtiRdatabuild::write_analysis_set()`. Set `ANALYSIS_SET` to its name
        and leave `DATASET` as "study", because a set is always cut from the
        study dataset. The report shows its exclusions in order.

     A name under `additional_datasets:` is not an analysis set, so putting it
     in `ANALYSIS_SET` stops with "No analysis set". Whichever route is taken,
     the report prints the dataset, file and dimensions it read, so a reader
     never has to ask which data a table describes. -->

```{r}
#| label: data
# The checksum of every dataset in manifest.yaml is checked before anything is
# read, so a result can name the data that produced it. It stops on a mismatch.
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
# EDIT: the built dataset this job reads. "study" is the file named by built:
# in _study.yml; any other name must be declared under additional_datasets:,
# such as a column subset written for R. Analysis sets are derived from
# "study" only, so another dataset needs ANALYSIS_SET <- NULL.
DATASET <- "study"
# EDIT: the analysis set this job reads: the columns and excluded rows declared
# under analysis_sets: in _study.yml and written once with
# hvtiRdatabuild::write_analysis_set(). A stale set stops and names the call
# that refreshes it. Use NULL to describe the whole built cohort instead.
ANALYSIS_SET <- "eda"
if (is.null(ANALYSIS_SET)) {
  d <- read_built(dataset = DATASET)
} else if (!identical(DATASET, "study")) {
  stop("An analysis set is written from the study dataset, not `", DATASET,
       "`. Set ANALYSIS_SET <- NULL to read `", DATASET, "` whole.", call. = FALSE)
} else {
  d <- hvtiRdatabuild::read_analysis_set(ANALYSIS_SET)
  knitr::kable(attr(d, "attrition"), caption = paste0(
    "Analysis set `", ANALYSIS_SET, "`: exclusions, in order"
  ))
}
cat("Data read: ", if (is.null(ANALYSIS_SET)) {
  paste0("dataset `", DATASET, "` (",
         basename(hvtiRutilities::built_path(dataset = DATASET)), ")")
} else {
  paste0("analysis set `", ANALYSIS_SET, "` of the study dataset")
}, ", ", nrow(d), " rows, ", ncol(d), " columns.\n", sep = "")
```

```{r}
#| label: spec
# EDIT: the variables to report, grouped. Each name is a section heading and
# becomes a row group in the table, in this order. These are the /* Demography */
# style banners from the SAS varlist; carry them across unchanged so the R table
# reads like the SAS one. Commented-out SAS blocks stay out.
GROUPS <- list(
  Demography = c("age", "female"),
  Symptoms   = c("nyha_pr")
)
# EDIT: NULL for the overall table. A column name reproduces the `%macro skip`
# by-group sections (by=, byvalue=): one column per level and a p-value column.
BY <- NULL
# EDIT: force a bucket where the data guess wrong, e.g. an ordinal score you
# want summarised as continuous: list(continuous = c("nyha_pr")). list() for none.
OVERRIDES <- list()
# EDIT: NULL skips the correlation variant (dc.tables.ods_<topic>.sas). To run
# it, name the numeric columns, the anchor to correlate them against (NULL for
# every pair), and an optional stratum:
#   list(vars = c("glu_pr", "creat_pr"), with = "a1c_pr", by = "a1c_grp")
CORR <- NULL
```

```{r}
#| label: helpers
classify_buckets <- function(d, vars, overrides = list()) {
  vars <- unique(as.character(vars))
  unknown <- setdiff(vars, names(d))
  if (length(unknown)) stop("Unknown table variable(s): ", paste(unknown, collapse = ", "), call. = FALSE)
  allowed <- c("continuous", "binary", "categorical")
  if (!is.list(overrides)) {
    stop("overrides must be a list named continuous, binary, or categorical.", call. = FALSE)
  }
  override_names <- names(overrides)
  if (length(overrides) &&
        (is.null(override_names) || anyNA(override_names) || any(!nzchar(override_names)))) {
    stop("Every override bucket must be named continuous, binary, or categorical.", call. = FALSE)
  }
  if (anyDuplicated(override_names)) {
    stop("Override bucket names must be unique.", call. = FALSE)
  }
  if (length(setdiff(override_names, allowed))) {
    stop("overrides must be a list named continuous, binary, or categorical.", call. = FALSE)
  }
  forced <- unlist(overrides, use.names = FALSE)
  if (anyDuplicated(forced)) stop("A variable may appear in only one override bucket.", call. = FALSE)
  if (length(setdiff(forced, vars))) {
    stop("Override variable(s) are not in vars: ", paste(setdiff(forced, vars), collapse = ", "), call. = FALSE)
  }

  is_binary <- vapply(d[vars], function(x) {
    if (is.logical(x)) return(TRUE)
    values <- unique(stats::na.omit(x))
    if (is.numeric(x)) return(setequal(values, c(0, 1)))
    if (is.factor(x)) {
      return(nlevels(x) == 2L && setequal(toupper(levels(x)), c("NO", "YES")))
    }
    if (is.character(x)) {
      return(length(values) == 2L && setequal(toupper(values), c("NO", "YES")))
    }
    FALSE
  }, logical(1))
  is_continuous <- vapply(d[vars], function(x) {
    is.numeric(x) && length(unique(x[!is.na(x)])) > 6L
  }, logical(1))
  buckets <- list(
    continuous = vars[!is_binary & is_continuous],
    binary = vars[is_binary],
    categorical = vars[!is_binary & !is_continuous]
  )
  for (bucket in intersect(allowed, names(overrides))) {
    move <- as.character(overrides[[bucket]])
    buckets <- lapply(buckets, setdiff, y = move)
    buckets[[bucket]] <- unique(c(buckets[[bucket]], move))
  }
  buckets
}
```

## Table

```{r}
#| label: table
b <- classify_buckets(d, unlist(GROUPS, use.names = FALSE), OVERRIDES)
tbl <- hv_tbl_summary(
  d, by = BY, groups = GROUPS,
  continuous = b$continuous, binary = b$binary, categorical = b$categorical,
  continuous_stat = "both", percentiles = c(15, 85),
  compare = if (is.null(BY)) "none" else "pvalue"
)
tbl
hv_man_table_save(
  hv_man_table(tbl),
  file = set_path("descriptive", paste0(ENDPOINT, "-", TYPE, "-dc-tables.docx"))
)
```

## Correlations

```{r}
#| label: correlation
#| results: asis
if (is.null(CORR)) {
  cat("Correlation variant not requested (`CORR <- NULL`).\n")
} else {
  ct <- hv_correlation_table(d, vars = CORR$vars, with = CORR$with)
  conf_level <- attr(ct, "conf_level")
  if (!is.null(CORR$by)) {
    st <- hv_correlation_table(d, vars = CORR$vars, with = CORR$with, by = CORR$by)
    ct[[CORR$by]] <- "Overall"
    ct <- rbind(ct[names(st)], st)
    attr(ct, "conf_level") <- conf_level
  }
  print(knitr::kable(ct, caption = sprintf(
    "Spearman and Pearson, %d%% Fisher interval", round(100 * attr(ct, "conf_level"))
  )))
  cm <- hv_correlation_matrix(d, vars = unique(c(CORR$with, CORR$vars)))
  fname <- "dc-tables-correlation-matrix.png"
  png(set_path("descriptive", fname), width = 10, height = 10, units = "in", res = 150)
  print(plot(cm) + theme_hv_manuscript())
  invisible(dev.off())
  cat("\n![Scatter-plot matrix](", file.path(paste0(ENDPOINT, "-", TYPE), fname), ")\n", sep = "")
}
```
