---
title: "Trends over operation year"
# Carried in the file rather than inherited from _quarto.yml. A file meant to be
# COPIED should not depend on settings from the directory it happens to sit in:
# rendered as a standalone document the project's format block never applies,
# which is how server-rendered reports end up with sibling _files/ trees instead
# of being self-contained.
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    page-layout: full
    grid:
      body-width: 2000px
      sidebar-width: 250px
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (an older graphs/dp.trends job) and list
     the figures it drew. A trends job usually draws several, and each one
     becomes one entry in TRENDS below. -->

Replaces `graphs/dp.trends`: list the trend figures it draws.

A `dc-trends` job shows how the cohort changed over the years of operation: the
share of patients with a characteristic, or the level of a measurement, year by
year, with a smooth through it. It **describes** the cohort. It fits no model and
tests nothing, so a trend here is a prompt for an analysis, not a result.

The engine is `hvtiPlotR::hv_trends()`. It takes one row per patient and computes
the per-year summary itself, so this job never aggregates by hand. The
`group_by()`/`summarize()` blocks and the per-series `smooth.spline()` calls that
older trends jobs carry are exactly what it replaces.

Two neighbors are **not** this job. Procedure volume per year is `dp-procs`, and
the distribution of a single variable is `dc-variable`.

```{r}
#| label: setup
# unnumbered: loads packages and checks versions only
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRtemplates:::.find_study_root(if (is.null(.in)) getwd() else dirname(.in))
.provenance_data <- list()
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(hvtiRutilities)
  library(hvtiPlotR)
  library(ggplot2)
})

# 2.7.7, not 2.7.6. This job reports how many rows each figure could not draw,
# and hv_trends() only counts them (meta$n_missing) from part-way through 2.7.6:
# the field arrived without a version bump, so a package reporting 2.7.6 may or
# may not have it. 2.7.7 is the first version that certainly does.
if (utils::packageVersion("hvtiPlotR") < "2.7.7") {
  stop("This job needs hvtiPlotR >= 2.7.7; ", utils::packageVersion("hvtiPlotR"),
       " is installed. Update it, then re-render.", call. = FALSE)
}
```

```{r}
#| label: guard-edits
#| results: asis
# The markers in this file name work a study author still has to do, and a job
# that still contains one has not been finished. This chunk is what makes that
# TRUE rather than merely stated: an unedited trends job would otherwise render
# green over the placeholder columns and the placeholder year origin below.
#
# knitr::current_input() is NULL outside a render, when a study author steps
# through chunks in RStudio, and there is no file to scan then, so this is a
# no-op in that case.
#
# The token is BUILT, not written literally, and that is not stylistic. Quarto
# knits through an intermediate and current_input() returns THAT file, so the
# scan reads a copy of this chunk along with everything else: a literal
# grep("<token>", .src) here matches its own source line, and the guard then
# fires on every render, finished or not. Measured, not theorised -- a probe
# found three markers in a file containing two. A guard that cries wolf on a
# finished job is a guard that gets deleted, which returns us to issue #27.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    # Report the marker TEXT. Line numbers here are the intermediate's, not
    # this .qmd's, so they need not match what the author sees in an editor.
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    # A job renders as a draft by default, so an author can iterate on a
    # working report and remove markers as they go. HVTI_TEMPLATE_STRICT
    # turns the draft into a stop for a final render. Only unset, 0, false
    # and no mean "not strict": an unrecognized value stops on purpose, so a
    # mistyped switch is seen rather than quietly ignored, which is the safe
    # direction to be wrong in.
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      # The banner is not optional. A draft render that looks like a finished
      # one is the same defect with an extra step: the warning scrolls past in
      # a log, while the .html is the artifact that gets sent to someone.
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

```{r}
#| label: guard-partial
#| results: asis
# unnumbered: a callout, printed only when part of the job is left out
# To render a job you have not finished, leave a chunk out with the chunk
# option skip, giving the reason in quotes, or call hvtiRtemplates::stop_here()
# in a chunk to leave out everything below it. A draft lists each one here; a
# final render refuses them, as it refuses an EDIT marker. ?stop_here has more.
hvtiRtemplates:::.guard_partial(knitr::current_input())
```

<!-- The two values below are written by add_job() from the subject/type it
     was called with, which are also the filename's first two fields. Every
     artifact path in this document is computed from them.

     For an EDA job the set key is (subject, eda). SUBJECT names WHAT is
     described: `cohort` for the whole study, or a label for a sub-cohort such
     as `mv_repair`. TYPE is always `eda`. Every EDA job for one subject then
     shares a set, so its trends, follow-up and EDA figures land
     together under graphs/<subject>-eda/. Decided 2026-09-10, because a trends
     plot describes a cohort rather than analyzing an endpoint.

     Do NOT hand-edit these declarations: the render-time guard below checks
     them against this file's own name and stops if they disagree. -->

```{r}
#| label: set
SUBJECT <- "cohort"
TYPE    <- "eda"

# `add_job()` writes SUBJECT/TYPE from the same values it put in this file's
# name, but a hand-edited declaration can drift from it afterward. set_path()
# below resolves from the declarations, not the filename, so a drifted
# declaration would silently write into ANOTHER set's artifact directory --
# exactly the collision the (subject, type) key exists to prevent, re-entered
# through the body instead of the name.
#
# knitr::current_input() is NULL outside a render (a study author running
# chunks interactively in RStudio), and there is no filename to check against
# yet, so this is a no-op in that case rather than a spurious error.
.current <- knitr::current_input()
if (!is.null(.current)) {
  # The name is template first, <prefix>[.<qualifier>].<subject>.<type>, or
  # <subject>-<type>-<prefix>[-<qualifier>] for a job scaffolded before
  # 2026-10. .job_name_fields() reads subject and type from either, whatever
  # the extension: Quarto knits through an intermediate, so
  # `knitr::current_input()` names the `.rmarkdown` file here, not the `.qmd`.
  .fields <- hvtiRtemplates:::.job_name_fields(.current)
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject, "', type '",
         .name_type, "'), but declares SUBJECT = \"", SUBJECT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

# Resolve a path inside this set's artifact directory. `kind` is the artifact
# folder -- "estimates" for serialized results, "graphs" for figures. The set
# directory sits one layer under the kind and never two, which is the whole
# layout rule.
#
# Created on first use rather than up front, so a job that writes nothing leaves
# no empty directories behind.
set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}

# Saves a figure as <name>.png and <name>.pdf in this set's graphs/ folder (or `kind`'s),
# under the SAVE_FIGURES and FIGURES study choices.
save_figure <- function(plot, name, width = 6, height = 4, kind = "graphs", linked = FALSE) {
  hvtiRtemplates:::.save_figure(plot, set_path(kind, paste0(name, ".png")), width, height,
                                SAVE_FIGURES, FIGURES, linked)
}
```

## Study choices

Set the values in this chunk before rendering.

```{r}
#| label: edit-study-choices
# EDIT: the registered dataset this job reads ("built" is the study dataset).
# MIGRATE-BEGIN: dc-trends-data
DATASET <- "built"
# MIGRATE-END: dc-trends-data

# EDIT: an hvtiRdatabuild analysis set, or NULL to read the whole dataset.
ANALYSIS_SET <- NULL

# EDIT: rows to keep, dplyr::filter() style, or NULL to keep every row:
#   WHERE <- quote(age >= 18)
#   WHERE <- rlang::exprs(age >= 18, hx_chf == 1)
# WHERE runs as the data are read, before `year` is derived below, so filter on
# the date column rather than on `year`.
WHERE <- NULL

# EDIT: the patient identifier. Without "ccfid" the job uses MRN, then eMRN;
# name another column, such as "randid", if the study uses one.
ID <- "ccfid"

# EDIT: what makes a row unique; one row per patient unless repeated measures
# add their visit time or date, for example KEY <- c(ID, "iv_echo").
KEY <- ID

# Optional, and needs no edit: NULL reads the cohort alone. To join one
# registered ancillary dataset (echoes, labs), name it in JOIN. The cohort
# above decides the patients, one row each; the joined records of other
# patients are dropped and counted in the data table.
#   JOIN_VARS: the cohort columns each joined row carries; NULL carries all,
#     and a column both datasets have stops, so list only those the job needs.
#   REDUCE: NULL keeps a row per joined record, keyed on that dataset's key;
#     list(rule = "first", by = "echo_date") keeps one row per patient ("last",
#     or "nearest" with to = a cohort date column). WHERE on a joined column
#     filters the records first, so "last" with WHERE echo_type == "TTE"
#     keeps each patient's last TTE. A tie stops: picking one record
#     silently would be a hidden choice; by = c("echo_date", "echo_seq")
#     breaks it. A job that counts or models
#     patients, one row each, needs REDUCE: NULL would count every record.
#   JOIN_KEY: overrides the joined dataset's registered key.
JOIN <- NULL
JOIN_VARS <- NULL
REDUCE <- NULL
JOIN_KEY <- NULL

# EDIT: one entry per figure. `cols` are the columns drawn together: several
# columns become several series in one figure, which is how a set of 0/1
# indicators such as NYHA class I to IV is shown. `kind` says how to read them:
#   "percent"     0/1 or logical, drawn as the percent of patients with it
#   "continuous"  a measurement, drawn on its own scale
# `labels` names the series in the legend (default: the column names). Fixing
# `ylim` and `ybreaks` keeps figures for different subgroups comparable; either
# may be set alone, and `ylim` zooms without dropping any patient.
# `summary` is the per-year point drawn under the smooth. Percent trends must
# use "mean" (the default); continuous trends may use "mean" or "median".
# MIGRATE-BEGIN: dc-trends-trends
TRENDS <- list(
  chf = list(cols = "hx_chf", kind = "percent", labels = "Heart failure",
             ylab = "Patients (%)", ylim = c(0, 100), ybreaks = seq(0, 100, 20)),
  lvmass = list(cols = "plvmassi", kind = "continuous", labels = "LV mass index",
                ylab = "LV mass index", ylim = NULL, ybreaks = NULL)
)
# MIGRATE-END: dc-trends-trends

# EDIT: optional. x-axis breaks shared by every figure, e.g. seq(1985, 2025, 5).
# NULL lets ggplot choose.
# MIGRATE-BEGIN: dc-trends-xbreaks
XBREAKS <- NULL
# MIGRATE-END: dc-trends-xbreaks

# EDIT: optional. Each entry is a filter, given as a function of the data that
# returns TRUE for the rows to keep. Every trend is drawn once per entry with
# the same axes, so the figures can be read side by side. The default draws the
# whole cohort only. A second entry might be
#   complex = function(d) d$complex == 1
# MIGRATE-BEGIN: dc-trends-subgroups
SUBGROUPS <- list(all = function(d) rep(TRUE, nrow(d)))
# MIGRATE-END: dc-trends-subgroups

# Each figure is saved to graphs/ as a PNG (for Word) and a PDF (for the publisher).
# SAVE_FIGURES <- FALSE saves neither; FIGURES keeps only the figures whose names
# start with one of its entries, e.g. FIGURES <- c("hp-survival"). The names are the
# file names, listed for each template in the templates README.
SAVE_FIGURES <- TRUE
FIGURES <- NULL
```

## Data

```{r}
#| label: tbl-data
#| tbl-cap: "The data this job read"
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
.cfg <- study_config(start = .root)
job_data <- hvtiRtemplates::read_job_data(.cfg, dataset = DATASET, analysis_set = ANALYSIS_SET,
                                          where = WHERE, id = ID, key = KEY, join = JOIN, join_vars = JOIN_VARS,
                                          reduce = REDUCE, join_key = JOIN_KEY)
d <- job_data$data
.provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(job_data$provenance),
                      if (!is.null(job_data$provenance_join)) list(job_data$provenance_join))
knitr::kable(job_data$record, col.names = c("Data", ""))
```

```{r}
#| label: data-attrition
#| results: asis
# unnumbered: its child chunk carries its own label and caption
if (!is.null(job_data$attrition)) {
  .fence <- strrep("`", 3)
  cat(knitr::knit_child(text = c(
    paste0(.fence, "{r}"), "#| label: tbl-data-attrition",
    paste0("#| tbl-cap: ", encodeString(paste0("Analysis set `", ANALYSIS_SET, "`: exclusions, in order"), quote = "\"")),
    "knitr::kable(job_data$attrition)", .fence
  ), envir = environment(), quiet = TRUE), sep = "\n")
}
```

```{r}
#| label: edit-year
# EDIT: the calendar year of operation, as a WHOLE year. hv_trends puts one
# summary point at every distinct value of x, so x must be the year itself.
# Years since an origin are an interval, fractional like every iv_ column, and
# origin plus a fractional interval would give one "annual" point per patient.
# Older jobs kept the two apart for this reason: they smoothed on the interval
# and averaged on an integer year column. If the dataset carries the calendar
# year, use that column; otherwise take the whole years of the interval and add
# the origin. The origin differs between studies, 1985 in one and 2000 in
# another, and a wrong one moves every figure along the x-axis, which is why
# the range is checked and printed below.
# MIGRATE-BEGIN: dc-trends-year
d$year <- floor(d$iv_opyrs) + 1985
# MIGRATE-END: dc-trends-year
```

```{r}
#| label: tbl-year-check
#| tbl-cap: "Operation years: patients, missing years, and the first and last year"
if (!is.numeric(d$year)) {
  stop("`year` is not numeric (", class(d$year)[1L], "). Build it from a ",
       "numeric column, such as years since an origin plus that origin.",
       call. = FALSE)
}
if (all(is.na(d$year))) {
  stop("`year` is missing for every patient, so there is nothing to plot. ",
       "Check the column it is built from.", call. = FALSE)
}
if (any(d$year != round(d$year), na.rm = TRUE)) {
  stop("`year` must be a whole calendar year. hv_trends draws one summary ",
       "point per distinct value, so a fractional year draws one point per ",
       "patient. Use floor() on a years-since-origin interval, or a ",
       "calendar-year column.", call. = FALSE)
}
# A plausible window, not a precise one. Its job is to catch the wrong-origin
# mistake, which lands decades or centuries out, not to police real data.
.this_year <- as.integer(format(Sys.Date(), "%Y"))
.out <- which(d$year < 1900 | d$year > .this_year + 1)
if (length(.out)) {
  stop(length(.out), " patient(s) have an operation year outside 1900 to ",
       .this_year + 1, " (for example ", d$year[.out[1L]], "). Check the ",
       "origin added to the years-since-origin column.", call. = FALSE)
}

knitr::kable(data.frame(
  quantity = c("patients", "year missing", "first year", "last year"),
  value    = c(nrow(d), sum(is.na(d$year)),
               min(d$year, na.rm = TRUE), max(d$year, na.rm = TRUE))
))
```

```{r}
#| label: helpers
# unnumbered: defines helpers only
# One row per patient per series: the shape hv_trends() takes. Several columns
# become several series through `series`, which hv_trends() groups on.
trend_long <- function(d, spec) {
  absent <- setdiff(spec$cols, names(d))
  if (length(absent)) {
    stop("TRENDS names column(s) not in the data: ",
         paste(absent, collapse = ", "), call. = FALSE)
  }
  labels <- if (is.null(spec$labels)) spec$cols else spec$labels
  if (length(labels) != length(spec$cols)) {
    stop("TRENDS entry has ", length(spec$cols), " column(s) but ",
         length(labels), " label(s).", call. = FALSE)
  }
  if (!spec$kind %in% c("percent", "continuous")) {
    stop("TRENDS `kind` must be \"percent\" or \"continuous\", not \"",
         spec$kind, "\".", call. = FALSE)
  }
  summary_fn <- if (is.null(spec$summary)) "mean" else spec$summary
  if (identical(spec$kind, "percent") && !identical(summary_fn, "mean")) {
    stop("A percent trend must use summary = \"mean\"; a median of a 0/1 ",
         "indicator is not a prevalence.", call. = FALSE)
  }
  parts <- lapply(seq_along(spec$cols), function(i) {
    v <- d[[spec$cols[i]]]
    if (identical(spec$kind, "percent")) {
      if (is.logical(v)) v <- as.numeric(v)
      if (!is.numeric(v)) {
        stop("`", spec$cols[i], "` is kind \"percent\" but is ",
             class(v)[1L], ". A percent column must be numeric or logical ",
             "and hold only 0/1 values.", call. = FALSE)
      }
      # A 1/2 or 0/1/9 code times 100 draws a plausible and wrong figure, so
      # anything other than 0/1 stops here instead of being scaled.
      bad <- setdiff(unique(v[!is.na(v)]), c(0, 1))
      if (length(bad)) {
        stop("`", spec$cols[i], "` is kind \"percent\" but holds ",
             paste(utils::head(bad, 5L), collapse = ", "),
             ". A percent column must be 0/1 or logical.", call. = FALSE)
      }
      v <- 100 * v
    }
    data.frame(year = d$year, value = v, series = labels[i],
               stringsAsFactors = FALSE)
  })
  out <- do.call(rbind, parts)
  out$series <- factor(out$series, levels = labels)
  out
}
```

## Figures

```{r}
#| label: figures
#| results: asis
# unnumbered: each child chunk below carries its own label and caption
# This document sits in descriptive/ and its figures are written to graphs/
# through set_path(), so the image link is the figure's path relative to this
# document, worked out at render time, as dc-gfup's is.
.fence <- strrep("`", 3)
.seen <- new.env()
.child <- function(label, caption, code) {
  label <- gsub("(^-+|-+$)", "", gsub("[^a-z0-9]+", "-", tolower(label)))
  if (!is.null(.seen[[label]])) stop("Two outputs would share the label ", label, ".", call. = FALSE)
  .seen[[label]] <- TRUE
  .opt <- paste0("#| ", sub("-.*$", "", label), "-cap: ", encodeString(caption, quote = "\""))
  cat(knitr::knit_child(text = c(paste0(.fence, "{r}"), paste0("#| label: ", label), .opt, code, .fence),
                        envir = parent.frame(), quiet = TRUE), sep = "\n")
  cat("\n")
}
figure_link <- function(file) {
  xfun::relative_path(file, if (exists(".in") && !is.null(.in)) dirname(.in) else getwd(), error = FALSE)
}
for (sg in names(SUBGROUPS)) {
  keep <- SUBGROUPS[[sg]](d)
  # One logical per patient, or the filter is wrong. R would silently recycle
  # a length-1 or short vector through d[keep, ] and select the wrong patients,
  # and a filter on a missing column returns logical(0).
  if (!is.logical(keep) || length(keep) != nrow(d)) {
    stop("SUBGROUPS entry `", sg, "` must return one TRUE/FALSE per patient (",
         nrow(d), "); it returned ", length(keep), " value(s) of type ",
         typeof(keep), ".", call. = FALSE)
  }
  keep[is.na(keep)] <- FALSE
  dd <- d[keep, , drop = FALSE]
  if (nrow(dd) == 0L) {
    stop("SUBGROUPS entry `", sg, "` selects no patients.", call. = FALSE)
  }
  cat("\n### ", sg, " (n = ", nrow(dd), ")\n\n", sep = "")
  for (nm in names(TRENDS)) {
    spec <- TRENDS[[nm]]
    tr <- hv_trends(trend_long(dd, spec), x_col = "year", y_col = "value",
                    group_col = if (length(spec$cols) == 1L) NULL else "series",
                    summary_fn = if (is.null(spec$summary)) "mean" else spec$summary)
    # Older jobs dropped incomplete rows with na.omit() and said nothing.
    # hv_trends() counts them, and the count is reported with every figure.
    cat("**", nm, "**: ", tr$meta$n_obs - tr$meta$n_missing, " of ",
        tr$meta$n_obs, " rows drawn; ", tr$meta$n_missing,
        " missing the year or the value.\n\n", sep = "")
    p <- plot(tr) +
      labs(x = "Year of operation", y = spec$ylab, color = NULL, shape = NULL) +
      theme_hv_manuscript()
    # Limits through coord_cartesian(), not scale limits. A scale limit DROPS
    # the patients outside it before the smooth is fitted, so the curve would
    # stop passing through the yearly points hv_trends() computed from all of
    # them, and any stretch of the fitted line past the limit would vanish.
    # Breaks and limits are applied independently: either may be set alone.
    if (!is.null(spec$ybreaks)) p <- p + scale_y_continuous(breaks = spec$ybreaks)
    if (!is.null(spec$ylim)) p <- p + coord_cartesian(ylim = spec$ylim)
    if (!is.null(XBREAKS)) p <- p + scale_x_continuous(breaks = XBREAKS)
    fname <- paste0("dc-trends-", nm, "-", sg, ".png")
    file <- save_figure(p, sub("[.]png$", "", fname), width = 8, height = 6, linked = TRUE)
    .link <- figure_link(file)
    .child(paste("fig trends", nm, sg), paste0("Trend in ", nm, " by year of operation, ", sg),
           "knitr::include_graphics(.link, error = FALSE)")
  }
}
```

The smooth is LOESS, `hv_trends()`'s default, with `span = 0.75`. Older jobs used
`smooth.spline(df = 6)`, which is not reproduced: the two differ at the ends of
the year range, where data is thinnest, so compare against an older figure by its
shape and not point by point. `plot(tr, span = ...)` changes the smoothing.

```{r}
#| label: provenance
#| echo: false
#| results: asis
cat(hvtiRtemplates:::.embed_provenance(
  .in,
  data = .provenance_data,
  extra = list(
    subject = SUBJECT,
    type = TYPE
  )
))
```
