---
title: "Descriptive tables"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (descriptive/dc.tables.*) and say whether it
     was the overall tables, the by-group tables, or the correlation variant. -->

Replaces `descriptive/dc.tables`: say which variant.

A `dc-tables` job is the formatted descriptive table: every variable the study
reports, grouped under the section headings the SAS `%desc_tab` lists carried as
comments, categorical as `n (%)` and continuous in the summary form you select.
The default is median (15th, 85th percentile).
`%desc_tab` wrote the categorical and continuous tables as two
RTF files; `hv_tbl_summary()` writes one table with both. It describes; with
`BY` set it separates the groups into columns. `COMPARE` chooses whether to
test them; a p-value here is a prompt for an analysis, not a result.

```{r}
#| label: setup
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRutilities::study_root(if (is.null(.in)) getwd() else dirname(.in))
.provenance_data <- list()
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(hvtiRutilities)
  library(hvtiRtables)
  library(hvtiPlotR)
  library(ggplot2)
})
if (utils::packageVersion("hvtiRdatabuild") < "0.2.1") {
  stop("This job needs hvtiRdatabuild >= 0.2.1 for read_analysis_set(); ",
       utils::packageVersion("hvtiRdatabuild"), " is installed.", call. = FALSE)
}
if (utils::packageVersion("hvtiRtables") < "1.0.1") {
  stop("This job needs hvtiRtables >= 1.0.1 for hv_correlation_table(); ",
       utils::packageVersion("hvtiRtables"), " is installed.", call. = FALSE)
}
if (utils::packageVersion("hvtiPlotR") < "2.7.14") {
  stop("This job needs hvtiPlotR >= 2.7.14 for hv_correlation_matrix(); ",
       utils::packageVersion("hvtiPlotR"), " is installed.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
# The markers in this file name work a study author still has to do, and a job
# that still contains one has not been finished. This chunk is what makes that
# TRUE rather than merely stated: an unedited job would otherwise render green
# over placeholder columns.
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src  <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

<!-- The two values below are written by add_job() from the subject/type it
     was called with, which are also the filename's first two fields. Every
     artifact path in this document is computed from them. -->

```{r}
#| label: set
SUBJECT <- "cohort"
TYPE    <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type     <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject, "', type '",
         .name_type, "'), but declares SUBJECT = \"", SUBJECT, "\", TYPE = \"", TYPE,
         "\". Fix the declaration or the filename before rendering.", call. = FALSE)
  }
}

set_path <- function(kind, file) {
  dir <- file.path(hvtiRutilities::study_dir(kind, .root),
                   paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(dir)) dir.create(dir, recursive = TRUE)
  file.path(dir, file)
}
```

## Study choices

Set the values in this chunk before rendering.

```{r}
#| label: study-choices
# EDIT: the built dataset this job reads. "study" is the file named by built:
# in _study.yml; any other name must be declared under additional_datasets:,
# such as a column subset written for R. Analysis sets are derived from
# "study" only, so another dataset needs ANALYSIS_SET <- NULL.
# EDIT: the analysis set this job reads: the columns and excluded rows declared
# under analysis_sets: in _study.yml and written once with
# hvtiRdatabuild::write_analysis_set(). A stale set stops and names the call
# that refreshes it. Use NULL to describe the whole built cohort instead.
# MIGRATE-BEGIN: dc-tables-data
DATASET <- "study"
ANALYSIS_SET <- "eda"
# MIGRATE-END: dc-tables-data

# EDIT: the variables to report, grouped. Each name is a section heading and
# becomes a row group in the table, in this order. These are the /* Demography */
# style banners from the SAS varlist; carry them across unchanged so the R table
# reads like the SAS one. Commented-out SAS blocks stay out.
# MIGRATE-BEGIN: dc-tables-config
GROUPS <- list(
  Demography = c("age", "female"),
  Symptoms   = c("nyha_pr")
)
# EDIT: NULL for the overall table. A column name reproduces the `%macro skip`
# by-group sections (by=): one column per observed level. Review byvalue=
# against the data before accepting the columns; COMPARE controls testing.
BY <- NULL
# All three NULL values use automatic classification for a new, unmigrated job.
# Migrated jobs declare every row explicitly; character(0) is an empty bucket.
CONTINUOUS <- NULL
BINARY <- NULL
CATEGORICAL <- NULL
COMPARE <- "none"             # EDIT: choose whether to compare groups
CONTINUOUS_STAT <- "median"   # EDIT: choose the continuous summary
PERCENTILES <- c(15, 85)       # EDIT: choose the reported percentiles
ABBREVIATIONS <- character(0) # EDIT: define abbreviations used in the table
WORD_FILE <- "dc-tables.docx"  # EDIT: choose the Word filename within this set
# MIGRATE-END: dc-tables-config
# EDIT: force a bucket where the data guess wrong, e.g. an ordinal score you
# want summarised as continuous: list(continuous = c("nyha_pr")). list() for none.
OVERRIDES <- list()
# EDIT: NULL skips the correlation variant (dc.tables.ods_<topic>.sas). To run
# it, name the numeric columns, the anchor to correlate them against (NULL for
# every pair), and an optional stratum:
#   list(vars = c("glu_pr", "creat_pr"), with = "a1c_pr", by = "a1c_grp")
CORR <- NULL
```

## Data

<!-- How this job finds its data. Read this before touching the chunk below.

     A study can offer a job data three ways, and `_study.yml` says which exist:

     1. The study dataset: the file named by `built:`. `DATASET <- "study"`
        with `ANALYSIS_SET <- NULL`.
     2. An additional dataset: any file declared under `additional_datasets:`,
        for example a column subset written for R. Set `DATASET` to its
        name there and `ANALYSIS_SET <- NULL`. It is read as it stands: no
        rows are excluded.
     3. An analysis set: declared under `analysis_sets:` with the columns kept
        and the rows excluded, then written once with
        `hvtiRdatabuild::write_analysis_set()`. Set `ANALYSIS_SET` to its name
        and leave `DATASET` as "study", because a set is always cut from the
        study dataset. The report shows its exclusions in order.

     A name under `additional_datasets:` is not an analysis set, so putting it
     in `ANALYSIS_SET` stops with "No analysis set". Whichever route is taken,
     the report prints the dataset, file and dimensions it read, so a reader
     never has to ask which data a table describes. -->

```{r}
#| label: data
# The checksum of every dataset in manifest.yaml is checked before anything is
# read, so a result can name the data that produced it. It stops on a mismatch.
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
if (is.null(ANALYSIS_SET)) {
  if (!is.character(DATASET) || length(DATASET) != 1L || is.na(DATASET) || !nzchar(DATASET)) {
    stop("DATASET must be set to one dataset registered in _study.yml, such as \"study\". ",
         "In a migrated job, see the migration report beside this job.", call. = FALSE)
  }
  .cfg <- study_config(start = .root)
  .data_read <- hvtiRtemplates:::.provenance_read(
    DATASET, .cfg, function() read_built(cfg = .cfg, dataset = DATASET)
  )
  d <- .data_read$value
  .provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(.data_read$record))
} else if (!identical(DATASET, "study")) {
  stop("An analysis set is written from the study dataset, not `", DATASET,
       "`. Set ANALYSIS_SET <- NULL to read `", DATASET, "` whole.", call. = FALSE)
} else {
  .cfg <- study_config(start = .root)
  .data_read <- hvtiRtemplates:::.provenance_read(
    "study", .cfg,
    function() hvtiRdatabuild::read_analysis_set(ANALYSIS_SET, cfg = .cfg),
    role = paste0("analysis_set:", ANALYSIS_SET)
  )
  d <- .data_read$value
  .provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(.data_read$record))
  knitr::kable(attr(d, "attrition"), caption = paste0(
    "Analysis set `", ANALYSIS_SET, "`: exclusions, in order"
  ))
}
cat("Data read: ", if (is.null(ANALYSIS_SET)) {
  paste0("dataset `", DATASET, "` (",
         basename(hvtiRutilities::built_path(dataset = DATASET)), ")")
} else {
  paste0("analysis set `", ANALYSIS_SET, "` of the study dataset")
}, ", ", nrow(d), " rows, ", ncol(d), " columns.\n", sep = "")
```


```{r}
#| label: helpers
classify_buckets <- function(d, vars, overrides = list()) {
  vars <- unique(as.character(vars))
  unknown <- setdiff(vars, names(d))
  if (length(unknown)) stop("Unknown table variable(s): ", paste(unknown, collapse = ", "), call. = FALSE)
  allowed <- c("continuous", "binary", "categorical")
  if (!is.list(overrides)) {
    stop("overrides must be a list named continuous, binary, or categorical.", call. = FALSE)
  }
  override_names <- names(overrides)
  if (length(overrides) &&
        (is.null(override_names) || anyNA(override_names) || any(!nzchar(override_names)))) {
    stop("Every override bucket must be named continuous, binary, or categorical.", call. = FALSE)
  }
  if (anyDuplicated(override_names)) {
    stop("Override bucket names must be unique.", call. = FALSE)
  }
  if (length(setdiff(override_names, allowed))) {
    stop("overrides must be a list named continuous, binary, or categorical.", call. = FALSE)
  }
  forced <- unlist(overrides, use.names = FALSE)
  if (anyDuplicated(forced)) stop("A variable may appear in only one override bucket.", call. = FALSE)
  if (length(setdiff(forced, vars))) {
    stop("Override variable(s) are not in vars: ", paste(setdiff(forced, vars), collapse = ", "), call. = FALSE)
  }

  is_binary <- vapply(d[vars], function(x) {
    if (is.logical(x)) return(TRUE)
    values <- unique(stats::na.omit(x))
    if (is.numeric(x)) return(setequal(values, c(0, 1)))
    if (is.factor(x)) {
      return(nlevels(x) == 2L && setequal(toupper(levels(x)), c("NO", "YES")))
    }
    if (is.character(x)) {
      return(length(values) == 2L && setequal(toupper(values), c("NO", "YES")))
    }
    FALSE
  }, logical(1))
  is_continuous <- vapply(d[vars], function(x) {
    is.numeric(x) && length(unique(x[!is.na(x)])) > 6L
  }, logical(1))
  buckets <- list(
    continuous = vars[!is_binary & is_continuous],
    binary = vars[is_binary],
    categorical = vars[!is_binary & !is_continuous]
  )
  for (bucket in intersect(allowed, names(overrides))) {
    move <- as.character(overrides[[bucket]])
    buckets <- lapply(buckets, setdiff, y = move)
    buckets[[bucket]] <- unique(c(buckets[[bucket]], move))
  }
  buckets
}
```

## Table

```{r}
#| label: table
explicit <- list(continuous = CONTINUOUS, binary = BINARY, categorical = CATEGORICAL)
if (all(vapply(explicit, is.null, logical(1)))) {
  b <- classify_buckets(d, unlist(GROUPS, use.names = FALSE), OVERRIDES)
  CONTINUOUS <- b$continuous
  BINARY <- b$binary
  CATEGORICAL <- b$categorical
} else {
  if (any(vapply(explicit, is.null, logical(1))) || length(OVERRIDES)) {
    stop("Declare all three buckets and leave OVERRIDES empty for explicit classification.", call. = FALSE)
  }
  rows <- unlist(GROUPS, use.names = FALSE)
  classified <- unlist(explicit, use.names = FALSE)
  if (anyDuplicated(rows) || anyDuplicated(classified) || !setequal(rows, classified)) {
    stop("Every grouped row must appear in exactly one explicit bucket.", call. = FALSE)
  }
  invisible(classify_buckets(d, rows, explicit))
}
tbl <- hvtiRtables::hv_tbl_summary(
  d, by = BY, groups = GROUPS, continuous = CONTINUOUS,
  binary = BINARY, categorical = CATEGORICAL, compare = COMPARE,
  percentiles = PERCENTILES, continuous_stat = CONTINUOUS_STAT
)
tbl
ft <- hvtiRtables::hv_man_table(tbl)
if (length(WORD_FILE) != 1L || is.na(WORD_FILE) ||
      !grepl("^[A-Za-z0-9_][A-Za-z0-9_.-]*[.]docx$", WORD_FILE)) {
  stop("WORD_FILE must be a plain .docx filename within this set.", call. = FALSE)
}
word_path <- file.path(study_dir("documents", .root), paste0(SUBJECT, "-", TYPE), WORD_FILE)
dir.create(dirname(word_path), recursive = TRUE, showWarnings = FALSE)
hvtiRtables::hv_man_table_save(ft, word_path, abbreviations = ABBREVIATIONS)
check <- hvtiRtables::hv_check_docx(word_path)
if (nrow(check)) stop("Word output failed the CORR structural check", call. = FALSE)
```

## Correlations

```{r}
#| label: correlation
#| results: asis
if (is.null(CORR)) {
  cat("Correlation variant not requested (`CORR <- NULL`).\n")
} else {
  ct <- hv_correlation_table(d, vars = CORR$vars, with = CORR$with)
  conf_level <- attr(ct, "conf_level")
  if (!is.null(CORR$by)) {
    st <- hv_correlation_table(d, vars = CORR$vars, with = CORR$with, by = CORR$by)
    ct[[CORR$by]] <- "Overall"
    ct <- rbind(ct[names(st)], st)
    attr(ct, "conf_level") <- conf_level
  }
  print(knitr::kable(ct, caption = sprintf(
    "Spearman and Pearson, %d%% Fisher interval", round(100 * attr(ct, "conf_level"))
  )))
  cm <- hv_correlation_matrix(d, vars = unique(c(CORR$with, CORR$vars)))
  fname <- "dc-tables-correlation-matrix.png"
  png(set_path("descriptive", fname), width = 10, height = 10, units = "in", res = 150)
  print(plot(cm) + theme_hv_manuscript())
  invisible(dev.off())
  cat("\n![Scatter-plot matrix](", file.path(paste0(SUBJECT, "-", TYPE), fname), ")\n", sep = "")
}
```

```{r}
#| label: provenance
#| echo: false
#| results: asis
cat(hvtiRtemplates:::.embed_provenance(
  .in,
  data = .provenance_data,
  extra = list(
    subject = SUBJECT,
    type = TYPE
  )
))
```
