---
title: "EDA report"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the EDA report this replaces, or delete this line for a new study. -->

Replaces an older EDA report.

A `dp-eda` job is the whole data-checking report over a new build, in one
render: an overview of every column, goodness of follow-up, then continuous
variables, categorical variables as percentages and categorical variables as
counts, each with its table. It is written for the biostatistician choosing
variables for an analysis set, not for a manuscript.

Each section is also a job of its own, and draws through the same function:
the follow-up section is `dp-gfup` (`hvtiPlotR::hv_followup_panels()`) with
the tables of `dc-gfup` (`hvtiRutilities::followup_check()`), and the other
three are `dp-postage` (`hvtiPlotR::hv_eda_pages()`). Given the same choices, a
section here is the same figure as the standalone job, so this report and a
standalone job cannot disagree about the data.

```{r}
#| label: setup
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRutilities::study_root(if (is.null(.in)) getwd() else dirname(.in))
.provenance_data <- list()
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(hvtiRutilities)
  library(hvtiPlotR)
  library(ggplot2)
})
if (utils::packageVersion("hvtiPlotR") < "2.7.17") {
  stop("This job needs hvtiPlotR >= 2.7.17 for hv_eda_pages() and hv_followup_panels(); ",
       utils::packageVersion("hvtiPlotR"), " is installed.", call. = FALSE)
}
if (!requireNamespace("patchwork", quietly = TRUE)) {
  stop("This job needs the patchwork package. Install it, then re-render.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

```{r}
#| label: set
SUBJECT <- "cohort"
TYPE    <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject,
         "', type '", .name_type, "'), but declares SUBJECT = \"", SUBJECT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```

## Study choices

Set the values in this chunk before rendering. Left as they are, they draw the
all-deaths follow-up panel and every column of the `eda` analysis set.

```{r}
#| label: study-choices
# EDIT: the built dataset this job reads. "study" is the file named by built:
# in _study.yml; any other name must be declared under additional_datasets:,
# such as a column subset written for R. Analysis sets are derived from
# "study" only, so another dataset needs ANALYSIS_SET <- NULL.
# EDIT: the analysis set this job reads: declare its columns and excluded rows
# under analysis_sets: in _study.yml, then write it with
# hvtiRdatabuild::write_analysis_set(). Use NULL only when the whole built
# cohort is the intended population.
DATASET <- "study"
ANALYSIS_SET <- "eda"

# Follow-up. These are dp-gfup's choices, and mean what they mean there.
# EDIT: the years-since-origin interval to the operation, and that origin. The
# origin differs between studies, and a wrong one slides every point along the
# x-axis without any other symptom, which is why the operation years are
# checked and printed below.
OPYRS <- "iv_opyrs"
ORIGIN_YEAR <- 1990

# EDIT: the close date of follow-up, as.Date("YYYY-MM-DD"). NULL estimates it
# as the latest operation date plus follow-up in the data, which is the date
# follow-up is known to reach, not the date it was closed; the report says
# which one it drew.
CLOSE_DATE <- NULL

# EDIT: one entry per death panel. `status` is the 1/0 death indicator and
# `time` its follow-up in years. Each panel also gets dc-gfup's tables, over
# the same two columns. A study that ascertains deaths two ways draws both,
# for example
#   systematic = list(status = "deads", time = "iv_deads", title = "Systematic deaths")
PANELS <- list(
  all = list(status = "dead", time = "iv_dead", title = "All deaths")
)

# EDIT: optional event panels, one entry per non-fatal event. `event` and
# `time` are the event's indicator and interval; `death` and `death_time` the
# death it competes with. A flagged event counts only when it comes strictly
# before death, so ties go to death. list() draws none. For example
#   reop = list(event = "ev_reop", time = "iv_reop", death = "dead",
#               death_time = "iv_dead", label = "Reoperation")
EVENTS <- list()

# Variables. These are dp-postage's choices, and mean what they mean there.
# X_VAR is the reference time every panel is drawn against.
# VARIABLES is NULL to draw every column, or the columns to draw, in page
# order. NULL leaves out X_VAR, EXCLUDE and any column that looks like an
# identifier or a date, and the report lists what it left out; name such a
# column here to draw it anyway. The follow-up columns are drawn too, as
# dp-postage draws them, so a section here stays the standalone job's figure.
# A named column that is not in the data stops the render, and the error
# lists every missing name at once.
# SECTIONS is any of "followup", "continuous", "percent" and "count". The
# report keeps this order whatever order they are named in, so the report
# reads the same across studies.
# ALPHA is point transparency, for the follow-up figures and the continuous
# panels: 0.5 lets a dense cohort read as a distribution rather than a blob.
X_VAR <- "year"                    # EDIT: reference time or year
VARIABLES <- NULL                  # EDIT: NULL draws every column, or name them
EXCLUDE <- character(0)
GRID_NCOL <- 4L
GRID_NROW <- 4L
UNIQUE_LIMIT <- 6L
SECTIONS <- c("followup", "continuous", "percent", "count")
ALPHA <- 0.5

# Red is dead and blue is alive, in every job. A non-fatal event is green.
# These are ColorBrewer Set1's red, blue and green, as dp-gfup draws them.
COLOURS <- c(alive = "#377EB8", dead = "#E41A1C", event = "#4DAF4A")
```

## Data

<!-- How this job finds its data. Read this before touching the chunk below.

     A study can offer a job data three ways, and `_study.yml` says which exist:

     1. The study dataset: the file named by `built:`. `DATASET <- "study"`
        with `ANALYSIS_SET <- NULL`.
     2. An additional dataset: any file declared under `additional_datasets:`,
        for example a column subset written for R. Set `DATASET` to its
        name there and `ANALYSIS_SET <- NULL`. It is read as it stands: no
        rows are excluded.
     3. An analysis set: declared under `analysis_sets:` with the columns kept
        and the rows excluded, then written once with
        `hvtiRdatabuild::write_analysis_set()`. Set `ANALYSIS_SET` to its name
        and leave `DATASET` as "study", because a set is always cut from the
        study dataset. The report shows its exclusions in order.

     A name under `additional_datasets:` is not an analysis set, so putting it
     in `ANALYSIS_SET` stops with "No analysis set". Whichever route is taken,
     the report prints the dataset, file and dimensions it read, so a reader
     never has to ask which data a table describes. -->

```{r}
#| label: data
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
if (is.null(ANALYSIS_SET)) {
  if (!is.character(DATASET) || length(DATASET) != 1L || is.na(DATASET) || !nzchar(DATASET)) {
    stop("DATASET must be set to one dataset registered in _study.yml, such as \"study\". ",
         "In a migrated job, see the migration report beside this job.", call. = FALSE)
  }
  .cfg <- study_config(start = .root)
  .data_read <- hvtiRtemplates:::.provenance_read(
    DATASET, .cfg, function() read_built(cfg = .cfg, dataset = DATASET)
  )
  d <- .data_read$value
  .provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(.data_read$record))
} else if (!identical(DATASET, "study")) {
  stop("An analysis set is written from the study dataset, not `", DATASET,
       "`. Set ANALYSIS_SET <- NULL to read `", DATASET, "` whole.", call. = FALSE)
} else {
  if (!requireNamespace("hvtiRdatabuild", quietly = TRUE) || utils::packageVersion("hvtiRdatabuild") < "0.2.1") {
    stop("This job needs hvtiRdatabuild >= 0.2.1 for read_analysis_set().", call. = FALSE)
  }
  .cfg <- study_config(start = .root)
  .analysis_set_path <- file.path(study_dir("datasets", .root), paste0(ANALYSIS_SET, ".parquet"))
  .data_read <- hvtiRtemplates:::.provenance_file_read(
    paste0("analysis_set:", ANALYSIS_SET), .analysis_set_path, .cfg,
    function() hvtiRdatabuild::read_analysis_set(ANALYSIS_SET, cfg = .cfg),
    role = paste0("analysis_set:", ANALYSIS_SET)
  )
  d <- .data_read$value
  .provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(.data_read$record))
  knitr::kable(attr(d, "attrition"), caption = paste0(
    "Analysis set `", ANALYSIS_SET, "`: exclusions, in order"
  ))
}
cat("Data read: ", if (is.null(ANALYSIS_SET)) {
  paste0("dataset `", DATASET, "` (",
         basename(hvtiRutilities::built_path(dataset = DATASET)), ")")
} else {
  paste0("analysis set `", ANALYSIS_SET, "` of the study dataset")
}, ", ", nrow(d), " rows, ", ncol(d), " columns.\n", sep = "")
```


```{r}
#| label: spec
# The follow-up choices are checked by hv_followup_panels() below; these are
# dp-postage's checks of the variable choices, unchanged.
column_names <- function(x) is.character(x) && !anyNA(x) && all(nzchar(x)) && !anyDuplicated(x)
if (!column_names(X_VAR) || length(X_VAR) != 1L) stop("X_VAR must name one column.", call. = FALSE)
if (!is.null(VARIABLES) && (!column_names(VARIABLES) || !length(VARIABLES))) {
  stop("VARIABLES must be NULL or name distinct columns.", call. = FALSE)
}
if (!column_names(EXCLUDE)) stop("EXCLUDE must name distinct columns.", call. = FALSE)
positive_whole <- function(x) is.numeric(x) && length(x) == 1L && !is.na(x) && is.finite(x) && x > 0 && x == floor(x)
for (value in list(GRID_NCOL, GRID_NROW, UNIQUE_LIMIT)) {
  if (!positive_whole(value)) stop("Grid dimensions and UNIQUE_LIMIT must be positive whole numbers.", call. = FALSE)
}
all_sections <- c("followup", "continuous", "percent", "count")
if (!column_names(SECTIONS) || !length(SECTIONS) || !all(SECTIONS %in% all_sections)) {
  stop("SECTIONS must be one or more of \"followup\", \"continuous\", \"percent\" and \"count\".", call. = FALSE)
}
if (!is.numeric(ALPHA) || length(ALPHA) != 1L || is.na(ALPHA) || ALPHA < 0 || ALPHA > 1) {
  stop("ALPHA must be one number from 0 to 1.", call. = FALSE)
}
unknown <- setdiff(c(X_VAR, VARIABLES, EXCLUDE), names(d))
if (length(unknown)) stop("Unknown EDA column(s): ", paste(unknown, collapse = ", "), call. = FALSE)
looks_like_id <- function(v) {
  grepl("(^|_)(id|identifier|date|datetime)($|_)|_dt$", v, ignore.case = TRUE) |
    vapply(d[v], function(x) inherits(x, c("Date", "POSIXt")), logical(1L))
}
if (is.null(VARIABLES)) {
  candidates <- setdiff(names(d), c(X_VAR, EXCLUDE))
  left_out <- candidates[looks_like_id(candidates)]
  VARIABLES <- setdiff(candidates, left_out)
  if (length(left_out)) {
    cat("Left out as identifier or date columns (name them in VARIABLES to draw them):",
        paste(left_out, collapse = ", "), "\n")
  }
} else {
  VARIABLES <- setdiff(VARIABLES, EXCLUDE)
  suspect <- looks_like_id(VARIABLES)
  if (any(suspect)) {
    warning("Selected likely identifier or date field(s): ", paste(VARIABLES[suspect], collapse = ", "), call. = FALSE)
  }
}
if (!length(VARIABLES)) stop("No EDA variables remain after exclusions.", call. = FALSE)

# Labels come from the dataset; a variable without one is shown by its name.
.labels <- label_map(d)
labels <- stats::setNames(.labels$label, .labels$key)

show_table <- function(x, caption, digits = getOption("digits")) {
  cat("\n", knitr::kable(x, caption = caption, digits = digits, row.names = FALSE), "\n", sep = "\n")
}
# Link each saved figure relative to wherever this document sits. Quarto
# rewrites an absolute path into a broken relative one and then cannot embed
# the image, and a fixed "../graphs" would break for a job kept in a subfolder.
# error = FALSE returns the path unchanged when no relative path exists (two
# Windows drives), so the figure is still saved rather than the render stopping.
figure_link <- function(file) {
  xfun::relative_path(file, if (exists(".in") && !is.null(.in)) dirname(.in) else getwd(), error = FALSE)
}
figure_files <- character(0)
```

## Overview

Every column of the data this report read, in dataset order, with its type,
label and share missing. A column the sections below leave out is still
listed here.

```{r}
#| label: overview-contents
#| results: asis
contents <- hvtiRutilities::proc_contents(d, order = "varnum")
show_table(contents$variables[, c("variable", "type", "label", "n_unique", "pct_missing")],
           paste0("Contents: ", nrow(d), " rows, ", ncol(d), " columns"))
```

```{r}
#| label: gfup-heading
#| results: asis
#| eval: !expr '"followup" %in% SECTIONS'
cat("\n## Goodness of follow-up\n\n")
cat("A point below the diagonal is follow-up short of what the close date allows. ",
    "A dead patient's point is where the death was recorded, so it is expected below ",
    "the line; it is the **blue** points drifting below it that need an explanation.\n\n", sep = "")
```

```{r}
#| label: gfup-window
#| results: asis
#| eval: !expr '"followup" %in% SECTIONS'
# hvtiPlotR::hv_followup_panels() checks every panel and event against the data,
# names every missing column at once, refuses a name used twice and the
# wrong-origin mistake, and works out the window the panels share. dp-gfup
# calls it with the same arguments, so these are that job's panels.
fp <- hvtiPlotR::hv_followup_panels(d, opyrs_col = OPYRS, origin_year = ORIGIN_YEAR,
                                    panels = PANELS, events = EVENTS, close_date = CLOSE_DATE)
close_date <- fp$meta$close_date
# Name the edit point a reader can find in this file, not the function's argument.
close_source <- if (is.null(CLOSE_DATE)) fp$meta$close_source else "set in CLOSE_DATE"
show_table(data.frame(
  quantity = c("patients", "operation year missing", "first operation", "last operation",
               "close date", "close date is"),
  value = c(fp$meta$n_obs, fp$meta$n_opyrs_missing, format(fp$meta$first_operation),
            format(fp$meta$study_end), format(close_date), close_source)
), "Operation years and the close date")
```

```{r}
#| label: gfup-panels
#| results: asis
#| eval: !expr '"followup" %in% SECTIONS'
# Each death panel gets dc-gfup's tables over its own two columns, through
# hvtiRutilities::followup_check(), the function dc-gfup calls. No identifier is
# selected, so the report carries no patient-level column; run dc-gfup to review
# the suspicious rows themselves.
plots <- plot(fp, alpha = ALPHA)
panel_counts <- list()
for (i in seq_len(nrow(fp$data))) {
  row <- fp$data[i, ]
  nm <- row$panel
  panel_counts[[nm]] <- list(drawn = row$n_drawn, excluded = row$n_excluded)
  cat("\n### ", row$title, "\n\n", row$n_drawn, " patients drawn; ", row$n_excluded,
      " missing a column this panel needs.\n\n", sep = "")
  if (row$type == "followup") {
    fc <- hvtiRutilities::followup_check(d, event = PANELS[[nm]]$status, followup = PANELS[[nm]]$time)
    show_table(fc$cohort, "Cohort: event, censored and missing event status")
    show_table(fc$intervals, "Follow-up interval: missing, negative and zero values")
    p <- plots[[nm]] +
      scale_colour_manual(values = c(Alive = COLOURS[["alive"]], Dead = COLOURS[["dead"]])) +
      scale_shape_manual(values = c(Alive = 1, Dead = 4))
  } else {
    levels <- c("No event", row$title, "Death")
    p <- plots[[nm]] +
      scale_colour_manual(values = stats::setNames(COLOURS[c("alive", "event", "dead")], levels)) +
      scale_shape_manual(values = stats::setNames(c(1, 2, 4), levels))
  }
  p <- p + labs(x = "Year of operation", y = "Follow-up (years)", colour = NULL, shape = NULL) +
    theme_hv_manuscript()
  file <- set_path("graphs", paste0("dp-eda-gfup-", nm, ".png"))
  ggplot2::ggsave(file, p, width = 7, height = 6, units = "in", dpi = 150)
  cat("![", row$title, "](", figure_link(file), ")\n\n", sep = "")
  figure_files <- c(figure_files, file)
}
```

```{r}
#| label: sections
# One section of hv_eda_pages() pages with its table, drawn exactly as
# dp-postage draws it. One categorical table serves the percent and the count
# sections, shown under whichever comes first, because a table can carry both
# columns where a figure needs two. Missing values are counted as a level, so
# a percentage is out of every patient, not only those with a value.
titles <- c(continuous = "Continuous variables", percent = "Categorical variables, percent",
            count = "Categorical variables, counts")
freq_section <- intersect(c("percent", "count"), SECTIONS)[1L]
draw_section <- function(section) {
  sec <- hvtiPlotR::hv_eda_pages(d, x_col = X_VAR, section = section, vars = VARIABLES,
                                 labels = labels, unique_limit = UNIQUE_LIMIT)
  cat("\n## ", titles[[section]], " (", sec$meta$n_vars, ")\n\n", sep = "")
  if (!sec$meta$n_vars) {
    cat("No selected variable belongs to this section.\n\n")
    return(invisible(character(0)))
  }
  vars <- sec$data$variable
  if (section == "continuous") {
    stats <- hvtiRutilities::proc_means(d, vars = vars, stats = c("n", "nmiss", "mean", "std", "min",
                                                                  "p25", "median", "p75", "max"))
    show_table(stats, "Continuous variables", digits = 2)
  } else if (identical(section, freq_section)) {
    freq <- do.call(rbind, lapply(vars, function(v) {
      f <- hvtiRutilities::proc_freq(d, v, missing = TRUE)
      level <- as.character(f[[1L]])
      data.frame(variable = v, level = ifelse(is.na(level), "(missing)", level),
                 n = f$Frequency, percent = round(f$Percent, 1))
    }))
    show_table(freq, "Categorical variables, missing counted")
  }
  pages <- plot(sec, ncol = GRID_NCOL, nrow = GRID_NROW, alpha = ALPHA)
  files <- character(0)
  for (i in seq_along(pages)) {
    file <- set_path("graphs", sprintf("dp-eda-%s-page-%02d.png", section, i))
    ggplot2::ggsave(file, pages[[i]] & theme_hv_manuscript(base_size = 8),
                    width = 11, height = 8.5, units = "in", dpi = 150)
    on_page <- attr(pages[[i]], "variables")
    cat("\n### ", titles[[section]], ", page ", i, "\n\n![", paste(labels[on_page], collapse = ", "),
        "](", figure_link(file), ")\n\n", sep = "")
    files <- c(files, file)
  }
  invisible(files)
}
```

```{r}
#| label: cont-pages
#| results: asis
#| eval: !expr '"continuous" %in% SECTIONS'
figure_files <- c(figure_files, draw_section("continuous"))
```

```{r}
#| label: pct-pages
#| results: asis
#| eval: !expr '"percent" %in% SECTIONS'
figure_files <- c(figure_files, draw_section("percent"))
```

```{r}
#| label: cnt-pages
#| results: asis
#| eval: !expr '"count" %in% SECTIONS'
figure_files <- c(figure_files, draw_section("count"))
```

```{r}
#| label: provenance
#| echo: false
#| results: asis
cat(hvtiRtemplates:::.embed_provenance(
  .in,
  data = .provenance_data,
  extra = list(
    subject = SUBJECT,
    type = TYPE,
    sections = SECTIONS,
    followup = if ("followup" %in% SECTIONS) {
      list(origin_year = ORIGIN_YEAR, close_date = format(close_date),
           close_date_source = close_source, panels = panel_counts)
    }
  )
))
```
