---
title: "EDA report"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the EDA report this replaces, or delete this line for a new study. -->

Replaces an older EDA report.

A `dp-eda` job is the whole data-checking report over a new build, in one
render: an overview of every column, goodness of follow-up, then continuous
variables, categorical variables as percentages and categorical variables as
counts, each with its table. It is written for the biostatistician choosing
variables for an analysis set, not for a manuscript.

Each section is also a job of its own, and draws through the same function:
the follow-up section is `dp-gfup` (`hvtiPlotR::hv_followup_panels()`) with
the tables of `dc-gfup` (`hvtiRutilities::followup_check()`), and the other
three are `dp-postage` (`hvtiPlotR::hv_eda_pages()`). Given the same choices, a
section here is the same figure as the standalone job, so this report and a
standalone job cannot disagree about the data.

```{r}
#| label: setup
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRtemplates:::.find_study_root(if (is.null(.in)) getwd() else dirname(.in))
.provenance_data <- list()
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(hvtiRutilities)
  library(hvtiPlotR)
  library(ggplot2)
})
if (utils::packageVersion("hvtiPlotR") < "2.8.0") {
  stop("This job needs hvtiPlotR >= 2.8.0 for hv_eda_pages(), hv_followup_panels() and scale_color_hv(); ",
       utils::packageVersion("hvtiPlotR"), " is installed.", call. = FALSE)
}
if (utils::packageVersion("hvtiRutilities") < "1.4.2") {
  stop("This job needs hvtiRutilities >= 1.4.2 for followup_check() and study_abbreviations(); ",
       utils::packageVersion("hvtiRutilities"), " is installed.", call. = FALSE)
}
if (!requireNamespace("patchwork", quietly = TRUE)) {
  stop("This job needs the patchwork package. Install it, then re-render.", call. = FALSE)
}
```

```{r}
#| label: guard-edits
#| results: asis
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

```{r}
#| label: set
SUBJECT <- "cohort"
TYPE    <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject,
         "', type '", .name_type, "'), but declares SUBJECT = \"", SUBJECT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```

## Study choices

Set the values in this chunk before rendering. Left as they are, they draw the
all-deaths follow-up panel and every column of the study dataset.

```{r}
#| label: edit-study-choices
# EDIT: the registered dataset this job reads ("study" is the built dataset).
# MIGRATE-BEGIN: dp-eda-data
DATASET <- "study"
# MIGRATE-END: dp-eda-data

# EDIT: an hvtiRdatabuild analysis set, or NULL to read the whole dataset.
ANALYSIS_SET <- NULL

# EDIT: rows to keep, dplyr::filter() style, or NULL to keep every row:
#   WHERE <- quote(age >= 18)
#   WHERE <- rlang::exprs(age >= 18, hx_chf == 1)
WHERE <- NULL

# EDIT: the patient identifier. Without "ccfid" the job uses MRN, then eMRN;
# name another column, such as "randid", if the study uses one.
ID <- "ccfid"

# EDIT: what makes a row unique; one row per patient unless repeated measures
# add their visit time or date, for example KEY <- c(ID, "iv_echo").
KEY <- ID

# Follow-up. These are dp-gfup's choices, and mean what they mean there.
# EDIT: the years-since-origin interval to the operation, and that origin. The
# origin differs between studies, and a wrong one slides every point along the
# x-axis without any other symptom, which is why the operation years are
# checked and printed below.
OPYRS <- "iv_opyrs"
ORIGIN_YEAR <- 1990

# EDIT: the close date of follow-up, as.Date("YYYY-MM-DD"). NULL estimates it
# as the latest operation date plus follow-up in the data, which is the date
# follow-up is known to reach, not the date it was closed; the report says
# which one it drew.
CLOSE_DATE <- NULL

# EDIT: one entry per death panel. `status` is the 1/0 death indicator and
# `time` its follow-up in years. Each panel also gets dc-gfup's tables, over
# the same two columns. A study that ascertains deaths two ways draws both,
# for example
#   systematic = list(status = "deads", time = "iv_deads", title = "Systematic deaths")
PANELS <- list(
  all = list(status = "dead", time = "iv_dead", title = "All deaths")
)

# EDIT: optional event panels, one entry per non-fatal event. `event` and
# `time` are the event's indicator and interval; `death` and `death_time` the
# death it competes with. A flagged event counts only when it comes strictly
# before death, so ties go to death. list() draws none. For example
#   reop = list(event = "ev_reop", time = "iv_reop", death = "dead",
#               death_time = "iv_dead", label = "Reoperation")
EVENTS <- list()

# Variables. These are dp-postage's choices, and mean what they mean there.
# X_VAR is the reference time every panel is drawn against.
# VARIABLES is NULL to draw every column, or the columns to draw, in page
# order. NULL leaves out X_VAR, EXCLUDE and any column that looks like an
# identifier or a date, and the report lists what it left out; name such a
# column here to draw it anyway. The job's ID and KEY columns are never drawn,
# even when named here. The follow-up columns are drawn too, as
# dp-postage draws them, so a section here stays the standalone job's figure.
# A named column that is not in the data stops the render, and the error
# lists every missing name at once.
# SECTIONS is any of "followup", "continuous", "percent" and "count". The
# report keeps this order whatever order they are named in, so the report
# reads the same across studies.
# ALPHA is point transparency, for the follow-up figures and the continuous
# panels: 0.5 lets a dense cohort read as a distribution rather than a blob.
# MIGRATE-BEGIN: dp-eda-variables
X_VAR <- "year"                    # EDIT: reference time or year
VARIABLES <- NULL                  # EDIT: NULL draws every column, or name them
EXCLUDE <- character(0)
GRID_NCOL <- 4L
GRID_NROW <- 4L
UNIQUE_LIMIT <- 6L
SECTIONS <- c("followup", "continuous", "percent", "count")
ALPHA <- 0.5
# MIGRATE-END: dp-eda-variables

# EDIT: NULL draws the house colors through hvtiPlotR::scale_color_hv(): dead
# vermillion, alive blue and a non-fatal event green, all colorblind safe. Name
# all three to choose your own; c(alive = "#377EB8", dead = "#E41A1C",
# event = "#4DAF4A") is ColorBrewer Set1, what these jobs drew before 1.2.3.
COLORS <- NULL

# EDIT: the longest label the figure pages and their captions show, in
# characters. Longer labels are shortened so that two labels that differ never
# read the same: a shared heading is abbreviated, then the label is cut and
# marked. Inf shows every label whole. Tables keep full labels, so a reader can
# look up what a shortened one stands for.
LABEL_MAX <- 40

# EDIT: this job's own abbreviations, beyond the study's list, as
# c("Phrase" = "Abbrev"); NA drops one the study or group list supplies. NULL
# uses the study's list alone. The study's list is the abbreviations: block in
# _study.yml, over the group default shipped in hvtiRutilities.
ABBREVIATIONS <- NULL
```

## Data

<!-- How this job finds its data. Read this before touching the chunk below.

     A study can offer a job data three ways, and `_study.yml` says which exist:

     1. The study dataset: the file named by `built:`. `DATASET <- "study"`
        with `ANALYSIS_SET <- NULL`.
     2. An additional dataset: any file declared under `additional_datasets:`,
        for example a column subset written for R. Set `DATASET` to its
        name there and `ANALYSIS_SET <- NULL`. It is read as it stands: only
        `WHERE` removes rows.
     3. An analysis set: declared under `analysis_sets:` with the columns kept
        and the rows excluded, then written once with
        `hvtiRdatabuild::write_analysis_set()`. Set `ANALYSIS_SET` to its name
        and leave `DATASET` as "study", because a set is always cut from the
        study dataset. The report shows its exclusions in order.

     A name under `additional_datasets:` is not an analysis set, so putting it
     in `ANALYSIS_SET` stops with "No analysis set". Whichever route is taken,
     the report prints what it read, the identifier it used, the rows each
     `WHERE` condition removed and the rows and patients kept, so a reader
     never has to ask which data a table describes. -->

```{r}
#| label: data
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
.cfg <- study_config(start = .root)
job_data <- hvtiRtemplates::read_job_data(.cfg, dataset = DATASET, analysis_set = ANALYSIS_SET,
                                          where = WHERE, id = ID, key = KEY)
d <- job_data$data
.provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(job_data$provenance))
knitr::kable(job_data$record, col.names = c("Data", ""), caption = "The data this job read")
if (!is.null(job_data$attrition)) {
  knitr::kable(job_data$attrition, caption = paste0("Analysis set `", ANALYSIS_SET, "`: exclusions, in order"))
}
```


```{r}
#| label: spec
# The follow-up choices are checked by hv_followup_panels() below; these are
# dp-postage's checks of the variable choices, unchanged.
column_names <- function(x) is.character(x) && !anyNA(x) && all(nzchar(x)) && !anyDuplicated(x)
if (!column_names(X_VAR) || length(X_VAR) != 1L) stop("X_VAR must name one column.", call. = FALSE)
if (!is.null(VARIABLES) && (!column_names(VARIABLES) || !length(VARIABLES))) {
  stop("VARIABLES must be NULL or name distinct columns.", call. = FALSE)
}
if (!column_names(EXCLUDE)) stop("EXCLUDE must name distinct columns.", call. = FALSE)
positive_whole <- function(x) is.numeric(x) && length(x) == 1L && !is.na(x) && is.finite(x) && x > 0 && x == floor(x)
for (value in list(GRID_NCOL, GRID_NROW, UNIQUE_LIMIT)) {
  if (!positive_whole(value)) stop("Grid dimensions and UNIQUE_LIMIT must be positive whole numbers.", call. = FALSE)
}
all_sections <- c("followup", "continuous", "percent", "count")
if (!column_names(SECTIONS) || !length(SECTIONS) || !all(SECTIONS %in% all_sections)) {
  stop("SECTIONS must be one or more of \"followup\", \"continuous\", \"percent\" and \"count\".", call. = FALSE)
}
if (!is.numeric(ALPHA) || length(ALPHA) != 1L || is.na(ALPHA) || ALPHA < 0 || ALPHA > 1) {
  stop("ALPHA must be one number from 0 to 1.", call. = FALSE)
}
if (!is.numeric(LABEL_MAX) || length(LABEL_MAX) != 1L || is.na(LABEL_MAX) || LABEL_MAX < 4) {
  stop("LABEL_MAX must be one number, at least 4, or Inf to show every label whole.", call. = FALSE)
}
if (!is.null(ABBREVIATIONS) && (!(is.character(ABBREVIATIONS) || is.list(ABBREVIATIONS)) ||
                                  (length(ABBREVIATIONS) && is.null(names(ABBREVIATIONS))))) {
  stop('ABBREVIATIONS must be NULL or name each abbreviation by its phrase, c("Phrase" = "Abbrev").', call. = FALSE)
}
unknown <- setdiff(c(X_VAR, VARIABLES, EXCLUDE), names(d))
if (length(unknown)) stop("Unknown EDA column(s): ", paste(unknown, collapse = ", "), call. = FALSE)
# The job's ID and KEY columns are never drawn: they are the columns the data
# step used, so a ccfid, or the MRN it fell back to, is caught whatever its case.
# MRN and eMRN are already dropped at read, unless one is the ID. The name rule
# catches other identifier columns: ccfid and patientid have no "_" before "id",
# and a bare "id$" would also take carotid and steroid. The date rule and the
# data rule catch columns that should not be drawn: dates, and text in which
# every value differs, once there are ten or more of them.
.job_selection <- attr(job_data$record, "selection")
is_job_key <- function(v) tolower(v) %in% tolower(c(.job_selection$id, .job_selection$key))
looks_like_id <- function(v) {
  is_job_key(v) |
    grepl("^(ccf|pat|patient|study|subject|record|case)_?(id|num|no)$|^e?mrn$", v, ignore.case = TRUE) |
    grepl("(^|_)(date|datetime)($|_)|_dt$", v, ignore.case = TRUE) |
    vapply(d[v], function(x) {
      seen <- x[!is.na(x)]
      inherits(x, c("Date", "POSIXt")) ||
        ((is.character(x) || is.factor(x)) && length(seen) >= 10L && !anyDuplicated(seen))
    }, logical(1L))
}
if (is.null(VARIABLES)) {
  candidates <- setdiff(names(d), c(X_VAR, EXCLUDE))
  left_out <- candidates[looks_like_id(candidates)]
  VARIABLES <- setdiff(candidates, left_out)
  if (length(left_out)) {
    cat("Left out as identifier or date columns (name them in VARIABLES to draw them):",
        paste(left_out, collapse = ", "), "\n")
  }
} else {
  VARIABLES <- setdiff(VARIABLES, EXCLUDE)
  if (any(is_job_key(VARIABLES))) {
    warning("Not drawn, as the job's ID or KEY: ", paste(VARIABLES[is_job_key(VARIABLES)], collapse = ", "),
            call. = FALSE)
    VARIABLES <- VARIABLES[!is_job_key(VARIABLES)]
  }
  suspect <- looks_like_id(VARIABLES)
  if (any(suspect)) {
    warning("Selected likely identifier or date field(s): ", paste(VARIABLES[suspect], collapse = ", "), call. = FALSE)
  }
}
if (!length(VARIABLES)) stop("No EDA variables remain after exclusions.", call. = FALSE)

# Labels come from the dataset; a variable without one is shown by its name.
# Shortened labels stay distinct, using the job's, the study's and the group's
# abbreviations, merged and checked by study_abbreviations().
.abbreviations <- hvtiRutilities::study_abbreviations(.cfg, extra = ABBREVIATIONS)
.labels <- label_map(d, label_max = LABEL_MAX, abbreviations = .abbreviations)
labels <- stats::setNames(.labels$label, .labels$key)
# The abbreviations these variables' labels show, as a key under the section.
# Only shortened labels count: a label that says "SP" in its own words is not
# using the key's SP.
abbreviation_key <- function(vars) {
  used <- attr(.labels, "abbreviations")
  row <- match(vars, .labels$key)
  shown <- .labels$label[row][.labels$label[row] != .labels$label_full[row]]
  hit <- vapply(used$abbreviation, function(a) {
    any(grepl(paste0("(?<![[:alnum:]])", gsub("([][{}()+*^$|\\\\?.])", "\\\\\\1", a), "(?![[:alnum:]])"), shown, perl = TRUE))
  }, logical(1L))
  if (any(hit)) {
    cat("\nAbbreviations: ", paste(used$abbreviation[hit], used$expansion[hit], sep = " = ", collapse = "; "), ".\n\n",
        sep = "")
  }
}

show_table <- function(x, caption, digits = getOption("digits")) {
  cat("\n", knitr::kable(x, caption = caption, digits = digits, row.names = FALSE), "\n", sep = "\n")
}
# Link each saved figure relative to wherever this document sits. Quarto
# rewrites an absolute path into a broken relative one and then cannot embed
# the image, and a fixed "../graphs" would break for a job kept in a subfolder.
# error = FALSE returns the path unchanged when no relative path exists (two
# Windows drives), so the figure is still saved rather than the render stopping.
figure_link <- function(file) {
  xfun::relative_path(file, if (exists(".in") && !is.null(.in)) dirname(.in) else getwd(), error = FALSE)
}
figure_files <- character(0)
```

## Overview

Every column of the data this report read, in dataset order, with its type,
label and share missing. A column the sections below leave out is still
listed here, unless it looks like an identifier: those are named below the
table and not described.

```{r}
#| label: overview-contents
#| results: asis
# The same rule the sections use, applied whatever VARIABLES says, less its date
# half: a date's share missing is worth seeing, an identifier's is not.
is_date <- vapply(d, inherits, logical(1L), c("Date", "POSIXt")) |
  grepl("(^|_)(date|datetime)($|_)|_dt$", names(d), ignore.case = TRUE)
id_cols <- names(d)[looks_like_id(names(d)) & !is_date]
contents <- hvtiRutilities::proc_contents(d, order = "varnum")
shown <- contents$variables[!contents$variables$variable %in% id_cols, ]
show_table(shown[, c("variable", "type", "label", "n_unique", "pct_missing")],
           paste0("Contents: ", nrow(d), " rows, ", ncol(d), " columns"))
if (length(id_cols)) cat("Identifier columns, not described:", paste(id_cols, collapse = ", "), "\n")
```

```{r}
#| label: gfup-heading
#| results: asis
#| eval: !expr '"followup" %in% SECTIONS'
cat("\n## Goodness of follow-up\n\n")
cat("A point below the diagonal is follow-up short of what the close date allows. ",
    "A dead patient's point is where the death was recorded, so it is expected below ",
    "the line; it is the **blue** points drifting below it that need an explanation.\n\n", sep = "")
```

```{r}
#| label: gfup-window
#| results: asis
#| eval: !expr '"followup" %in% SECTIONS'
# hvtiPlotR::hv_followup_panels() checks every panel and event against the data,
# names every missing column at once, refuses a name used twice and the
# wrong-origin mistake, and works out the window the panels share. dp-gfup
# calls it with the same arguments, so these are that job's panels.
fp <- hvtiPlotR::hv_followup_panels(d, opyrs_col = OPYRS, origin_year = ORIGIN_YEAR,
                                    panels = PANELS, events = EVENTS, close_date = CLOSE_DATE)
close_date <- fp$meta$close_date
# Name the edit point a reader can find in this file, not the function's argument.
close_source <- if (is.null(CLOSE_DATE)) fp$meta$close_source else "set in CLOSE_DATE"
show_table(data.frame(
  quantity = c("patients", "operation year missing", "first operation", "last operation",
               "close date", "close date is"),
  value = c(fp$meta$n_obs, fp$meta$n_opyrs_missing, format(fp$meta$first_operation),
            format(fp$meta$study_end), format(close_date), close_source)
), "Operation years and the close date")
```

```{r}
#| label: gfup-panels
#| results: asis
#| eval: !expr '"followup" %in% SECTIONS'
# Each death panel gets dc-gfup's tables over its own two columns, through
# hvtiRutilities::followup_check(), the function dc-gfup calls. No identifier is
# selected, so the report carries no patient-level column; run dc-gfup to review
# the suspicious rows themselves.
if (exists("COLOURS", inherits = FALSE)) stop("COLOURS is now COLORS: rename it in edit-study-choices ",
                                              "(study-choices in a job made before that rename).", call. = FALSE)
if (!is.null(COLORS) && !(is.character(COLORS) && !anyNA(COLORS) && length(COLORS) == 3L &&
                            setequal(names(COLORS), c("alive", "dead", "event")))) {
  stop("COLORS must be NULL or name exactly alive, dead and event.", call. = FALSE)
}
# Dead or Death is the event and Alive or No event the censored level, so a
# non-fatal event takes the house rule's next color, green.
status_colors <- function(levels) {
  if (is.null(COLORS)) return(scale_color_hv(event = c("Dead", "Death"), censored = c("Alive", "No event")))
  keys <- if (length(levels) == 2L) c("alive", "dead") else c("alive", "event", "dead")
  scale_color_manual(values = stats::setNames(COLORS[keys], levels))
}
plots <- plot(fp, alpha = ALPHA)
panel_counts <- list()
for (i in seq_len(nrow(fp$data))) {
  row <- fp$data[i, ]
  nm <- row$panel
  panel_counts[[nm]] <- list(drawn = row$n_drawn, excluded = row$n_excluded)
  cat("\n### ", row$title, "\n\n", row$n_drawn, " patients drawn; ", row$n_excluded,
      " missing a column this panel needs.\n\n", sep = "")
  if (row$type == "followup") {
    fc <- hvtiRutilities::followup_check(d, event = PANELS[[nm]]$status, followup = PANELS[[nm]]$time)
    show_table(fc$cohort, "Cohort: event, censored and missing event status")
    show_table(fc$intervals, "Follow-up interval: missing, negative and zero values")
    p <- plots[[nm]] +
      status_colors(c("Alive", "Dead")) +
      scale_shape_manual(values = c(Alive = 1, Dead = 4))
  } else {
    levels <- c("No event", row$title, "Death")
    p <- plots[[nm]] +
      status_colors(levels) +
      scale_shape_manual(values = stats::setNames(c(1, 2, 4), levels))
  }
  p <- p + labs(x = "Year of operation", y = "Follow-up (years)", color = NULL, shape = NULL) +
    theme_hv_manuscript()
  file <- set_path("graphs", paste0("dp-eda-gfup-", nm, ".png"))
  ggplot2::ggsave(file, p, width = 7, height = 6, units = "in", dpi = 150)
  cat("![", row$title, "](", figure_link(file), ")\n\n", sep = "")
  figure_files <- c(figure_files, file)
}
```

```{r}
#| label: sections
# One section of hv_eda_pages() pages with its table, drawn exactly as
# dp-postage draws it. One categorical table serves the percent and the count
# sections, shown under whichever comes first, because a table can carry both
# columns where a figure needs two. Missing values are counted as a level, so
# a percentage is out of every patient, not only those with a value.
titles <- c(continuous = "Continuous variables", percent = "Categorical variables, percent",
            count = "Categorical variables, counts")
freq_section <- intersect(c("percent", "count"), SECTIONS)[1L]
draw_section <- function(section) {
  sec <- hvtiPlotR::hv_eda_pages(d, x_col = X_VAR, section = section, vars = VARIABLES,
                                 labels = labels, unique_limit = UNIQUE_LIMIT)
  cat("\n## ", titles[[section]], " (", sec$meta$n_vars, ")\n\n", sep = "")
  if (!sec$meta$n_vars) {
    cat("No selected variable belongs to this section.\n\n")
    return(invisible(character(0)))
  }
  vars <- sec$data$variable
  if (section == "continuous") {
    stats <- hvtiRutilities::proc_means(d, vars = vars, stats = c("n", "nmiss", "mean", "std", "min",
                                                                  "p25", "median", "p75", "max"))
    show_table(stats, "Continuous variables", digits = 2)
  } else if (identical(section, freq_section)) {
    freq <- do.call(rbind, lapply(vars, function(v) {
      f <- hvtiRutilities::proc_freq(d, v, missing = TRUE)
      level <- as.character(f[[1L]])
      data.frame(variable = v, level = ifelse(is.na(level), "(missing)", level),
                 n = f$Frequency, percent = round(f$Percent, 1))
    }))
    show_table(freq, "Categorical variables, missing counted")
  }
  pages <- plot(sec, ncol = GRID_NCOL, nrow = GRID_NROW, alpha = ALPHA)
  files <- character(0)
  for (i in seq_along(pages)) {
    file <- set_path("graphs", sprintf("dp-eda-%s-page-%02d.png", section, i))
    ggplot2::ggsave(file, pages[[i]] & scale_fill_hv() & theme_hv_manuscript(base_size = 8),
                    width = 11, height = 8.5, units = "in", dpi = 150)
    on_page <- attr(pages[[i]], "variables")
    cat("\n### ", titles[[section]], ", page ", i, "\n\n![", paste(labels[on_page], collapse = ", "),
        "](", figure_link(file), ")\n\n", sep = "")
    files <- c(files, file)
  }
  abbreviation_key(vars)
  invisible(files)
}
```

```{r}
#| label: cont-pages
#| results: asis
#| eval: !expr '"continuous" %in% SECTIONS'
figure_files <- c(figure_files, draw_section("continuous"))
```

```{r}
#| label: pct-pages
#| results: asis
#| eval: !expr '"percent" %in% SECTIONS'
figure_files <- c(figure_files, draw_section("percent"))
```

```{r}
#| label: cnt-pages
#| results: asis
#| eval: !expr '"count" %in% SECTIONS'
figure_files <- c(figure_files, draw_section("count"))
```

```{r}
#| label: provenance
#| echo: false
#| results: asis
cat(hvtiRtemplates:::.embed_provenance(
  .in,
  data = .provenance_data,
  extra = list(
    subject = SUBJECT,
    type = TYPE,
    label_max = LABEL_MAX,
    abbreviations = data.frame(phrase = names(.abbreviations), abbreviation = unname(.abbreviations),
                               source = attr(.abbreviations, "source")),
    sections = SECTIONS,
    followup = if ("followup" %in% SECTIONS) {
      list(origin_year = ORIGIN_YEAR, close_date = format(close_date),
           close_date_source = close_source, panels = panel_counts)
    }
  )
))
```
