---
title: "EDA postage stamps"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (tp.dp.EDA_barplots_scatterplots*.R). -->

Replaces an older EDA bar-plot and scatter-plot sweep.

A `dp-postage` job is the data-checking sweep over a new build: one small panel
per variable against operation year, pages of them, with a table beside each
section. It is for finding coding errors, drift and missingness before any
analysis. It is not a figure for a manuscript.

The sweep has three sections, and `SECTIONS` picks which to draw: continuous
variables as points, then categorical variables as percentages, then the same
categorical variables as counts. The percent and count pages bin the year the
same way, so their bars line up. The pages come from `hvtiPlotR::hv_eda_pages()`,
the same function the EDA report draws through, so a section here is the same
figure as that section of the report.

```{r}
#| label: setup
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRutilities::study_root(if (is.null(.in)) getwd() else dirname(.in))
.provenance_data <- list()
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(hvtiRutilities)
  library(hvtiPlotR)
  library(ggplot2)
})
if (utils::packageVersion("hvtiPlotR") < "2.7.16") {
  stop("This job needs hvtiPlotR >= 2.7.16 for hv_eda_pages(); ",
       utils::packageVersion("hvtiPlotR"), " is installed.", call. = FALSE)
}
if (!requireNamespace("patchwork", quietly = TRUE)) {
  stop("This job needs the patchwork package. Install it, then re-render.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

```{r}
#| label: set
SUBJECT <- "cohort"
TYPE    <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject,
         "', type '", .name_type, "'), but declares SUBJECT = \"", SUBJECT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```

## Study choices

Set the values in this chunk before rendering.

```{r}
#| label: study-choices
# EDIT: the built dataset this job reads. "study" is the file named by built:
# in _study.yml; any other name must be declared under additional_datasets:,
# such as a column subset written for R. Analysis sets are derived from
# "study" only, so another dataset needs ANALYSIS_SET <- NULL.
# EDIT: the analysis set this job reads: declare its columns and excluded rows
# under analysis_sets: in _study.yml, then write it with
# hvtiRdatabuild::write_analysis_set(). Use NULL only when the whole built
# cohort is the intended population.
# VARIABLES is NULL to draw every column, or the columns to draw, in page
# order. NULL leaves out X_VAR, EXCLUDE and any column that looks like an
# identifier or a date, and the report lists what it left out; name such a
# column here to draw it anyway. A named column that is not in the data stops
# the render, and the error lists every missing name at once.
# SECTIONS is any of "continuous", "percent" and "count", in the order to
# draw them. ALPHA is point transparency for continuous panels: 0.5 lets a dense
# cohort read as a distribution rather than a blob.
# MIGRATE-BEGIN: dp-postage-config
DATASET <- "study"
ANALYSIS_SET <- "eda"
X_VAR <- "year"                    # EDIT: reference time or year
VARIABLES <- NULL                  # EDIT: NULL draws every column, or name them
EXCLUDE <- character(0)
GRID_NCOL <- 4L
GRID_NROW <- 4L
UNIQUE_LIMIT <- 6L
SECTIONS <- c("continuous", "percent", "count")
ALPHA <- 0.5
# MIGRATE-END: dp-postage-config

```

## Data

<!-- How this job finds its data. Read this before touching the chunk below.

     A study can offer a job data three ways, and `_study.yml` says which exist:

     1. The study dataset: the file named by `built:`. `DATASET <- "study"`
        with `ANALYSIS_SET <- NULL`.
     2. An additional dataset: any file declared under `additional_datasets:`,
        for example a column subset written for R. Set `DATASET` to its
        name there and `ANALYSIS_SET <- NULL`. It is read as it stands: no
        rows are excluded.
     3. An analysis set: declared under `analysis_sets:` with the columns kept
        and the rows excluded, then written once with
        `hvtiRdatabuild::write_analysis_set()`. Set `ANALYSIS_SET` to its name
        and leave `DATASET` as "study", because a set is always cut from the
        study dataset. The report shows its exclusions in order.

     A name under `additional_datasets:` is not an analysis set, so putting it
     in `ANALYSIS_SET` stops with "No analysis set". Whichever route is taken,
     the report prints the dataset, file and dimensions it read, so a reader
     never has to ask which data a table describes. -->

```{r}
#| label: data
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
if (is.null(ANALYSIS_SET)) {
  if (!is.character(DATASET) || length(DATASET) != 1L || is.na(DATASET) || !nzchar(DATASET)) {
    stop("DATASET must be set to one dataset registered in _study.yml, such as \"study\". ",
         "In a migrated job, see the migration report beside this job.", call. = FALSE)
  }
  .cfg <- study_config(start = .root)
  .data_read <- hvtiRtemplates:::.provenance_read(
    DATASET, .cfg, function() read_built(cfg = .cfg, dataset = DATASET)
  )
  d <- .data_read$value
  .provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(.data_read$record))
} else if (!identical(DATASET, "study")) {
  stop("An analysis set is written from the study dataset, not `", DATASET,
       "`. Set ANALYSIS_SET <- NULL to read `", DATASET, "` whole.", call. = FALSE)
} else {
  if (!requireNamespace("hvtiRdatabuild", quietly = TRUE) || utils::packageVersion("hvtiRdatabuild") < "0.2.1") {
    stop("This job needs hvtiRdatabuild >= 0.2.1 for read_analysis_set().", call. = FALSE)
  }
  .cfg <- study_config(start = .root)
  .analysis_set_path <- file.path(study_dir("datasets", .root), paste0(ANALYSIS_SET, ".parquet"))
  .data_read <- hvtiRtemplates:::.provenance_file_read(
    paste0("analysis_set:", ANALYSIS_SET), .analysis_set_path, .cfg,
    function() hvtiRdatabuild::read_analysis_set(ANALYSIS_SET, cfg = .cfg),
    role = paste0("analysis_set:", ANALYSIS_SET)
  )
  d <- .data_read$value
  .provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(.data_read$record))
  knitr::kable(attr(d, "attrition"), caption = paste0(
    "Analysis set `", ANALYSIS_SET, "`: exclusions, in order"
  ))
}
cat("Data read: ", if (is.null(ANALYSIS_SET)) {
  paste0("dataset `", DATASET, "` (",
         basename(hvtiRutilities::built_path(dataset = DATASET)), ")")
} else {
  paste0("analysis set `", ANALYSIS_SET, "` of the study dataset")
}, ", ", nrow(d), " rows, ", ncol(d), " columns.\n", sep = "")
```

```{r}
#| label: spec
# Variable order is page order. Character and factor columns become bars.
# Numeric columns become bars only for a small set of non-negative integers;
# other numeric columns become points.
column_names <- function(x) is.character(x) && !anyNA(x) && all(nzchar(x)) && !anyDuplicated(x)
if (!column_names(X_VAR) || length(X_VAR) != 1L) stop("X_VAR must name one column.", call. = FALSE)
if (!is.null(VARIABLES) && (!column_names(VARIABLES) || !length(VARIABLES))) {
  stop("VARIABLES must be NULL or name distinct columns.", call. = FALSE)
}
if (!column_names(EXCLUDE)) stop("EXCLUDE must name distinct columns.", call. = FALSE)
positive_whole <- function(x) is.numeric(x) && length(x) == 1L && !is.na(x) && is.finite(x) && x > 0 && x == floor(x)
for (value in list(GRID_NCOL, GRID_NROW, UNIQUE_LIMIT)) {
  if (!positive_whole(value)) stop("Grid dimensions and UNIQUE_LIMIT must be positive whole numbers.", call. = FALSE)
}
if (!column_names(SECTIONS) || !length(SECTIONS) || !all(SECTIONS %in% c("continuous", "percent", "count"))) {
  stop("SECTIONS must be one or more of \"continuous\", \"percent\" and \"count\".", call. = FALSE)
}
if (!is.numeric(ALPHA) || length(ALPHA) != 1L || is.na(ALPHA) || ALPHA < 0 || ALPHA > 1) {
  stop("ALPHA must be one number from 0 to 1.", call. = FALSE)
}
unknown <- setdiff(c(X_VAR, VARIABLES, EXCLUDE), names(d))
if (length(unknown)) stop("Unknown EDA column(s): ", paste(unknown, collapse = ", "), call. = FALSE)
# A name rule and a data rule. The names catch ccfid and patientid, which have no
# "_" before "id"; a bare "id$" would also take carotid and steroid. The data rule
# catches an identifier under a name nobody listed: text in which every value
# differs, once there are ten or more of them.
looks_like_id <- function(v) {
  grepl("(^|_)(id|identifier|date|datetime)($|_)|_dt$", v, ignore.case = TRUE) |
    grepl("^((ccf|pat|patient|study|subject|record|case)_?(id|num|no)|mrn)$", v, ignore.case = TRUE) |
    vapply(d[v], function(x) {
      seen <- x[!is.na(x)]
      inherits(x, c("Date", "POSIXt")) ||
        ((is.character(x) || is.factor(x)) && length(seen) >= 10L && !anyDuplicated(seen))
    }, logical(1L))
}
if (is.null(VARIABLES)) {
  candidates <- setdiff(names(d), c(X_VAR, EXCLUDE))
  left_out <- candidates[looks_like_id(candidates)]
  VARIABLES <- setdiff(candidates, left_out)
  if (length(left_out)) {
    cat("Left out as identifier or date columns (name them in VARIABLES to draw them):",
        paste(left_out, collapse = ", "), "\n")
  }
} else {
  VARIABLES <- setdiff(VARIABLES, EXCLUDE)
  suspect <- looks_like_id(VARIABLES)
  if (any(suspect)) {
    warning("Selected likely identifier or date field(s): ", paste(VARIABLES[suspect], collapse = ", "), call. = FALSE)
  }
}
if (!length(VARIABLES)) stop("No EDA variables remain after exclusions.", call. = FALSE)
```

## Pages

Each section opens with its table: summary statistics for the continuous
variables, and one frequency table for the categorical variables, shown once
under whichever categorical section comes first. Missing values are counted as a
level, so a percentage is out of every patient, not only those with a value.

```{r}
#| label: pages
#| results: asis
# Labels come from the dataset; a variable without one is shown by its name.
.labels <- label_map(d)
labels <- stats::setNames(.labels$label, .labels$key)
titles <- c(continuous = "Continuous variables", percent = "Categorical variables, percent",
            count = "Categorical variables, counts")
page_files <- character(0)
freq_shown <- FALSE
for (section in SECTIONS) {
  sec <- hvtiPlotR::hv_eda_pages(d, x_col = X_VAR, section = section, vars = VARIABLES,
                                 labels = labels, unique_limit = UNIQUE_LIMIT)
  cat("\n## ", titles[[section]], " (", sec$meta$n_vars, ")\n\n", sep = "")
  if (!sec$meta$n_vars) {
    cat("No selected variable belongs to this section.\n\n")
    next
  }
  vars <- sec$data$variable
  if (section == "continuous") {
    stats <- hvtiRutilities::proc_means(d, vars = vars, stats = c("n", "nmiss", "mean", "std", "min",
                                                                  "p25", "median", "p75", "max"))
    cat(knitr::kable(stats, digits = 2, row.names = FALSE, caption = "Continuous variables"), sep = "\n")
  } else if (!freq_shown) {
    freq <- do.call(rbind, lapply(vars, function(v) {
      f <- hvtiRutilities::proc_freq(d, v, missing = TRUE)
      level <- as.character(f[[1L]])
      data.frame(variable = v, level = ifelse(is.na(level), "(missing)", level),
                 n = f$Frequency, percent = round(f$Percent, 1))
    }))
    cat(knitr::kable(freq, row.names = FALSE, caption = "Categorical variables, missing counted"), sep = "\n")
    freq_shown <- TRUE
  }
  pages <- plot(sec, ncol = GRID_NCOL, nrow = GRID_NROW, alpha = ALPHA)
  for (i in seq_along(pages)) {
    file <- set_path("graphs", sprintf("dp-postage-%s-page-%02d.png", section, i))
    ggplot2::ggsave(file, pages[[i]] & theme_hv_manuscript(base_size = 8),
                    width = 11, height = 8.5, units = "in", dpi = 150)
    on_page <- attr(pages[[i]], "variables")
    # Link relative to wherever this document sits. Quarto rewrites an absolute
    # path into a broken relative one and then cannot embed the image, and a
    # fixed "../graphs" would break for a job kept in a subfolder. error = FALSE
    # returns the path unchanged when no relative path exists (two Windows
    # drives), so the page is still saved rather than the render stopping.
    link <- xfun::relative_path(file, if (exists(".in") && !is.null(.in)) dirname(.in) else getwd(),
                                error = FALSE)
    cat("\n### Page ", i, "\n\n![", paste(labels[on_page], collapse = ", "), "](", link, ")\n\n", sep = "")
    page_files <- c(page_files, file)
  }
}
invisible(page_files)
```

```{r}
#| label: provenance
#| echo: false
#| results: asis
cat(hvtiRtemplates:::.embed_provenance(
  .in,
  data = .provenance_data,
  extra = list(
    subject = SUBJECT,
    type = TYPE
  )
))
```
