---
title: "EDA postage stamps"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (tp.dp.EDA_barplots_scatterplots*.R). -->

Replaces an older EDA bar-plot and scatter-plot sweep.

A `dp-postage` job is the data-checking sweep over a new build: one small panel
per variable against operation year, bars for categorical variables and points
for continuous ones, pages of them. It is for finding coding errors, drift and
missingness before any analysis. It is not a figure for a manuscript.

```{r}
#| label: setup
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(hvtiRutilities)
  library(hvtiPlotR)
  library(ggplot2)
})
if (utils::packageVersion("hvtiRdatabuild") < "0.2.1") {
  stop("This job needs hvtiRdatabuild >= 0.2.1 for read_analysis_set(); ",
       utils::packageVersion("hvtiRdatabuild"), " is installed.", call. = FALSE)
}
if (utils::packageVersion("hvtiPlotR") < "2.7.14") {
  stop("This job needs hvtiPlotR >= 2.7.14 for hv_eda(); ",
       utils::packageVersion("hvtiPlotR"), " is installed.", call. = FALSE)
}
if (!requireNamespace("patchwork", quietly = TRUE)) {
  stop("This job needs the patchwork package. Install it, then re-render.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

```{r}
#| label: set
ENDPOINT <- "cohort"
TYPE     <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint,
         "', type '", .name_type, "'), but declares ENDPOINT = \"", ENDPOINT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```

## Data

<!-- How this job finds its data. Read this before touching the chunk below.

     A study can offer a job data three ways, and `_study.yml` says which exist:

     1. The study dataset: the file named by `built:`. `DATASET <- "study"`
        with `ANALYSIS_SET <- NULL`.
     2. An additional dataset: any file declared under `additional_datasets:`,
        for example a column subset written for R. Set `DATASET` to its
        name there and `ANALYSIS_SET <- NULL`. It is read as it stands: no
        rows are excluded.
     3. An analysis set: declared under `analysis_sets:` with the columns kept
        and the rows excluded, then written once with
        `hvtiRdatabuild::write_analysis_set()`. Set `ANALYSIS_SET` to its name
        and leave `DATASET` as "study", because a set is always cut from the
        study dataset. The report shows its exclusions in order.

     A name under `additional_datasets:` is not an analysis set, so putting it
     in `ANALYSIS_SET` stops with "No analysis set". Whichever route is taken,
     the report prints the dataset, file and dimensions it read, so a reader
     never has to ask which data a table describes. -->

```{r}
#| label: data
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
# EDIT: the built dataset this job reads. "study" is the file named by built:
# in _study.yml; any other name must be declared under additional_datasets:,
# such as a column subset written for R. Analysis sets are derived from
# "study" only, so another dataset needs ANALYSIS_SET <- NULL.
DATASET <- "study"
# EDIT: the analysis set this job reads: declare its columns and excluded rows
# under analysis_sets: in _study.yml, then write it with
# hvtiRdatabuild::write_analysis_set(). Use NULL only when the whole built
# cohort is the intended population.
ANALYSIS_SET <- "eda"
if (is.null(ANALYSIS_SET)) {
  d <- read_built(dataset = DATASET)
} else if (!identical(DATASET, "study")) {
  stop("An analysis set is written from the study dataset, not `", DATASET,
       "`. Set ANALYSIS_SET <- NULL to read `", DATASET, "` whole.", call. = FALSE)
} else {
  d <- hvtiRdatabuild::read_analysis_set(ANALYSIS_SET)
  knitr::kable(attr(d, "attrition"), caption = paste0(
    "Analysis set `", ANALYSIS_SET, "`: exclusions, in order"
  ))
}
cat("Data read: ", if (is.null(ANALYSIS_SET)) {
  paste0("dataset `", DATASET, "` (",
         basename(hvtiRutilities::built_path(dataset = DATASET)), ")")
} else {
  paste0("analysis set `", ANALYSIS_SET, "` of the study dataset")
}, ", ", nrow(d), " rows, ", ncol(d), " columns.\n", sep = "")
```

```{r}
#| label: spec
# EDIT: the x-axis for every panel, usually operation year.
X_COL <- "year"
# EDIT: the variables to sweep, named by column with a display label. Order is
# page order. Character and factor columns become bars. Numeric columns become
# bars only when they have at most 6 distinct non-missing values and every value
# is a non-negative whole number no greater than 6; all other numeric columns
# become scatter panels. A column therefore does not need a type here.
VARS <- c(age = "Age (years)", female = "Female")
# Panels per page. 3 x 3 keeps a label readable on a printed page.
NCOL <- 3L
NROW <- 3L
```

## Pages

```{r}
#| label: pages
#| results: asis
unknown <- setdiff(c(X_COL, names(VARS)), names(d))
if (length(unknown)) {
  stop("Unknown EDA column(s): ", paste(unknown, collapse = ", "), call. = FALSE)
}
plots <- lapply(names(VARS), function(v) {
  plot(hv_eda(d, x_col = X_COL, y_col = v, y_label = VARS[[v]])) +
    theme_hv_manuscript(base_size = 8)
})
per_page <- NCOL * NROW
pages <- split(plots, ceiling(seq_along(plots) / per_page))
for (i in seq_along(pages)) {
  fname <- sprintf("dp-postage-%02d.png", i)
  png(set_path("descriptive", fname), width = 11, height = 8.5, units = "in", res = 150)
  print(patchwork::wrap_plots(pages[[i]], ncol = NCOL, nrow = NROW))
  invisible(dev.off())
  cat("\n![Page ", i, "](", file.path(paste0(ENDPOINT, "-", TYPE), fname), ")\n", sep = "")
}
```
