---
title: "EDA postage stamps"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (tp.dp.EDA_barplots_scatterplots*.R). -->

Replaces an older EDA bar-plot and scatter-plot sweep.

A `dp-postage` job is the data-checking sweep over a new build: one small panel
per variable against operation year, bars for categorical variables and points
for continuous ones, pages of them. It is for finding coding errors, drift and
missingness before any analysis. It is not a figure for a manuscript.

```{r}
#| label: setup
# The study root is the nearest directory above this file holding _study.yml,
# so the job renders the same from the Render button, quarto render, or
# render_job(), at any depth, with no path in this document to edit.
.in <- knitr::current_input(dir = TRUE)
.root <- hvtiRutilities::study_root(if (is.null(.in)) getwd() else dirname(.in))
.provenance_data <- list()
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(hvtiRutilities)
  library(hvtiPlotR)
  library(ggplot2)
})
if (utils::packageVersion("hvtiPlotR") < "2.7.14") {
  stop("This job needs hvtiPlotR >= 2.7.14 for hv_eda(); ",
       utils::packageVersion("hvtiPlotR"), " is installed.", call. = FALSE)
}
if (!requireNamespace("patchwork", quietly = TRUE)) {
  stop("This job needs the patchwork package. Install it, then re-render.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_STRICT")) %in% c("", "0", "false", "no")) {
      warning(.msg, "\nRendering as a draft; the banner goes when the last marker does. ",
              "Set HVTI_TEMPLATE_STRICT to 1, true or yes to make this stop.", call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Unresolved markers remain. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, "\nThis render stops because HVTI_TEMPLATE_STRICT is '",
           Sys.getenv("HVTI_TEMPLATE_STRICT"), "'. Unset it, or set it to 0, false ",
           "or no, to render a draft instead.", call. = FALSE)
    }
  }
}
```

```{r}
#| label: set
SUBJECT <- "cohort"
TYPE    <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_subject <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_subject, SUBJECT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (subject '", .name_subject,
         "', type '", .name_type, "'), but declares SUBJECT = \"", SUBJECT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root),
                 paste0(SUBJECT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```

## Study choices

Set the values in this chunk before rendering.

```{r}
#| label: study-choices
# EDIT: the built dataset this job reads. "study" is the file named by built:
# in _study.yml; any other name must be declared under additional_datasets:,
# such as a column subset written for R. Analysis sets are derived from
# "study" only, so another dataset needs ANALYSIS_SET <- NULL.
# EDIT: the analysis set this job reads: declare its columns and excluded rows
# under analysis_sets: in _study.yml, then write it with
# hvtiRdatabuild::write_analysis_set(). Use NULL only when the whole built
# cohort is the intended population.
# MIGRATE-BEGIN: dp-postage-config
DATASET <- "study"
ANALYSIS_SET <- "eda"
X_VAR <- "year"                    # EDIT: reference time or year
VARIABLES <- c("age", "female")     # EDIT: variables shown as panels
EXCLUDE <- character(0)
GRID_NCOL <- 4L
GRID_NROW <- 4L
UNIQUE_LIMIT <- 6L
SHOW_PERCENT <- FALSE
# MIGRATE-END: dp-postage-config

```

## Data

<!-- How this job finds its data. Read this before touching the chunk below.

     A study can offer a job data three ways, and `_study.yml` says which exist:

     1. The study dataset: the file named by `built:`. `DATASET <- "study"`
        with `ANALYSIS_SET <- NULL`.
     2. An additional dataset: any file declared under `additional_datasets:`,
        for example a column subset written for R. Set `DATASET` to its
        name there and `ANALYSIS_SET <- NULL`. It is read as it stands: no
        rows are excluded.
     3. An analysis set: declared under `analysis_sets:` with the columns kept
        and the rows excluded, then written once with
        `hvtiRdatabuild::write_analysis_set()`. Set `ANALYSIS_SET` to its name
        and leave `DATASET` as "study", because a set is always cut from the
        study dataset. The report shows its exclusions in order.

     A name under `additional_datasets:` is not an analysis set, so putting it
     in `ANALYSIS_SET` stops with "No analysis set". Whichever route is taken,
     the report prints the dataset, file and dimensions it read, so a reader
     never has to ask which data a table describes. -->

```{r}
#| label: data
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
if (is.null(ANALYSIS_SET)) {
  if (!is.character(DATASET) || length(DATASET) != 1L || is.na(DATASET) || !nzchar(DATASET)) {
    stop("DATASET must be set to one dataset registered in _study.yml, such as \"study\". ",
         "In a migrated job, see the migration report beside this job.", call. = FALSE)
  }
  .cfg <- study_config(start = .root)
  .data_read <- hvtiRtemplates:::.provenance_read(
    DATASET, .cfg, function() read_built(cfg = .cfg, dataset = DATASET)
  )
  d <- .data_read$value
  .provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(.data_read$record))
} else if (!identical(DATASET, "study")) {
  stop("An analysis set is written from the study dataset, not `", DATASET,
       "`. Set ANALYSIS_SET <- NULL to read `", DATASET, "` whole.", call. = FALSE)
} else {
  if (!requireNamespace("hvtiRdatabuild", quietly = TRUE) || utils::packageVersion("hvtiRdatabuild") < "0.2.1") {
    stop("This job needs hvtiRdatabuild >= 0.2.1 for read_analysis_set().", call. = FALSE)
  }
  .cfg <- study_config(start = .root)
  .data_read <- hvtiRtemplates:::.provenance_read(
    "study", .cfg,
    function() hvtiRdatabuild::read_analysis_set(ANALYSIS_SET, cfg = .cfg),
    role = paste0("analysis_set:", ANALYSIS_SET)
  )
  d <- .data_read$value
  .provenance_data <- c(if (exists(".provenance_data")) .provenance_data else list(), list(.data_read$record))
  knitr::kable(attr(d, "attrition"), caption = paste0(
    "Analysis set `", ANALYSIS_SET, "`: exclusions, in order"
  ))
}
cat("Data read: ", if (is.null(ANALYSIS_SET)) {
  paste0("dataset `", DATASET, "` (",
         basename(hvtiRutilities::built_path(dataset = DATASET)), ")")
} else {
  paste0("analysis set `", ANALYSIS_SET, "` of the study dataset")
}, ", ", nrow(d), " rows, ", ncol(d), " columns.\n", sep = "")
```

```{r}
#| label: spec
# Variable order is page order. Labels come from the registered dataset.
# Character and factor columns become bars. Numeric columns become bars only
# for a small set of non-negative integers; other numeric columns become points.
column_names <- function(x) is.character(x) && !anyNA(x) && all(nzchar(x)) && !anyDuplicated(x)
if (!column_names(X_VAR) || length(X_VAR) != 1L) stop("X_VAR must name one column.", call. = FALSE)
if (!column_names(VARIABLES) || !length(VARIABLES)) stop("VARIABLES must name distinct columns.", call. = FALSE)
if (!column_names(EXCLUDE)) stop("EXCLUDE must name distinct columns.", call. = FALSE)
positive_whole <- function(x) is.numeric(x) && length(x) == 1L && !is.na(x) && is.finite(x) && x > 0 && x == floor(x)
for (value in list(GRID_NCOL, GRID_NROW, UNIQUE_LIMIT)) {
  if (!positive_whole(value)) stop("Grid dimensions and UNIQUE_LIMIT must be positive whole numbers.", call. = FALSE)
}
if (!is.logical(SHOW_PERCENT) || length(SHOW_PERCENT) != 1L || is.na(SHOW_PERCENT)) {
  stop("SHOW_PERCENT must be TRUE or FALSE.", call. = FALSE)
}
unknown <- setdiff(c(X_VAR, VARIABLES, EXCLUDE), names(d))
if (length(unknown)) stop("Unknown EDA column(s): ", paste(unknown, collapse = ", "), call. = FALSE)
VARIABLES <- setdiff(VARIABLES, EXCLUDE)
if (!length(VARIABLES)) stop("No EDA variables remain after exclusions.", call. = FALSE)
suspect <- grepl("(^|_)(id|identifier|date|datetime)($|_)|_dt$", VARIABLES, ignore.case = TRUE) |
  vapply(d[VARIABLES], function(x) inherits(x, c("Date", "POSIXt")), logical(1L))
if (any(suspect)) {
  warning("Selected likely identifier or date field(s): ", paste(VARIABLES[suspect], collapse = ", "), call. = FALSE)
}
```

## Pages

```{r}
#| label: helpers
.postage_pages <- function(plots, ncol, nrow) {
  for (value in list(ncol, nrow)) {
    if (!positive_whole(value)) stop("Postage grid dimensions must be positive whole numbers.", call. = FALSE)
  }
  unname(split(plots, ceiling(seq_along(plots) / (ncol * nrow))))
}
```

```{r}
#| label: pages
#| results: asis
plots <- lapply(VARIABLES, function(variable) {
  plot(hvtiPlotR::hv_eda(d, x_col = X_VAR, y_col = variable,
                         y_label = get_label(label_map(d), variable),
                         unique_limit = UNIQUE_LIMIT, show_percent = SHOW_PERCENT)) +
    theme_hv_manuscript(base_size = 8)
})
pages <- .postage_pages(plots, GRID_NCOL, GRID_NROW)
page_plots <- lapply(pages, patchwork::wrap_plots, ncol = GRID_NCOL, nrow = GRID_NROW)
page_files <- character(length(pages))
for (i in seq_along(pages)) {
  page_files[i] <- set_path("graphs", sprintf("dp-postage-page-%02d.png", i))
  ggplot2::ggsave(page_files[i], page_plots[[i]], width = 11, height = 8.5, units = "in", dpi = 150)
}
knitr::include_graphics(page_files)
```

```{r}
#| label: provenance
#| echo: false
#| results: asis
cat(hvtiRtemplates:::.embed_provenance(
  .in,
  data = .provenance_data,
  extra = list(
    subject = SUBJECT,
    type = TYPE
  )
))
```
