---
title: "EDA postage stamps"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (tp.dp.EDA_barplots_scatterplots*.R). -->

Replaces an older EDA bar-plot and scatter-plot sweep.

A `dp-postage` job is the data-checking sweep over a new build: one small panel
per variable against operation year, bars for categorical variables and points
for continuous ones, pages of them. It is for finding coding errors, drift and
missingness before any analysis. It is not a figure for a manuscript.

```{r}
#| label: setup
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(hvtiRutilities)
  library(hvtiPlotR)
  library(ggplot2)
})
if (utils::packageVersion("hvtiPlotR") < "2.7.14") {
  stop("This job needs hvtiPlotR >= 2.7.14 for hv_eda(); ",
       utils::packageVersion("hvtiPlotR"), " is installed.", call. = FALSE)
}
if (!requireNamespace("patchwork", quietly = TRUE)) {
  stop("This job needs the patchwork package. Install it, then re-render.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it.\nTo render a partly-worked draft anyway, set ",
      "HVTI_TEMPLATE_DRAFT to 1, true or yes -- the report then carries a ",
      "draft banner. Any other value, including 0, leaves this guard strict."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_DRAFT")) %in% c("1", "true", "yes")) {
      warning(.msg, call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Rendered with `HVTI_TEMPLATE_DRAFT` set. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, call. = FALSE)
    }
  }
}
```

```{r}
#| label: set
ENDPOINT <- "cohort"
TYPE     <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint,
         "', type '", .name_type, "'), but declares ENDPOINT = \"", ENDPOINT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(hvtiRutilities::study_dir(kind, .root), paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```

## Data

```{r}
#| label: data
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
# EDIT: select the registered dataset or analysis set this job reads.
# For an analysis set, declare its columns and excluded rows
# under analysis_sets: in _study.yml, then write it with
# hvtiRdatabuild::write_analysis_set(). Alternatively, set ANALYSIS_SET to NULL
# and select DATASET ("study" for the whole built cohort).
# MIGRATE-BEGIN: dp-postage-config
DATASET <- NULL
ANALYSIS_SET <- "eda"
X_VAR <- "year"                    # EDIT: reference time or year
VARIABLES <- c("age", "female")     # EDIT: variables shown as panels
EXCLUDE <- character(0)
GRID_NCOL <- 4L
GRID_NROW <- 4L
UNIQUE_LIMIT <- 6L
SHOW_PERCENT <- FALSE
# MIGRATE-END: dp-postage-config
if (is.null(DATASET) == is.null(ANALYSIS_SET)) {
  stop("Select exactly one of DATASET and ANALYSIS_SET.", call. = FALSE)
}
.selection <- if (is.null(DATASET)) ANALYSIS_SET else DATASET
if (!is.character(.selection) || length(.selection) != 1L || is.na(.selection) || !nzchar(.selection)) {
  stop("DATASET or ANALYSIS_SET must be one non-empty name.", call. = FALSE)
}
if (!is.null(DATASET)) {
  cfg <- study_config(.root)
  d <- read_built(cfg, dataset = DATASET)
} else {
  if (!requireNamespace("hvtiRdatabuild", quietly = TRUE) || utils::packageVersion("hvtiRdatabuild") < "0.2.1") {
    stop("This job needs hvtiRdatabuild >= 0.2.1 for read_analysis_set().", call. = FALSE)
  }
  d <- hvtiRdatabuild::read_analysis_set(ANALYSIS_SET)
  knitr::kable(attr(d, "attrition"), caption = paste0(
    "Analysis set `", ANALYSIS_SET, "`: exclusions, in order"
  ))
}
```

```{r}
#| label: spec
# Variable order is page order. Labels come from the registered dataset.
# Character and factor columns become bars. Numeric columns become bars only
# for a small set of non-negative integers; other numeric columns become points.
column_names <- function(x) is.character(x) && !anyNA(x) && all(nzchar(x)) && !anyDuplicated(x)
if (!column_names(X_VAR) || length(X_VAR) != 1L) stop("X_VAR must name one column.", call. = FALSE)
if (!column_names(VARIABLES) || !length(VARIABLES)) stop("VARIABLES must name distinct columns.", call. = FALSE)
if (!column_names(EXCLUDE)) stop("EXCLUDE must name distinct columns.", call. = FALSE)
positive_whole <- function(x) is.numeric(x) && length(x) == 1L && !is.na(x) && is.finite(x) && x > 0 && x == floor(x)
for (value in list(GRID_NCOL, GRID_NROW, UNIQUE_LIMIT)) {
  if (!positive_whole(value)) stop("Grid dimensions and UNIQUE_LIMIT must be positive whole numbers.", call. = FALSE)
}
if (!is.logical(SHOW_PERCENT) || length(SHOW_PERCENT) != 1L || is.na(SHOW_PERCENT)) {
  stop("SHOW_PERCENT must be TRUE or FALSE.", call. = FALSE)
}
unknown <- setdiff(c(X_VAR, VARIABLES, EXCLUDE), names(d))
if (length(unknown)) stop("Unknown EDA column(s): ", paste(unknown, collapse = ", "), call. = FALSE)
VARIABLES <- setdiff(VARIABLES, EXCLUDE)
if (!length(VARIABLES)) stop("No EDA variables remain after exclusions.", call. = FALSE)
suspect <- grepl("(^|_)(id|identifier|date|datetime)($|_)|_dt$", VARIABLES, ignore.case = TRUE) |
  vapply(d[VARIABLES], function(x) inherits(x, c("Date", "POSIXt")), logical(1L))
if (any(suspect)) {
  warning("Selected likely identifier or date field(s): ", paste(VARIABLES[suspect], collapse = ", "), call. = FALSE)
}
```

## Pages

```{r}
#| label: helpers
.postage_pages <- function(plots, ncol, nrow) {
  for (value in list(ncol, nrow)) {
    if (!positive_whole(value)) stop("Postage grid dimensions must be positive whole numbers.", call. = FALSE)
  }
  unname(split(plots, ceiling(seq_along(plots) / (ncol * nrow))))
}
```

```{r}
#| label: pages
#| results: asis
plots <- lapply(VARIABLES, function(variable) {
  plot(hvtiPlotR::hv_eda(d, x_col = X_VAR, y_col = variable,
                         y_label = get_label(label_map(d), variable),
                         unique_limit = UNIQUE_LIMIT, show_percent = SHOW_PERCENT)) +
    theme_hv_manuscript(base_size = 8)
})
pages <- .postage_pages(plots, GRID_NCOL, GRID_NROW)
page_plots <- lapply(pages, patchwork::wrap_plots, ncol = GRID_NCOL, nrow = GRID_NROW)
page_files <- character(length(pages))
for (i in seq_along(pages)) {
  page_files[i] <- set_path("graphs", sprintf("dp-postage-page-%02d.png", i))
  ggplot2::ggsave(page_files[i], page_plots[[i]], width = 11, height = 8.5, units = "in", dpi = 150)
}
knitr::include_graphics(page_files)
```
