---
title: "EDA postage stamps"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: default
    embed-resources: true
---

<!-- EDIT: name the job this replaces (tp.dp.EDA_barplots_scatterplots*.R). -->

Replaces an older EDA bar-plot and scatter-plot sweep.

A `dp-postage` job is the data-checking sweep over a new build: one small panel
per variable against operation year, bars for categorical variables and points
for continuous ones, pages of them. It is for finding coding errors, drift and
missingness before any analysis. It is not a figure for a manuscript.

```{r}
#| label: setup
.root <- if (file.exists("_quarto.yml")) "." else ".."
if (!file.exists(file.path(.root, "_quarto.yml"))) {
  stop("Neither . nor .. contains _quarto.yml, so the project root cannot be ",
       "resolved. Render this from the project root or from the job directory.",
       call. = FALSE)
}
for (f in list.files(file.path(.root, "R"), pattern = "[.]R$", full.names = TRUE)) source(f)
suppressPackageStartupMessages({
  library(hvtiRutilities)
  library(hvtiPlotR)
  library(ggplot2)
})
if (utils::packageVersion("hvtiPlotR") < "2.7.14") {
  stop("This job needs hvtiPlotR >= 2.7.14 for hv_eda(); ",
       utils::packageVersion("hvtiPlotR"), " is installed.", call. = FALSE)
}
if (!requireNamespace("patchwork", quietly = TRUE)) {
  stop("This job needs the patchwork package. Install it, then re-render.", call. = FALSE)
}
```

```{r}
#| label: edit-guard
#| results: asis
.tok <- paste0("ED", "IT", ":")
.cur <- knitr::current_input()
if (!is.null(.cur)) {
  .src <- readLines(.cur, warn = FALSE)
  .hits <- grep(.tok, .src, fixed = TRUE)
  if (length(.hits)) {
    .msg <- paste0(
      length(.hits), " unresolved ", .tok, " marker(s) remain in this job:\n",
      paste0("  - ", trimws(substr(.src[.hits], 1L, 96L)), collapse = "\n"),
      "\nA job that still contains one has not been finished. Work each ",
      "marker and delete it.\nTo render a partly-worked draft anyway, set ",
      "HVTI_TEMPLATE_DRAFT to 1, true or yes -- the report then carries a ",
      "draft banner. Any other value, including 0, leaves this guard strict."
    )
    if (tolower(Sys.getenv("HVTI_TEMPLATE_DRAFT")) %in% c("1", "true", "yes")) {
      warning(.msg, call. = FALSE)
      cat("\n::: {.callout-important title=\"DRAFT -- this job is unfinished\"}\n")
      cat("Rendered with `HVTI_TEMPLATE_DRAFT` set. **The numbers below are not",
          "a result.**\n\n```\n", .msg, "\n```\n", sep = "")
      cat(":::\n\n")
    } else {
      stop(.msg, call. = FALSE)
    }
  }
}
```

```{r}
#| label: set
ENDPOINT <- "cohort"
TYPE     <- "eda"

.current <- knitr::current_input()
if (!is.null(.current)) {
  .fields <- strsplit(sub("[.][^.]+$", "", basename(.current)), "-", fixed = TRUE)[[1L]]
  .name_endpoint <- if (length(.fields) >= 1L) .fields[[1L]] else NA_character_
  .name_type <- if (length(.fields) >= 2L) .fields[[2L]] else NA_character_
  if (!identical(.name_endpoint, ENDPOINT) || !identical(.name_type, TYPE)) {
    stop("This file is named '", .current, "' (endpoint '", .name_endpoint,
         "', type '", .name_type, "'), but declares ENDPOINT = \"", ENDPOINT,
         "\", TYPE = \"", TYPE, "\". Fix the declaration or the filename before rendering.",
         call. = FALSE)
  }
}

set_path <- function(kind, file) {
  d <- file.path(.root, kind, paste0(ENDPOINT, "-", TYPE))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
  file.path(d, file)
}
```

## Data

```{r}
#| label: data
hvtiRutilities::verify_manifest(file.path(.root, "manifest.yaml"))
# EDIT: the analysis set this job reads: declare its columns and excluded rows
# under analysis_sets: in _study.yml, then write it with
# hvtiRdatabuild::write_analysis_set(). Replace both lines with d <- read_built()
# only when the whole built cohort is the intended population.
ANALYSIS_SET <- "eda"
d <- hvtiRdatabuild::read_analysis_set(ANALYSIS_SET)
knitr::kable(attr(d, "attrition"), caption = paste0(
  "Analysis set `", ANALYSIS_SET, "`: exclusions, in order"
))
```

```{r}
#| label: spec
# EDIT: the x-axis for every panel, usually operation year.
X_COL <- "year"
# EDIT: the variables to sweep, named by column with a display label. Order is
# page order. hv_eda() decides bar or scatter per variable (6 distinct values or
# fewer is categorical), so a column does not need a type here.
VARS <- c(age = "Age (years)", female = "Female")
# Panels per page. 3 x 3 keeps a label readable on a printed page.
NCOL <- 3L
NROW <- 3L
```

## Pages

```{r}
#| label: pages
#| results: asis
unknown <- setdiff(c(X_COL, names(VARS)), names(d))
if (length(unknown)) {
  stop("Unknown EDA column(s): ", paste(unknown, collapse = ", "), call. = FALSE)
}
plots <- lapply(names(VARS), function(v) {
  plot(hv_eda(d, x_col = X_COL, y_col = v, y_label = VARS[[v]])) +
    theme_hv_manuscript(base_size = 8)
})
per_page <- NCOL * NROW
pages <- split(plots, ceiling(seq_along(plots) / per_page))
for (i in seq_along(pages)) {
  fname <- sprintf("dp-postage-%02d.png", i)
  png(set_path("descriptive", fname), width = 11, height = 8.5, units = "in", res = 150)
  print(patchwork::wrap_plots(pages[[i]], ncol = NCOL, nrow = NROW))
  invisible(dev.off())
  cat("\n![Page ", i, "](", file.path(paste0(ENDPOINT, "-", TYPE), fname), ")\n", sep = "")
}
```
